### 2021-07-15 analysis/report/effectiveness_report.Rmd
add rmd report; output rds plots; yaml tweaks

---
title: "Comparative vaccine effectiveness (ChAdOx1 versus BNT162b2) in Health and Social Care workers"
output:
  html_document:
    code_folding: hide
    keep_md: yes
  pdf_document: default
  md_document:
    variant: gfm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library('tidyverse')
library('here')
library('glue')
library('lubridate')
library('gt')
library('patchwork')

output_dir <- here("output")

pct <- scales::label_percent(accuracy=0.1)

```

# Introduction

This study compares the effectiveness of the *BNT162b2* mRNA (Pfizer-BioNTech) and *ChAdOx1* (Oxford-AstraZeneca) vaccines amongst health and social care workers (HCWs). 
This group was chosen as they were prioritised for vaccination due to high occupational exposure and during the period where both vaccines were available (4 January onwards) it is expected that the actual vaccine brand received will not be strongly influenced by other determinants of COVID-19-related outcomes (i.e., approximately random treatment allocation and therefore reduced confounding).

The code and data for this report can be found at the OpenSafely [comparative-ve-research GitHub repository](https://github.com/opensafely/comparative-ve-research).

# Methods

## Study population

People meeting the following criteria are included:

* HCWs who have received at least one dose of BNT162b2 or ChAdOx1. 
* Registered at a GP practice using TPP's SystmOne clinical information system on the day of vaccination.
* Aged 18-65.
* Vaccinated on or after 4 January 2021, when both vaccine brands were being administered. 
* Not Clinically Extremely Vulnerable (CEV), as set out by government guidance, at the time of vaccination.

Study participants are followed up from vaccination date until the first of:

* The outcome of interest
* Death or deregistration
* Second vaccine dose
* 31 May 2021

### Identifying vaccinated HCWs

Those vaccinated as part of England's national vaccination programme (for example not including vaccine trial participants) are asked whether they work in health or social care. This information is sent to OpenSAFELY-TPP from NHS Digital's COVID-19 data store.

Note -- many of those flagged as HCWs do not have a vaccination record, which shouldn't be the case if the question was asked as part of the vaccination process.

## Study measures

### Exposure

The vaccination brand, BNT162b2 or ChAdOx1. This is available in the GP record directly, via the National Immunisation Management System (NIMS).

### Outcomes

* SARS-CoV-2 infection, as identified via SGSS records. Both PCR and LFT tests are included.
* Unplanned COVID-19-related hospitalisation via HES records. ICD-10 codes are used to identify COVID-19-related admissions. 
* Unplanned COVID-19-related ICU admission via HES records. ICD-10 codes are used to identify COVID-19-related admissions. 
* COVID-19-related death via death registration records. With COVID-19 ICD-10 codes anywhere on the death certificate (i.e., as an underlying or contributing cause of death)

### Covariates

Patient characteristics used for adjustment include: age, sex, deprivation, ethnicity, region, clinically “at risk” (but not clinically extremely vulnerable) as per national prioritisation guidelines, asthma, number of previous SARS-CoV-2 tests (via SGSS), rurality, evidence of prior covid infection (positive test or COVID-19 hospitalisation), number of recorded comorbidities, severe mental illness. All variables are ascertained as at the time of vaccination.

## Statistical Analysis

Time-dependent Cox models are used to estimate the relative hazard of each outcome for ChAdOx1 versus BNT162b2 vaccine recipients.
Modelled covariates include age, sex, ethnicity, deprivation, evidence of prior covid infection, asthma, clinically “at risk”, number of recorded comorbidities and severe mental illness.
NUTS1 regions are used as a stratification variable to account for geographical variation in event rates for the outcomes of interest (primarily driven by changes in local infection rates).

Comparative effectiveness will be estimated at weekly post-vaccination intervals to account for the potential differences in vaccine protection over time between the two brands.

Two time-scales are used for triagulation:

* On the *calendar time-scale*, time zero is 4 January 2021, and people only contribute follow-up time once they have been vaccinated (i.e., delayed study entry).

* On the *treatment time-scale*, time zero is the date of vaccination. Calendar-time is included as an additional covariate, modelled as a restricted cubic spline.

# Results

Note all counts below 6 are redacted.

## Flowchart

```{r flowchart, echo=FALSE}

flowchart <- read_csv(here("output", "data", "flowchart.csv"))

flowchart %>%
  select(
    Criteria = criteria,
    N = n,
    `N excluded` = n_exclude,
    `% excluded` = pct_exclude,
    `% remaining` = pct_all
  ) %>%
  gt() %>%
  fmt_percent(
    columns = starts_with(c("%")),
    decimals = 1
  ) %>%
  fmt_number(
    columns = starts_with(c("N")),
    decimals= 0
  ) %>%
  fmt_missing(
    everything(),
    missing_text="-"
  )

```

## Baseline demographics

```{r table1, echo=FALSE}
read_rds(here("output", "descriptive", "tables", "table1.rds"))
```

## Event rates

```{r table_irr, echo=FALSE}
read_rds(here("output", "descriptive", "tables", "table_irr.rds"))
```

## Unadjusted survival curves

```{r, curves, echo=FALSE, message=FALSE, out.width='90%'}

print_plots <- function(outcome, outcome_descr){
  km_image <- here("output", "descriptive", "km", glue("plot_survival_ci_{outcome}.png"))
  calendar_image <- here("output", outcome, "calendar", glue("forest_plot_cox.png"))
  timesincevax_image <- here("output", outcome, "timesincevax", glue("forest_plot_cox.png"))

  km_plot <- read_rds(here("output", "descriptive", "km", glue("plot_survival_ci_{outcome}.rds")))
  calendar_plot <- read_rds(here("output", outcome, "calendar", glue("forest_plot_cox.rds")))
  timesincevax_plot <- read_rds(here("output", outcome, "timesincevax", glue("forest_plot_cox.rds")))

  #cat(glue("### {outcome_descr}"), "  \n")

  effects_plot <- wrap_plots(calendar_plot, timesincevax_plot, nrow=1, guides='collect')

  wrap_plot(
    km_plot+theme(legend.position = c(0.1,0.1), legend.justification = c(0,0)), 
    effects_plot, 
    ncol=1,
  ) + 
  plot_annotation(title = '{outcome_descr}')

}

print_plots("test", "SARS-CoV-2 test")
print_plots("postest", "SARS-CoV-2 positive test")
print_plots("emergency", "A&E attendances")
print_plots("covidadmitted", "COVID-19 hospital admission")
print_plots("covidcc", "COVID-19 critical care admission")

```

### 2021-07-15 analysis/descriptive/km.R

@@ -116,6 +116,8 @@ data_tte <- data_cohort %>%
       TRUE ~ NA_character_
     ),
 
+    vax1_date = vax1_date-1, # assume vaccination occurs at the start of the day, and all other events occur at the end of the day.
+
     # time to last follow up day
     tte_enddate = tte(vax1_date, end_date, end_date),
 
@@ -214,7 +216,7 @@ get_colour_scales <- function(colour_type = "qual"){
   if(colour_type == "qual"){
     list(
       scale_color_brewer(type="qual", palette="Set1", na.value="grey"),
-      scale_fill_brewer(type="qual", palette="Set1", guide=FALSE, na.value="grey")
+      scale_fill_brewer(type="qual", palette="Set1", guide="none", na.value="grey")
       #ggthemes::scale_color_colorblind(),
       #ggthemes::scale_fill_colorblind(guide=FALSE),
       #rcartocolor::scale_color_carto_d(palette = "Safe"),
@@ -260,7 +262,7 @@ ggplot_surv <- function(.surv_data, colour_var, colour_name, colour_type="qual",
       x="Days since vaccination",
       y="Event-free rate",
       colour=colour_name,
-      title=title
+      title=NULL
     )+
     theme_minimal(base_size=9)+
     theme(
@@ -340,7 +342,13 @@ plot_combinations %>%
   ) %>%
   pwalk(ggsave) %>%
   mutate(filename = str_replace(filename, ".png", ".svg")) %>%
-  pwalk(ggsave)
+  pwalk(ggsave) %>%
+  transmute(
+    x = plot,
+    path = file.path(path, str_replace(filename, ".svg", ".rds")),
+    compress="gz"
+  ) %>%
+  pwalk(write_rds)
 
 
 plot_combinations %>%
@@ -355,5 +363,11 @@ plot_combinations %>%
   ) %>%
   pwalk(ggsave) %>%
   mutate(filename = str_replace(filename, ".png", ".svg")) %>%
-  pwalk(ggsave)
+  pwalk(ggsave) %>%
+  transmute(
+    x = plot,
+    path = file.path(path, str_replace(filename, ".svg", ".rds")),
+    compress="gz"
+  ) %>%
+  pwalk(write_rds)
 

### 2021-07-15 analysis/descriptive/table1.R

@@ -38,13 +38,13 @@ gbl_vars <- jsonlite::fromJSON(
 #list2env(gbl_vars, globalenv())
 
 ## create output directories ----
-fs::dir_create(here::here("output", "descriptive", "tables"))
+fs::dir_create(here("output", "descriptive", "tables"))
 
 
 ## import metadata ----
 var_labels <- read_rds(here("output", "data", "metadata_labels.rds"))
 
-list_formula <- read_rds(here::here("output", "data", "metadata_formulas.rds"))
+list_formula <- read_rds(here("output", "data", "metadata_formulas.rds"))
 list2env(list_formula, globalenv())
 
 
@@ -56,10 +56,10 @@ data_cohort <- read_rds(here("output", "data", "data_cohort.rds"))
 tab_summary_baseline <- data_cohort %>%
   select(
     all_of(names(var_labels)),
-    -age, -region, -stp
+    -age, -region, -stp, -vax1_type
   ) %>%
   tbl_summary(
-    by = vax1_type,
+    by = vax1_type_descr,
     label=unname(var_labels[names(.)])
   )  %>%
   modify_footnote(starts_with("stat_") ~ NA)
@@ -70,8 +70,9 @@ tab_csv <- tab_summary_baseline$table_body
 names(tab_csv) <- tab_summary_baseline$table_header$label
 tab_csv <- tab_csv[, (!tab_summary_baseline$table_header$hide | tab_summary_baseline$table_header$label=="variable")]
 
-gtsave(as_gt(tab_summary_baseline), here::here("output", "descriptive", "tables", "table1.html"))
-write_csv(tab_csv, here::here("output", "descriptive", "tables", "table1.csv"))
+write_rds(tab_summary_baseline, here("output", "descriptive", "tables", "table1.rds"))
+gtsave(as_gt(tab_summary_baseline), here("output", "descriptive", "tables", "table1.html"))
+write_csv(tab_csv, here("output", "descriptive", "tables", "table1.csv"))
 
 
 
@@ -92,6 +93,6 @@ tab_region_csv <- tab_summary_region$table_body
 names(tab_region_csv) <- tab_summary_region$table_header$label
 tab_region_csv <- tab_region_csv[, (!tab_summary_region$table_header$hide | tab_summary_region$table_header$label=="variable")]
 
-gtsave(as_gt(tab_summary_region), here::here("output", "descriptive", "tables", "table1_regions.html"))
[…]

### 2021-07-15 analysis/descriptive/table_irr.R

@@ -363,6 +363,7 @@ tab_summary <- data_summary %>%
     columns = "timesincevax"
   )
 
+write_rds(tab_summary, here("output", "descriptive", "tables", "table_irr.rds"))
 gtsave(tab_summary, here("output", "descriptive", "tables", "table_irr.html"))
 
 

### 2021-07-15 analysis/models/model_cox.R

@@ -72,6 +72,7 @@ data_cohort <- read_rds(here("output", "data", "data_cohort.rds"))
 # redo tte variables to indclude censoring date (ie use na.censor=FALSE)
 data_tte <- data_cohort %>%
   mutate(
+    vax1_date = vax1_date-1, # assume vaccination occurs at the start of the day, and all other events occur at the end of the day.
 
     # time to study end date
     tte_enddate = tte(vax1_date, end_date, end_date),

### 2021-07-15 analysis/models/report_cox.R

@@ -3,7 +3,7 @@
 # This script:
 # imports fitted MSMs
 # calculates robust CIs taking into account patient-level clustering
-# outputs forest plots for the primary vaccine-outcome relationship
+# outputs effect plots for the primary vaccine-outcome relationship
 # outputs plots showing model-estimated spatio-temporal trends
 #
 # The script should only be run via an action in the project.yaml only
@@ -99,7 +99,7 @@ write_csv(tidy_summary, path = here::here("output", outcome, timescale, glue::gl
 
 
 if(timescale == "calendar"){
-  coxmod_forest_data <- tidy_summary %>%
+  coxmod_effect_data <- tidy_summary %>%
     filter(str_detect(term, fixed("vax1_az"))) %>%
     mutate(
       term=str_replace(term, pattern=fixed("vax1_az:timesincevax"), ""),
@@ -109,7 +109,7 @@ if(timescale == "calendar"){
 }
 
 if(timescale=="timesincevax"){
-  coxmod_forest_data <- tidy_summary %>%
+  coxmod_effect_data <- tidy_summary %>%
     filter(str_detect(term, fixed("timesincevax")) | str_detect(term, fixed("vax1_az"))) %>%
     mutate(
       term=str_replace(term, pattern=fixed("vax1_az:strata(timesincevax)"), ""),
@@ -117,7 +117,7 @@ if(timescale=="timesincevax"){
     )
 }
 
-coxmod_forest_data <- coxmod_forest_data %>%
+coxmod_effect_data <- coxmod_effect_data %>%
   mutate(
     term_left = as.numeric(str_extract(term, "^\\d+"))-1,
     term_right = as.numeric(str_extract(term, "\\d+$"))-1,
@@ -125,8 +125,8 @@ coxmod_forest_data <- coxmod_forest_data %>%
     term_midpoint = term_left + (term_right+1-term_left)/2
   )
 
-coxmod_forest <-
-  ggplot(data = coxmod_forest_data) +
+coxmod_effect <-
+  ggplot(data = coxmod_effect_data) +
   geom_point(aes(y=estimate, x=term_midpoint, colour=model_name), position = position_dodge(width = 1.8))+
   geom_linerange(aes(ymin=conf.low, ymax=conf.high, x=term_midpoint, colour=model_name), position = position_dodge(width = 1.8))+
   geom_hline(aes(yintercept=1), colour='grey')+
[…]

### 2021-07-15 analysis/process/data_process.R

@@ -29,7 +29,7 @@ gbl_vars <- jsonlite::fromJSON(
 
 # output processed data to rds ----
 
-dir.create(here("output", "data"), showWarnings = FALSE, recursive=TRUE)
+fs::dir_create(here("output", "data"))
 
 
 # process ----
@@ -107,6 +107,14 @@ data_processed <- data_extract %>%
       right=FALSE
     ),
 
+    sex = fct_case_when(
+      sex == "F" ~ "Female",
+      sex == "M" ~ "Male",
+      #sex == "I" ~ "Inter-sex",
+      #sex == "U" ~ "Unknown",
+      TRUE ~ NA_character_
+    ),
+
     ethnicity_combined = if_else(is.na(ethnicity), ethnicity_6_sus, ethnicity),
 
     ethnicity_combined = fct_case_when(
@@ -154,6 +162,12 @@ data_processed <- data_extract %>%
     vax1_type = case_when(
       pmin(covid_vax_az_1_date, as.Date("2030-01-01"), na.rm=TRUE) <= pmin(covid_vax_pfizer_1_date, as.Date("2030-01-01"), na.rm=TRUE) ~ "az",
       pmin(covid_vax_az_1_date, as.Date("2030-01-01"), na.rm=TRUE) > pmin(covid_vax_pfizer_1_date, as.Date("2030-01-01"), na.rm=TRUE) ~ "pfizer",
+      TRUE ~ NA_character_
+    ),
+
+    vax1_type_descr = fct_case_when(
+      vax1_type == "pfizer" ~ "BNT162b2",
+      vax1_type == "az" ~ "ChAdOx1",
       TRUE ~ NA_character_
     ),
 

### 2021-07-15 analysis/process/data_selection.R

@@ -13,11 +13,13 @@ library('tidyverse')
 library('here')
 library('glue')
 
+source(here("analysis", "lib", "utility_functions.R"))
+
 ## import command-line arguments ----
 args <- commandArgs(trailingOnly=TRUE)
 
 ## create output directories ----
-dir.create(here("output", "data"), showWarnings = FALSE, recursive=TRUE)
+fs::dir_create(here("output", "data"))
 
 ## Import processed data ----
 
@@ -62,7 +64,6 @@ data_flowchart <- data_criteria %>%
     c3_nopriorvaccine = c2_notcev & (no_prior_vaccine),
     c4_enddate = c3_nopriorvaccine & vax_beforeenddate,
     c5_4jan = c4_enddate & vax1_4janonwards,
-
   ) %>%
   summarise(
     across(.fns=sum)
@@ -77,5 +78,15 @@ data_flowchart <- data_criteria %>%
     pct_exclude = n_exclude/lag(n),
     pct_all = n / first(n),
     pct_step = n / lag(n),
+    crit = str_extract(criteria, "^c\\d+"),
+    criteria = fct_case_when(
+      crit == "c0" ~ "All HCWs aged 16-65",
+      crit == "c1" ~ "  with no missing info for age, IMD, ethnicity, region",
+      crit == "c2" ~ "  who are not clinically extremely vulnerable",
+      crit == "c3" ~ "  with no vaccination record prior to the national roll-out",
+      crit == "c4" ~ "  with vaccination on or before the study end date",
+      crit == "c5" ~ "  with vaccination on or after 4 January 2021",
+      TRUE ~ NA_character_
+    )
   )
 write_csv(data_flowchart, here("output", "data", "flowchart.csv"))

### 2021-07-15 analysis/process/design.R

@@ -99,6 +99,7 @@ write_rds(list_formula, here("output", "data", "metadata_formulas.rds"))
 ## variable labels
 var_labels <- list(
   vax1_type ~ "Vaccine type",
+  vax1_type_descr ~ "Vaccine type",
   age ~ "Age",
   ageband ~ "Age",
   sex ~ "Sex",

### 2021-07-15 create-project.R

@@ -93,6 +93,9 @@ action_report <- function(
     run = glue("r:latest analysis/models/report_cox.R"),
     arguments = c(outcome, timescale),
     needs = list("design", glue("model_{outcome}_{timescale}")),
+    highly_sensitive = list(
+      rds = glue("output/{outcome}/{timescale}/estimates*.rds")
+    ),
     moderately_sensitive = list(
       svg = glue("output/{outcome}/{timescale}/forest*.svg"),
       png = glue("output/{outcome}/{timescale}/forest*.png"),
@@ -209,6 +212,9 @@ actions_list <- splice(
     name = "descr_table1",
     run = "r:latest analysis/descriptive/table1.R",
     needs = list("design", "data_selection"),
+    highly_sensitive = list(
+      rds = "output/descriptive/tables/table1*.rds"
+    ),
     moderately_sensitive = list(
       html = "output/descriptive/tables/table1*.html",
       csv = "output/descriptive/tables/table1*.csv"
@@ -230,6 +236,9 @@ actions_list <- splice(
     run = "r:latest analysis/descriptive/table_irr.R",
     arguments = c("output/data/data_processed.rds", "output/data_properties"),
     needs = list("design", "data_selection"),
+    highly_sensitive = list(
+      rds = "output/descriptive/tables/table_irr.rds"
+    ),
     moderately_sensitive = list(
       html = "output/descriptive/tables/table_irr.html",
       csv = "output/descriptive/tables/table_irr.csv"
@@ -241,6 +250,9 @@ actions_list <- splice(
     run = "r:latest analysis/descriptive/km.R",
     arguments = c("output/data/data_processed.rds", "output/data_properties"),
     needs = list("design", "data_selection"),
+    highly_sensitive = list(
+      rds = "output/descriptive/km/plot_survival*.rds"
+    ),
     moderately_sensitive = list(
       png = "output/descriptive/km/plot_survival*.png",
       svg = "output/descriptive/km/plot_survival*.svg"
@@ -266,8 +278,29 @@ actions_list <- splice(
   action_model("covidadmitted", "timesincevax"),
   action_report("covidadmitted", "timesincevax"),
   action_model("covidadmitted", "calendar"),
-  action_report("covidadmitted", "calendar")
+  action_report("covidadmitted", "calendar"),
+
+  # action_model("covidcc", "timesincevax"),
[…]

### 2021-07-15 analysis/models/report_cox.R
use deprecated path argument

@@ -166,7 +166,7 @@ coxmod_effect <-
 coxmod_effect
 ## save plot
 
-write_rds(coxmod_effect, file=here::here("output", outcome, timescale, glue::glue("forest_plot_cox.rds")))
+write_rds(coxmod_effect, path=here::here("output", outcome, timescale, glue::glue("forest_plot_cox.rds")))
 ggsave(filename=here::here("output", outcome, timescale, glue::glue("forest_plot_cox.svg")), coxmod_effect, width=20, height=15, units="cm")
 ggsave(filename=here::here("output", outcome, timescale, glue::glue("forest_plot_cox.png")), coxmod_effect, width=20, height=15, units="cm")
 

### 2021-07-15 analysis/models/report_cox.R
rename outputs

@@ -95,7 +95,7 @@ add_column(outcome = outcome, .before=1)
 
 if(removeobs) rm(coxmod0, coxmod1, coxmod2, coxmod3)
 
-write_csv(tidy_summary, path = here::here("output", outcome, timescale, glue::glue("estimates_cox.csv")))
+write_csv(tidy_summary, path = here::here("output", outcome, timescale, glue::glue("effect_estimates_cox.csv")))
 
 
 if(timescale == "calendar"){
@@ -166,9 +166,9 @@ coxmod_effect <-
 coxmod_effect
 ## save plot
 
-write_rds(coxmod_effect, path=here::here("output", outcome, timescale, glue::glue("forest_plot_cox.rds")))
-ggsave(filename=here::here("output", outcome, timescale, glue::glue("forest_plot_cox.svg")), coxmod_effect, width=20, height=15, units="cm")
-ggsave(filename=here::here("output", outcome, timescale, glue::glue("forest_plot_cox.png")), coxmod_effect, width=20, height=15, units="cm")
+write_rds(coxmod_effect, path=here::here("output", outcome, timescale, glue::glue("effect_plot_cox.rds")))
+ggsave(filename=here::here("output", outcome, timescale, glue::glue("effect_plot_cox.svg")), coxmod_effect, width=20, height=15, units="cm")
+ggsave(filename=here::here("output", outcome, timescale, glue::glue("effect_plot_cox.png")), coxmod_effect, width=20, height=15, units="cm")
 
 
 

### 2021-07-15 create-project.R

@@ -94,12 +94,12 @@ action_report <- function(
     arguments = c(outcome, timescale),
     needs = list("design", glue("model_{outcome}_{timescale}")),
     highly_sensitive = list(
-      rds = glue("output/{outcome}/{timescale}/estimates*.rds")
+      rds = glue("output/{outcome}/{timescale}/effect*.rds")
     ),
     moderately_sensitive = list(
-      svg = glue("output/{outcome}/{timescale}/forest*.svg"),
-      png = glue("output/{outcome}/{timescale}/forest*.png"),
-      csv = glue("output/{outcome}/{timescale}/estimates*.csv")
+      svg = glue("output/{outcome}/{timescale}/effect*.svg"),
+      png = glue("output/{outcome}/{timescale}/effect*.png"),
+      csv = glue("output/{outcome}/{timescale}/effect*.csv")
     )
   )
 }

### 2021-07-15 analysis/report/effectiveness_report.Rmd
update rmd

@@
 
-## Unadjusted survival curves
+## Comparative effectiveness
 
-```{r, curves, echo=FALSE, message=FALSE, out.width='90%'}
+The plots below show unadjusted Kaplan-Meier curves, and the *ChAdOx1* versus *BNT162b2* hazard ratios for each model.
+
+```{r, curves, echo=FALSE, message=FALSE, warning=FALSE, out.width='90%'}
 
@@
   km_image <- here("output", "descriptive", "km", glue("plot_survival_ci_{outcome}.png"))
-  calendar_image <- here("output", outcome, "calendar", glue("forest_plot_cox.png"))
-  timesincevax_image <- here("output", outcome, "timesincevax", glue("forest_plot_cox.png"))
+  calendar_image <- here("output", outcome, "calendar", glue("effect_plot_cox.png"))
+  timesincevax_image <- here("output", outcome, "timesincevax", glue("effect_plot_cox.png"))
 
   km_plot <- read_rds(here("output", "descriptive", "km", glue("plot_survival_ci_{outcome}.rds")))
-  calendar_plot <- read_rds(here("output", outcome, "calendar", glue("forest_plot_cox.rds")))
-  timesincevax_plot <- read_rds(here("output", outcome, "timesincevax", glue("forest_plot_cox.rds")))
+  km_plot <- km_plot + theme(legend.position = c(0.1,0.1), legend.justification = c(0,0))
+
+  calendar_plot <- read_rds(here("output", outcome, "calendar", glue("effect_plot_cox.rds"))) + 
+    labs(title=NULL)+
+    coord_cartesian(ylim=c(0.25,4))+
+    theme(text = element_text(size=10))
+
+  timesincevax_plot <- read_rds(here("output", outcome, "timesincevax", glue("effect_plot_cox.rds"))) + 
+    labs(title=NULL)+
+    coord_cartesian(ylim=c(0.25,4))+
+    theme(text = element_text(size=10))
 
@@
 
-  effects_plot <- wrap_plots(calendar_plot, timesincevax_plot, nrow=1, guides='collect')
+  effects_plot <- wrap_plots(calendar_plot, timesincevax_plot, nrow=1)#, guides='collect')
 
-  wrap_plot(
-    km_plot+theme(legend.position = c(0.1,0.1), legend.justification = c(0,0)), 
+  wrap_plots(
+   km_plot,
    effects_plot,
-    ncol=1,
+   ncol=1
   ) +
-  plot_annotation(title = '{outcome_descr}')
+  plot_annotation(title = outcome_descr)
 
@@
 print_plots("postest", "SARS-CoV-2 positive test")
-print_plots("emergency", "A&E attendances")
+print_plots("emergency", "A&E attendance")
 print_plots("covidadmitted", "COVID-19 hospital admission")
-print_plots("covidcc", "COVID-19 critical care admission")
+#print_plots("covidcc", "COVID-19 critical care admission")
 

### 2021-07-15 analysis/process/design.R
use fs

@@ -10,7 +10,7 @@
 library('tidyverse')
 library('here')
 ## create output directories ----
-dir.create(here("output", "data"), showWarnings = FALSE, recursive=TRUE)
+fs::dir_create(here("output", "data"))
 
 ## Import processed data ----
 

### 2021-07-15 analysis/report/effectiveness_report.Rmd
tweak dates

@@
 
-pct <- scales::label_percent(accuracy=0.1)
+gbl_vars <- jsonlite::fromJSON(
+  txt=here("analysis", "global-variables.json")
+)
 
@@
 This study compares the effectiveness of the *BNT162b2* mRNA (Pfizer-BioNTech) and *ChAdOx1* (Oxford-AstraZeneca) vaccines amongst health and social care workers (HCWs). 
-This group was chosen as they were prioritised for vaccination due to high occupational exposure and during the period where both vaccines were available (4 January onwards) it is expected that the actual vaccine brand received will not be strongly influenced by other determinants of COVID-19-related outcomes (i.e., approximately random treatment allocation and therefore reduced confounding).
+This group was chosen as they were prioritised for vaccination due to high occupational exposure and during the period where both vaccines were available (`r format(as.Date(gbl_vars$start_date_az), "%d %B %Y")` onwards) it is expected that the actual vaccine brand received will not be strongly influenced by other determinants of COVID-19-related outcomes (i.e., approximately random treatment allocation and therefore reduced confounding).
 
@@
 * Aged 18-65.
-* Vaccinated on or after 4 January 2021, when both vaccine brands were being administered. 
+* Vaccinated on or after `r format(as.Date(gbl_vars$start_date_az), "%d %B %Y")`, when both vaccine brands were being administered. 
 * Not Clinically Extremely Vulnerable (CEV), as set out by government guidance, at the time of vaccination.
@@
 * Second vaccine dose
-* 31 May 2021
+* The study end date, `r format(as.Date(gbl_vars$end_date), "%d %B %Y")`
 
@@
 
-* On the *calendar time-scale*, time zero is 4 January 2021, and people only contribute follow-up time once they have been vaccinated (i.e., delayed study entry).
+* On the *calendar time-scale*, time zero is `r format(as.Date(gbl_vars$start_date_az), "%d %B %Y")`, and people only contribute follow-up time once they have been vaccinated (i.e., delayed study entry).
 

### 2021-07-20 analysis/report/effectiveness_report.Rmd
output rds effects; redo report plots

@@
   km_image <- here("output", "descriptive", "km", glue("plot_survival_ci_{outcome}.png"))
-  calendar_image <- here("output", outcome, "calendar", glue("effect_plot_cox.png"))
-  timesincevax_image <- here("output", outcome, "timesincevax", glue("effect_plot_cox.png"))
+  calendar_image <- here("output", outcome, "calendar", glue("effect_plot.png"))
+  timesincevax_image <- here("output", outcome, "timesincevax", glue("effect_plot.png"))
 
@@
 
-  calendar_plot <- read_rds(here("output", outcome, "calendar", glue("effect_plot_cox.rds"))) + 
-    labs(title=NULL)+
-    coord_cartesian(ylim=c(0.25,4))+
-    theme(text = element_text(size=10))
+  calendar_effects <- read_rds(here("output", outcome, "calendar", glue("effect_estimates.rds"))) %>%
+    mutate(timescale = "Calendar time")
+  timesincevax_effects <- read_rds(here("output", outcome, "timesincevax", glue("effect_estimates.rds"))) %>%
+    mutate(timescale = "Vaccinated time")
 
-  timesincevax_plot <- read_rds(here("output", outcome, "timesincevax", glue("effect_plot_cox.rds"))) + 
-    labs(title=NULL)+
-    coord_cartesian(ylim=c(0.25,4))+
-    theme(text = element_text(size=10))
+  effects <- bind_rows(calendar_effects, timesincevax_effects)
 
-  #cat(glue("### {outcome_descr}"), "  \n")
+  effects_plot <-
+    ggplot(data = effects) +
+    geom_point(aes(y=estimate, x=term_midpoint, colour=model_name), position = position_dodge(width = 1.8))+
+    geom_linerange(aes(ymin=conf.low, ymax=conf.high, x=term_midpoint, colour=model_name), position = position_dodge(width = 1.8))+
+    geom_hline(aes(yintercept=1), colour='grey')+
+    facet_grid(cols=vars(timescale))+
+    scale_x_continuous(breaks=unique(effects$term_left), limits=c(min(effects$term_left), max(effects$term_right)+1), expand = c(0, 0))+
+    scale_y_log10(
+    breaks=c(0.1, 0.25, 0.33, 0.5, 0.75, 1, 1.33, 2, 3, 4, 10)
+    #sec.axis = dup_axis(name="favours\n<- BNT162b2 / ChAdOx1 ->", breaks = NULL)
+  )+
+    scale_colour_brewer(type="qual", palette="Set2", guide=guide_legend(ncol=2))+
+    labs(
+      y="Hazard ratio\nChAdOx1 / BNT162b2",
+      x="Days since vaccination",
+      colour=NULL
+     ) +
+    theme_bw()+
+    theme(
+      panel.border = element_blank(),
+      axis.line.y = element_line(colour = "black"),
 
-  effects_plot <- wrap_plots(calendar_plot, timesincevax_plot, nrow=1)#, guides='collect')
+      panel.grid.minor.x = element_blank(),
+      panel.grid.minor.y = element_blank(),
+      strip.background = element_blank(),
+      strip.placement = "outside",
+      strip.text.y.left = element_text(angle = 0),
+
+      panel.spacing = unit(0.8, "lines"),
+
+      plot.title = element_text(hjust = 0),
+      plot.title.position = "plot",
+      plot.caption.position = "plot",
+      plot.caption = element_text(hjust = 0, face= "italic"),
+
+      legend.position = "bottom"
+     ) +
+   NULL
+
+  # 
+  # calendar_plot <- read_rds(here("output", outcome, "calendar", glue("effect_plot.rds"))) + 
+  #   labs(title=NULL)+
+  #   coord_cartesian(ylim=c(0.25,4))+
+  #   theme(text = element_text(size=10))
+  # 
+  # timesincevax_plot <- read_rds(here("output", outcome, "timesincevax", glue("effect_plot.rds"))) + 
+  #   labs(title=NULL)+
+  #   coord_cartesian(ylim=c(0.25,4))+
+  #   theme(text = element_text(size=10))
+  # 
+  cat(glue("### {outcome_descr}"), "  \n")
+
+  #effects_plot <- wrap_plots(calendar_plot, timesincevax_plot, nrow=1)#, guides='collect')
 

### 2021-07-20 analysis/models/report_cox.R

@@ -95,7 +95,7 @@ add_column(outcome = outcome, .before=1)
 
 if(removeobs) rm(coxmod0, coxmod1, coxmod2, coxmod3)
 
-write_csv(tidy_summary, path = here::here("output", outcome, timescale, glue::glue("effect_estimates_cox.csv")))
+write_csv(tidy_summary, path = here::here("output", outcome, timescale, glue::glue("effect_estimates.csv")))
 
 
 if(timescale == "calendar"){
@@ -124,6 +124,8 @@ coxmod_effect_data <- coxmod_effect_data %>%
     term_right = if_else(is.na(term_right), max(term_left)+6, term_right),
     term_midpoint = term_left + (term_right+1-term_left)/2
   )
+
+write_rds(coxmod_effect_data, path = here::here("output", outcome, timescale, glue::glue("effect_estimates.rds")))
 
 coxmod_effect <-
   ggplot(data = coxmod_effect_data) +
@@ -166,9 +168,9 @@ coxmod_effect <-
 coxmod_effect
 ## save plot
 
-write_rds(coxmod_effect, path=here::here("output", outcome, timescale, glue::glue("effect_plot_cox.rds")))
-ggsave(filename=here::here("output", outcome, timescale, glue::glue("effect_plot_cox.svg")), coxmod_effect, width=20, height=15, units="cm")
-ggsave(filename=here::here("output", outcome, timescale, glue::glue("effect_plot_cox.png")), coxmod_effect, width=20, height=15, units="cm")
+write_rds(coxmod_effect, path=here::here("output", outcome, timescale, glue::glue("effect_plot.rds")))
+ggsave(filename=here::here("output", outcome, timescale, glue::glue("effect_plot.svg")), coxmod_effect, width=20, height=15, units="cm")
+ggsave(filename=here::here("output", outcome, timescale, glue::glue("effect_plot.png")), coxmod_effect, width=20, height=15, units="cm")
 
 
 

### 2021-07-23 analysis/report/effectiveness_report.Rmd
improve axes

@@
 
-```{r, curves, echo=FALSE, message=FALSE, warning=FALSE, out.width='90%'}
+```{r, curves, echo=FALSE, message=FALSE, warning=FALSE, out.width='90%', results='asis'}
 
@@
     scale_y_log10(
-    breaks=c(0.1, 0.25, 0.33, 0.5, 0.75, 1, 1.33, 2, 3, 4, 10)
-    #sec.axis = dup_axis(name="favours\n<- BNT162b2 / ChAdOx1 ->", breaks = NULL)
+    breaks=c(0.1, 0.125, 0.25, 0.33, 0.5, 0.75, 1, 1.33, 2, 3, 4, 8,10),
+    sec.axis = dup_axis(name="favours\n <- BNT162b2 / ChAdOx1 ->", breaks = NULL)
   )+
@@
     labs(
-      y="Hazard ratio\nChAdOx1 / BNT162b2",
+      y="Hazard ratio",#\nChAdOx1 / BNT162b2",
       x="Days since vaccination",
@@
   # 
-  cat(glue("### {outcome_descr}"), "  \n")
+  #cat(glue("### {outcome_descr}"), "  \n")
 
   #effects_plot <- wrap_plots(calendar_plot, timesincevax_plot, nrow=1)#, guides='collect')
-
-  wrap_plots(
-   km_plot,
-   effects_plot,
-   ncol=1
-  ) +
-  plot_annotation(title = outcome_descr)
+  cat("  \n\n")
+  print(glue("### {outcome_descr}"), "  \n")
+  print(km_plot)
+  print(effects_plot)
+  cat("  \n\n")
+  # wrap_plots(
+  #  km_plot + labs(title=outcome_descr),
+  #  effects_plot,
+  #  ncol=1
+  # ) +
+  # plot_annotation(title = outcome_descr)
 

### 2021-07-23 analysis/descriptive/km.R

@@ -48,39 +48,6 @@ ceiling_any <- function(x, to=1){
 }
 
 
-# # function to extract total plot height minus panel height
-# plotHeight <- function(plot, unit){
-#   grob <- ggplot2::ggplotGrob(plot)
-#   grid::convertHeight(gtable::gtable_height(grob), unitTo=unit, valueOnly=TRUE)
-# }
-#
-# # function to extract total plot width minus panel height
-# plotWidth <- function(plot, unit){
-#   grob <- ggplot2::ggplotGrob(plot)
-#   grid::convertWidth(gtable::gtable_width(grob), unitTo=unit, valueOnly=TRUE)
-# }
-#
-# # function to extract total number of bars plot (strictly this is the number of rows in the build of the plot data)
-# plotNbars <- function(plot){
-#   length(unique(ggplot2::ggplot_build(plot)$data[[1]]$x))
-# }
-#
-# # function to extract total number of bars plot (strictly this is the number of rows in the build of the plot data)
-# plotNfacetrows <- function(plot){
-#   length(levels(ggplot2::ggplot_build(plot)$data[[1]]$PANEL))
-# }
-#
-# plotNyscales <- function(plot){
-#   length(ggplot2::ggplot_build(plot)$layout$panel_scales_y[[1]]$range$range)
-# }
-#
-#
-# # function to extract total number of panels
-# plotNpanelrows <- function(plot){
-#   length(unique(ggplot2::ggplot_build(plot)$layout$layout$ROW))
-# }
-
-
 ## define theme ----
 
 plot_theme <-

### 2021-07-23 analysis/report/effectiveness_report.Rmd
minor text edits

@@
 
-Comparative effectiveness will be estimated at weekly post-vaccination intervals to account for the potential differences in vaccine protection over time between the two brands.
+Comparative effectiveness is estimated at weekly post-vaccination intervals to account for the potential differences in vaccine protection over time between the two brands.
 
-Two time-scales are used for triagulation:
+Two time-scales are used for triangulation:
 
@@
 
-* On the *treatment time-scale*, time zero is the date of vaccination. Calendar-time is included as an additional covariate, modelled as a restricted cubic spline.
+* On the *treatment time-scale*, time zero is the date of vaccination.
+Calendar-time is included as an additional covariate using restricted cubic spline with knots at the first and second tertile, plus boundary knots.
 
@@
 
-Note all counts below 6 are redacted.
+Note all counts below 6 are redacted and Kaplan-Meier survival estimates are rounded for disclosure control.
 

### 2021-08-09 analysis/report/effectiveness_report.Rmd
update report

@@
 
-```{r flowchart, echo=FALSE}
+```{r flowchart, echo=FALSE, warning=FALSE, message=FALSE}
 
@@
 
-print_plots <- function(outcome, outcome_descr){
+print_plots_km <- function(outcome){
+
   km_image <- here("output", "descriptive", "km", glue("plot_survival_ci_{outcome}.png"))
-  calendar_image <- here("output", outcome, "calendar", glue("effect_plot.png"))
-  timesincevax_image <- here("output", outcome, "timesincevax", glue("effect_plot.png"))
 
   km_plot <- read_rds(here("output", "descriptive", "km", glue("plot_survival_ci_{outcome}.rds")))
-  km_plot <- km_plot + theme(legend.position = c(0.1,0.1), legend.justification = c(0,0))
 
-  calendar_effects <- read_rds(here("output", outcome, "calendar", glue("effect_estimates.rds"))) %>%
+  print(km_plot)
+  cat("  \n\n")
+
+}
+
+print_plots_effects <- function(outcome, modeltype){
+
+  suffix <- if(modeltype=="plr"){
+    "_pw"
+  } else{
+    ""
+  }
+
+  calendar_effects <- read_rds(here("output", "models", outcome, "calendar", glue("report{modeltype}_effects{suffix}.rds"))) %>%
     mutate(timescale = "Calendar time")
-  timesincevax_effects <- read_rds(here("output", outcome, "timesincevax", glue("effect_estimates.rds"))) %>%
+  timesincevax_effects <- read_rds(here("output", "models", outcome, "timesincevax", glue("report{modeltype}_effects{suffix}.rds"))) %>%
     mutate(timescale = "Vaccinated time")
 
-  effects <- bind_rows(calendar_effects, timesincevax_effects)
+  tidy_effects <- bind_rows(calendar_effects, timesincevax_effects)
 
-  effects_plot <-
-    ggplot(data = effects) +
-    geom_point(aes(y=estimate, x=term_midpoint, colour=model_name), position = position_dodge(width = 1.8))+
-    geom_linerange(aes(ymin=conf.low, ymax=conf.high, x=term_midpoint, colour=model_name), position = position_dodge(width = 1.8))+
+  plot_effects <-
+    ggplot(data = tidy_effects) +
+    geom_point(aes(y=exp(estimate), x=term_midpoint, colour=model_name), position = position_dodge(width = 1.8))+
+    geom_linerange(aes(ymin=exp(conf.low), ymax=exp(conf.high), x=term_midpoint, colour=model_name), position = position_dodge(width = 1.8))+
     geom_hline(aes(yintercept=1), colour='grey')+
     facet_grid(cols=vars(timescale))+
-    scale_x_continuous(breaks=unique(effects$term_left), limits=c(min(effects$term_left), max(effects$term_right)+1), expand = c(0, 0))+
+    scale_x_continuous(breaks=unique(tidy_effects$term_left), limits=c(min(tidy_effects$term_left), max(tidy_effects$term_right)+1), expand = c(0, 0))+
     scale_y_log10(
@@
 
-  # 
-  # calendar_plot <- read_rds(here("output", outcome, "calendar", glue("effect_plot.rds"))) + 
-  #   labs(title=NULL)+
-  #   coord_cartesian(ylim=c(0.25,4))+
-  #   theme(text = element_text(size=10))
-  # 
-  # timesincevax_plot <- read_rds(here("output", outcome, "timesincevax", glue("effect_plot.rds"))) + 
-  #   labs(title=NULL)+
-  #   coord_cartesian(ylim=c(0.25,4))+
-  #   theme(text = element_text(size=10))
-  # 
-  #cat(glue("### {outcome_descr}"), "  \n")
-
-  #effects_plot <- wrap_plots(calendar_plot, timesincevax_plot, nrow=1)#, guides='collect')
+  print(plot_effects)
   cat("  \n\n")
-  print(glue("### {outcome_descr}"), "  \n")
-  print(km_plot)
-  print(effects_plot)
-  cat("  \n\n")
-  # wrap_plots(
-  #  km_plot + labs(title=outcome_descr),
-  #  effects_plot,
-  #  ncol=1
-  # ) +
-  # plot_annotation(title = outcome_descr)
-
 }
 
-print_plots("test", "SARS-CoV-2 test")
-print_plots("postest", "SARS-CoV-2 positive test")
-print_plots("emergency", "A&E attendance")
-print_plots("covidadmitted", "COVID-19 hospital admission")
-#print_plots("covidcc", "COVID-19 critical care admission")
+cat("  \n\n")
+print(glue("### SARS-CoV-2 test"), "  \n")
+print(glue("#### Unadjusted"), "  \n")
+print_plots_km("test")
+print(glue("#### Cox"), "  \n")
+print_plots_effects("test", "cox")
+print(glue("#### PLR"), "  \n")
+print_plots_effects("test", "plr")
+
+cat("  \n\n")
+print(glue("### SARS-CoV-2 positive test"), "  \n")
+print(glue("#### Unadjusted"), "  \n")
+print_plots_km("postest")
+print(glue("#### Cox"), "  \n")
+print_plots_effects("postest", "cox")
+print(glue("#### PLR"), "  \n")
+print_plots_effects("postest", "plr")
+
+cat("  \n\n")
+print(glue("### A&E attendance"), "  \n")
+print(glue("#### Unadjusted"), "  \n")
+print_plots_km("emergency")
+print(glue("#### Cox"), "  \n")
+print_plots_effects("emergency", "cox")
+print(glue("#### PLR"), "  \n")
+print_plots_effects("emergency", "plr")
+
+cat("  \n\n")
+print(glue("### COVID-19 hospital admission"), "  \n")
+print(glue("#### Unadjusted"), "  \n")
+print_plots_km("covidadmitted")
+print(glue("#### Cox"), "  \n")
+print_plots_effects("covidadmitted", "cox")
+print(glue("#### PLR"), "  \n")
+print_plots_effects("covidadmitted", "plr")
 

### 2021-08-09 create-project.R

@@ -344,6 +344,7 @@ actions_list <- splice(
   # action_model("covidcc", "calendar", "plr", 50000),
   # action_report("covidcc", "calendar", "plr"),
 
+  comment("# # # # # # # # # # # # # # # # # # #", "Report", "# # # # # # # # # # # # # # # # # # #"),
 
   action(
     name = "rmd_report",
@@ -355,7 +356,8 @@ actions_list <- splice(
       as.list(glue("report_{outcome}_{timescale}_{modeltype}", outcome=rep(c("test", "postest", "emergency", "covidadmitted"),each=2), timescale = rep(c("timesincevax", "calendar"), 4), modeltype="cox"))
     ),
     moderately_sensitive = list(
-      html = "output/effectiveness_report.html"
+      html = "output/effectiveness_report.html",
+      md = "output/effectiveness_report.md"
     )
   )
 )
@@ -384,12 +386,12 @@ as.yaml(project_list, indent=2) %>%
 
 names(actions_list) %>% tibble(action=.) %>%
   mutate(
-    model = str_detect(action, "model"),
-    model_number = cumsum(model)
+    model = action==""  & lag(action!="", 1, TRUE),
+    model_number = cumsum(model),
   ) %>%
   group_by(model_number) %>%
   summarise(
-    sets = paste(action, collapse=" ")
+    sets = str_trim(paste(action, collapse=" "))
   ) %>% pull(sets) %>%
   paste(collapse="\n") %>%
   writeLines(here("actions.txt"))

### 2021-08-10 analysis/report/effectiveness_report.Rmd
render md, html, and keep figures

@@
 output:
-  html_document:
-    code_folding: hide
-    keep_md: yes
-  pdf_document: default
-  md_document:
-    variant: gfm
+  github_document:
+    keep_html: TRUE
+  #  code_folding: hide
+    #keep_md: yes
+  #pdf_document: default
+  # md_document:
+  #   variant: gfm
+  #html_document
 ---
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
 
@@
 output_dir <- here("output")
+
+fs::dir_create(output_dir, "report", "figures")
+
+knitr::opts_chunk$set(
+  echo = TRUE,
+  fig.path = paste0(fs::path(output_dir, "report", "figures"), "/")
+)
 

### 2021-08-10 create-project.R

@@ -348,18 +348,32 @@ actions_list <- splice(
 
   action(
     name = "rmd_report",
-    run = glue("r:latest -e {q}", q = single_quote('rmarkdown::render("analysis/report/effectiveness_report.Rmd", knit_root_dir = "/workspace", output_dir="/workspace/output")') ),
+    run = glue(
+      "r:latest -e {q}",
+      q = single_quote('rmarkdown::render("analysis/report/effectiveness_report.Rmd",  knit_root_dir = "/workspace",  output_dir = "/workspace/output/report", output_format = c("rmarkdown::github_document")   )')
+    ),
     needs = splice(
       "design", "data_selection",
       "descr_table1", "descr_irr",
       "descr_km",
-      as.list(glue("report_{outcome}_{timescale}_{modeltype}", outcome=rep(c("test", "postest", "emergency", "covidadmitted"),each=2), timescale = rep(c("timesincevax", "calendar"), 4), modeltype="cox"))
+      as.list(
+        glue_data(
+             .x = expand_grid(
+               outcome = c("test", "postest", "emergency", "covidadmitted"),
+               modeltype = c("cox", "plr"),
+               timescale = c("timesincevax", "calendar")
+             ),
+             "report_{outcome}_{timescale}_{modeltype}"
+        )
+      )
     ),
     moderately_sensitive = list(
-      html = "output/effectiveness_report.html",
-      md = "output/effectiveness_report.md"
+      html = "output/report/effectiveness_report.html",
+      md = "output/report/effectiveness_report.md",
+      figures = "output/report/figures/*.png"
     )
   )
+
 )
 
 
@@ -395,3 +409,4 @@ names(actions_list) %>% tibble(action=.) %>%
   ) %>% pull(sets) %>%
   paste(collapse="\n") %>%
   writeLines(here("actions.txt"))
+

### 2021-08-10 analysis/report/effectiveness_report.Rmd
add cumulative incidence currves to markdown report

@@
 
+print_plots_cmlinc <- function(outcome, spline){
+
+  calendar_curve <- read_rds(here("output", "models", outcome, "calendar", glue("reportplr_cmlinccurves_{spline}.rds"))) %>%
+    mutate(timescale = "Calendar time")
+  timesincevax_curve <- read_rds(here("output", "models", outcome, "timesincevax", glue("reportplr_cmlinccurves_{spline}.rds"))) %>%
+    mutate(timescale = "Vaccinated time")
+
+  curves <- bind_rows(calendar_curve, timesincevax_curve)
+
+  plot_cmlinc <-
+    ggplot(curves %>% filter(model=="3"))+
+    geom_step(aes(x=tstop, y=1-survival, group=vax1_az_descr, colour=vax1_az_descr))+
+    facet_grid(cols=vars(timescale))+
+    scale_x_continuous(
+      breaks = seq(0,7*52,by=14),
+      expand = expansion(0)
+    )+
+    scale_y_continuous(
+      expand = expansion(0)
+    )+
+    scale_colour_brewer(type="qual", palette="Set2")+
+    labs(
+      x="Days since first dose",
+      y="Cumulative risk",
+      colour=NULL
+    )+
+    theme_bw()+
+    theme(
+      legend.position=c(0.05,.95),
+      legend.justification = c(0,1),
+      axis.text.x.top=element_text(hjust=0),
+
+      panel.border = element_blank(),
+      axis.line.y = element_line(colour = "black"),
+
+      panel.grid.minor.x = element_blank(),
+      panel.grid.minor.y = element_blank(),
+      strip.background = element_blank(),
+      strip.placement = "outside",
+      strip.text.y.left = element_text(angle = 0),
+
+      panel.spacing = unit(0.8, "lines"),
+
+    )+
+    NULL
+
+  print(plot_cmlinc)
+  cat("  \n\n")
+}
+
 cat("  \n\n")
@@
 print_plots_effects("test", "plr")
+print_plots_cmlinc("test", "ns")
 
@@
 print_plots_effects("postest", "plr")
+print_plots_cmlinc("postest", "ns")
 
@@
 print_plots_effects("emergency", "plr")
+print_plots_cmlinc("emergency", "ns")
 
@@
 print_plots_effects("covidadmitted", "plr")
+print_plots_cmlinc("covidadmitted", "ns")
 

### 2021-08-10 analysis/models/report_plr.R

@@ -265,6 +265,8 @@ cml_inc <- ggplot(curves %>% filter(model=="3"))+
     axis.text.x.top=element_text(hjust=0)
   )
 
+write_rds(curves, here("output", "models", outcome, timescale, glue("reportplr_cmlinccurves_pw.rds")), compress="gz")
+
 ggsave(filename=here("output", "models", outcome, timescale, glue("reportplr_cmlincplot_pw.svg")), cml_inc, width=20, height=15, units="cm")
 ggsave(filename=here("output", "models", outcome, timescale, glue("reportplr_cmlincplot_pw.png")), cml_inc, width=20, height=15, units="cm")
 
@@ -390,6 +392,8 @@ cml_inc <- ggplot(curves %>% filter(model=="3"))+
     axis.text.x.top=element_text(hjust=0)
   )
 
+write_rds(curves, here("output", "models", outcome, timescale, glue("reportplr_cmlinccurves_ns.rds")), compress="gz")
+
 ggsave(filename=here("output", "models", outcome, timescale, glue("reportplr_cmlincplot_ns.svg")), cml_inc, width=20, height=15, units="cm")
 ggsave(filename=here("output", "models", outcome, timescale, glue("reportplr_cmlincplot_ns.png")), cml_inc, width=20, height=15, units="cm")
 
@@ -447,7 +451,7 @@ plotplr <-
   ) +
   NULL
 
-
+write_rds(plotplr, path=here("output", "models", outcome, timescale, glue("reportplr_effectsplot_ns.rds")))
 ggsave(filename=here("output", "models", outcome, timescale, glue("reportplr_effectsplot_ns.svg")), plotplr, width=20, height=15, units="cm")
 ggsave(filename=here("output", "models", outcome, timescale, glue("reportplr_effectsplot_ns.png")), plotplr, width=20, height=15, units="cm")
 

### 2021-08-10 analysis/seconddose/graphs.R
remove testing variables; change redaction colour

@@ -108,9 +108,9 @@ data_tte <- data_cohort %>%
 
 plot_2dose <- function(var, var_descr, weekly){
 
-  weekly <- TRUE
-  var <- "ageband"
-  var_descr <- "Age group"
+  # weekly <- TRUE
+  # var <- "ageband"
+  # var_descr <- "Age group"
 
   resolution <- ifelse(weekly, "weekly", "daily")
 
@@ -165,7 +165,7 @@ plot_2dose <- function(var, var_descr, weekly){
     )+
     scale_y_continuous(limits=c(14,18*7), breaks=seq(0,365,14))+
     scale_fill_gradientn(
-      colours=c("white", "lightgrey", "lightgrey", "darkblue"),
+      colours=c("transparent", "aliceblue", "aliceblue", "darkblue"),
       values=scales::rescale(c(0,1,redact_threshold,max(data_freq$n, na.rm=TRUE))),
       na.value = "grey50"
     )+

### 2021-08-10 analysis/report/effectiveness_report_vaccinatedtime.Rmd
add vaxtime-only report

--- 2021-08-10 analysis/report/effectiveness_report.Rmd
@@
 
-fs::dir_create(output_dir, "report", "figures")
+fs::dir_create(output_dir, "report", "figures-vaxtime")
 
@@
   echo = TRUE,
-  fig.path = paste0(fs::path(output_dir, "report", "figures"), "/")
+  fig.path = paste0(fs::path(output_dir, "report", "figures-vaxtime"), "/")
 )
@@
 Modelled covariates include age, sex, ethnicity, deprivation, evidence of prior covid infection, asthma, clinically “at risk”, number of recorded comorbidities and severe mental illness.
-NUTS1 regions are used as a stratification variable to account for geographical variation in event rates for the outcomes of interest (primarily driven by changes in local infection rates).
+NHS regions are used as a stratification variable to account for geographical variation in event rates for the outcomes of interest (primarily driven by changes in local infection rates).
 
 Comparative effectiveness is estimated at weekly post-vaccination intervals to account for the potential differences in vaccine protection over time between the two brands. 
-
-Two time-scales are used for triangulation:
-
-* On the *calendar time-scale*, time zero is `r format(as.Date(gbl_vars$start_date_az), "%d %B %Y")`, and people only contribute follow-up time once they have been vaccinated (i.e., delayed study entry).
-
-* On the *treatment time-scale*, time zero is the date of vaccination.
-Calendar-time is included as an additional covariate using restricted cubic spline with knots at the first and second tertile, plus boundary knots.
+Time zero is the date of vaccination. Calendar-time is included as an additional covariate using restricted cubic spline with knots at the first and second tertile, plus boundary knots.
 
@@
 
-```{r, curves, echo=FALSE, message=FALSE, warning=FALSE, out.width='90%', results='asis'}
+```{r, curves_vaxtime, echo=FALSE, message=FALSE, warning=FALSE, out.width='90%', results='asis'}
 
@@
 
-  calendar_effects <- read_rds(here("output", "models", outcome, "calendar", glue("report{modeltype}_effects{suffix}.rds"))) %>%
-    mutate(timescale = "Calendar time")
-  timesincevax_effects <- read_rds(here("output", "models", outcome, "timesincevax", glue("report{modeltype}_effects{suffix}.rds"))) %>%
-    mutate(timescale = "Vaccinated time")
-
-  tidy_effects <- bind_rows(calendar_effects, timesincevax_effects)
+  tidy_effects <- read_rds(here("output", "models", outcome, "timesincevax", glue("report{modeltype}_effects{suffix}.rds")))
 
@@
     geom_hline(aes(yintercept=1), colour='grey')+
-    facet_grid(cols=vars(timescale))+
     scale_x_continuous(breaks=unique(tidy_effects$term_left), limits=c(min(tidy_effects$term_left), max(tidy_effects$term_right)+1), expand = c(0, 0))+
@@
 
-  calendar_curve <- read_rds(here("output", "models", outcome, "calendar", glue("reportplr_cmlinccurves_{spline}.rds"))) %>%
-    mutate(timescale = "Calendar time")
-  timesincevax_curve <- read_rds(here("output", "models", outcome, "timesincevax", glue("reportplr_cmlinccurves_{spline}.rds"))) %>%
-    mutate(timescale = "Vaccinated time")
-
-  curves <- bind_rows(calendar_curve, timesincevax_curve)
+  curves <- read_rds(here("output", "models", outcome, "timesincevax", glue("reportplr_cmlinccurves_{spline}.rds")))
 
   plot_cmlinc <-
-    ggplot(curves %>% filter(model=="3"))+
+    ggplot(curves)+
     geom_step(aes(x=tstop, y=1-survival, group=vax1_az_descr, colour=vax1_az_descr))+
-    facet_grid(cols=vars(timescale))+
     scale_x_continuous(
@@
     )+
-    scale_colour_brewer(type="qual", palette="Set2")+
+    scale_colour_brewer(type="qual", palette="Set2", guide=guide_legend(ncol=2))+
     labs(
@@
       panel.grid.minor.y = element_blank(),
-      strip.background = element_blank(),
-      strip.placement = "outside",
-      strip.text.y.left = element_text(angle = 0),
 

### 2021-08-10 create-project.R

@@ -372,6 +372,33 @@ actions_list <- splice(
       md = "output/report/effectiveness_report.md",
       figures = "output/report/figures/*.png"
     )
+  ),
+
+  action(
+    name = "rmd_report_vaxtime",
+    run = glue(
+      "r:latest -e {q}",
+      q = single_quote('rmarkdown::render("analysis/report/effectiveness_report_vaccinatedtime.Rmd",  knit_root_dir = "/workspace",  output_dir = "/workspace/output/report", output_format = c("rmarkdown::github_document")   )')
+    ),
+    needs = splice(
+      "design", "data_selection",
+      "descr_table1", "descr_irr",
+      "descr_km",
+      as.list(
+        glue_data(
+          .x = expand_grid(
+            outcome = c("test", "postest", "emergency", "covidadmitted"),
+            modeltype = c("cox", "plr")
+          ),
+          "report_{outcome}_timesincevax_{modeltype}"
+        )
+      )
+    ),
+    moderately_sensitive = list(
+      html = "output/report/effectiveness_report_vaccinatedtime.html",
+      md = "output/report/effectiveness_report_vaccinatedtime.md",
+      figures = "output/report/figures-vaxtime/*.png"
+    )
   )
 
 )

### 2021-08-11 analysis/report/effectiveness_report.Rmd
update report

@@
 
-### Covariates
+## Statistical Analysis
 
-Patient characteristics used for adjustment include: age, sex, deprivation, ethnicity, region, clinically “at risk” (but not clinically extremely vulnerable) as per national prioritisation guidelines, asthma, number of previous SARS-CoV-2 tests (via SGSS), rurality, evidence of prior covid infection (positive test or COVID-19 hospitalisation), number of recorded comorbidities, severe mental illness.
+The aim is to estimate comparative vaccine effectiveness, i.e., the relative hazard of each outcome for ChAdOx1 versus BNT162b2 vaccine recipients.
+The effect is permitted to vary by time since vaccination, to account for the potential differences in vaccine protection over time between the two brands.
+
+Patient characteristics used for adjustment include: age, sex, deprivation, ethnicity, NHS region, clinically “at risk” (but not clinically extremely vulnerable) as per national prioritisation guidelines, asthma, number of previous SARS-CoV-2 tests (via SGSS), rurality, evidence of prior covid infection (positive test or COVID-19 hospitalisation), number of recorded comorbidities, severe mental illness.
 All variables are ascertained as at the time of vaccination.
 
-## Statistical Analysis
+NHS regions are used as a stratification variable to account for geographical variation in event rates for the outcomes of interest (primarily driven by changes in local infection rates).
 
-Time-dependent Cox models are used to estimate the relative hazard of each outcome for ChAdOx1 versus BNT162b2 vaccine recipients.
-Modelled covariates include age, sex, ethnicity, deprivation, evidence of prior covid infection, asthma, clinically “at risk”, number of recorded comorbidities and severe mental illness.
-NUTS1 regions are used as a stratification variable to account for geographical variation in event rates for the outcomes of interest (primarily driven by changes in local infection rates).
-
-Comparative effectiveness is estimated at weekly post-vaccination intervals to account for the potential differences in vaccine protection over time between the two brands.
-
-Two time-scales are used for triangulation:
+### Time-dependent Cox models
+Time-dependent Cox models are used to estimate the time-varying effects for each vaccine type. Two time-scales are used for triangulation:
 
 * On the *calendar time-scale*, time zero is `r format(as.Date(gbl_vars$start_date_az), "%d %B %Y")`, and people only contribute follow-up time once they have been vaccinated (i.e., delayed study entry).
+Vaccine type, time since vaccination in weekly intervals, and the interaction between these two terms are included to estimate time-varying vaccine effects for each vaccine type.
+In `R`'s `coxph` function, this is modelled using `vaccine_type*week_since_first_dose`
 
 * On the *treatment time-scale*, time zero is the date of vaccination.
-Calendar-time is included as an additional covariate using restricted cubic spline with knots at the first and second tertile, plus boundary knots.
+Calendar-time is included as an additional covariate, in weekly intervals, using restricted cubic spline with knots at the first and second tertile, plus boundary knots.
+Here, follow-up time is stratified into weekly intervals with the vaccine-specific effect estimated within each time-strata.
+In `R`'s `coxph` function, this is modelled using `vaccine_type:strata(week_since_first_dose)`.
+
+### Pooled Logistic Regression models
+
+We emulate the Cox models described above using pooled logistic regression (PLR), with the outcome risk estimated at daily intervals.
+This serves two purposes.
+Firstly, a continuous-time estimate of comparative effectiveness, as opposed to the piecewise-linear approximation above, can be obtained using flexible parametric splines.
+Secondly, the absolute cumulative risk of each outcome of interest for each vaccine type, marginalised over the population, can be obtained using the parametric g-formula.
+These models include a restricted cubic spline with 3 degrees of freedom on the timescale of interest, interacted with vaccine type.
+Cumulative risk is then calculated as the average risk for each day of follow-up predicted by the PLR model, under the (complimentary counterfactual) assumptions that every patient received the BNT162b2 vaccine or that every patient received the ChAdOx1 vaccine.
+
+The person-time dataset needed to fit the PLR model is large and leads to infeasible RAM requirements / computation time.
+To deal with this, a sampling strategy is used such that all those who experienced an outcome and a random sample of 50000 who did not experience the ouotcome are selected for the models.
+The models are weighted to recover the characteristics of the complete dataset.
+This weighting is also applied to the average cumulative risk.
 
@@
 
-The plots below show unadjusted Kaplan-Meier curves, and the *ChAdOx1* versus *BNT162b2* hazard ratios for each model.
+The plots below show:
 
-```{r, curves, echo=FALSE, message=FALSE, warning=FALSE, out.width='90%', results='asis'}
+* Unadjusted Kaplan-Meier curves
+* *ChAdOx1* versus *BNT162b2* piecewise-linear hazard ratios estimated using the Cox models
+* *ChAdOx1* versus *BNT162b2* continuous hazard ratios estimated using the PLR models
+* Marginalised cumulative risk for *ChAdOx1* and *BNT162b2* estimated using the PLR models
+
+```{r, curves, echo=FALSE, message=FALSE, warning=FALSE, out.width='80%', results='asis'}
 
@@
 
-print_plots_cmlinc <- function(outcome, spline){
+print_plots_spline <- function(outcome){
 
-  calendar_curve <- read_rds(here("output", "models", outcome, "calendar", glue("reportplr_cmlinccurves_{spline}.rds"))) %>%
+  calendar_effects <- read_rds(here("output", "models", outcome, "calendar", glue("reportplr_effects_ns.rds"))) %>%
     mutate(timescale = "Calendar time")
-  timesincevax_curve <- read_rds(here("output", "models", outcome, "timesincevax", glue("reportplr_cmlinccurves_{spline}.rds"))) %>%
+  timesincevax_effects <- read_rds(here("output", "models", outcome, "timesincevax", glue("reportplr_effects_ns.rds"))) %>%
+    mutate(timescale = "Vaccinated time")
+
+  spline <- bind_rows(calendar_effects, timesincevax_effects)
+
+  plot_spline <-
+    ggplot(spline)+
+    geom_hline(aes(yintercept=1), colour='grey')+
+    geom_line(aes(x=tstop-1, y=hr, colour=model_name))+
+    geom_ribbon(aes(x=tstop-1, ymin=hr.ll, ymax=hr.ul, fill=model_name), alpha=0.2, colour="transparent")+
+    facet_grid(cols=vars(timescale))+
+    scale_x_continuous(
+      breaks = seq(0,7*52,by=14),
+      expand = expansion(0)
+    )+
+    scale_y_log10(
+      breaks=c(1/8, 1/6, 0.25, 0.33, 0.5, 0.67, 1, 1.5, 2, 3, 4, 6, 8),
+      sec.axis = dup_axis(name="favours\n <- BNT162b2 / ChAdOx1 ->", breaks = NULL)
+    )+
+    scale_colour_brewer(type="qual", palette="Set2", guide=guide_legend(ncol=1))+
+    scale_fill_brewer(type="qual", palette="Set2", guide="none")+
+    labs(
+      x = "Days since first dose",
+      y = "Hazard ratio",
+      colour = NULL, fill=NULL
+    )+
+    theme_bw()+
+    theme(
+      panel.border = element_blank(),
+      axis.line.y = element_line(colour = "black"),
+
+      panel.grid.minor.y = element_blank(),
+      strip.background = element_blank(),
+      strip.placement = "outside",
+      strip.text.y.left = element_text(angle = 0),
+
+      panel.spacing = unit(0.8, "lines"),
+
+      plot.title = element_text(hjust = 0),
+      plot.title.position = "plot",
+      plot.caption.position = "plot",
+      plot.caption = element_text(hjust = 0, face= "italic"),
+
+      legend.position = "bottom"
+     ) +
+   NULL
+
+  print(plot_spline)
+  cat("  \n\n")
+}
+
+print_plots_cmlinc <- function(outcome){
+
+  calendar_curve <- read_rds(here("output", "models", outcome, "calendar", glue("reportplr_cmlinccurves_ns.rds"))) %>%
+    mutate(timescale = "Calendar time")
+  timesincevax_curve <- read_rds(here("output", "models", outcome, "timesincevax", glue("reportplr_cmlinccurves_ns.rds"))) %>%
     mutate(timescale = "Vaccinated time")
@@
 print_plots_effects("test", "plr")
-print_plots_cmlinc("test", "ns")
+print_plots_spline("test")
+print_plots_cmlinc("test")
 
@@
 print_plots_effects("postest", "cox")
-print(glue("#### PLR"), "  \n")
+print(glue("#### PLR, peicewise"), "  \n")
 print_plots_effects("postest", "plr")
-print_plots_cmlinc("postest", "ns")
+print(glue("#### PLR, spline"), "  \n")
+print_plots_spline("postest")
+print_plots_cmlinc("postest")
 
@@
 print_plots_effects("emergency", "cox")
-print(glue("#### PLR"), "  \n")
+print(glue("#### PLR, peicewise"), "  \n")
 print_plots_effects("emergency", "plr")
-print_plots_cmlinc("emergency", "ns")
+print(glue("#### PLR, spline"), "  \n")
+print_plots_spline("emergency")
+print_plots_cmlinc("emergency")
 
@@
 print_plots_effects("covidadmitted", "cox")
-print(glue("#### PLR"), "  \n")
+print(glue("#### PLR, peicewise"), "  \n")
 print_plots_effects("covidadmitted", "plr")
-print_plots_cmlinc("covidadmitted", "ns")
+print(glue("#### PLR, spline"), "  \n")
+print_plots_spline("covidadmitted")
+print_plots_cmlinc("covidadmitted")
 

### 2021-08-11 analysis/lib/utility_functions.R

@@ -258,14 +258,14 @@ tidy_custom.glm  <- function(model, conf.int=TRUE, conf.level=0.95, exponentiate
   output
 }
 
-plr.predict <- function(x, clcov, newdata){
+plr.predict <- function(model, clcov, newdata){
   # from https://stackoverflow.com/questions/3790116/using-clustered-covariance-matrix-in-predict-lm
-  if(missing(newdata)){ newdata <- x$model }
-  tt <- terms(x)
+  if(missing(newdata)){ newdata <- model$model }
+  tt <- terms(model)
   Terms <- delete.response(tt)
   m.mat <- model.matrix(Terms, data=newdata)
-  m.coef <- x$coef
-  fit <- as.vector(m.mat %*% x$coef)
+  m.coef <- model$coef
+  fit <- as.vector(m.mat %*% model$coef)
   se.fit <- sqrt( rowSums(m.mat * (m.mat %*% clcov)) )
   return(list(fit=fit, se.fit=se.fit))
 }

### 2021-08-11 analysis/models/report_plr.R

@@ -22,7 +22,7 @@ if(length(args)==0){
   # use for interactive testing
   removeobs <- FALSE
   outcome <- "postest"
-  timescale <- "timesincevax"
+  timescale <- "calendar"
 } else {
   removeobs <- TRUE
   outcome <- args[[1]]
@@ -272,8 +272,6 @@ ggsave(filename=here("output", "models", outcome, timescale, glue("reportplr_cml
 
 
 
-
-
 # # # # # # # # # # # # # # # # # # # # # # # # # # # #
 # import spline models ----
 # # # # # # # # # # # # # # # # # # # # # # # # # # # #
@@ -342,6 +340,7 @@ effectsplr <-
     tidy_plr_ns %>% group_by(model_name, model) %>% summarise() %>% ungroup(), by="model"
   )
 
+write_rds(effectsplr, path=here("output", "models", outcome, timescale, glue("reportplr_effects_ns.rds")))
 
 plotplr <-
   ggplot(effectsplr)+
@@ -380,14 +379,7 @@ ggsave(filename=here("output", "models", outcome, timescale, glue("reportplr_eff
 
 ## risk-adjusted survival curves ----
 
-# no CIs / standard errors because difficult todo!
-
-#vcovCL0 <- read_rds(here("output", "models", outcome, timescale, glue("modelplr_vcov0_ns.csv")))
-#vcovCL1 <- read_rds(here("output", "models", outcome, timescale, glue("modelplr_vcov1_ns.csv")))
-#vcovCL2 <- read_rds(here("output", "models", outcome, timescale, glue("modelplr_vcov2_ns.csv")))
-#vcovCL3 <- read_rds(here("output", "models", outcome, timescale, glue("modelplr_vcov3_ns.csv")))
-
-#plr.predict(plrmod0, vcovCL0, mutate(data_plr, vax1_az=1L))
+# no CIs / standard errors yet because difficult to do!
 
 survival_az <- data_plr %>%
   mutate(vax1_az=1L) %>%
@@ -402,7 +394,6 @@ survival_az <- data_plr %>%
     outcome_prob2=predict(plrmod2, newdata=., type="response"),
     outcome_prob3=predict(plrmod3, newdata=., type="response"),
   )
-
 
 survival_pfizer <- data_plr %>%
   mutate(vax1_az=0L) %>%

### 2021-08-11 analysis/report/effectiveness_report.Rmd
colours

@@
     )+
-    scale_colour_brewer(type="qual", palette="Set2")+
+    scale_colour_brewer(type="qual", palette="Set1", na.value="grey")+
+    scale_fill_brewer(type="qual", palette="Set1", guide="none", na.value="grey")+
     labs(

### 2021-08-12 analysis/descriptive/km.R
add any admission date

@@ -110,6 +110,9 @@ data_tte <- data_cohort %>%
     tte_emergency = tte(vax1_date-1, emergency_date, censor_date, na.censor=FALSE),
     ind_emergency = censor_indicator(emergency_date, censor_date),
 
+    tte_admitted = tte(vax1_date-1, admitted_date, censor_date, na.censor=FALSE),
+    ind_admitted = censor_indicator(admitted_date, censor_date),
+
     tte_covidadmitted = tte(vax1_date-1, covidadmitted_date, censor_date, na.censor=FALSE),
     ind_covidadmitted = censor_indicator(covidadmitted_date, censor_date),
 
@@ -254,6 +257,7 @@ metadata_outcomes <- tribble(
   "test", "SARS-CoV-2 test",
   "postest", "Positive test",
   "emergency", "A&E attendance",
+  "admitted", "Unpanned admission",
   "covidadmitted", "COVID-19 admission",
   "covidcc", "COVID-19 critical care",
   "coviddeath", "COVID-19 death",

### 2021-08-12 analysis/descriptive/table_irr.R

@@ -78,6 +78,9 @@ data_tte <- data_cohort %>%
     tte_emergency = tte(vax1_date-1, emergency_date, censor_date, na.censor=TRUE),
     ind_emergency = censor_indicator(emergency_date, censor_date),
 
+    tte_admitted = tte(vax1_date-1, admitted_date, censor_date, na.censor=TRUE),
+    ind_admitted = censor_indicator(admitted_date, censor_date),
+
     tte_covidadmitted = tte(vax1_date-1, covidadmitted_date, censor_date, na.censor=TRUE),
     ind_covidadmitted = censor_indicator(covidadmitted_date, censor_date),
 
@@ -122,6 +125,7 @@ data_cox_split <- tmerge(
   test = event(tte_test),
   postest = event(tte_postest),
   emergency = event(tte_emergency),
+  admitted = event(tte_admitted),
   covidadmitted = event(tte_covidadmitted),
   covidcc = event(tte_covidcc),
   coviddeath = event(tte_coviddeath),
@@ -132,6 +136,7 @@ data_cox_split <- tmerge(
   status_test = tdc(tte_test),
   status_postest = tdc(tte_postest),
   status_emergency = tdc(tte_emergency),
+  status_admitted = tdc(tte_admitted),
   status_covidadmitted = tdc(tte_covidadmitted),
   status_covidcc = tdc(tte_covidcc),
   status_coviddeath = tdc(tte_coviddeath),
@@ -248,14 +253,16 @@ data_summary <- local({
   temp1 <- pt_summary(data_cox_split, "test")
   temp2 <- pt_summary(data_cox_split, "postest")
   temp3 <- pt_summary(data_cox_split, "emergency")
-  temp4 <- pt_summary(data_cox_split, "covidadmitted")
-  temp5 <- pt_summary(data_cox_split, "covidcc")
-  temp6 <- pt_summary(data_cox_split, "coviddeath")
-  temp7 <- pt_summary(data_cox_split, "noncoviddeath")
-  temp8 <- pt_summary(data_cox_split, "death")
+  temp4 <- pt_summary(data_cox_split, "admitted")
+  temp5 <- pt_summary(data_cox_split, "covidadmitted")
+  temp6 <- pt_summary(data_cox_split, "covidcc")
+  temp7 <- pt_summary(data_cox_split, "coviddeath")
+  temp8 <- pt_summary(data_cox_split, "noncoviddeath")
[…]

### 2021-08-12 analysis/process/design.R

@@ -21,6 +21,7 @@ metadata_outcomes <- tribble(
   "test", "covid_test_date", "SARS-CoV-2 test",
   "postest", "positive_test_date", "Positive SARS-CoV-2 test",
   "emergency", "emergency_date", "A&E attendance",
+  "admitted", "admitted_date", "Unplanned hospitalisation",
   "covidadmitted", "covidadmitted_date", "COVID-19 hospitalisation",
   "covidcc", "covidcc_date", "COVID-19 critical care",
   "coviddeath", "coviddeath_date", "COVID-19 death",

### 2021-08-12 create-project.R

@@ -324,6 +324,15 @@ actions_list <- splice(
   action_model("emergency", "calendar", "plr", 50000),
   action_report("emergency", "calendar", "plr"),
 
+  comment("### Unplanned Hospital admission"),
+  action_model("admitted", "timesincevax", "cox"),
+  action_report("admitted", "timesincevax", "cox"),
+  action_model("admitted", "calendar", "cox"),
+  action_report("admitted", "calendar", "cox"),
+  action_model("admitted", "timesincevax", "plr", 50000),
+  action_report("admitted", "timesincevax", "plr"),
+  action_model("admitted", "calendar", "plr", 50000),
+  action_report("admitted", "calendar", "plr"),
 
   comment("###  COVID-19 hospital admission"),
   action_model("covidadmitted", "timesincevax", "cox"),

### 2021-08-12 analysis/descriptive/km.R
change end of follow up date to max 14 weeks post-vaccine

@@ -86,7 +86,7 @@ data_tte <- data_cohort %>%
     # assume vaccination occurs at the start of the day, and all other events occur at the end of the day.
     # so use vax1_date - 1
 
-    censor_date = pmin(end_date, dereg_date, death_date, covid_vax_any_2_date, na.rm=TRUE),
+    censor_date = pmin(vax1_date - 1 + (7*14), end_date, dereg_date, death_date, covid_vax_any_2_date, na.rm=TRUE),
 
     # time to last follow up day
     tte_enddate = tte(vax1_date-1, end_date, end_date),

### 2021-08-12 analysis/descriptive/table_irr.R

@@ -61,7 +61,7 @@ data_tte <- data_cohort %>%
     vax1_type,
     end_date,
 
-    censor_date = pmin(end_date, dereg_date, death_date, covid_vax_any_2_date, na.rm=TRUE),
+    censor_date = pmin(vax1_date - 1 + (7*14), end_date, dereg_date, death_date, covid_vax_any_2_date, na.rm=TRUE),
 
     # time to last follow up day
     tte_enddate = tte(vax1_date-1, end_date, end_date),

### 2021-08-12 analysis/models/model_cox.R

@@ -75,7 +75,7 @@ data_tte <- data_cohort %>%
   mutate(
 
     outcome_date = .[[glue("{outcome_var}")]],
-    censor_date = pmin(end_date, dereg_date, death_date, covid_vax_any_2_date, na.rm=TRUE),
+    censor_date = pmin(vax1_date - 1 + (7*14), end_date, dereg_date, death_date, covid_vax_any_2_date, na.rm=TRUE),
 
     # assume vaccination occurs at the start of the day, and all other events occur at the end of the day.
     tte_censor = tte(vax1_date-1, censor_date, censor_date, na.censor=TRUE),

### 2021-08-12 analysis/models/model_plr.R

@@ -78,7 +78,7 @@ data_cohort <- read_rds(here("output", "data", "data_cohort.rds"))
 data_tte <- data_cohort %>%
   mutate(
     outcome_date = .[[glue("{outcome_var}")]],
-    censor_date = pmin(end_date, dereg_date, death_date, covid_vax_any_2_date, na.rm=TRUE),
+    censor_date = pmin(vax1_date - 1 + (7*14), end_date, dereg_date, death_date, covid_vax_any_2_date, na.rm=TRUE),
 
     tte_censor = tte(vax1_date-1, censor_date, censor_date, na.censor=TRUE),
     ind_censor = censor_indicator(censor_date, censor_date),

### 2021-08-12 analysis/process/data_selection.R

@@ -39,7 +39,6 @@ data_criteria <- data_processed %>%
     vax1_afterstartdate = vax1_date >= start_date_az,
     vax1_azpfizer = vax1_type %in% c("az", "pfizer"),
 
-
     include = (
       has_age & has_sex & has_imd & has_ethnicity & has_region &
         #has_follow_up_previous_year &

### 2021-08-12 analysis/report/effectiveness_report.Rmd
fix dual timescales by only including date of vaccination as a baseline covariate in the vaccination timescale models.
calendar-time PLR models still probably wrong!

@@
 * On the *treatment time-scale*, time zero is the date of vaccination.
-Calendar-time is included as an additional covariate, in weekly intervals, using restricted cubic spline with knots at the first and second tertile, plus boundary knots.
+Date of vaccination is included as an additional baseline covariate using a restricted cubic spline with knots at the first and second tertile, plus boundary knots.
 Here, follow-up time is stratified into weekly intervals with the vaccine-specific effect estimated within each time-strata.

### 2021-08-12 analysis/models/model_cox.R

@@ -166,44 +166,44 @@ if(timescale=="calendar"){
 # - post-vax follow-up is already overlapping, so can use az/pfizer : weekly strata
 # - need to adjust for calendar time
 if(timescale=="timesincevax"){
-  # one row per patient per follow-up calendar day
-  calendar_time <- data_cox %>%
-    select(patient_id, vax1_day, tte_stop) %>%
-    mutate(
-      calendar_day = map2(vax1_day, tte_stop, ~.x:(.y+.x))
-    ) %>%
-    unnest(c(calendar_day)) %>%
-    mutate(
-      #calendar_week = paste0("week ", str_pad(floor(calendar_day/7), 2, pad = "0")),
-      calendar_week = floor(calendar_day/7),
-      treatment_day = calendar_day - vax1_day,
-      treatment_week = floor(treatment_day/7)
-    )
-
-  # one row per patient per follow-up calendar week
-  calendar_week <- calendar_time %>%
-    group_by(patient_id, calendar_week) %>%
-    filter(first(calendar_day)==calendar_day) %>%
-    ungroup()
-
-  # one row per patient per follow-up post-vax week
-  treatment_week <- calendar_time %>%
-    group_by(patient_id, treatment_week) %>%
-    filter(first(treatment_day)==treatment_day) %>%
-    ungroup()
-
-  data_cox <- data_cox %>%
-    tmerge(
-      data1 = .,
-      data2 = treatment_week,
-      id = patient_id,
-      calendar_day = tdc(treatment_day, calendar_day)
-    )
+  # # one row per patient per follow-up calendar day
+  # calendar_time <- data_cox %>%
+  #   select(patient_id, vax1_day, tte_stop) %>%
+  #   mutate(
+  #     calendar_day = map2(vax1_day, tte_stop, ~.x:(.y+.x))
+  #   ) %>%
+  #   unnest(c(calendar_day)) %>%
+  #   mutate(
+  #     #calendar_week = paste0("week ", str_pad(floor(calendar_day/7), 2, pad = "0")),
+  #     calendar_week = floor(calendar_day/7),
+  #     treatment_day = calendar_day - vax1_day,
+  #     treatment_week = floor(treatment_day/7)
+  #   )
+  #
+  # # one row per patient per follow-up calendar week
+  # calendar_week <- calendar_time %>%
+  #   group_by(patient_id, calendar_week) %>%
+  #   filter(first(calendar_day)==calendar_day) %>%
[…]

### 2021-08-12 analysis/models/model_plr.R

@@ -162,10 +162,14 @@ if(timescale=="calendar"){
 }
 if(timescale=="timesincevax"){
   formula_timescale <- . ~ . + ns(log(tstop), 3) # spline for timescale only
-  formula_spacetime <- . ~ . + ns(tstop_calendar, 3)*region # spline for space-time adjustments
+  formula_spacetime <- . ~ . + ns(vax1_day, 3)*region # spline for space-time adjustments
 }
 
-
+## NOTE
+# calendar-time PLR models are still probably wrong!
+# as time-varying splines for both vaccination time and calendar time are included.
+# This wrongly allows for estimation of the probability of, eg, the risk of death
+# in february for someone vaccinated on 1 march
 
 ### piecewise formulae ----
 ### estimand

### 2021-08-12 analysis/models/report_plr.R

@@ -67,10 +67,15 @@ if(timescale=="calendar"){
   formula_spacetime <- . ~ . + ns(tstop_calendar, 3)*region # spline for space-time adjustments
 }
 if(timescale=="timesincevax"){
-  formula_timescale <- . ~ . +  ns(log(tstop), 3) # spline for timescale only
-  formula_spacetime <- . ~ . + ns(tstop_calendar, 3)*region # spline for space-time adjustments
+  formula_timescale <- . ~ . + ns(log(tstop), 3) # spline for timescale only
+  formula_spacetime <- . ~ . + ns(vax1_day, 3)*region # spline for space-time adjustments
 }
 
+## NOTE
+# calendar-time PLR models are still probably wrong!
+# as time-varying splines for both vaccination time and calendar time are included.
+# This wrongly allows for estimation of the probability of, eg, the risk of death
+# in february for someone vaccinated on 1 march
 
 ### piecewise formulae ----
 ### estimand
@@ -91,6 +96,7 @@ formula0_ns <- formula_vaxonly_ns
 formula1_ns <- formula_vaxonly_ns %>% update(formula_spacetime)
 formula2_ns <- formula_vaxonly_ns %>% update(formula_spacetime) %>% update(formula_demog)
 formula3_ns <- formula_vaxonly_ns %>% update(formula_spacetime) %>% update(formula_demog) %>% update (formula_comorbs)
+
 
 # Import processed data ----
 

### 2021-08-12 analysis/process/design.R
add extra follow-up week to cuts

@@ -82,7 +82,7 @@ formula_all_rhsvars <- update(1 ~ 1, formula_exposure) %>%
   update(formula_secular) %>%
   update(formula_secular_region)
 
-postvaxcuts <- c(0, 7, 14, 21, 28, 35)
+postvaxcuts <- c(0, 7, 14, 21, 28, 35, 42)
 
 list_formula <- lst(
   formula_exposure,

### 2021-08-17 analysis/lib/utility_functions.R
add se for response

@@ -258,19 +258,36 @@ tidy_custom.glm  <- function(model, conf.int=TRUE, conf.level=0.95, exponentiate
   output
 }
 
-plr.predict <- function(model, clcov, newdata){
+plr.predict <- function(model, clcov, newdata, type=c("link", "response")){
+
   # from https://stackoverflow.com/questions/3790116/using-clustered-covariance-matrix-in-predict-lm
+
   if(missing(newdata)){ newdata <- model$model }
-  tt <- terms(model)
+  tt <- terms(model) # this helpfully grabs the correct spline basis from the model, rather than recalculating based on `newdata`
   Terms <- delete.response(tt)
   m.mat <- model.matrix(Terms, data=newdata)
   m.coef <- model$coef
-  fit <- as.vector(m.mat %*% model$coef)
-  se.fit <- sqrt( rowSums(m.mat * (m.mat %*% clcov)) )
-  return(list(fit=fit, se.fit=se.fit))
+
+  nu <- as.vector(m.mat %*% m.coef) # n x 1
+
+  if(type=="link"){
+    fit <- nu
+    fit.se <- sqrt( rowSums(m.mat * (m.mat %*% clcov)) )
+  }
+
+  if(type=="response"){
+
+    logit2nu <- (exp(nu)/((1+exp(nu))^2)) # n x 1
+    deriv_nu <- t(m.mat) * c(logit2nu)  # k x n
+
+    stopifnot("term indices are not aligned" = rownames(clcov)==names(deriv_nu))
+
+    fit <- plogis(nu)
+    fit.se <- sqrt(rowSums((t(deriv_nu) %*% vcov) * t(deriv_nu)))
+  }
+
+  return(list(fit=fit, fit.se=fit.se))
 }
-
-
 
 
 # functions for IRR confidence intervals ----

### 2021-08-17 analysis/models/model_plr.R
weights implicitly from model data

@@ -249,7 +249,7 @@ plr_process <- function(plrmod, number, cluster, splinetype){
 plrmod0 <- parglm(
   formula = formula0_pw,
   data = data_plr,
-  weights = data_plr$sample_weights,
+  weights = sample_weights,
   family = binomial,
   control = parglmparams,
   na.action = "na.fail",
@@ -261,11 +261,12 @@ summary0 <- plr_process(
 )
 if(removeobs){remove(plrmod0)}
 
+
 ## model 1 - minimally adjusted vaccination effect model, stratification by region only
 plrmod1 <- parglm(
   formula = formula1_pw,
   data = data_plr,
-  weights = data_plr$sample_weights,
+  weights = sample_weights,
   family = binomial,
   control = parglmparams,
   na.action = "na.fail",
@@ -281,7 +282,7 @@ if(removeobs){remove(plrmod1)}
 plrmod2 <- parglm(
   formula = formula2_pw,
   data = data_plr,
-  weights = data_plr$sample_weights,
+  weights = sample_weights,
   family = binomial,
   control = parglmparams,
   na.action = "na.fail",
@@ -297,7 +298,7 @@ if(removeobs){remove(plrmod2)}
 plrmod3 <- parglm(
   formula = formula3_pw,
   data = data_plr,
-  weights = data_plr$sample_weights,
+  weights = sample_weights,
   family = binomial,
   control = parglmparams,
   na.action = "na.fail",
@@ -332,7 +333,7 @@ write_rds(model_tidy, here::here("output", "models", outcome, timescale, glue("m
 plrmod0 <- parglm(
   formula = formula0_ns,
   data = data_plr,
-  weights = data_plr$sample_weights,
+  weights = sample_weights,
   family = binomial,
   control = parglmparams,
   na.action = "na.fail",
@@ -347,7 +348,7 @@ if(removeobs){remove(plrmod0)}
 plrmod1 <- parglm(
   formula = formula1_ns,
   data = data_plr,
-  weights = data_plr$sample_weights,
+  weights = sample_weights,
   family = binomial,
   control = parglmparams,
   na.action = "na.fail",
@@ -362,7 +363,7 @@ if(removeobs){remove(plrmod1)}
 plrmod2 <- parglm(
   formula = formula2_ns,
   data = data_plr,
-  weights = data_plr$sample_weights,
+  weights = sample_weights,
   family = binomial,
   control = parglmparams,
[…]

### 2021-08-17 analysis/models/report_plr.R
add confidence intervals for adjusted survival curves

@@ -385,81 +385,149 @@ ggsave(filename=here("output", "models", outcome, timescale, glue("reportplr_eff
 
 ## risk-adjusted survival curves ----
 
-# no CIs / standard errors yet because difficult to do!
+cmlinc_variance <- function(model, vcov, newdata, id, time, weights){
 
-survival_az <- data_plr %>%
-  mutate(vax1_az=1L) %>%
-  transmute(
-    patient_id,
-    tstop,
-    tstop_calendar,
-    vax1_az,
-    sample_weights,
-    outcome_prob0=predict(plrmod0, newdata=., type="response"),
-    outcome_prob1=predict(plrmod1, newdata=., type="response"),
-    outcome_prob2=predict(plrmod2, newdata=., type="response"),
-    outcome_prob3=predict(plrmod3, newdata=., type="response"),
-  )
+  # calculate variance of adjusted survival / adjusted cumulative incidence
+  # needs model object, cluster-vcov, input data, model weights, and indices for patient and time
 
-survival_pfizer <- data_plr %>%
-  mutate(vax1_az=0L) %>%
-  transmute(
-    patient_id,
-    tstop,
-    tstop_calendar,
-    vax1_az,
-    sample_weights,
-    outcome_prob0=predict(plrmod0, newdata=., type="response"),
-    outcome_prob1=predict(plrmod1, newdata=., type="response"),
-    outcome_prob2=predict(plrmod2, newdata=., type="response"),
-    outcome_prob3=predict(plrmod3, newdata=., type="response"),
+  if(missing(newdata)){ newdata <- model$model }
+  tt <- terms(model) # this helpfully grabs the correct spline basis from the model, rather than recalculating based on `newdata`
+  Terms <- delete.response(tt)
+  m.mat <- model.matrix(Terms, data=newdata)
+  m.coef <- model$coef
+
+  N <- nrow(m.mat)
+  K <- length(m.coef)
+
+  # log-odds, nu_t, at time t
+  nu <- m.coef %*% t(m.mat) # t_i x 1
+  # part of partial derivative
+  pdc <- (exp(nu)/((1+exp(nu))^2)) # t_i x 1
+  # summand for partial derivative of P_t(theta_t | X_t), for each time t and term k
+
[…]

### 2021-08-17 analysis/report/effectiveness_report.Rmd
add any-admitted to report

@@
 cat("  \n\n")
+print(glue("### Any unplanned hospital admission"), "  \n")
+print(glue("#### Unadjusted"), "  \n")
+print_plots_km("admitted")
+print(glue("#### Cox"), "  \n")
+print_plots_effects("admitted", "cox")
+print(glue("#### PLR, peicewise"), "  \n")
+print_plots_effects("admitted", "plr")
+print(glue("#### PLR, spline"), "  \n")
+print_plots_spline("admitted")
+print_plots_cmlinc("admitted")
+
+cat("  \n\n")
 print(glue("### COVID-19 hospital admission"), "  \n")

### 2021-08-18 create-project.R
add timescales comaprison action

@@ -117,6 +117,13 @@ defaults_list <- list(
 ## actions ----
 actions_list <- splice(
 
+  comment("# # # # # # # # # # # # # # # # # # #",
+          "DO NOT EDIT project.yaml DIRECTLY",
+          "This file is created by create-project.R",
+          "Edit and run create-project.R to update the project.yaml",
+          "# # # # # # # # # # # # # # # # # # #"
+          ),
+
 
   comment("# # # # # # # # # # # # # # # # # # #", "Temp second dose info", "# # # # # # # # # # # # # # # # # # #"),
   action(
@@ -297,6 +304,9 @@ actions_list <- splice(
     name = "descr_vaxdate",
     run = "r:latest analysis/descriptive/vax_date.R",
     needs = list("design", "data_selection"),
+    highly_sensitive = list(
+      rds = "output/descriptive/vaxdate/*.rds"
+    ),
     moderately_sensitive = list(
       png = "output/descriptive/vaxdate/*.png"
     )
@@ -372,7 +382,7 @@ actions_list <- splice(
   # action_model("covidcc", "calendar", "plr", 50000),
   # action_report("covidcc", "calendar", "plr"),
 
-  comment("# # # # # # # # # # # # # # # # # # #", "Report", "# # # # # # # # # # # # # # # # # # #"),
+  comment("# # # # # # # # # # # # # # # # # # #", "Reports", "# # # # # # # # # # # # # # # # # # #"),
 
   action(
     name = "rmd_report",
@@ -383,15 +393,11 @@ actions_list <- splice(
     needs = splice(
       "design", "data_selection",
       "descr_table1", "descr_irr",
-      "descr_km",
+      "descr_km", "descr_vaxdate",
       as.list(
-        glue_data(
-             .x = expand_grid(
-               outcome = c("test", "postest", "emergency", "admitted", "covidadmitted"),
-               modeltype = c("cox", "plr"),
-               timescale = c("timesincevax", "calendar")
-             ),
-             "report_{outcome}_{timescale}_{modeltype}"
+        glue(
+             outcome = c("test", "postest", "emergency", "admitted", "covidadmitted"),
+             "report_{outcome}_timesincevax_plr"
         )
       )
     ),
@@ -403,29 +409,30 @@ actions_list <- splice(
   ),
 
   action(
-    name = "rmd_report_vaxtime",
+    name = "rmd_report_timescales",
     run = glue(
       "r:latest -e {q}",
[…]

### 2021-08-18 analysis/descriptive/vax_date.R
better plots

@@ -42,70 +42,45 @@ cumulvax <- data_cohort %>%
   summarise(
     n=n()
   ) %>%
+  ungroup() %>%
+  add_row(
+    vax1_type_descr = unique(.$vax1_type_descr),
+    vax1_date = min(.$vax1_date) - 1,
+    n=0,
+    .before=1
+  ) %>%
   group_by(vax1_type_descr) %>%
-  arrange(vax1_date) %>%
   mutate(
     cumuln = cumsum(n)
-  )
+  ) %>%
+  arrange(vax1_type_descr, vax1_date)
 
 
 plot_stack <-
   ggplot(cumulvax)+
   geom_area(
     aes(
-      x=vax1_date, y=cumuln,
+      x=vax1_date+1, y=cumuln,
       group=vax1_type_descr,
       fill=vax1_type_descr
     ),
     alpha=0.5
   )+
   scale_x_date(
-    breaks = seq(min(cumulvax$vax1_date),max(cumulvax$vax1_date)+1,by=14),
-    limits = c(lubridate::floor_date((min(cumulvax$vax1_date)), "1 month"), NA),
+    breaks = seq(min(cumulvax$vax1_date),max(cumulvax$vax1_date)+1,by=14)+1,
+    limits = c(lubridate::floor_date(min(cumulvax$vax1_date), "1 month"), NA),
     labels = scales::date_format("%d/%m"),
     expand = expansion(0),
     sec.axis = sec_axis(
       trans = ~as.Date(.),
       breaks=as.Date(seq(floor_date(min(cumulvax$vax1_date), "month"), ceiling_date(max(cumulvax$vax1_date), "month"),by="month")),
-      labels = scales::date_format("%b %y")
+      labels = scales::date_format("%B %y")
     )
   )+
   scale_fill_brewer(type="qual", palette="Set1")+
   labs(
     x="Date",
-    y="Vaccination status",
-    colour=NULL,
-    fill=NULL,
-    alpha=NULL
-  ) +
-  theme_minimal()+
-  theme(legend.position = "bottom")
-
-
-plot_step <-
-  ggplot(cumulvax)+
-  geom_step(
-    aes(
-      x=vax1_date, y=cumuln,
-      group=vax1_type_descr,
-      colour=vax1_type_descr
-    )
-  )+
-  scale_x_date(
-    breaks = seq(min(cumulvax$vax1_date),max(cumulvax$vax1_date)+1,by=14),
[…]

### 2021-08-18 analysis/report/effectiveness_report.Rmd
add timescales comparison; update risk-adjusted survival plots

--- 2021-08-12 analysis/report/effectiveness_report.Rmd
@@
 
-Note -- many of those flagged as HCWs do not have a vaccination record, which shouldn't be the case if the question was asked as part of the vaccination process.
+Note -- many of those flagged as HCWs do not have a vaccination record, which shouldn't be the case if the question was asked as part of the vaccination process. This needs further investigation.
 
@@
 Patient characteristics used for adjustment include: age, sex, deprivation, ethnicity, NHS region, clinically “at risk” (but not clinically extremely vulnerable) as per national prioritisation guidelines, asthma, number of previous SARS-CoV-2 tests (via SGSS), rurality, evidence of prior covid infection (positive test or COVID-19 hospitalisation), number of recorded comorbidities, severe mental illness.
-All variables are ascertained as at the time of vaccination.
+All characteristics are ascertained as at the time of vaccination.
 
-NHS regions are used as a stratification variable to account for geographical variation in event rates for the outcomes of interest (primarily driven by changes in local infection rates).
+NHS regions are used as a stratification variable to account for geographical variation in event rates for the outcomes of interest, for example due to changes in local infection rates.
 
 ### Time-dependent Cox models
-Time-dependent Cox models are used to estimate the time-varying effects for each vaccine type. Two time-scales are used for triangulation:
-
-* On the *calendar time-scale*, time zero is `r format(as.Date(gbl_vars$start_date_az), "%d %B %Y")`, and people only contribute follow-up time once they have been vaccinated (i.e., delayed study entry).
-Vaccine type, time since vaccination in weekly intervals, and the interaction between these two terms are included to estimate time-varying vaccine effects for each vaccine type.
-In `R`'s `coxph` function, this is modelled using `vaccine_type*week_since_first_dose`
-
-* On the *treatment time-scale*, time zero is the date of vaccination.
+Time-dependent Cox models are used to estimate the time-varying effects for each vaccine type.
+Time zero is the date of vaccination, i.e., using a treatment time-scale.
+Outcome events occurring on the same day as vaccination are included in follow-up.
 Date of vaccination is included as an additional baseline covariate using a restricted cubic spline with knots at the first and second tertile, plus boundary knots.
@@
 
-We emulate the Cox models described above using pooled logistic regression (PLR), with the outcome risk estimated at daily intervals.
-This serves two purposes.
-Firstly, a continuous-time estimate of comparative effectiveness, as opposed to the piecewise-linear approximation above, can be obtained using flexible parametric splines.
-Secondly, the absolute cumulative risk of each outcome of interest for each vaccine type, marginalised over the population, can be obtained using the parametric g-formula.
-These models include a restricted cubic spline with 3 degrees of freedom on the timescale of interest, interacted with vaccine type.
-Cumulative risk is then calculated as the average risk for each day of follow-up predicted by the PLR model, under the (complimentary counterfactual) assumptions that every patient received the BNT162b2 vaccine or that every patient received the ChAdOx1 vaccine.
+We emulate the Cox model described above using pooled logistic regression (PLR), with the outcome risk estimated at daily intervals.
+These models include a restricted cubic spline with 3 degrees of freedom on the log of the timescale (time since vaccination), interacted with vaccine type.
+The PLR models serve two purposes.
+Firstly, a continuous-time estimate of comparative effectiveness, as opposed to the piecewise-linear approximation above, is obtained.
+Secondly, the risk-adjusted survival rates for each vaccine type are obtained using the parametric g-formula.
+This is the average risk for each day of follow-up predicted by the PLR model, under the (complimentary counterfactual) assumption that every patient received the BNT162b2 vaccine or that every patient received the ChAdOx1 vaccine.
+
+Confidence intervals for the risk-adjusted survival rates are calculated using the delta method (i.e, a first-order Taylor series approximation of the variance of the cumulative incidence).
 
 The person-time dataset needed to fit the PLR model is large and leads to infeasible RAM requirements / computation time.
-To deal with this, a sampling strategy is used such that all those who experienced an outcome and a random sample of 50000 who did not experience the ouotcome are selected for the models.
+To deal with this, a sampling strategy is used such that all those who experienced an outcome and a random sample of 50000 who did not experience the outcome are selected for the models.
 The models are weighted to recover the characteristics of the complete dataset.
-This weighting is also applied to the average cumulative risk.
+This weighting is also applied to the risk-adjusted survival curves.
 
@@
 
-Note all counts below 6 are redacted and Kaplan-Meier survival estimates are rounded for disclosure control.
+Note all counts below 6 are redacted and survival estimates are rounded for disclosure control.
 
@@
 
+## Vaccination dates
+
+```{r, vaxdate, echo=FALSE, message=FALSE, warning=FALSE, out.width='80%', results='asis'}
+  vaxdate_plot <- read_rds(here("output", "descriptive", "vaxdate", "plot_vaxdate_stack.rds"))
+  print(vaxdate_plot)
+  cat("  \n\n")
+```
+
 ## Event rates
@@
 
-* Unadjusted Kaplan-Meier curves
-* *ChAdOx1* versus *BNT162b2* piecewise-linear hazard ratios estimated using the Cox models
-* *ChAdOx1* versus *BNT162b2* continuous hazard ratios estimated using the PLR models
-* Marginalised cumulative risk for *ChAdOx1* and *BNT162b2* estimated using the PLR models
+* *ChAdOx1* versus *BNT162b2* hazard ratio splines
+* Risk-adjusted survival curves for *ChAdOx1* and *BNT162b2*
 
-```{r, curves, echo=FALSE, message=FALSE, warning=FALSE, out.width='80%', results='asis'}
+```{r, curves, echo=FALSE, message=FALSE, warning=FALSE, out.width='90%', fig.height=6, results='asis'}
 
-print_plots_km <- function(outcome){
+plots_spline <- function(outcome){
 
-  km_image <- here("output", "descriptive", "km", glue("plot_survival_ci_{outcome}.png"))
+  timesincevax_effects <- read_rds(here("output", "models", outcome, "timesincevax", glue("reportplr_effects_ns.rds"))) %>%
+    mutate(
+      timescale = "Vaccinated time",
+      model_name = fct_inorder(str_wrap(model_name, 15))
+    )
 
-  km_plot <- read_rds(here("output", "descriptive", "km", glue("plot_survival_ci_{outcome}.rds")))
-
-  print(km_plot)
-  cat("  \n\n")
-
-}
-
-print_plots_effects <- function(outcome, modeltype){
-
-  suffix <- if(modeltype=="plr"){
-    "_pw"
-  } else{
-    ""
-  }
-
-  calendar_effects <- read_rds(here("output", "models", outcome, "calendar", glue("report{modeltype}_effects{suffix}.rds"))) %>%
-    mutate(timescale = "Calendar time")
-  timesincevax_effects <- read_rds(here("output", "models", outcome, "timesincevax", glue("report{modeltype}_effects{suffix}.rds"))) %>%
-    mutate(timescale = "Vaccinated time")
-
-  tidy_effects <- bind_rows(calendar_effects, timesincevax_effects)
-
-  plot_effects <-
-    ggplot(data = tidy_effects) +
-    geom_point(aes(y=exp(estimate), x=term_midpoint, colour=model_name), position = position_dodge(width = 1.8))+
-    geom_linerange(aes(ymin=exp(conf.low), ymax=exp(conf.high), x=term_midpoint, colour=model_name), position = position_dodge(width = 1.8))+
+  plot_spline <-
+    ggplot(timesincevax_effects)+
     geom_hline(aes(yintercept=1), colour='grey')+
-    facet_grid(cols=vars(timescale))+
-    scale_x_continuous(breaks=unique(tidy_effects$term_left), limits=c(min(tidy_effects$term_left), max(tidy_effects$term_right)+1), expand = c(0, 0))+
+    geom_line(aes(x=tstop-1, y=hr))+
+    geom_ribbon(aes(x=tstop-1, ymin=hr.ll, ymax=hr.ul), alpha=0.2, colour="transparent")+
+    facet_grid(cols=vars(model_name))+
+    scale_x_continuous(
+      breaks = seq(0,7*52,by=14),
+      expand = expansion(0)
+    )+
     scale_y_log10(
-    breaks=c(0.1, 0.125, 0.25, 0.33, 0.5, 0.75, 1, 1.33, 2, 3, 4, 8,10),
-    sec.axis = dup_axis(name="favours\n <- BNT162b2 / ChAdOx1 ->", breaks = NULL)
+      breaks=c(0.125, 0.167, 0.25, 0.33, 0.5, 0.67, 1, 1.5, 2, 3, 4, 6, 8),
+      sec.axis = dup_axis(name="Lower hazard after \n <- BNT162b2 / ChAdOx1 ->", breaks = NULL)
     )+
-    scale_colour_brewer(type="qual", palette="Set2", guide=guide_legend(ncol=2))+
     labs(
-      y="Hazard ratio",#\nChAdOx1 / BNT162b2",
-      x="Days since vaccination",
-      colour=NULL
-     ) +
+      x = "Days since first dose",
+      y = "Hazard ratio"
+    )+
     theme_bw()+
@@
       strip.background = element_blank(),
-      strip.placement = "outside",
-      strip.text.y.left = element_text(angle = 0),
 
@@
       plot.caption.position = "plot",
-      plot.caption = element_text(hjust = 0, face= "italic"),
+      plot.caption = element_text(hjust = 0, face= "italic")
 
-      legend.position = "bottom"
      ) +
    NULL
-
-  print(plot_effects)
-  cat("  \n\n")
+  plot_spline
+  #print(plot_spline)
+  #cat("  \n\n")
 }
 
-print_plots_spline <- function(outcome){
+plots_adjsurv <- function(outcome){
 
-  calendar_effects <- read_rds(here("output", "models", outcome, "calendar", glue("reportplr_effects_ns.rds"))) %>%
-    mutate(timescale = "Calendar time")
-  timesincevax_effects <- read_rds(here("output", "models", outcome, "timesincevax", glue("reportplr_effects_ns.rds"))) %>%
+  timesincevax_curve <- read_rds(here("output", "models", outcome, "timesincevax", glue("reportplr_adjustedsurvival_ns.rds"))) %>%
     mutate(timescale = "Vaccinated time")
 
-  spline <- bind_rows(calendar_effects, timesincevax_effects)
-
-  plot_spline <-
-    ggplot(spline)+
-    geom_hline(aes(yintercept=1), colour='grey')+
-    geom_line(aes(x=tstop-1, y=hr, colour=model_name))+
-    geom_ribbon(aes(x=tstop-1, ymin=hr.ll, ymax=hr.ul, fill=model_name), alpha=0.2, colour="transparent")+
-    facet_grid(cols=vars(timescale))+
-    scale_x_continuous(
-      breaks = seq(0,7*52,by=14),
-      expand = expansion(0)
-    )+
-    scale_y_log10(
-      breaks=c(1/8, 1/6, 0.25, 0.33, 0.5, 0.67, 1, 1.5, 2, 3, 4, 6, 8),
-      sec.axis = dup_axis(name="favours\n <- BNT162b2 / ChAdOx1 ->", breaks = NULL)
-    )+
-    scale_colour_brewer(type="qual", palette="Set2", guide=guide_legend(ncol=1))+
-    scale_fill_brewer(type="qual", palette="Set2", guide="none")+
-    labs(
-      x = "Days since first dose",
-      y = "Hazard ratio",
-      colour = NULL, fill=NULL
-    )+
-    theme_bw()+
-    theme(
-      panel.border = element_blank(),
-      axis.line.y = element_line(colour = "black"),
-
-      panel.grid.minor.y = element_blank(),
-      strip.background = element_blank(),
-      strip.placement = "outside",
-      strip.text.y.left = element_text(angle = 0),
-
-      panel.spacing = unit(0.8, "lines"),
-
-      plot.title = element_text(hjust = 0),
-      plot.title.position = "plot",
-      plot.caption.position = "plot",
-      plot.caption = element_text(hjust = 0, face= "italic"),
-
-      legend.position = "bottom"
-     ) +
-   NULL
-
-  print(plot_spline)
-  cat("  \n\n")
-}
-
-print_plots_cmlinc <- function(outcome){
-
-  calendar_curve <- read_rds(here("output", "models", outcome, "calendar", glue("reportplr_cmlinccurves_ns.rds"))) %>%
-    mutate(timescale = "Calendar time")
-  timesincevax_curve <- read_rds(here("output", "models", outcome, "timesincevax", glue("reportplr_cmlinccurves_ns.rds"))) %>%
-    mutate(timescale = "Vaccinated time")
-
-  curves <- bind_rows(calendar_curve, timesincevax_curve)
-
-  plot_cmlinc <-
-    ggplot(curves %>% filter(model=="3"))+
-    geom_step(aes(x=tstop, y=1-survival, group=vax1_az_descr, colour=vax1_az_descr))+
-    facet_grid(cols=vars(timescale))+
+  plot_adjsurv <-
+    ggplot(timesincevax_curve)+
+    geom_step(aes(x=tstop, y=survival, group=vax1_az_descr, colour=vax1_az_descr))+
+    geom_rect(aes(xmin=lag(tstop, 1, 0), xmax=tstop, ymin=survival.ll, ymax=survival.ul, group=vax1_az_descr, fill=vax1_az_descr), alpha=0.1)+
+    facet_grid(cols=vars(model_name))+
     scale_x_continuous(
@@
       x="Days since first dose",
-      y="Cumulative risk",
+      y="Adjusted survival",
       colour=NULL
@@
     theme(
-      legend.position=c(0.05,.95),
-      legend.justification = c(0,1),
+      legend.position="bottom",
+      legend.direction = "horizontal",
       axis.text.x.top=element_text(hjust=0),
@@
       strip.background = element_blank(),
-      strip.placement = "outside",
-      strip.text.y.left = element_text(angle = 0),
+      strip.text = element_blank(),
 
@@
     NULL
+  plot_adjsurv
 
-  print(plot_cmlinc)
-  cat("  \n\n")
+  #print(plot_adjsurv)
+  #cat("  \n\n")
 }
@@
 print(glue("### SARS-CoV-2 test"), "  \n")
-print(glue("#### Unadjusted"), "  \n")
-print_plots_km("test")
-print(glue("#### Cox"), "  \n")
-print_plots_effects("test", "cox")
-print(glue("#### PLR"), "  \n")
-print_plots_effects("test", "plr")
-print_plots_spline("test")
-print_plots_cmlinc("test")
+wrap_plots(plots_spline("test"), plots_adjsurv("test"), nrow=2)
 
@@
 print(glue("### SARS-CoV-2 positive test"), "  \n")
-print(glue("#### Unadjusted"), "  \n")
-print_plots_km("postest")
-print(glue("#### Cox"), "  \n")
-print_plots_effects("postest", "cox")
-print(glue("#### PLR, peicewise"), "  \n")
-print_plots_effects("postest", "plr")
-print(glue("#### PLR, spline"), "  \n")
-print_plots_spline("postest")
-print_plots_cmlinc("postest")
+wrap_plots(plots_spline("postest"), plots_adjsurv("postest"), nrow=2)
 
@@
 print(glue("### A&E attendance"), "  \n")
-print(glue("#### Unadjusted"), "  \n")
-print_plots_km("emergency")
-print(glue("#### Cox"), "  \n")
-print_plots_effects("emergency", "cox")
-print(glue("#### PLR, peicewise"), "  \n")
-print_plots_effects("emergency", "plr")
-print(glue("#### PLR, spline"), "  \n")
-print_plots_spline("emergency")
-print_plots_cmlinc("emergency")
+wrap_plots(plots_spline("emergency"), plots_adjsurv("emergency"), nrow=2)
+
+cat("  \n\n")
+print(glue("### Any unplanned hospital admission"), "  \n")
+wrap_plots(plots_spline("admitted"), plots_adjsurv("admitted"), nrow=2)
 
@@
 print(glue("### COVID-19 hospital admission"), "  \n")
-print(glue("#### Unadjusted"), "  \n")
-print_plots_km("covidadmitted")
-print(glue("#### Cox"), "  \n")
-print_plots_effects("covidadmitted", "cox")
-print(glue("#### PLR, peicewise"), "  \n")
-print_plots_effects("covidadmitted", "plr")
-print(glue("#### PLR, spline"), "  \n")
-print_plots_spline("covidadmitted")
-print_plots_cmlinc("covidadmitted")
+wrap_plots(plots_spline("covidadmitted"), plots_adjsurv("covidadmitted"), nrow=2)
 

### 2021-08-18 analysis/report/effectiveness_report_comparetimescales.Rmd
add timescales comparison; update risk-adjusted survival plots

---
title: "Comparative vaccine effectiveness (ChAdOx1 versus BNT162b2) in Health and Social Care workers"
output:
  github_document:
    keep_html: TRUE
  #  code_folding: hide
    #keep_md: yes
  #pdf_document: default
  # md_document:
  #   variant: gfm
  #html_document
---

```{r setup, include=FALSE}

library('tidyverse')
library('here')
library('glue')
library('lubridate')
library('gt')
library('patchwork')

output_dir <- here("output")

fs::dir_create(output_dir, "report", "figures")

knitr::opts_chunk$set(
  echo = TRUE,
  fig.path = paste0(fs::path(output_dir, "report", "figures_timescales"), "/")
)

gbl_vars <- jsonlite::fromJSON(
  txt=here("analysis", "global-variables.json")
)

```

# Introduction

This study compares the effectiveness of the *BNT162b2* mRNA (Pfizer-BioNTech) and *ChAdOx1* (Oxford-AstraZeneca) vaccines amongst health and social care workers (HCWs). 
This group was chosen as they were prioritised for vaccination due to high occupational exposure and during the period where both vaccines were available (`r format(as.Date(gbl_vars$start_date_az), "%d %B %Y")` onwards) it is expected that the actual vaccine brand received will not be strongly influenced by other determinants of COVID-19-related outcomes (i.e., approximately random treatment allocation and therefore reduced confounding).

The code and data for this report can be found at the OpenSafely [comparative-ve-research GitHub repository](https://github.com/opensafely/comparative-ve-research).

# Methods

## Study population

People meeting the following criteria are included:

* HCWs who have received at least one dose of BNT162b2 or ChAdOx1. 
* Registered at a GP practice using TPP's SystmOne clinical information system on the day of vaccination.
* Aged 18-65.
* Vaccinated on or after `r format(as.Date(gbl_vars$start_date_az), "%d %B %Y")`, when both vaccine brands were being administered. 
* Not Clinically Extremely Vulnerable (CEV), as set out by government guidance, at the time of vaccination.

Study participants are followed up from vaccination date until the first of:

* The outcome of interest
* Death or deregistration
* Second vaccine dose
* The study end date, `r format(as.Date(gbl_vars$end_date), "%d %B %Y")`

### Identifying vaccinated HCWs

Those vaccinated as part of England's national vaccination programme (for example not including vaccine trial participants) are asked whether they work in health or social care. This information is sent to OpenSAFELY-TPP from NHS Digital's COVID-19 data store.

Note -- many of those flagged as HCWs do not have a vaccination record, which shouldn't be the case if the question was asked as part of the vaccination process.

## Study measures

### Exposure

The vaccination brand, BNT162b2 or ChAdOx1. This is available in the GP record directly, via the National Immunisation Management System (NIMS).

### Outcomes

* SARS-CoV-2 infection, as identified via SGSS records. Both PCR and LFT tests are included.
* Unplanned COVID-19-related hospitalisation via HES records. ICD-10 codes are used to identify COVID-19-related admissions. 
* Unplanned COVID-19-related ICU admission via HES records. ICD-10 codes are used to identify COVID-19-related admissions. 
* COVID-19-related death via death registration records. With COVID-19 ICD-10 codes anywhere on the death certificate (i.e., as an underlying or contributing cause of death)

## Statistical Analysis

The aim is to estimate comparative vaccine effectiveness, i.e., the relative hazard of each outcome for ChAdOx1 versus BNT162b2 vaccine recipients. The effect is permitted to vary by time since vaccination, to account for the potential differences in vaccine protection over time between the two brands.

Patient characteristics used for adjustment include: age, sex, deprivation, ethnicity, NHS region, clinically “at risk” (but not clinically extremely vulnerable) as per national prioritisation guidelines, asthma, number of previous SARS-CoV-2 tests (via SGSS), rurality, evidence of prior covid infection (positive test or COVID-19 hospitalisation), number of recorded comorbidities, severe mental illness. All variables are ascertained as at the time of vaccination.

NHS regions are used as a stratification variable to account for geographical variation in event rates for the outcomes of interest (primarily driven by changes in local infection rates).

### Time-dependent Cox models
Time-dependent Cox models are used to estimate the time-varying effects for each vaccine type. Two time-scales are used for triangulation:

* On the *calendar time-scale*, time zero is `r format(as.Date(gbl_vars$start_date_az), "%d %B %Y")`, and people only contribute follow-up time once they have been vaccinated (i.e., delayed study entry). Vaccine type, time since vaccination in weekly intervals, and the interaction between these two terms are included to estimate time-varying vaccine effects for each vaccine type. In `R`'s `coxph` function, this is modelled using `vaccine_type*week_since_first_dose`

* On the *treatment time-scale*, time zero is the date of vaccination. Date of vaccination is included as an additional baseline covariate using a restricted cubic spline with knots at the first and second tertile, plus boundary knots. Here, follow-up time is stratified into weekly intervals with the vaccine-specific effect estimated within each time-strata. In `R`'s `coxph` function, this is modelled using `vaccine_type:strata(week_since_first_dose)`.

### Pooled Logistic Regression models

We emulate the Cox models described above using pooled logistic regression (PLR), with the outcome risk estimated at daily intervals. This serves two purposes. Firstly, a continuous-time estimate of comparative effectiveness, as opposed to the piecewise-linear approximation above, can be obtained using flexible parametric splines. Secondly, the absolute cumulative risk of each outcome of interest for each vaccine type, marginalised over the population, can be obtained using the parametric g-formula. These models include a restricted cubic spline with 3 degrees of freedom on the timescale of interest, interacted with vaccine type. Cumulative risk is then calculated as the average risk for each day of follow-up predicted by the PLR model, under the (complimentary counterfactual) assumptions that every patient received the BNT162b2 vaccine or that every patient received the ChAdOx1 vaccine.

The person-time dataset needed to fit the PLR model is large and leads to infeasible RAM requirements / computation time. To deal with this, a sampling strategy is used such that all those who experienced an outcome and a random sample of 50000 who did not experience the ouotcome are selected for the models. The models are weighted to recover the characteristics of the complete dataset. This weighting is also applied to the average cumulative risk.

# Results

Note all counts below 6 are redacted and Kaplan-Meier survival estimates are rounded for disclosure control.

## Flowchart

```{r flowchart, echo=FALSE, warning=FALSE, message=FALSE}

flowchart <- read_csv(here("output", "data", "flowchart.csv"))

flowchart %>%
  select(
    Criteria = criteria,
    N = n,
    `N excluded` = n_exclude,
    `% excluded` = pct_exclude,
    `% remaining` = pct_all
  ) %>%
  gt() %>%
  fmt_percent(
    columns = starts_with(c("%")),
    decimals = 1
  ) %>%
  fmt_number(
    columns = starts_with(c("N")),
    decimals= 0
  ) %>%
  fmt_missing(
    everything(),
    missing_text="-"
  )

```

## Baseline demographics

```{r table1, echo=FALSE}
read_rds(here("output", "descriptive", "tables", "table1.rds"))
```

## Event rates

```{r table_irr, echo=FALSE}
read_rds(here("output", "descriptive", "tables", "table_irr.rds"))
```

## Comparative effectiveness

The plots below show:

* Unadjusted Kaplan-Meier curves
* *ChAdOx1* versus *BNT162b2* piecewise-linear hazard ratios estimated using the Cox models
* *ChAdOx1* versus *BNT162b2* continuous hazard ratios estimated using the PLR models
* Marginalised cumulative risk for *ChAdOx1* and *BNT162b2* estimated using the PLR models

```{r, curves, echo=FALSE, message=FALSE, warning=FALSE, out.width='80%', results='asis'}

print_plots_km <- function(outcome){

  km_image <- here("output", "descriptive", "km", glue("plot_survival_ci_{outcome}.png"))

  km_plot <- read_rds(here("output", "descriptive", "km", glue("plot_survival_ci_{outcome}.rds")))

  print(km_plot)
  cat("  \n\n")

}

print_plots_effects <- function(outcome, modeltype){

  suffix <- if(modeltype=="plr"){
    "_pw"
  } else{
    ""
  }

  calendar_effects <- read_rds(here("output", "models", outcome, "calendar", glue("report{modeltype}_effects{suffix}.rds"))) %>%
    mutate(timescale = "Calendar time")
  timesincevax_effects <- read_rds(here("output", "models", outcome, "timesincevax", glue("report{modeltype}_effects{suffix}.rds"))) %>%
    mutate(timescale = "Vaccinated time")

  tidy_effects <- bind_rows(calendar_effects, timesincevax_effects)

  plot_effects <-
    ggplot(data = tidy_effects) +
    geom_point(aes(y=exp(estimate), x=term_midpoint, colour=model_name), position = position_dodge(width = 1.8))+
    geom_linerange(aes(ymin=exp(conf.low), ymax=exp(conf.high), x=term_midpoint, colour=model_name), position = position_dodge(width = 1.8))+
    geom_hline(aes(yintercept=1), colour='grey')+
    facet_grid(cols=vars(timescale))+
    scale_x_continuous(breaks=unique(tidy_effects$term_left), limits=c(min(tidy_effects$term_left), max(tidy_effects$term_right)+1), expand = c(0, 0))+
    scale_y_log10(
    breaks=c(0.1, 0.125, 0.25, 0.33, 0.5, 0.75, 1, 1.33, 2, 3, 4, 8,10),
    sec.axis = dup_axis(name="favours\n <- BNT162b2 / ChAdOx1 ->", breaks = NULL)
  )+
    scale_colour_brewer(type="qual", palette="Set2", guide=guide_legend(ncol=2))+
    labs(
      y="Hazard ratio",#\nChAdOx1 / BNT162b2",
      x="Days since vaccination",
      colour=NULL
     ) +
    theme_bw()+
    theme(
      panel.border = element_blank(),
      axis.line.y = element_line(colour = "black"),

      panel.grid.minor.x = element_blank(),
      panel.grid.minor.y = element_blank(),
      strip.background = element_blank(),
      strip.placement = "outside",
      strip.text.y.left = element_text(angle = 0),

      panel.spacing = unit(0.8, "lines"),

      plot.title = element_text(hjust = 0),
      plot.title.position = "plot",
      plot.caption.position = "plot",
      plot.caption = element_text(hjust = 0, face= "italic"),

      legend.position = "bottom"
     ) +
   NULL

  print(plot_effects)
  cat("  \n\n")
}

print_plots_spline <- function(outcome){

  calendar_effects <- read_rds(here("output", "models", outcome, "calendar", glue("reportplr_effects_ns.rds"))) %>%
    mutate(timescale = "Calendar time")
  timesincevax_effects <- read_rds(here("output", "models", outcome, "timesincevax", glue("reportplr_effects_ns.rds"))) %>%
    mutate(timescale = "Vaccinated time")

  spline <- bind_rows(calendar_effects, timesincevax_effects)

  plot_spline <-
    ggplot(spline)+
    geom_hline(aes(yintercept=1), colour='grey')+
    geom_line(aes(x=tstop-1, y=hr, colour=model_name))+
    geom_ribbon(aes(x=tstop-1, ymin=hr.ll, ymax=hr.ul, fill=model_name), alpha=0.2, colour="transparent")+
    facet_grid(cols=vars(timescale))+
    scale_x_continuous(
      breaks = seq(0,7*52,by=14),
      expand = expansion(0)
    )+
    scale_y_log10(
      breaks=c(1/8, 1/6, 0.25, 0.33, 0.5, 0.67, 1, 1.5, 2, 3, 4, 6, 8),
      sec.axis = dup_axis(name="favours\n <- BNT162b2 / ChAdOx1 ->", breaks = NULL)
    )+
    scale_colour_brewer(type="qual", palette="Set2", guide=guide_legend(ncol=1))+
    scale_fill_brewer(type="qual", palette="Set2", guide="none")+
    labs(
      x = "Days since first dose",
      y = "Hazard ratio",
      colour = NULL, fill=NULL
    )+
    theme_bw()+
    theme(
      panel.border = element_blank(),
      axis.line.y = element_line(colour = "black"),

      panel.grid.minor.y = element_blank(),
      strip.background = element_blank(),
      strip.placement = "outside",
      strip.text.y.left = element_text(angle = 0),

      panel.spacing = unit(0.8, "lines"),

      plot.title = element_text(hjust = 0),
      plot.title.position = "plot",
      plot.caption.position = "plot",
      plot.caption = element_text(hjust = 0, face= "italic"),

      legend.position = "bottom"
     ) +
   NULL

  print(plot_spline)
  cat("  \n\n")
}

print_plots_adjsurv <- function(outcome){

  calendar_curve <- read_rds(here("output", "models", outcome, "calendar", glue("reportplr_adjustedsurvival_ns.rds"))) %>%
    mutate(timescale = "Calendar time")
  timesincevax_curve <- read_rds(here("output", "models", outcome, "timesincevax", glue("reportplr_adjustedsurvival_ns.rds"))) %>%
    mutate(timescale = "Vaccinated time")

  curves <- bind_rows(calendar_curve, timesincevax_curve)

  plot_cmlinc <-
    ggplot(curves %>% filter(model=="3"))+
    geom_step(aes(x=tstop, y=survival, group=vax1_az_descr, colour=vax1_az_descr))+
    geom_rect(aes(xmin=lag(tstop, 1, 0), xmax=tstop, ymin=survival.ll, ymax=survival.ul, group=vax1_az_descr, fill=vax1_az_descr), alpha=0.1)+
    facet_grid(cols=vars(timescale))+
    scale_x_continuous(
      breaks = seq(0,7*52,by=14),
      expand = expansion(0)
    )+
    scale_y_continuous(
      expand = expansion(0)
    )+
    scale_colour_brewer(type="qual", palette="Set1", na.value="grey")+
    scale_fill_brewer(type="qual", palette="Set1", guide="none", na.value="grey")+
    labs(
      x="Days since first dose",
      y="Adjusted survival",
      colour=NULL
    )+
    theme_bw()+
    theme(
      legend.position=c(0.05,.95),
      legend.justification = c(0,1),
      axis.text.x.top=element_text(hjust=0),

      panel.border = element_blank(),
      axis.line.y = element_line(colour = "black"),

      panel.grid.minor.x = element_blank(),
      panel.grid.minor.y = element_blank(),
      strip.background = element_blank(),
      strip.placement = "outside",
      strip.text.y.left = element_text(angle = 0),

      panel.spacing = unit(0.8, "lines"),

    )+
    NULL

  print(plot_cmlinc)
  cat("  \n\n")
}

cat("  \n\n")
print(glue("### SARS-CoV-2 test"), "  \n")
print(glue("#### Unadjusted"), "  \n")
print_plots_km("test")
print(glue("#### Cox"), "  \n")
print_plots_effects("test", "cox")
print(glue("#### PLR, piecewise"), "  \n")
print_plots_effects("test", "plr")
print(glue("#### PLR, spline"), "  \n")
print_plots_spline("test")
print_plots_adjsurv("test")

cat("  \n\n")
print(glue("### SARS-CoV-2 positive test"), "  \n")
print(glue("#### Unadjusted"), "  \n")
print_plots_km("postest")
print(glue("#### Cox"), "  \n")
print_plots_effects("postest", "cox")
print(glue("#### PLR, piecewise"), "  \n")
print_plots_effects("postest", "plr")
print(glue("#### PLR, spline"), "  \n")
print_plots_spline("postest")
print_plots_adjsurv("postest")

cat("  \n\n")
print(glue("### A&E attendance"), "  \n")
print(glue("#### Unadjusted"), "  \n")
print_plots_km("emergency")
print(glue("#### Cox"), "  \n")
print_plots_effects("emergency", "cox")
print(glue("#### PLR, piecewise"), "  \n")
print_plots_effects("emergency", "plr")
print(glue("#### PLR, spline"), "  \n")
print_plots_spline("emergency")
print_plots_adjsurv("emergency")

cat("  \n\n")
print(glue("### Any unplanned hospital admission"), "  \n")
print(glue("#### Unadjusted"), "  \n")
print_plots_km("admitted")
print(glue("#### Cox"), "  \n")
print_plots_effects("admitted", "cox")
print(glue("#### PLR, piecewise"), "  \n")
print_plots_effects("admitted", "plr")
print(glue("#### PLR, spline"), "  \n")
print_plots_spline("admitted")
print_plots_adjsurv("admitted")

cat("  \n\n")
print(glue("### COVID-19 hospital admission"), "  \n")
print(glue("#### Unadjusted"), "  \n")
print_plots_km("covidadmitted")
print(glue("#### Cox"), "  \n")
print_plots_effects("covidadmitted", "cox")
print(glue("#### PLR, piecewise"), "  \n")
print_plots_effects("covidadmitted", "plr")
print(glue("#### PLR, spline"), "  \n")
print_plots_spline("covidadmitted")
print_plots_adjsurv("covidadmitted")

```

### 2021-08-18 analysis/models/model_plr.R
update model formulae

@@ -159,11 +159,20 @@ formula_outcome <- outcome_event ~ 1
 if(timescale=="calendar"){
   formula_timescale <- . ~ . + ns(tstop_calendar, 3) # spline for timescale only
   formula_spacetime <- . ~ . + ns(tstop_calendar, 3)*region # spline for space-time adjustments
+
+  formula_timesincevax_pw <- . ~ . + vax1_az * timesincevax_pw
+  formula_timesincevax_ns <- . ~ . + vax1_az * ns(log(tstop), 3)
+
 }
 if(timescale=="timesincevax"){
   formula_timescale <- . ~ . + ns(log(tstop), 3) # spline for timescale only
   formula_spacetime <- . ~ . + ns(vax1_day, 3)*region # spline for space-time adjustments
+
+  formula_timesincevax_pw <- . ~ . + vax1_az + vax1_az:timesincevax_pw
+  formula_timesincevax_ns <- . ~ . + vax1_az + vax1_az:ns(log(tstop), 3)
+
 }
+
 
 ## NOTE
 # calendar-time PLR models are still probably wrong!
@@ -173,7 +182,6 @@ if(timescale=="timesincevax"){
 
 ### piecewise formulae ----
 ### estimand
-formula_timesincevax_pw <- . ~ . + vax1_az*timesincevax_pw
 formula_vaxonly_pw <- formula_outcome  %>% update(formula_timesincevax_pw) %>% update(formula_timescale)
 
 formula0_pw <- formula_vaxonly_pw
@@ -183,7 +191,6 @@ formula3_pw <- formula_vaxonly_pw %>% update(formula_spacetime) %>% update(formu
 
 ### natural cubic spline formulae ----
 ### estimands
-formula_timesincevax_ns <- . ~ . + vax1_az*ns(log(tstop), 3)
 formula_vaxonly_ns <- formula_outcome  %>% update(formula_timesincevax_ns) %>% update(formula_timescale)
 
 formula0_ns <- formula_vaxonly_ns

### 2021-08-18 analysis/models/report_plr.R

@@ -65,10 +65,18 @@ formula_outcome <- outcome_event ~ 1
 if(timescale=="calendar"){
   formula_timescale <- . ~ . + ns(tstop_calendar, 3) # spline for timescale only
   formula_spacetime <- . ~ . + ns(tstop_calendar, 3)*region # spline for space-time adjustments
+
+  formula_timesincevax_pw <- . ~ . + vax1_az * timesincevax_pw
+  formula_timesincevax_ns <- . ~ . + vax1_az * ns(log(tstop), 3)
+
 }
 if(timescale=="timesincevax"){
   formula_timescale <- . ~ . + ns(log(tstop), 3) # spline for timescale only
   formula_spacetime <- . ~ . + ns(vax1_day, 3)*region # spline for space-time adjustments
+
+  formula_timesincevax_pw <- . ~ . + vax1_az + vax1_az:timesincevax_pw
+  formula_timesincevax_ns <- . ~ . + vax1_az + vax1_az:ns(log(tstop), 3)
+
 }
 
 ## NOTE
@@ -79,7 +87,6 @@ if(timescale=="timesincevax"){
 
 ### piecewise formulae ----
 ### estimand
-formula_timesincevax_pw <- . ~ . + vax1_az*timesincevax_pw
 formula_vaxonly_pw <- formula_outcome  %>% update(formula_timesincevax_pw) %>% update(formula_timescale)
 
 formula0_pw <- formula_vaxonly_pw
@@ -89,7 +96,6 @@ formula3_pw <- formula_vaxonly_pw %>% update(formula_spacetime) %>% update(formu
 
 ### natural cubic spline formulae ----
 ### estimands
-formula_timesincevax_ns <- . ~ . + vax1_az*ns(log(tstop), 3)
 formula_vaxonly_ns <- formula_outcome  %>% update(formula_timesincevax_ns) %>% update(formula_timescale)
 
 formula0_ns <- formula_vaxonly_ns

### 2021-08-23 analysis/descriptive/km.R
hard-code last follow-up date; add extra cut

@@ -82,7 +82,7 @@ data_tte <- data_cohort %>%
     # assume vaccination occurs at the start of the day, and all other events occur at the end of the day.
     # so use vax1_date - 1
 
-    censor_date = pmin(vax1_date - 1 + (7*14), end_date, dereg_date, death_date, covid_vax_any_2_date, na.rm=TRUE),
+    censor_date = pmin(vax1_date - 1 + lastfupday, end_date, dereg_date, death_date, covid_vax_any_2_date, na.rm=TRUE),
 
     # time to last follow up day
     tte_enddate = tte(vax1_date-1, end_date, end_date),

### 2021-08-23 analysis/descriptive/table_irr.R

@@ -61,7 +61,7 @@ data_tte <- data_cohort %>%
     vax1_type,
     end_date,
 
-    censor_date = pmin(vax1_date - 1 + (7*14), end_date, dereg_date, death_date, covid_vax_any_2_date, na.rm=TRUE),
+    censor_date = pmin(vax1_date - 1 + lastfupday, end_date, dereg_date, death_date, covid_vax_any_2_date, na.rm=TRUE),
 
     # time to last follow up day
     tte_enddate = tte(vax1_date-1, end_date, end_date),

### 2021-08-23 analysis/models/model_cox.R

@@ -75,7 +75,7 @@ data_tte <- data_cohort %>%
   mutate(
 
     outcome_date = .[[glue("{outcome_var}")]],
-    censor_date = pmin(vax1_date - 1 + (7*14), end_date, dereg_date, death_date, covid_vax_any_2_date, na.rm=TRUE),
+    censor_date = pmin(vax1_date - 1 + lastfupday, end_date, dereg_date, death_date, covid_vax_any_2_date, na.rm=TRUE),
 
     # assume vaccination occurs at the start of the day, and all other events occur at the end of the day.
     tte_censor = tte(vax1_date-1, censor_date, censor_date, na.censor=TRUE),

### 2021-08-23 analysis/models/model_plr.R

@@ -78,7 +78,7 @@ data_cohort <- read_rds(here("output", "data", "data_cohort.rds"))
 data_tte <- data_cohort %>%
   mutate(
     outcome_date = .[[glue("{outcome_var}")]],
-    censor_date = pmin(vax1_date - 1 + (7*14), end_date, dereg_date, death_date, covid_vax_any_2_date, na.rm=TRUE),
+    censor_date = pmin(vax1_date - 1 + lastfupday, end_date, dereg_date, death_date, covid_vax_any_2_date, na.rm=TRUE),
 
     tte_censor = tte(vax1_date-1, censor_date, censor_date, na.censor=TRUE),
     ind_censor = censor_indicator(censor_date, censor_date),

### 2021-08-23 analysis/models/report_cox.R

@@ -88,7 +88,7 @@ effectscox <- effectscox %>%
   mutate(
     term_left = as.numeric(str_extract(term, "^\\d+"))-1,
     term_right = as.numeric(str_extract(term, "\\d+$"))-1,
-    term_right = if_else(is.na(term_right), max(term_left)+6, term_right),
+    term_right = if_else(is.na(term_right), lastfupday, term_right),
     term_midpoint = term_left + (term_right+1-term_left)/2
   )
 

### 2021-08-23 analysis/models/report_plr.R

@@ -128,7 +128,7 @@ effectsplr <- tidy_plr_pw %>%
     pw=fct_inorder(pw),
     term_left = as.numeric(str_extract(pw, "^\\d+"))-1,
     term_right = as.numeric(str_extract(pw, "\\d+$"))-1,
-    term_right = if_else(is.na(term_right), max(term_left, na.rm=TRUE)+6, term_right),
+    term_right = if_else(is.na(term_right), lastfupday, term_right),
     term_midpoint = term_left + (term_right+1-term_left)/2
   )
 

### 2021-08-23 analysis/process/design.R

@@ -82,7 +82,10 @@ formula_all_rhsvars <- update(1 ~ 1, formula_exposure) %>%
   update(formula_secular) %>%
   update(formula_secular_region)
 
-postvaxcuts <- c(0, 7, 14, 21, 28, 35, 42)
+postvaxcuts <- c(0, 7, 14, 21, 28, 35, 42, 70)
+
+lastfupday <- 14*7
+
 
 list_formula <- lst(
   formula_exposure,
@@ -91,7 +94,8 @@ list_formula <- lst(
   formula_secular,
   formula_secular_region,
   formula_all_rhsvars,
-  postvaxcuts
+  postvaxcuts,
+  lastfupday
 )
 
 write_rds(list_formula, here("output", "data", "metadata_formulas.rds"))

### 2021-08-23 analysis/report/effectiveness_report_comparetimescales.Rmd
better plots

@@
     facet_grid(cols=vars(timescale))+
-    scale_x_continuous(breaks=unique(tidy_effects$term_left), limits=c(min(tidy_effects$term_left), max(tidy_effects$term_right)+1), expand = c(0, 0))+
+    scale_x_continuous(
+      breaks = seq(0, 7*52, 14), 
+      limits = c(min(tidy_effects$term_left), 14*7), 
+      expand = c(0, 0)
+    )+
     scale_y_log10(
-    breaks=c(0.1, 0.125, 0.25, 0.33, 0.5, 0.75, 1, 1.33, 2, 3, 4, 8,10),
+    breaks=c(0.125, 0.167, 0.25, 0.333, 0.5, 0.667, 1, 1.5, 2, 3, 4, 6, 8),
     sec.axis = dup_axis(name="favours\n <- BNT162b2 / ChAdOx1 ->", breaks = NULL)
@@
     scale_y_log10(
-      breaks=c(1/8, 1/6, 0.25, 0.33, 0.5, 0.67, 1, 1.5, 2, 3, 4, 6, 8),
+      breaks=c(0.125, 0.167, 0.25, 0.333, 0.5, 0.667, 1, 1.5, 2, 3, 4, 6, 8),
       sec.axis = dup_axis(name="favours\n <- BNT162b2 / ChAdOx1 ->", breaks = NULL)

### 2021-08-23 analysis/report/effectiveness_report.Rmd
better plots

@@
     scale_y_log10(
-      breaks=c(0.125, 0.167, 0.25, 0.33, 0.5, 0.67, 1, 1.5, 2, 3, 4, 6, 8),
+      breaks=c(0.125, 0.167, 0.25, 0.333, 0.5, 0.667, 1, 1.5, 2, 3, 4, 6, 8),
       sec.axis = dup_axis(name="Lower hazard after \n <- BNT162b2 / ChAdOx1 ->", breaks = NULL)
@@
     scale_y_continuous(
-      expand = expansion(0)
+      expand = expansion(0),
+      limits = c(NA,1L)
     )+
@@
     )+
-    theme_bw()+
+    theme_minimal()+
     theme(
@@
       axis.line.y = element_line(colour = "black"),
+      axis.line.x.top = element_line(colour = "black"),
 

### 2021-08-26 analysis/report/effectiveness_report.Rmd
make report self-contained

@@
 output:
-  github_document:
-    keep_html: TRUE
+  html_document:
+    self_contained: TRUE
+    keep_md: TRUE
   #  code_folding: hide
@@
   #   variant: gfm
-  #html_document
+  github_document:
+    keep_html: TRUE
 ---
@@
 
-This study compares the effectiveness of the *BNT162b2* mRNA (Pfizer-BioNTech) and *ChAdOx1* (Oxford-AstraZeneca) vaccines amongst health and social care workers (HCWs). 
-This group was chosen as they were prioritised for vaccination due to high occupational exposure and during the period where both vaccines were available (`r format(as.Date(gbl_vars$start_date_az), "%d %B %Y")` onwards) it is expected that the actual vaccine brand received will not be strongly influenced by other determinants of COVID-19-related outcomes (i.e., approximately random treatment allocation and therefore reduced confounding).
+The COVID-19 global pandemic has prompted the rapid development and delivery of vaccines to combat the disease.
+Following demonstration of high safety and efficacy against symptomatic and severe disease in phase-III randomised clinical trials (RCTs), two vaccines have been approved and widely administered as part of the national vaccination programme in the United Kingdom: the Pfizer-BioNTech BNT162b2 mRNA COVID-19 vaccine (*BNT162b2*) and the Oxford-AstraZeneca ChAdOx1 nCoV-19 vaccine (*ChAdOx1*).
+To date, there have been no RCTs that have directly compared these two vaccines to estimate the relative efficacy against COVID-19 infection and disease in the same cohort.
+The concurrent roll-out of BNT162b2 and ChAdOx1 within the same UK population provides an opportunity to emulate such a trial using observational data.
+
+COVID-19 vaccination in the UK has been prioritised based on the risk of infection and subsequent severity of disease.
+Patient-facing health and social care workers (HCWs) were amongst the first groups eligible for vaccination due to the high occupational exposure to the SARS-CoV-2 virus, and many were vaccinated during the period where both vaccines were widely used (`r format(as.Date(gbl_vars$start_date_az), "%d %B %Y")` onwards).
+This study assessed the comparative effectiveness of one dose of ChAdOx1 with one dose of BNT162b2 in HCWs, using the OpenSAFELY-TPP database covering around 40% of England's population.
 
@@
 
-People meeting the following criteria are included:
+Vaccinated HCWs were included in the study if:
 
-* HCWs who have received at least one dose of BNT162b2 or ChAdOx1. 
-* Registered at a GP practice using TPP's SystmOne clinical information system on the day of vaccination.
-* Aged 18-65.
-* Vaccinated on or after `r format(as.Date(gbl_vars$start_date_az), "%d %B %Y")`, when both vaccine brands were being administered. 
-* Not Clinically Extremely Vulnerable (CEV), as set out by government guidance, at the time of vaccination.
+* registered at a GP practice using TPP's SystmOne clinical information system on the day of they received their first dose of BNT162b2 or ChAdOx1; 
+* the date of vaccination was between `r format(as.Date(gbl_vars$start_date_az), "%d %B")` and `r format(as.Date(gbl_vars$lastvax_date), "%d %B %Y")` (`r as.Date(gbl_vars$lastvax_date) - as.Date(gbl_vars$start_date_az)` days), a period when both vaccines were being administered widely;
+* aged between 18 and 64 inclusive; 
+* not classed as Clinically Extremely Vulnerable, as set out by government guidance, at the time of vaccination; 
+* information on sex, ethnicity, deprivation, and geographical region was known.
 
-Study participants are followed up from vaccination date until the first of:
+Study participants were followed up for 14 weeks from the date of the first dose, or until the first of:
 
-* The outcome of interest
-* Death or deregistration
-* Second vaccine dose
-* The study end date, `r format(as.Date(gbl_vars$end_date), "%d %B %Y")`
+* death or deregistration;
+* Second vaccine dose;
+* The study end date, `r format(as.Date(gbl_vars$end_date), "%d %B %Y")`.
 
@@
 Note -- many of those flagged as HCWs do not have a vaccination record, which shouldn't be the case if the question was asked as part of the vaccination process. This needs further investigation.
-
-## Study measures
-
-### Exposure
-
-The vaccination brand, BNT162b2 or ChAdOx1. This is available in the GP record directly, via the National Immunisation Management System (NIMS).
 

### 2021-08-26 create-project.R

@@ -388,7 +388,7 @@ actions_list <- splice(
     name = "rmd_report",
     run = glue(
       "r:latest -e {q}",
-      q = single_quote('rmarkdown::render("analysis/report/effectiveness_report.Rmd",  knit_root_dir = "/workspace",  output_dir = "/workspace/output/report", output_format = c("rmarkdown::github_document")   )')
+      q = single_quote('rmarkdown::render("analysis/report/effectiveness_report.Rmd",  knit_root_dir = "/workspace",  output_dir = "/workspace/output/report", output_format = c("html_document")   )')
     ),
     needs = splice(
       "design", "data_selection",
@@ -403,8 +403,8 @@ actions_list <- splice(
     ),
     moderately_sensitive = list(
       html = "output/report/effectiveness_report.html",
-      md = "output/report/effectiveness_report.md",
-      figures = "output/report/figures/*.png"
+      md = "output/report/effectiveness_report.md"#,
+      #figures = "output/report/figures/*.png"
     )
   ),
 

### 2021-08-26 analysis/process/data_selection.R
simplify flowchart

@@ -21,6 +21,12 @@ args <- commandArgs(trailingOnly=TRUE)
 ## create output directories ----
 fs::dir_create(here("output", "data"))
 
+
+# import globally defined repo variables from
+gbl_vars <- jsonlite::read_json(
+  path=here("analysis", "global-variables.json")
+)
+
 ## Import processed data ----
 
 data_processed <- read_rds(here("output", "data", "data_processed.rds"))
@@ -64,13 +70,10 @@ write_rds(data_cohort, here("output", "data", "data_cohort.rds"), compress="gz")
 
 data_flowchart <- data_criteria %>%
   transmute(
-    c0_all = TRUE,
+    c0_all = vax1_azpfizer & vax1_afterstartdate & vax1_beforelastvaxdate,
     #c1_1yearfup = c0_all & (has_follow_up_previous_year),
     c1_notmissing = c0_all & (has_age & has_sex & has_imd & has_ethnicity & has_region),
-    c2_notcev = c1_notmissing & not_cev,
-    c3_lastvaxdate = c2_notcev & vax1_beforelastvaxdate,
-    c4_startdate = c3_lastvaxdate & vax1_afterstartdate,
-    c5_azpfizer = c4_startdate & vax1_azpfizer
+    c2_notcev = c1_notmissing & not_cev
   ) %>%
   summarise(
     across(.fns=sum)
@@ -87,12 +90,12 @@ data_flowchart <- data_criteria %>%
     pct_step = n / lag(n),
     crit = str_extract(criteria, "^c\\d+"),
     criteria = fct_case_when(
-      crit == "c0" ~ "All vaccinated HCWs aged 16-65",
+      crit == "c0" ~ "All HCWs aged 18-64\n  receiving first dose of BNT162b2 or ChAdOx1\n  between 4 January and 28 February 2021",
       crit == "c1" ~ "  with no missing demographic information",
       crit == "c2" ~ "  who are not clinically extremely vulnerable",
-      crit == "c3" ~ "  with vaccination on or before recruitment end date",
-      crit == "c4" ~ "  with vaccination on or after recruitment start date",
-      crit == "c5" ~ "  with Pfizer/BNT or Oxford/AZ vaccine",
+      #crit == "c3" ~ "  with vaccination on or before recruitment end date",
+      #crit == "c4" ~ "  with vaccination on or after recruitment start date",
+      #crit == "c5" ~ "  with Pfizer/BNT or Oxford/AZ vaccine",
       TRUE ~ NA_character_
     )
   )

### 2021-08-28 analysis/report/effectiveness_manuscript.Rmd
add manuscript

---
title: "Comparative vaccine effectiveness (ChAdOx1 versus BNT162b2) in Health and Social Care workers"
output:
  html_document:
    self_contained: TRUE
    keep_md: TRUE
  github_document:
    keep_html: TRUE
  # md_document:
  #   variant: gfm
---

```{r setup, include=FALSE}

library('tidyverse')
library('here')
library('glue')
library('lubridate')
library('gt')
library('patchwork')
library('scales')
source(here("analysis", "lib", "utility_functions.R"))

output_dir <- here("output")

fs::dir_create(output_dir, "report", "figures")

knitr::opts_chunk$set(
  echo = TRUE,
  fig.path = paste0(fs::path(output_dir, "report", "figures"), "/")
)

gbl_vars <- jsonlite::fromJSON(
  txt=here("analysis", "global-variables.json")
)

data_cohort <- read_rds(here("output", "data", "data_cohort.rds"))

flowchart <- read_csv(here("output", "data", "flowchart.csv"))

metadata_outcomes <- read_rds(here("output", "data", "metadata_outcomes.rds"))

list_formula <- read_rds(here("output", "data", "metadata_formulas.rds"))
list2env(list_formula, globalenv())

```

# Introduction

The COVID-19 global pandemic has prompted the rapid development and delivery of vaccines to combat the disease. Following demonstration of high safety and efficacy against symptomatic and severe disease in phase-III randomised clinical trials (RCTs), two vaccines have been approved and widely administered as part of the national vaccination programme in the United Kingdom: the Pfizer-BioNTech BNT162b2 mRNA COVID-19 vaccine (*BNT162b2*) and the Oxford-AstraZeneca ChAdOx1 nCoV-19 vaccine (*ChAdOx1*). Post-authorisation assessment of vaccine effectiveness using observational data is necessary to monitor the success of such programmes as, invariably, target populations and settings differ substantially from those of trials. To date, there have been no RCTs that have directly compared the BNT162b2 and ChAdOx1 vaccines to estimate the relative efficacy against COVID-19 infection and disease in the same population. The concurrent roll-out of these vaccines across the UK provides an opportunity to emulate such a trial using observational data.

COVID-19 vaccination in the UK has been prioritised based on the risk of infection and subsequent severity of disease. Patient-facing health and social care workers (HCWs) were amongst the first groups eligible for vaccination due to the high occupational exposure to the SARS-CoV-2 virus, and many were vaccinated during the period where both vaccines were widely used. This study assesses the comparative effectiveness of one dose of ChAdOx1 with one dose of BNT162b2 in HCWs, using the OpenSAFELY-TPP linked primary care database covering around 40% of England's population.

# Methods

## Study data

The OpenSAFELY-TPP database covers 24 million people registered at GP surgeries that use TPP's SystmOne software. This primary care data is linked (via patients NHS numbers) with A&E attendance and in-patient hospital spell records via NHS Digital’s Hospital Episode Statistics (HES), national coronavirus testing records via the Second Generation Surveillance System (SGSS), and national death registry records from the Office for National Statistics (ONS). Vaccination status is available in the GP record directly via the National Immunisation Management System (NIMS). HCW status is recorded for all vaccine recipients at the time of vaccination, and this information is sent to OpenSAFELY-TPP from NHS Digital's COVID-19 data store.

## Study population

We studied health and social care workers in England vaccinated with either BNT162b2 or ChAdOx1. This group was prioritised for vaccination at the start of the vaccine roll-out due to the high occupational exposure to the SARS-CoV-2 virus, and many were vaccinated during the period where both vaccines were widely used.

Vaccinated HCWs were included in the study if: they were registered at a GP practice using TPP's SystmOne clinical information system on the day of they received their first dose of BNT162b2 or ChAdOx1; the date of vaccination was between `r format(as.Date(gbl_vars$start_date_az), "%d %B")` and `r format(as.Date(gbl_vars$lastvax_date), "%d %B %Y")` (`r as.integer(as.Date(gbl_vars$lastvax_date) - as.Date(gbl_vars$start_date_az)) + 1` days), a period when both vaccines were being administered widely; they aged between 18 and 64 inclusive; not classed as Clinically Extremely Vulnerable, as set out by government guidance, at the time of vaccination; information on sex, ethnicity, deprivation, and geographical region was known.

Study participants were followed up for no more than 14 weeks from the day of the first dose. Follow-up was censored at: the study end date, `r format(as.Date(gbl_vars$end_date), "%d %B %Y")`; a second vaccine dose; death or de-registration.

## Outcomes

Three outcomes were defined: positive SARS-CoV-2 swab test, A&E attendances, and COVID-19 hospital admissions.

Positive SARS-CoV-2 tests were identified using SGSS records and based on swab date. Both polymerase chain reaction (PCR) and lateral flow tests are included. We did not differentiate between symptomatic and asymptomatic infection. COVID-19 hospital admissions were identified using HES in-patient hospital records. Admissions where the ICD-10 coded primary or non-primary "reason for admission" included U07.1 (“COVID-19, virus identified”) and U07.2 ( “COVID-19, virus not identified”) were included.

Although severe disease (such as requirement for intensive or critical care) and mortality were of interest there were too few events to investigate these outcomes fully. Unadjusted rates for these outcomes are reported descriptively. COVID-19 critical care admissions were identified via HES. COVID-19 deaths were identified using linked death registration data. Death with COVID-19 ICD-10 codes (as above) mentioned anywhere on the death certificate (i.e., as an underlying or contributing cause of death) were included.

## Additional variables

Patient characteristics used describe the cohort and for confounder adjustment include: age, sex (male or female), English Index of Multiple Deprivation (IMD, grouped by quintiles), ethnicity (Black, Mixed, South Asian, White, Other, as per the UK census), NHS region (East of England, Midlands, London, North East and Yorkshire, North West, South East, South West), number of conditions in the clinically “at risk” (but not clinically extremely vulnerable) classification, as per national prioritisation guidelines, <!--number of previous SARS-CoV-2 tests (via SGSS), rurality,-->evidence of prior SARS-CoV-2 infection (positive test or COVID-19 hospitalisation), learning disabilities, severe mental illness. All characteristics were ascertained as at the time of vaccination.

## Statistical Analysis

We estimated the relative effectiveness of a first dose of ChAdOx1 versus BNT162b2 using pooled logistic regression (PLR) models with the outcome risk estimated at daily intervals. The effect is permitted to vary by time since vaccination to account for the potential differences in vaccine protection over time between the two brands. The PLR models emulate a Cox model with continuous-time-dependent effects, and enables the estimation of risk-adjusted survival rates for each vaccine type using the parametric g-formula. This is one minus the average of the cumulative incidence for each day of follow-up predicted by the PLR model, under the (counterfactual) assumption that every person received the BNT162b2 vaccine or that every patient received the ChAdOx1 vaccine. Assuming adequate confounder adjustment, this replicates the cumulative incidence that would have been observed in an RCT comparing the two vaccines in the population under consideration. Standard errors for the PLR model estimates were obtained using the HC0 clustered sandwich estimator to account for within-person clustering. Confidence intervals for the risk-adjusted cumulative incidence were obtained using the delta method (i.e, a first-order Taylor series approximation of the variance).

Four models were fit for each outcome, with progressive adjustment for confounders: (1) An unadjusted model using a restricted cubic spline term with 4 degrees of freedom for time since vaccination, vaccine type, and their interaction; (2) Additionally adjusting for region-specific calendar-time effects, by including a restricted cubic spline term with 4 degrees of freedom for the date of vaccination and its interaction with region; (3) additionally adjusting for demographic characteristics; (4) additionally adjusting for clinical characteristics.

<!-- Events occurring on the same day as vaccination are included in follow-up. -->

PLR models are computationally expensive to fit, as the input dataset must be arranged as one row per person per day of follow-up. To manage this, a sampling strategy was used such that all those who experienced the event of interest are selected, and a random sample of 50000 who did not experience the event are selected. The models are weighted to recover the characteristics of the complete cohort. This weighting is also applied to the risk-adjusted survival curves. Sampling was based on a ranked hash of the person identifier to ensure the sampled non-events overlapped as much as possible across outcomes.

## Data and code

Data management and analyses were conducted in Python 3.8 and R version 4.0.2. All code is available for review and reuse at https://github.com/opensafely/comparative-ve-research.

# Results

Note all counts below 6 are redacted and survival estimates are rounded for disclosure control.

## Study population

```{r characteristics, echo=FALSE, warning=FALSE, message=FALSE}

data_cohort <- 
  data_cohort %>%
  mutate(
    censor_date = pmin(vax1_date - 1 + lastfupday, end_date, dereg_date, death_date, covid_vax_any_2_date, na.rm=TRUE),
    tte_censor = tte(vax1_date-1, censor_date, censor_date, na.censor=TRUE),

    postest = censor_indicator(positive_test_date, censor_date),
    emergency = censor_indicator(emergency_date, censor_date),
    admitted = censor_indicator(admitted_date, censor_date),
    covidadmitted = censor_indicator(covidadmitted_date, censor_date),
    covidcc = censor_indicator(covidcc_date, censor_date),
    coviddeath = censor_indicator(coviddeath_date, censor_date),
    death = censor_indicator(death_date, censor_date),
  )

baseline <- 
  data_cohort %>%
  summarise(
    n = n(),
    age_median = median(age),
    age_Q1 = quantile(age, 0.25),
    age_Q3 = quantile(age, 0.75),
    female = mean(sex=="Female"),
    fu_years = sum(tte_censor)/365.25,
    fu_median = median(tte_censor),
    priorinfection = mean(prior_covid_infection),

    postest = sum(postest),
    emergency = sum(emergency),
    admitted = sum(admitted),
    covidadmitted = sum(covidadmitted),
    coviddeath = sum(coviddeath),
    death = sum(death),
  )

baseline_az <- 
  data_cohort %>%
  filter(vax1_type=="az") %>%
  summarise(
    n = n(),
    age_median = median(age),
    age_Q1 = quantile(age, 0.25),
    age_Q3 = quantile(age, 0.75),
    female = mean(sex=="Female"),
    fu_years = sum(tte_censor)/365.25,
    fu_median = median(tte_censor),
    priorinfection = mean(prior_covid_infection),
  )

baseline_pfizer <- 
  data_cohort %>%
  filter(vax1_type=="pfizer") %>%
  summarise(
    n = n(),
    age_median = median(age),
    age_Q1 = quantile(age, 0.25),
    age_Q3 = quantile(age, 0.75),
    female = mean(sex=="Female"),
    fu_years = sum(tte_censor)/365.25,
    fu_median = median(tte_censor),
    priorinfection = mean(prior_covid_infection),
  )

```

A total of `r flowchart[1, "n"]` HCWs aged 18-64 receiving a first dose of BNT162b2 or ChAdOx1 between 4 January and 28 February 2021 and actively registered at at TPP practice were identified, with `r flowchart[[nrow(flowchart), "n"]]` (`r `label_percent(0.1)(flowchart[[nrow(flowchart), "pct_all"]])`) meeting the study eligibility criteria. `r baseline_pfizer$n` (`r `label_percent(0.1)(baseline_pfizer$n/baseline$n)`) were vaccinated with BNT162b2, contributing `r baseline_pfizer$fu_years` person-years of follow-up. For ChAdOx1, this was `r baseline_az$n` (`r label_percent(0.1)(baseline_az$n/baseline$n)`) and `r baseline_az$fu_years` person-years. Baseline characteristics are reported in Table 1. The cumulative proportion of vaccinated study participants by calendar time is reported in Figure 1.

### Table 0: Inclusion criteria
```{r flowchart, echo=FALSE, warning=FALSE, message=FALSE}
flowchart %>%
  select(
    Criteria = criteria,
    N = n,
    `N excluded` = n_exclude,
    `% excluded` = pct_exclude,
    `% remaining` = pct_all
  ) %>%
  gt() %>%
  fmt_percent(
    columns = starts_with(c("%")),
    decimals = 1
  ) %>%
  fmt_number(
    columns = starts_with(c("N")),
    decimals= 0
  ) %>%
  fmt_missing(
    everything(),
    missing_text="-"
  )

```

### Table 1: Baseline characteristics

```{r table1, echo=FALSE}
read_rds(here("output", "descriptive", "tables", "table1.rds"))
```

### Figure 1: Vaccination dates

```{r, vaxdate, echo=FALSE, message=FALSE, warning=FALSE, out.width='80%', results='asis'}
  vaxdate_plot <- read_rds(here("output", "descriptive", "vaxdate", "plot_vaxdate_stack.rds"))
  print(vaxdate_plot)
  cat("  \n\n")
```

## Event rates

Figure 2 shows event rates by vaccine type, and the unadjusted incidence ratio ratio. Over the duration of `r label_number(0.1)(baseline$fu_years)` person-years of follow-up there were `r baseline$postest` positive SARS-CoV-2 tests, `r baseline$covidadmitted` COVID-19 hospital admissions, and `r baseline$coviddeath` COVID-19 deaths.

### Figure 2: Event rates

```{r table_irr, echo=FALSE}
read_rds(here("output", "descriptive", "tables", "table_irr.rds"))
```

## Comparative effectiveness

Figure 3 shows, for each outcome based on the fully adjusted model, the time-dependent hazard ratio for ChAdOx1 versus BNT162b2, and one minus the risk-adjusted cumulative incidence<!--, and the absolute risk difference-->. Similar figures for the non-fully-adjusted models are presented in supplementary materials.

```{r, effects, echo=FALSE, message=FALSE, warning=FALSE, out.width='90%', fig.height=6, results='asis'}

  outcomes <- 
    c(
      "postest",
      "emergency",
      "covidadmitted"
    ) %>% 
    set_names(., .)

  cmlinc <-
    map_dfr(
      outcomes,
      ~read_rds(here("output", "models", .x, "timesincevax", glue("reportplr_adjustedsurvival_ns.rds"))),
      .id="outcome"
    ) %>%
    left_join(metadata_outcomes, by="outcome") %>%
    mutate(
      outcome_descr = fct_inorder(outcome_descr),
      model_name = fct_inorder(str_wrap(model_name, 15))
    ) %>%
    filter(model==3)

  plot_cmlinc <-
    ggplot(cmlinc)+
    geom_step(aes(x=tstop, y=survival, group=vax1_az_descr, colour=vax1_az_descr))+
    geom_rect(aes(xmin=lag(tstop, 1, 0), xmax=tstop, ymin=survival.ll, ymax=survival.ul, group=vax1_az_descr, fill=vax1_az_descr), alpha=0.1)+
    geom_hline(aes(yintercept=1), colour='black')+
    facet_grid(cols=vars(outcome_descr))+
    scale_x_continuous(
      breaks = seq(0,7*52,by=14),
      expand = expansion(0)
    )+
    scale_y_continuous(
      expand = expansion(0),
      limits = c(NA,1L)
    )+
    scale_colour_brewer(type="qual", palette="Set1", na.value="grey")+
    scale_fill_brewer(type="qual", palette="Set1", guide="none", na.value="grey")+
    labs(
      x=NULL,
      y="1 - cumulative incidence",
      colour=NULL
    )+
    theme_minimal()+
    theme(
      legend.position=c(0.9,0.05),
      legend.justification = c(1,0),
      legend.direction = "vertical",
      axis.text.x.top=element_text(hjust=0),

      panel.border = element_blank(),
      axis.line.y = element_line(colour = "black"),
      axis.line.x.top = element_line(colour = "black"),

      panel.grid.minor.x = element_blank(),
      panel.grid.minor.y = element_blank(),
      strip.background = element_blank(),

      panel.spacing = unit(0.8, "lines"),

    )+
    NULL

  hazard <-
    map_dfr(
      outcomes,
      ~read_rds(here("output", "models", .x, "timesincevax", glue("reportplr_effects_ns.rds"))),
      .id="outcome"
    ) %>%
    left_join(metadata_outcomes, by="outcome") %>%
    mutate(
      outcome_descr = fct_inorder(outcome_descr),
      model_name = fct_inorder(str_wrap(model_name, 15))
    ) %>%
    filter(model==3)

  plot_hazard <-
    ggplot(hazard)+
    geom_hline(aes(yintercept=1), colour='black')+
    geom_line(aes(x=tstop, y=hr), colour='orchid4')+
    geom_ribbon(aes(x=tstop, ymin=hr.ll, ymax=hr.ul), alpha=0.1, colour="transparent", fill='orchid4')+
    facet_grid(cols=vars(outcome_descr))+
    scale_x_continuous(
      limits = c(0, NA),
      breaks = seq(0,7*52,by=14),
      expand = expansion(0)
    )+
    scale_y_log10(
      #breaks=c(0.125, 0.167, 0.25, 0.333, 0.5, 0.667, 1, 1.5, 2, 3, 4, 6, 8),
      sec.axis = dup_axis(name="Lower hazard after \n <- BNT162b2 / ChAdOx1 ->", breaks = NULL)
    )+
    labs(
      x = "Days since first dose",
      y = "Hazard ratio"
    )+
    theme_bw()+
    theme(
      panel.border = element_blank(),
      axis.line.y.left = element_line(colour = "black"),

      panel.grid.minor.x = element_blank(),
      panel.grid.minor.y = element_blank(),
      strip.background = element_blank(),
      strip.text = element_blank(),

      panel.spacing = unit(0.8, "lines"),

      plot.title = element_text(hjust = 0),
      plot.title.position = "plot",
      plot.caption.position = "plot",
      plot.caption = element_text(hjust = 0, face= "italic")

     ) +
   NULL

  patchwork::wrap_plots(plot_cmlinc, plot_hazard, ncol=1)

  riskdifference <- 
    cmlinc  %>%
    group_by(model, model_name, outcome, outcome_descr, tstop) %>%
    arrange(tstop, model, outcome, vax1_az) %>%
    mutate(
      diff = survival - first(survival),
      diff_format = label_number(0.01, scale=1000)(diff)
    ) %>%
    filter(vax1_az==1)

```

# Discussion

This observational study of `r baseline$n` health and social care workers living in England found no substantial differences in rates of COVID-19-related events between those receiving a first dose of BNT162b2 or ChAdOx1. At 6 weeks post-vaccination, the absolute risk difference (ChAdOx1 - BNT162b2) for a positive SARS-CoV-2 test was `r filter(riskdifference, outcome=="postest", tstop==7*6)$diff_format` per 1000 people, and for COVID-19 hospital admissions was `r filter(riskdifference, outcome=="covidadmitted", tstop==7*6)$diff_format` per 1000 people. There is a clear leveling-off of the event rates after around 3 weeks, consistent with the expected time-to-onset of vaccine-induced immunity of around 2 weeks plus the delay from infection to a positive test.

## Strengths and weaknesses

This study used routinely-collected health records with comprehensive coverage of primary care, hospital admissions, COVID-19 testing, COVID-19 vaccination, and death registrations for over 300,000 health and social care workers living in England. This group were eligible for vaccination at the start of the UK's vaccination programme due to exposure to higher viral loads and the need to reduce enforced absences in frontline healthcare staff during a global pandemic. They are the only early vaccinees who are relatively young and healthy, and were vaccinated during a period where infection rates were high, thus providing one of the few opportunities to study healthy people at a time when event rates are sufficiently high to make meaningful comparisons between vaccines.

We were unable to study differences in the vaccines ability to reduce transmission, nor fully investigate differences in protection against severe disease. This latter limitation is thanks to clear protective benefits of both vaccines, reducing the absolute numbers of events, and therefore statistical power, in the studied cohort.

Despite adjustment for a range of potential confounders and the fairly arbitrary allocation of vaccine type, the possibility of unmeasured confounding remains. The cold-chain storage requirements of BNT162b2 meant that it is more likely to have be administered in Acute NHS Trusts and other large vaccination centres, and is thus a potential confounder for instance due to differences in exposure to risk across these settings. Though we adjusted for patient rurality, we were unable to directly account for occupational differences.

## Findings in context

To our knowledge, this is first study to assess effectiveness of BNT162b2 and ChAdOx1 vaccines in a head-to-head comparison.

A number of studies estimate COVID-19 vaccine effectiveness in observational data using unvaccinated controls. Inferring comparative effectiveness from these studies is possible by calculating the ratio of 1 minus the vaccine effectiveness for each vaccine.

* EAVE-II estimated VE in 18-64 age group in 5th week to be 92% for BNT162b2 and 100% for ChAdOx1, = HR=0.00
* EAVE-II estimated VE in 65-79 age group in 4th week to be 91% for BNT162b2 and 68% for ChAdOx1, = HR=3.56
* EAVE-II estimated VE in 80+ age group in 4th week to be 88% for BNT162b2 and 81% for ChAdOx1, = HR=1.58

are extremely vulnerable to confounding for example due to vaccine prioritisation and eligiblity policies. However, many of these biases can be bypassed when making direct comparisons between recipients of different vaccine types. The United Kingdom was amongst the first countries to begin a national vaccination programme and one of the few to administer both BNT162b2 and ChAdOx1 concurrently, thus allowing.

## Conclusion

This study found no substantial difference in the rates of COVID-19-related events following vaccination with BNT162b2 or ChAdOx1 in a cohort of health and social care workers in England. Further studies are needed to assess comparative effectiveness for newer, more prevalent variants

### 2021-08-28 create-project.R

@@ -403,8 +403,30 @@ actions_list <- splice(
     ),
     moderately_sensitive = list(
       html = "output/report/effectiveness_report.html",
-      md = "output/report/effectiveness_report.md"#,
-      #figures = "output/report/figures/*.png"
+      md = "output/report/effectiveness_report.md"
+    )
+  ),
+
+  action(
+    name = "rmd_manuscript",
+    run = glue(
+      "r:latest -e {q}",
+      q = single_quote('rmarkdown::render("analysis/report/effectiveness_manuscript.Rmd",  knit_root_dir = "/workspace",  output_dir = "/workspace/output/report", output_format = c("html_document")   )')
+    ),
+    needs = splice(
+      "design", "data_selection",
+      "descr_table1", "descr_irr",
+      "descr_km", "descr_vaxdate",
+      as.list(
+        glue(
+          outcome = c("test", "postest", "emergency", "admitted", "covidadmitted"),
+          "report_{outcome}_timesincevax_plr"
+        )
+      )
+    ),
+    moderately_sensitive = list(
+      html = "output/report/effectiveness_manuscript.html",
+      md = "output/report/effectiveness_manuscript.md"
     )
   ),
 
@@ -435,6 +457,8 @@ actions_list <- splice(
       figures = "output/report/figures_timescales/*.png"
     )
   )
+
+
 
 )
 

### 2021-08-28 analysis/codelists.py
update clinical info using primis codelists; add rurality

@@ -77,107 +77,190 @@ ethnicity_16 = codelist_from_csv(
     category_column="Grouping_16",
 )
 
+# 
+# solid_organ_transplantation = codelist_from_csv(
+#     "codelists/opensafely-solid-organ-transplantation.csv",
+#     system="ctv3",
+#     column="CTV3ID",
+# )
+# 
+# lung_cancer = codelist_from_csv(
+#     "codelists/opensafely-lung-cancer.csv", system="ctv3", column="CTV3ID",
+# )
+# haematological_cancer = codelist_from_csv(
+#     "codelists/opensafely-haematological-cancer.csv", system="ctv3", column="CTV3ID",
+# )
+# bone_marrow_transplant = codelist_from_csv(
+#     "codelists/opensafely-bone-marrow-transplant.csv", system="ctv3", column="CTV3ID",
+# )
+# cystic_fibrosis = codelist_from_csv(
+#     "codelists/opensafely-cystic-fibrosis.csv", system="ctv3", column="CTV3ID",
+# )
+# 
+# sickle_cell_disease = codelist_from_csv(
+#     "codelists/opensafely-sickle-cell-disease.csv", system="ctv3", column="CTV3ID",
+# )
+# 
+# permanent_immunosuppression = codelist_from_csv(
+#     "codelists/opensafely-permanent-immunosuppression.csv",
+#     system="ctv3",
+#     column="CTV3ID",
+# )
+# temporary_immunosuppression = codelist_from_csv(
+#     "codelists/opensafely-temporary-immunosuppression.csv",
+#     system="ctv3",
+#     column="CTV3ID",
+# )
+# chronic_cardiac_disease = codelist_from_csv(
+#     "codelists/opensafely-chronic-cardiac-disease.csv", system="ctv3", column="CTV3ID",
+# )
+# learning_disability = codelist_from_csv(
+#     "codelists/opensafely-learning-disabilities.csv",
+#     system="ctv3",
+#     column="CTV3Code",
+# )
+# downs_syndrome = codelist_from_csv(
+#     "codelists/opensafely-down-syndrome.csv",
+#     system="ctv3",
+#     column="code",
+# )
+# cerebral_palsy = codelist_from_csv(
+#     "codelists/opensafely-cerebral-palsy.csv",
+#     system="ctv3",
+#     column="code",
+# )
+# learning_disability_including_downs_syndrome_and_cerebral_palsy = combine_codelists(
[…]

### 2021-08-28 analysis/descriptive/table1.R

@@ -56,7 +56,7 @@ data_cohort <- read_rds(here("output", "data", "data_cohort.rds"))
 tab_summary_baseline <- data_cohort %>%
   select(
     all_of(names(var_labels)),
-    -age, -stp, -vax1_type, -dialysis
+    -age, -stp, -vax1_type
   ) %>%
   tbl_summary(
     by = vax1_type_descr,

### 2021-08-28 analysis/dummy/data_makedummy.R

@@ -236,30 +236,19 @@ sim_list = list(
     needs = "death_day"
   ),
 
-  chronic_cardiac_disease = bn_node( ~rbernoulli(n=1, p = 0.05)),
-  heart_failure = bn_node( ~rbernoulli(n=1, p = 0.05)),
-  other_heart_disease = bn_node( ~rbernoulli(n=1, p = 0.05)),
-  diabetes = bn_node( ~rbernoulli(n=1, p = 0.05)),
-  chronic_kidney_disease = bn_node( ~rbernoulli(n=1, p = 0.05)),
-  chronic_liver_disease = bn_node( ~rbernoulli(n=1, p = 0.05)),
-  current_copd = bn_node( ~rbernoulli(n=1, p = 0.05)),
-  LD_incl_DS_and_CP = bn_node( ~rbernoulli(n=1, p = 0.05)),
-  cystic_fibrosis = bn_node( ~rbernoulli(n=1, p = 0.05)),
-  other_resp_conditions = bn_node( ~rbernoulli(n=1, p = 0.05)),
-  lung_cancer = bn_node( ~rbernoulli(n=1, p = 0.05)),
-  haematological_cancer = bn_node( ~rbernoulli(n=1, p = 0.05)),
-  cancer_excl_lung_and_haem = bn_node( ~rbernoulli(n=1, p = 0.05)),
-  chemo_or_radio = bn_node( ~rbernoulli(n=1, p = 0.05)),
-  solid_organ_transplantation = bn_node( ~rbernoulli(n=1, p = 0.05)),
-  bone_marrow_transplant = bn_node( ~rbernoulli(n=1, p = 0.05)),
-  sickle_cell_disease = bn_node( ~rbernoulli(n=1, p = 0.05)),
-  permanant_immunosuppression = bn_node( ~rbernoulli(n=1, p = 0.05)),
-  temporary_immunosuppression = bn_node( ~rbernoulli(n=1, p = 0.05)),
-  asplenia = bn_node( ~rbernoulli(n=1, p = 0.05)),
-  dmards = bn_node( ~rbernoulli(n=1, p = 0.05)),
-  dementia = bn_node( ~rbernoulli(n=1, p = 0.05)),
-  other_neuro_conditions = bn_node( ~rbernoulli(n=1, p = 0.05)),
-  psychosis_schiz_bipolar = bn_node( ~rbernoulli(n=1, p = 0.05)),
[…]

### 2021-08-28 analysis/process/data_process.R

@@ -157,21 +157,30 @@ data_processed <- data_extract %>%
       TRUE ~ NA_character_
     ),
 
-    any_immunosuppression = (permanant_immunosuppression | asplenia | dmards | solid_organ_transplantation | sickle_cell_disease | temporary_immunosuppression | bone_marrow_transplant | chemo_or_radio),
+    rural_urban_group = fct_case_when(
+      rural_urban %in% c(1,2) ~ "Urban conurbation",
+      rural_urban %in% c(3,4) ~ "Urban city or town",
+      rural_urban %in% c(5,6,7,8) ~ "Rural town or village",
+      TRUE ~ NA_character_
+    ),
 
     multimorb =
-      (bmi %in% c("Obese II (35-39.9)", "Obese III (40+)")) +
-      (chronic_cardiac_disease | heart_failure | other_heart_disease) +
-      (dialysis) +
+      (sev_obesity) +
+      (chronic_heart_disease) +
+      (chronic_kidney_disease)+
       (diabetes) +
       (chronic_liver_disease)+
-      (current_copd | other_resp_conditions)+
-      (lung_cancer | haematological_cancer | cancer_excl_lung_and_haem)+
-      (any_immunosuppression)+
-      (dementia | other_neuro_conditions)+
-      (LD_incl_DS_and_CP)+
-      (psychosis_schiz_bipolar),
+      (chronic_resp_disease | asthma)+
+      (immunosuppressed | asplenia)+
+      (chronic_neuro_disease)#+
+      #(learndis)+
+      #(sev_mental),
+    ,
     multimorb = cut(multimorb, breaks = c(0, 1, 2, 3, Inf), labels=c("0", "1", "2", "3+"), right=FALSE),
+
+    # clinically at-risk group
+    cv = immunosuppressed | chronic_kidney_disease | chronic_resp_disease | diabetes | chronic_liver_disease |
+      chronic_neuro_disease | chronic_heart_disease | asplenia | learndis | sev_mental,
 
     prior_covid_infection = !is.na(prior_positive_test_date) | !is.na(prior_covidadmitted_date) | !is.na(prior_primary_care_covid_case_date),
 

### 2021-08-28 analysis/process/data_selection.R

@@ -90,7 +90,7 @@ data_flowchart <- data_criteria %>%
     pct_step = n / lag(n),
     crit = str_extract(criteria, "^c\\d+"),
     criteria = fct_case_when(
-      crit == "c0" ~ "All HCWs aged 18-64\n  receiving first dose of BNT162b2 or ChAdOx1\n  between 4 January and 28 February 2021",
+      crit == "c0" ~ "HCWs aged 18-64\n  receiving first dose of BNT162b2 or ChAdOx1\n  between 4 January and 28 February 2021",
       crit == "c1" ~ "  with no missing demographic information",
       crit == "c2" ~ "  who are not clinically extremely vulnerable",
       #crit == "c3" ~ "  with vaccination on or before recruitment end date",

### 2021-08-28 analysis/process/design.R

@@ -35,41 +35,12 @@ write_rds(metadata_outcomes, here("output", "data", "metadata_outcomes.rds"))
 
 formula_exposure <- . ~ . + timesincevax_pw
 #formula_demog <- . ~ . + age + I(age * age) + sex + imd + ethnicity
-formula_demog <- . ~ . + poly(age, degree=2, raw=TRUE) + sex + imd_Q5 + ethnicity_combined
+formula_demog <- . ~ . + poly(age, degree=2, raw=TRUE) + sex + imd_Q5 + ethnicity_combined + rural_urban_group
 formula_comorbs <- . ~ . +
-  bmi +
   prior_covid_infection +
-  #heart_failure +
-  #other_heart_disease +
-
-  #dialysis +
-  #diabetes +
-  #chronic_liver_disease +
-
-  #current_copd +
-  #cystic_fibrosis +
-  #other_resp_conditions +
-
-  #lung_cancer +
-  #haematological_cancer +
-  #cancer_excl_lung_and_haem +
-
-  #chemo_or_radio +
-  #solid_organ_transplantation +
-  #bone_marrow_transplant +
-  #sickle_cell_disease +
-  #permanant_immunosuppression +
-  #temporary_immunosuppression +
-  #asplenia +
-  #dmards +
-  #any_immunosuppression +
-
-  #other_neuro_conditions +
-
-  #LD_incl_DS_and_CP +
-  psychosis_schiz_bipolar +
-
-  multimorb
+  multimorb +
+  learndis +
+  sev_mental
 
 formula_region <- . ~ . + region
 formula_secular <- . ~ . + ns(tstop, df=4)
@@ -111,41 +82,23 @@ var_labels <- list(
   ethnicity_combined ~ "Ethnicity",
   imd_Q5 ~ "IMD",
   region ~ "Region",
+  rural_urban_group ~ "Rural/urban category",
   stp ~ "STP",
   vax1_day ~ "Day of vaccination",
 
-  bmi ~ "Body Mass Index",
+  sev_obesity ~ "Body Mass Index > 40 kg/m^2",
 
-  heart_failure ~ "Heart failure",
-  other_heart_disease ~ "Other heart disease",
-
-  #chronic_kidney_disease ~ "Chronic kidney disease",
+  chronic_heart_disease ~ "Chronic heart disease",
+  chronic_kidney_disease ~ "Chronic kidney disease",
   diabetes ~ "Diabetes",
   chronic_liver_disease ~ "Chronic liver disease",
[…]

### 2021-08-28 analysis/study_definition.py

@@ -725,43 +725,157 @@ study = StudyDefinition(
   ############################################################
   ######### CLINICAL CO-MORBIDITIES ##########################
   ############################################################
-  ### THESE CODELISTS ARE AVAILABLE ON CODELISTS.OPENSAFELY.ORG
-  #### AND CAN BE UPDATED. PLEASE REVIEW AND CHECK YOU ARE HAPPY
-  #### WITH THE CODELIST USED #################################
+  # From PRIMIS
   
   
-  chronic_cardiac_disease=patients.with_these_clinical_events(
-    codelists.chronic_cardiac_disease,
-    on_or_before="covid_vax_any_1_date - 1 days",
-    returning="binary_flag",
-    return_expectations={"incidence": 0.01},
-  ),
-  heart_failure=patients.with_these_clinical_events(
-    codelists.heart_failure,
-    on_or_before="covid_vax_any_1_date - 1 days",
-    returning="binary_flag",
-    return_expectations={"incidence": 0.01},
-  ),
-  other_heart_disease=patients.with_these_clinical_events(
-    codelists.other_heart_disease,
-    on_or_before="covid_vax_any_1_date - 1 days",
-    returning="binary_flag",
-    return_expectations={"incidence": 0.01},
+  asthma = patients.satisfying(
+    """
+      astadm OR
+      (ast AND astrxm1 AND astrxm2 AND astrxm3)
+      """,
+    # Asthma Admission codes
+    astadm=patients.with_these_clinical_events(
+      codelists.astadm,
+      returning="binary_flag",
+      on_or_before="covid_vax_any_1_date - 1 day",
+    ),
+    # Asthma Diagnosis code
+    ast = patients.with_these_clinical_events(
+      codelists.ast,
+      returning="binary_flag",
+      on_or_before="covid_vax_any_1_date - 1 day",
+    ),
+    # Asthma systemic steroid prescription code in month 1
+    astrxm1=patients.with_these_medications(
+      codelists.astrx,
+      returning="binary_flag",
+      between=["covid_vax_any_1_date - 31 days", "covid_vax_any_1_date - 1 days"],
+    ),
+    # Asthma systemic steroid prescription code in month 2
+    astrxm2=patients.with_these_medications(
+      codelists.astrx,
+      returning="binary_flag",
+      between=["covid_vax_any_1_date - 61 days", "covid_vax_any_1_date - 32 days"],
[…]

### 2021-08-28 analysis/models/report_plr.R
improve model matrix terms

@@ -308,10 +308,11 @@ get_HRspline <- function(.data, model, vcov, df){
   ## function to get AZ/pfizer hazard ratio spline over time since first dose
 
   tstop <- .data$tstop
-  spline <- ns(.data$tstop, df=df)
 
-  mat0 <- model.matrix(model, data = mutate(.data, vax1_az=0))
-  mat1 <- model.matrix(model, data = mutate(.data, vax1_az=1))
+  tt <- terms(model) # this helpfully grabs the correct spline basis from the model, rather than recalculating based on `newdata`
+  Terms <- delete.response(tt)
+  mat0 <- model.matrix(Terms, data=mutate(.data, vax1_az=0))
+  mat1 <- model.matrix(Terms, data=mutate(.data, vax1_az=1))
 
   tstop_distinct <- unique(tstop)
 
@@ -523,7 +524,7 @@ survival_pfizer <- ccf(data_plr, vax1_az=0)
 survival_az <- ccf(data_plr, vax1_az=1)
 
 curves <-
-  bind_rows(survival_az, survival_pfizer) %>%
+  bind_rows(survival_pfizer, survival_az) %>%
   mutate(
     model = as.integer(model),
     vax1_az_descr = if_else(vax1_az==1, "ChAdOx1", "BNT162b2")

### 2021-08-28 analysis/descriptive/vax_date.R
better plots

@@ -58,24 +58,31 @@ cumulvax <- data_cohort %>%
 
 plot_stack <-
   ggplot(cumulvax)+
-  geom_area(
+  geom_bar(
     aes(
-      x=vax1_date+1, y=cumuln,
+      x=vax1_date+0.5,
+      y=cumuln,
       group=vax1_type_descr,
       fill=vax1_type_descr
     ),
-    alpha=0.5
+    alpha=0.5,
+    position=position_stack(),
+    stat="identity",
+    width=1
   )+
   scale_x_date(
     breaks = seq(min(cumulvax$vax1_date),max(cumulvax$vax1_date)+1,by=14)+1,
     limits = c(lubridate::floor_date(min(cumulvax$vax1_date), "1 month"), NA),
-    labels = scales::date_format("%d/%m"),
-    expand = expansion(0),
+    labels = scales::label_date("%d/%m"),
+    expand = expansion(add=1),
     sec.axis = sec_axis(
       trans = ~as.Date(.),
       breaks=as.Date(seq(floor_date(min(cumulvax$vax1_date), "month"), ceiling_date(max(cumulvax$vax1_date), "month"),by="month")),
-      labels = scales::date_format("%B %y")
+      labels = scales::label_date("%B %y")
     )
+  )+
+  scale_y_continuous(
+    expand = expansion(c(0, NA))
   )+
   scale_fill_brewer(type="qual", palette="Set1")+
   labs(
@@ -88,7 +95,9 @@ plot_stack <-
   theme_minimal()+
   theme(
     axis.text.x.top=element_text(hjust=0),
-    legend.position = "bottom"
+    axis.ticks.x=element_line(),
+    legend.position = c(0.3,0.8),
+    legend.justification = c(1,0)
   )
 
 
@@ -99,17 +108,19 @@ plot_step <-
       x=vax1_date+1, y=cumuln,
       group=vax1_type_descr,
       colour=vax1_type_descr
-    )
+    ),
+    direction = "vh",
+    size=1
   )+
   scale_x_date(
     breaks = seq(min(cumulvax$vax1_date),max(cumulvax$vax1_date)+1,by=14)+1,
     limits = c(lubridate::floor_date((min(cumulvax$vax1_date)), "1 month"), NA),
-    labels = scales::date_format("%d/%m"),
-    expand = expansion(0),
+    labels = scales::label_date("%d/%m"),
+    expand = expansion(add=1),
     sec.axis = sec_axis(
[…]

### 2021-09-01 analysis/descriptive/table_irr.R
add fortnightly follow-up times

@@ -18,6 +18,7 @@ library('glue')
 library('survival')
 library('gt')
 library('gtsummary')
+library('scales')
 
 ## Import custom user functions from lib
 
@@ -109,8 +110,8 @@ if(removeobs) rm(data_cohort)
 postvax_time <- data_tte %>%
   select(patient_id, tte_censor) %>%
   mutate(
-    fup_day = list(postvaxcuts),
-    timesincevax = map(fup_day, ~droplevels(timesince_cut(.x+1, postvaxcuts)))
+    fup_day = list(postvaxcuts_2week),
+    timesincevax = map(fup_day, ~droplevels(timesince_cut_end(.x+1, postvaxcuts_2week)))
   ) %>%
   unnest(c(fup_day, timesincevax))
 
@@ -318,4 +319,53 @@ tab_summary <- data_summary %>%
 write_rds(tab_summary, here("output", "descriptive", "tables", "table_irr.rds"))
 gtsave(tab_summary, here("output", "descriptive", "tables", "table_irr.html"))
 
+
+tab_summary_simple <-
+  data_summary %>%
+  filter(
+    event %in% c(
+      "postest",
+      "emergency",
+      "covidadmitted",
+      "coviddeath",
+      "death"
+    )
+  ) %>%
+  transmute(
+    outcome_descr, timesincevax,
+    pfizer_q,
+    pfizer_rate_fmt = if_else(pfizer_rate<0.001, "<0.001", number_format(0.001)(pfizer_rate)),
+    pfizer_rate_fmt = if_else(pfizer_rate==0, "0", pfizer_rate_fmt),
+    az_q,
+    az_rate_fmt = if_else(az_rate<0.001, "<0.001", number_format(0.001)(az_rate)),
+    az_rate_fmt = if_else(az_rate==0, "0", az_rate_fmt),
+  ) %>%
+  gt(
+    groupname_col = "outcome_descr",
+  ) %>%
+  cols_label(
+    outcome_descr = "Outcome",
+    timesincevax = "Time since first dose",
+
+    pfizer_q = "BNT162b2\nEvents / person-years",
+    pfizer_rate_fmt = "BNT162b2\nIncidence",
+
+    az_q   = "ChAdOx1\nEvents / person-years",
+    az_rate_fmt = "ChAdOx1\nIncidence",
+  ) %>%
+  fmt_missing(
+    everything(),
+    missing_text="--"
+  ) %>%
+  cols_align(
+    align = "right",
+    columns = everything()
[…]

### 2021-09-01 analysis/lib/utility_functions.R

@@ -134,6 +134,31 @@ timesince_cut <- function(time_since, breaks, prefix=""){
   period
 }
 
+timesince_cut_end <- function(time_since, breaks, prefix=""){
+
+  # this function defines post-vaccination time-periods at `time_since`,
+  # delimited by `breaks`
+
+  # note, intervals are open on the left and closed on the right
+  # so at the exact time point the vaccination occurred, it will be classed as "pre-dose".
+
+  stopifnot("time_since should be strictly non-negative" = time_since>=0)
+  time_since <- as.numeric(time_since)
+  time_since <- if_else(!is.na(time_since), time_since, Inf)
+
+  lab_left <- breaks[-length(breaks)]+1
+  lab_right <- breaks[-1]
+  label <- paste0(lab_left, "-", lab_right)
+  labels <- paste0(prefix, label)
+
+  #labels0 <- cut(c(breaks, Inf), breaks_aug)
+  #labels <- paste0(prefix, c(prelabel, as.character(labels0[-1])))
+  period <- cut(time_since, breaks=breaks, labels=labels, include.lowest=FALSE)
+
+  period
+}
+
+
 # tidy functions for specific objects ----
 
 

### 2021-09-01 analysis/process/design.R

@@ -54,6 +54,7 @@ formula_all_rhsvars <- update(1 ~ 1, formula_exposure) %>%
   update(formula_secular_region)
 
 postvaxcuts <- c(0, 7, 14, 21, 28, 35, 42, 70)
+postvaxcuts_2week <- c(0, 14, 28, 42, 70, 84, 98)
 
 lastfupday <- 14*7
 
@@ -66,6 +67,7 @@ list_formula <- lst(
   formula_secular_region,
   formula_all_rhsvars,
   postvaxcuts,
+  postvaxcuts_2week,
   lastfupday
 )
 

### 2021-09-01 create-project.R

@@ -278,11 +278,11 @@ actions_list <- splice(
     arguments = c("output/data/data_processed.rds", "output/data_properties"),
     needs = list("design", "data_selection"),
     highly_sensitive = list(
-      rds = "output/descriptive/tables/table_irr.rds"
+      rds = "output/descriptive/tables/table_irr*.rds"
     ),
     moderately_sensitive = list(
-      html = "output/descriptive/tables/table_irr.html",
-      csv = "output/descriptive/tables/table_irr.csv"
+      html = "output/descriptive/tables/table_irr*.html",
+      csv = "output/descriptive/tables/table_irr*.csv"
     )
   ),
 

### 2021-09-01 analysis/report/draft-manuscript.Rmd
first rough draft of manuscript

--- 2021-08-28 analysis/report/effectiveness_manuscript.Rmd
@@
 output:
+  word_document: default
   html_document: 
-    self_contained: TRUE
-    keep_md: TRUE
-  github_document:
-    keep_html: TRUE
-  # md_document:
-  #   variant: gfm
+    keep_md: yes
+  bookdown::html_document2:
+    number_sections: false
+    toc: false
+bibliography: references.bib
+csl: https://raw.githubusercontent.com/citation-style-language/styles/master/ieee-transactions-on-medical-imaging.csl
+link-citations: yes
+zotero: true
 ---
@@
 list2env(list_formula, globalenv())
-
 ```
@@
 The COVID-19 global pandemic has prompted the rapid development and delivery of vaccines to combat the disease.
-Following demonstration of high safety and efficacy against symptomatic and severe disease in phase-III randomised clinical trials (RCTs), two vaccines have been approved and widely administered as part of the national vaccination programme in the United Kingdom: the Pfizer-BioNTech BNT162b2 mRNA COVID-19 vaccine (*BNT162b2*) and the Oxford-AstraZeneca ChAdOx1 nCoV-19 vaccine (*ChAdOx1*).
+Following demonstration of high safety and efficacy against symptomatic and severe disease in phase-III randomised clinical trials (RCTs), two vaccines have been approved and widely administered as part of the national vaccination programme in the United Kingdom: the Pfizer-BioNTech BNT162b2 mRNA COVID-19 vaccine @polack2020 (*BNT162b2*) and the Oxford-AstraZeneca ChAdOx1 nCoV-19 vaccine @voysey2021 (*ChAdOx1*).
 Post-authorisation assessment of vaccine effectiveness using observational data is necessary to monitor the success of such programmes as, invariably, target populations and settings differ substantially from those of trials.
 To date, there have been no RCTs that have directly compared the BNT162b2 and ChAdOx1 vaccines to estimate the relative efficacy against COVID-19 infection and disease in the same population.
-The concurrent roll-out of these vaccines across the UK provides an opportunity to emulate such a trial using observational data.
+The concurrent roll-out of these vaccines across the UK, combined with the country's well-developed electronic health record infrastructure, provides a rare opportunity to emulate such a trial using observational data.
 
-COVID-19 vaccination in the UK has been prioritised based on the risk of infection and subsequent severity of disease.
-Patient-facing health and social care workers (HCWs) were amongst the first groups eligible for vaccination due to the high occupational exposure to the SARS-CoV-2 virus, and many were vaccinated during the period where both vaccines were widely used.
+COVID-19 vaccination in the UK has been prioritised based on the risk of infection and subsequent severity of disease .
+Patient-facing health and social care workers (HCWs) were amongst the first groups eligible for vaccination due to the high occupational exposure to the SARS-CoV-2 virus, and many were vaccinated during the period where both vaccines were widely used @collaborative2021.
 This study assesses the comparative effectiveness of one dose of ChAdOx1 with one dose of BNT162b2 in HCWs, using the OpenSAFELY-TPP linked primary care database covering around 40% of England's population.
@@
 The OpenSAFELY-TPP database covers 24 million people registered at GP surgeries that use TPP's SystmOne software.
-This primary care data is linked (via patients NHS numbers) with A&E attendance and in-patient hospital spell records via NHS Digital’s Hospital Episode Statistics (HES), national coronavirus testing records via the Second Generation Surveillance System (SGSS), and national death registry records from the Office for National Statistics (ONS).
+This primary care data is linked (via NHS numbers) with A&E attendance and in-patient hospital spell records via NHS Digital's Hospital Episode Statistics (HES), national coronavirus testing records via the Second Generation Surveillance System (SGSS), and national death registry records from the Office for National Statistics (ONS).
 Vaccination status is available in the GP record directly via the National Immunisation Management System (NIMS).
@@
 
-Three outcomes were defined: positive SARS-CoV-2 swab test, A&E attendances, and COVID-19 hospital admissions.
-
+Three outcomes were defined.
 Positive SARS-CoV-2 tests were identified using SGSS records and based on swab date.
@@
 We did not differentiate between symptomatic and asymptomatic infection.
+A&E attendances were identified using HES emergency care records.
 COVID-19 hospital admissions were identified using HES in-patient hospital records.
-Admissions where the ICD-10 coded primary or non-primary "reason for admission" included U07.1 (“COVID-19, virus identified”) and U07.2 ( “COVID-19, virus not identified”) were included.
+Admissions where the ICD-10 coded primary or non-primary "reason for admission" included U07.1 ("COVID-19, virus identified") and U07.2 ( "COVID-19, virus not identified") were included @emergenc.
 
@@
 
-Patient characteristics used describe the cohort and for confounder adjustment include: age, sex (male or female), English Index of Multiple Deprivation (IMD, grouped by quintiles), ethnicity (Black, Mixed, South Asian, White, Other, as per the UK census), NHS region (East of England, Midlands, London, North East and Yorkshire, North West, South East, South West), number of conditions in the clinically “at risk” (but not clinically extremely vulnerable) classification, as per national prioritisation guidelines, <!--number of previous SARS-CoV-2 tests (via SGSS), rurality,-->evidence of prior SARS-CoV-2 infection (positive test or COVID-19 hospitalisation), learning disabilities, severe mental illness.
+Participant characteristics used describe the cohort and for confounder adjustment include: age, sex (male or female), English Index of Multiple Deprivation (IMD, grouped by quintiles), ethnicity (Black, Mixed, South Asian, White, Other, as per the UK census), NHS region (East of England, Midlands, London, North East and Yorkshire, North West, South East, South West), number of conditions in the clinically "at risk" (but not clinically extremely vulnerable) classification, as per national prioritisation guidelines, <!--number of previous SARS-CoV-2 tests (via SGSS),--> rurality, evidence of prior SARS-CoV-2 infection (positive test or COVID-19 hospitalisation), learning disabilities, severe mental illness.
 All characteristics were ascertained as at the time of vaccination.
@@
 
-We estimated the relative effectiveness of a first dose of ChAdOx1 versus BNT162b2 using pooled logistic regression (PLR) models with the outcome risk estimated at daily intervals.
+We estimated the relative effectiveness of a first dose of ChAdOx1 versus BNT162b2 using pooled logistic regression (PLR) models [@cupples1988a; @ngwa2016] with the outcome risk estimated at daily intervals.
 The effect is permitted to vary by time since vaccination to account for the potential differences in vaccine protection over time between the two brands.
-The PLR models emulate a Cox model with continuous-time-dependent effects, and enables the estimation of risk-adjusted survival rates for each vaccine type using the parametric g-formula.
-This is one minus the average of the cumulative incidence for each day of follow-up predicted by the PLR model, under the (counterfactual) assumption that every person received the BNT162b2 vaccine or that every patient received the ChAdOx1 vaccine.
-Assuming adequate confounder adjustment, this replicates the cumulative incidence that would have been observed in an RCT comparing the two vaccines in the population under consideration.
-Standard errors for the PLR model estimates were obtained using the HC0 clustered sandwich estimator to account for within-person clustering.
+The PLR models approximate Cox models with time-varying effects, and enables the estimation of risk-adjusted survival rates for each vaccine type using the parametric g-formula.
+This is one minus the average of the cumulative incidence for each day of follow-up predicted by the PLR model, under the (counterfactual) assumption that every participant received the BNT162b2 vaccine or that every participant received the ChAdOx1 vaccine.
+Assuming appropriate parameterisation of risk (i.e., adequate confounder adjustment), this replicates the cumulative incidence that would have been observed in an RCT comparing the two vaccines in the population under consideration.
+Standard errors for the PLR model were obtained using the HC0 clustered sandwich estimator to account for within-participant clustering.
 Confidence intervals for the risk-adjusted cumulative incidence were obtained using the delta method (i.e, a first-order Taylor series approximation of the variance).
@@
 The models are weighted to recover the characteristics of the complete cohort.
-This weighting is also applied to the risk-adjusted survival curves.
 Sampling was based on a ranked hash of the person identifier to ensure the sampled non-events overlapped as much as possible across outcomes.
@@
 
-Data management and analyses were conducted in Python 3.8 and R version 4.0.2. All code is available for review and reuse at https://github.com/opensafely/comparative-ve-research.
+Data management and analyses were conducted in Python 3.8 and R version 4.0.2.
+All code is available for review and reuse at <https://github.com/opensafely/comparative-ve-research>.
+No person-level data is shared.
+Any reported figures based on counts below 6 are redacted or rounded for disclosure control.
 
 # Results
-
-Note all counts below 6 are redacted and survival estimates are rounded for disclosure control.
 
@@
     priorinfection = mean(prior_covid_infection),
+    median_vaxday = first(start_date_az) + floor(median(vax1_day)) -1
   )
@@
     priorinfection = mean(prior_covid_infection),
+    median_vaxday = first(start_date_az) + floor(median(vax1_day)) - 1
   )
-
 ```
 
-A total of `r flowchart[1, "n"]` HCWs aged 18-64 receiving a first dose of BNT162b2 or ChAdOx1 between 4 January and 28 February 2021 and actively registered at at TPP practice were identified, with `r flowchart[[nrow(flowchart), "n"]]` (`r `label_percent(0.1)(flowchart[[nrow(flowchart), "pct_all"]])`) meeting the study eligibility criteria.
-`r baseline_pfizer$n` (`r `label_percent(0.1)(baseline_pfizer$n/baseline$n)`) were vaccinated with BNT162b2, contributing `r baseline_pfizer$fu_years` person-years of follow-up.
-For ChAdOx1, this was `r baseline_az$n` (`r label_percent(0.1)(baseline_az$n/baseline$n)`) and `r baseline_az$fu_years` person-years.
-Baseline characteristics are reported in Table 1.
+A total of `r flowchart[1, "n"]` HCWs aged 18-64 receiving a first dose of BNT162b2 or ChAdOx1 between 4 January and 28 February 2021 and actively registered at at TPP practice were identified, with `r flowchart[[nrow(flowchart), "n"]]` (`r label_percent(0.1)(flowchart[[nrow(flowchart), "pct_all"]])`) meeting the study eligibility criteria.
+`r baseline_pfizer$n` (`r label_percent(0.1)(baseline_pfizer$n/baseline$n)`) were vaccinated with BNT162b2, contributing `r label_number(0.1)(baseline_pfizer$fu_years)` person-years of follow-up.
+For ChAdOx1, this was `r baseline_az$n` (`r label_percent(0.1)(baseline_az$n/baseline$n)`) and `r label_number(0.1)(baseline_az$fu_years)` person-years.
+
+Baseline characteristics are reported in Table 1 and were largely well-balanced between recipients of each vaccine, though regional and temporal differences in the distribution of each vaccine are notable.
+BNT162b2 was on average administered earlier than ChAdOx1 (median day of vaccination `r format(baseline_pfizer$median_vaxday, "%d %B")` for BNT162b2, `r format(baseline_az$median_vaxday, "%d %B")` for ChadOx1).
+BNT162b2 was relatively more likely to be administered in the South and East of England, and ChAdOx1 the Midlands and Northern England.
+The proportion of each clinical condition is slightly higher in ChAdOx1 recipients, though consistently under a 0.6% percent-point difference.
+
 The cumulative proportion of vaccinated study participants by calendar time is reported in Figure 1.
 
-### Table 0: Inclusion criteria
+#### Table 0: Inclusion criteria
+
 ```{r flowchart, echo=FALSE, warning=FALSE, message=FALSE}
@@
   )
-
 ```
 
-### Table 1: Baseline characteristics
+#### Table 1: Baseline characteristics
 
-```{r table1, echo=FALSE}
+```{r table1, echo=FALSE, warning=FALSE, message=FALSE}
 read_rds(here("output", "descriptive", "tables", "table1.rds"))
@@
 
-### Figure 1: Vaccination dates
+#### Figure 1: Vaccination dates
 
@@
 
-### Figure 2: Event rates
+#### Table 2: Event rates
 
-```{r table_irr, echo=FALSE}
-read_rds(here("output", "descriptive", "tables", "table_irr.rds"))
+```{r table_irr, echo=FALSE, message=FALSE, warning=FALSE}
+read_rds(here("output", "descriptive", "tables", "table_irr_simple.rds"))
 ```
@@
 
-Figure 3 shows, for each outcome based on the fully adjusted model, the time-dependent hazard ratio for ChAdOx1 versus BNT162b2, and one minus the risk-adjusted cumulative incidence<!--, and the absolute risk difference-->.
-Similar figures for the non-fully-adjusted models are presented in supplementary materials.
-
-```{r, effects, echo=FALSE, message=FALSE, warning=FALSE, out.width='90%', fig.height=6, results='asis'}
+```{r, effects, echo=FALSE, message=FALSE, warning=FALSE}
 
@@
 
-  patchwork::wrap_plots(plot_cmlinc, plot_hazard, ncol=1)
-
   riskdifference <- 
@@
     mutate(
-      diff = survival - first(survival),
+      diff = (1-survival) - first((1-survival)),
       diff_format = label_number(0.01, scale=1000)(diff)
@@
 
+  plot_riskdiff <-
+    ggplot(riskdifference)+
+    geom_hline(aes(yintercept=0), colour='black')+
+    geom_step(aes(x=tstop, y=diff*1000), colour='orchid4')+
+    # geom_rect(aes(xmin=lag(tstop, 1, 0), xmax=tstop, ymin=diff.ll*1000, ymax=diff.ul*1000), alpha=0.1, colour="transparent", fill='orchid4')+
+    facet_grid(cols=vars(outcome_descr))+
+    scale_x_continuous(
+      limits = c(0, NA),
+      breaks = seq(0,7*52,by=14),
+      expand = expansion(0)
+    )+
+    scale_y_continuous(
+      expand = expansion(0),
+      sec.axis = dup_axis(name="Lower risk after \n <- BNT162b2 / ChAdOx1 ->", breaks = NULL)
+    )+
+    scale_colour_brewer(type="qual", palette="Set1", na.value="grey")+
+    scale_fill_brewer(type="qual", palette="Set1", guide="none", na.value="grey")+
+    labs(
+      x=NULL,
+      y="Risk difference\nper 1000",
+      colour=NULL
+    )+
+    theme_bw()+
+    theme(
+      panel.border = element_blank(),
+      axis.line.y.left = element_line(colour = "black"),
+
+      panel.grid.minor.x = element_blank(),
+      panel.grid.minor.y = element_blank(),
+      strip.background = element_blank(),
+      strip.text = element_blank(),
+
+      panel.spacing = unit(0.8, "lines"),
+
+      plot.title = element_text(hjust = 0),
+      plot.title.position = "plot",
+      plot.caption.position = "plot",
+      plot.caption = element_text(hjust = 0, face= "italic")
+
+     ) +
+   NULL
+
+```
+
+Figure 3 shows, for each outcome based on the fully adjusted model, the time-dependent hazard ratio for ChAdOx1 versus BNT162b2, and one minus the risk-adjusted cumulative incidence, and the absolute risk difference.
+Similar figures for the non-fully-adjusted models are presented in supplementary materials.
+
+At 6 weeks post-vaccination, the absolute risk difference (ChAdOx1 - BNT162b2) for a positive SARS-CoV-2 test was `r filter(riskdifference, outcome=="postest", tstop==7*6)$diff_format` per 1000 people, and for COVID-19 hospital admissions was `r filter(riskdifference, outcome=="covidadmitted", tstop==7*6)$diff_format` per 1000 people.
+
+#### Figure 2: Comparative effectivevess
+
+```{r comp,echo=FALSE, message=FALSE, out.width='90%', fig.height=8, results='asis'}
+patchwork::wrap_plots(plot_cmlinc, plot_riskdiff, plot_hazard, ncol=1)
 ```
@@
 
-This observational study of `r baseline$n` health and social care workers living in England found no substantial differences in rates of COVID-19-related events between those receiving a first dose of BNT162b2 or ChAdOx1.
-At 6 weeks post-vaccination, the absolute risk difference (ChAdOx1 - BNT162b2) for a positive SARS-CoV-2 test was `r filter(riskdifference, outcome=="postest", tstop==7*6)$diff_format` per 1000 people, and for COVID-19 hospital admissions was `r filter(riskdifference, outcome=="covidadmitted", tstop==7*6)$diff_format` per 1000 people.
+This observational study of almost 1/3 million health and social care workers living in England found no significant differences in rates of COVID-19-related events between those receiving a first dose of BNT162b2 or ChAdOx1.
+For hospitalisation, it was not possible to rule-out large differences in the risk due to uncertainty in the effect estimates.
 There is a clear leveling-off of the event rates after around 3 weeks, consistent with the expected time-to-onset of vaccine-induced immunity of around 2 weeks plus the delay from infection to a positive test.
@@
 
-This study used routinely-collected health records with comprehensive coverage of primary care, hospital admissions, COVID-19 testing, COVID-19 vaccination, and death registrations for over 300,000 health and social care workers living in England.
-This group were eligible for vaccination at the start of the UK's vaccination programme due to exposure to higher viral loads and the need to reduce enforced absences in frontline healthcare staff during a global pandemic.
-They are the only early vaccinees who are relatively young and healthy, and were vaccinated during a period where infection rates were high, thus providing one of the few opportunities to study healthy people at a time when event rates are sufficiently high to make meaningful comparisons between vaccines.
+We used routinely-collected health records with comprehensive coverage of primary care, hospital admissions, COVID-19 testing, COVID-19 vaccination, and death registrations to study vaccinated health and social care workers.
+This group were eligible for vaccination at the start of the UK's vaccination programme due to exposure to higher viral loads and the need to reduce enforced absences in essential healthcare workers during a global pandemic.
+They are the only early vaccinees who are relatively young and healthy, and were vaccinated during a period where infection rates were high and both vaccines were being administered.
+This provides a rare opportunity to study comparative effectiveness with sufficient power under conditions that, to some extent, approximate random vaccine allocation.
+However, some limitations remain.
 
-We were unable to study differences in the vaccines ability to reduce transmission, nor fully investigate differences in protection against severe disease.
-This latter limitation is thanks to clear protective benefits of both vaccines, reducing the absolute numbers of events, and therefore statistical power, in the studied cohort.
+We were unable to fully investigate differences in protection against severe disease, in large part thanks to clear protective benefits of both vaccines, reducing the absolute numbers of events, and therefore statistical power, in the studied cohort.
 
-Despite adjustment for a range of potential confounders and the fairly arbitrary allocation of vaccine type, the possibility of unmeasured confounding remains.
+Despite reasonable balance of vaccine allocation across baseline characteristics and adjustment for a range of potential confounders, the possibility of unmeasured confounding remains.
 The cold-chain storage requirements of BNT162b2 meant that it is more likely to have be administered in Acute NHS Trusts and other large vaccination centres, and is thus a potential confounder for instance due to differences in exposure to risk across these settings.
-Though we adjusted for patient rurality, we were unable to directly account for occupational differences.
+Though we adjusted for region, rurality, and deprivation, we were unable to directly account for occupational differences that may affect exposure risk and vaccine type.
+
+The dominant circulating variant during the period of study was the Alpha variant, which has since been supplanted by other strains, in particular the Delta variant which we did not consider here.
 
@@
 
-To our knowledge, this is first study to assess effectiveness of BNT162b2 and ChAdOx1 vaccines in a head-to-head comparison.
+A number of studies estimate COVID-19 vaccine effectiveness in observational data using unvaccinated controls [@cabezas2021; @vasileiou2021]. [describe pfizer versus AZ findings]
 
-A number of studies estimate COVID-19 vaccine effectiveness in observational data using unvaccinated controls.
-Inferring comparative effectiveness from these studies is possible by calculating the ratio of 1 minus the vaccine effectiveness for each vaccine.
+Such designs are extremely vulnerable to confounding for example due to vaccine prioritisation and eligiblity policies, as well as vaccine access and acceptance, that cause substantial imbalance between vaccinated and unvaccinated groups.
+Many of these biases can be bypassed when making direct comparisons between recipients of different vaccine types.
+To our knowledge, the present study is the first to assess effectiveness of the BNT162b2 and ChAdOx1 vaccines in a head-to-head comparison.
+We found...
 
-* EAVE-II estimated VE in 18-64 age group in 5th week to be 92% for BNT162b2 and 100% for ChAdOx1, = HR=0.00
-* EAVE-II estimated VE in 65-79 age group in 4th week to be 91% for BNT162b2 and 68% for ChAdOx1, = HR=3.56
-* EAVE-II estimated VE in 80+ age group in 4th week to be 88% for BNT162b2 and 81% for ChAdOx1, = HR=1.58
-
-are extremely vulnerable to confounding for example due to vaccine prioritisation and eligiblity policies.
-However, many of these biases can be bypassed when making direct comparisons between recipients of different vaccine types.
-The United Kingdom was amongst the first countries to begin a national vaccination programme and one of the few to administer both BNT162b2 and ChAdOx1 concurrently, thus allowing.
+<!--# Inferring comparative effectiveness from these studies is possible by calculating the ratio of 1 minus the vaccine effectiveness for each vaccine: EAVE-II estimated VE in 18-64 age group in 5th week to be 92% for BNT162b2 and 100% for ChAdOx1, = HR=0.00; EAVE-II estimated VE in 65-79 age group in 4th week to be 91% for BNT162b2 and 68% for ChAdOx1, = HR=3.56; EAVE-II estimated VE in 80+ age group in 4th week to be 88% for BNT162b2 and 81% for ChAdOx1, = HR=1.58.-->
 
@@
 This study found no substantial difference in the rates of COVID-19-related events following vaccination with BNT162b2 or ChAdOx1 in a cohort of health and social care workers in England.
-Further studies are needed to assess comparative effectiveness for newer, more prevalent variants
+Further studies are needed to assess comparative effectiveness in newer, more prevalent variants.
+
+## References

### 2021-09-01 create-project.R

@@ -411,7 +411,7 @@ actions_list <- splice(
     name = "rmd_manuscript",
     run = glue(
       "r:latest -e {q}",
-      q = single_quote('rmarkdown::render("analysis/report/effectiveness_manuscript.Rmd",  knit_root_dir = "/workspace",  output_dir = "/workspace/output/report", output_format = c("html_document")   )')
+      q = single_quote('rmarkdown::render("analysis/report/draft-manuscript.Rmd",  knit_root_dir = "/workspace",  output_dir = "/workspace/output/report", output_format = c("html_document")   )')
     ),
     needs = splice(
       "design", "data_selection",

### 2021-09-01 analysis/report/effectiveness_report_comparetimescales.Rmd
simplify timescale comparison report

@@
 
-# Introduction
+### Time-dependent Cox models
 
-This study compares the effectiveness of the *BNT162b2* mRNA (Pfizer-BioNTech) and *ChAdOx1* (Oxford-AstraZeneca) vaccines amongst health and social care workers (HCWs). 
-This group was chosen as they were prioritised for vaccination due to high occupational exposure and during the period where both vaccines were available (`r format(as.Date(gbl_vars$start_date_az), "%d %B %Y")` onwards) it is expected that the actual vaccine brand received will not be strongly influenced by other determinants of COVID-19-related outcomes (i.e., approximately random treatment allocation and therefore reduced confounding).
-
-The code and data for this report can be found at the OpenSafely [comparative-ve-research GitHub repository](https://github.com/opensafely/comparative-ve-research).
-
-# Methods
-
-## Study population
-
-People meeting the following criteria are included:
-
-* HCWs who have received at least one dose of BNT162b2 or ChAdOx1. 
-* Registered at a GP practice using TPP's SystmOne clinical information system on the day of vaccination.
-* Aged 18-65.
-* Vaccinated on or after `r format(as.Date(gbl_vars$start_date_az), "%d %B %Y")`, when both vaccine brands were being administered. 
-* Not Clinically Extremely Vulnerable (CEV), as set out by government guidance, at the time of vaccination.
-
-Study participants are followed up from vaccination date until the first of:
-
-* The outcome of interest
-* Death or deregistration
-* Second vaccine dose
-* The study end date, `r format(as.Date(gbl_vars$end_date), "%d %B %Y")`
-
-### Identifying vaccinated HCWs
-
-Those vaccinated as part of England's national vaccination programme (for example not including vaccine trial participants) are asked whether they work in health or social care.
-This information is sent to OpenSAFELY-TPP from NHS Digital's COVID-19 data store.
-
-Note -- many of those flagged as HCWs do not have a vaccination record, which shouldn't be the case if the question was asked as part of the vaccination process.
-
-## Study measures
-
-### Exposure
-
-The vaccination brand, BNT162b2 or ChAdOx1. This is available in the GP record directly, via the National Immunisation Management System (NIMS).
-
-### Outcomes
-
-* SARS-CoV-2 infection, as identified via SGSS records. Both PCR and LFT tests are included.
-* Unplanned COVID-19-related hospitalisation via HES records. ICD-10 codes are used to identify COVID-19-related admissions. 
-* Unplanned COVID-19-related ICU admission via HES records. ICD-10 codes are used to identify COVID-19-related admissions. 
-* COVID-19-related death via death registration records. With COVID-19 ICD-10 codes anywhere on the death certificate (i.e., as an underlying or contributing cause of death)
-
-## Statistical Analysis
-
-The aim is to estimate comparative vaccine effectiveness, i.e., the relative hazard of each outcome for ChAdOx1 versus BNT162b2 vaccine recipients.
-The effect is permitted to vary by time since vaccination, to account for the potential differences in vaccine protection over time between the two brands.
-
-Patient characteristics used for adjustment include: age, sex, deprivation, ethnicity, NHS region, clinically “at risk” (but not clinically extremely vulnerable) as per national prioritisation guidelines, asthma, number of previous SARS-CoV-2 tests (via SGSS), rurality, evidence of prior covid infection (positive test or COVID-19 hospitalisation), number of recorded comorbidities, severe mental illness.
-All variables are ascertained as at the time of vaccination.
-
-NHS regions are used as a stratification variable to account for geographical variation in event rates for the outcomes of interest (primarily driven by changes in local infection rates).
-
-### Time-dependent Cox models
 Time-dependent Cox models are used to estimate the time-varying effects for each vaccine type. Two time-scales are used for triangulation:
 
-* On the *calendar time-scale*, time zero is `r format(as.Date(gbl_vars$start_date_az), "%d %B %Y")`, and people only contribute follow-up time once they have been vaccinated (i.e., delayed study entry).
+-   On the *calendar time-scale*, time zero is `r format(as.Date(gbl_vars$start_date_az), "%d %B %Y")`, and people only contribute follow-up time once they have been vaccinated (i.e., delayed study entry).
 Vaccine type, time since vaccination in weekly intervals, and the interaction between these two terms are included to estimate time-varying vaccine effects for each vaccine type.
@@
 
-* On the *treatment time-scale*, time zero is the date of vaccination.
+-   On the *treatment time-scale*, time zero is the date of vaccination.
 Date of vaccination is included as an additional baseline covariate using a restricted cubic spline with knots at the first and second tertile, plus boundary knots.
@@
 
-## Flowchart
-
-```{r flowchart, echo=FALSE, warning=FALSE, message=FALSE}
-
-flowchart <- read_csv(here("output", "data", "flowchart.csv"))
-
-flowchart %>%
-  select(
-    Criteria = criteria,
-    N = n,
-    `N excluded` = n_exclude,
-    `% excluded` = pct_exclude,
-    `% remaining` = pct_all
-  ) %>%
-  gt() %>%
-  fmt_percent(
-    columns = starts_with(c("%")),
-    decimals = 1
-  ) %>%
-  fmt_number(
-    columns = starts_with(c("N")),
-    decimals= 0
-  ) %>%
-  fmt_missing(
-    everything(),
-    missing_text="-"
-  )
-
-```
-
-## Baseline demographics
-
-```{r table1, echo=FALSE}
-read_rds(here("output", "descriptive", "tables", "table1.rds"))
-```
-
 ## Event rates
@@
 ## Comparative effectiveness
-
-The plots below show:
-
-* Unadjusted Kaplan-Meier curves
-* *ChAdOx1* versus *BNT162b2* piecewise-linear hazard ratios estimated using the Cox models
-* *ChAdOx1* versus *BNT162b2* continuous hazard ratios estimated using the PLR models
-* Marginalised cumulative risk for *ChAdOx1* and *BNT162b2* estimated using the PLR models
 

### 2021-09-01 analysis/report/draft-manuscript.Rmd
update csl template; tweaks

@@
 output:
-  word_document: default
   html_document: 
     keep_md: yes
+  word_document: default
   bookdown::html_document2:
@@
 bibliography: references.bib
-csl: https://raw.githubusercontent.com/citation-style-language/styles/master/ieee-transactions-on-medical-imaging.csl
+csl: nature.csl # from https://raw.githubusercontent.com/citation-style-language/styles/master/nature.csl
 link-citations: yes
@@
 
-Baseline characteristics are reported in Table 1 and were largely well-balanced between recipients of each vaccine, though regional and temporal differences in the distribution of each vaccine are notable.
-BNT162b2 was on average administered earlier than ChAdOx1 (median day of vaccination `r format(baseline_pfizer$median_vaxday, "%d %B")` for BNT162b2, `r format(baseline_az$median_vaxday, "%d %B")` for ChadOx1).
-BNT162b2 was relatively more likely to be administered in the South and East of England, and ChAdOx1 the Midlands and Northern England.
-The proportion of each clinical condition is slightly higher in ChAdOx1 recipients, though consistently under a 0.6% percent-point difference.
-
-The cumulative proportion of vaccinated study participants by calendar time is reported in Figure 1.
-
 #### Table 0: Inclusion criteria
@@
 
+Baseline characteristics are reported in Table 1 and were largely well-balanced between recipients of each vaccine, though regional and temporal differences in the distribution of each vaccine are notable.
+BNT162b2 was on average administered earlier than ChAdOx1 (median day of vaccination `r format(baseline_pfizer$median_vaxday, "%d %B")` for BNT162b2, `r format(baseline_az$median_vaxday, "%d %B")` for ChadOx1).
+BNT162b2 was relatively more likely to be administered in the South and East of England, and ChAdOx1 the Midlands and Northern England.
+The proportion of each clinical condition is slightly higher in ChAdOx1 recipients, though consistently under a 0.6% percent-point difference.
+
 #### Table 1: Baseline characteristics
@@
 ```
+
+The cumulative proportion of vaccinated study participants by calendar time is reported in Figure 1.
 
@@
 
-Figure 2 shows event rates by vaccine type, and the unadjusted incidence ratio ratio.
-Over the duration of `r   label_number(0.1)(baseline$fu_years)` person-years of follow-up there were `r baseline$postest` positive SARS-CoV-2 tests, `r baseline$covidadmitted` COVID-19 hospital admissions, and `r baseline$coviddeath` COVID-19 deaths.
+Table 2 shows event rates by vaccine type, and the crude incidence.
+Over the duration of `r label_number(0.1)(baseline$fu_years)` person-years of follow-up there were `r baseline$postest` positive SARS-CoV-2 tests, `r baseline$emergency` A&E attendances, `r baseline$covidadmitted` COVID-19 hospital admissions, and `r baseline$coviddeath` COVID-19 deaths.
 
@@
 
-At 6 weeks post-vaccination, the absolute risk difference (ChAdOx1 - BNT162b2) for a positive SARS-CoV-2 test was `r filter(riskdifference, outcome=="postest", tstop==7*6)$diff_format` per 1000 people, and for COVID-19 hospital admissions was `r filter(riskdifference, outcome=="covidadmitted", tstop==7*6)$diff_format` per 1000 people.
+At 6 weeks post-vaccination, the absolute risk difference (ChAdOx1 - BNT162b2) for a positive SARS-CoV-2 test was `r filter(riskdifference, outcome=="postest", tstop==7*6)$diff_format` per 1000 people, for A&E attendances was `r filter(riskdifference, outcome=="emergency", tstop==7*6)$diff_format` and for COVID-19 hospital admissions was `r filter(riskdifference, outcome=="covidadmitted", tstop==7*6)$diff_format` per 1000 people.
 

### 2021-09-01 analysis/report/draft-manuscript.Rmd
update title

@@
 ---
-title: "Comparative vaccine effectiveness (ChAdOx1 versus BNT162b2) in Health and Social Care workers"
+title: "Comparative vaccine effectiveness of ChAdOx1 versus BNT162b2 in Health and Social Care workers in England"
 output:

### 2021-09-01 analysis/report/draft-manuscript.Rmd
self-contained render

--- 2021-09-01 analysis/report/draft-manuscript.Rmd
@@
     keep_md: yes
+    self_contained: TRUE
   word_document: default

### 2021-09-01 create-project.R
update filename

@@ -425,8 +425,8 @@ actions_list <- splice(
       )
     ),
     moderately_sensitive = list(
-      html = "output/report/effectiveness_manuscript.html",
-      md = "output/report/effectiveness_manuscript.md"
+      html = "output/report/draft-manuscript.html",
+      md = "output/report/draft-manuscript.md"
     )
   ),
 

### 2021-09-01 analysis/report/draft-manuscript.Rmd
temporarily remove citations

--- 2021-09-01 analysis/report/draft-manuscript.Rmd
@@
     toc: false
-bibliography: references.bib
-csl: nature.csl # from https://raw.githubusercontent.com/citation-style-language/styles/master/nature.csl
-link-citations: yes
-zotero: true
+#bibliography: references.bib
+#csl: nature.csl # from https://raw.githubusercontent.com/citation-style-language/styles/master/nature.csl
+#link-citations: yes
+#zotero: true
 ---
@@
 The COVID-19 global pandemic has prompted the rapid development and delivery of vaccines to combat the disease.
-Following demonstration of high safety and efficacy against symptomatic and severe disease in phase-III randomised clinical trials (RCTs), two vaccines have been approved and widely administered as part of the national vaccination programme in the United Kingdom: the Pfizer-BioNTech BNT162b2 mRNA COVID-19 vaccine @polack2020 (*BNT162b2*) and the Oxford-AstraZeneca ChAdOx1 nCoV-19 vaccine @voysey2021 (*ChAdOx1*).
+Following demonstration of high safety and efficacy against symptomatic and severe disease in phase-III randomised clinical trials (RCTs), two vaccines have been approved and widely administered as part of the national vaccination programme in the United Kingdom: the Pfizer-BioNTech BNT162b2 mRNA COVID-19 vaccine <!--# @polack2020 --> (*BNT162b2*) and the Oxford-AstraZeneca ChAdOx1 nCoV-19 vaccine <!--# @voysey2021 --> (*ChAdOx1*).
 Post-authorisation assessment of vaccine effectiveness using observational data is necessary to monitor the success of such programmes as, invariably, target populations and settings differ substantially from those of trials.
@@
 COVID-19 vaccination in the UK has been prioritised based on the risk of infection and subsequent severity of disease .
-Patient-facing health and social care workers (HCWs) were amongst the first groups eligible for vaccination due to the high occupational exposure to the SARS-CoV-2 virus, and many were vaccinated during the period where both vaccines were widely used @collaborative2021.
+Patient-facing health and social care workers (HCWs) were amongst the first groups eligible for vaccination due to the high occupational exposure to the SARS-CoV-2 virus, and many were vaccinated during the period where both vaccines were widely used <!--#  @collaborative2021 -->.
 This study assesses the comparative effectiveness of one dose of ChAdOx1 with one dose of BNT162b2 in HCWs, using the OpenSAFELY-TPP linked primary care database covering around 40% of England's population.
@@
 COVID-19 hospital admissions were identified using HES in-patient hospital records.
-Admissions where the ICD-10 coded primary or non-primary "reason for admission" included U07.1 ("COVID-19, virus identified") and U07.2 ( "COVID-19, virus not identified") were included @emergenc.
+Admissions where the ICD-10 coded primary or non-primary "reason for admission" included U07.1 ("COVID-19, virus identified") and U07.2 ( "COVID-19, virus not identified") were included <!--# @emergenc --> .
 
@@
 
-We estimated the relative effectiveness of a first dose of ChAdOx1 versus BNT162b2 using pooled logistic regression (PLR) models [@cupples1988a; @ngwa2016] with the outcome risk estimated at daily intervals.
+We estimated the relative effectiveness of a first dose of ChAdOx1 versus BNT162b2 using pooled logistic regression (PLR) models <!--# [@cupples1988a; @ngwa2016] --> with the outcome risk estimated at daily intervals.
 The effect is permitted to vary by time since vaccination to account for the potential differences in vaccine protection over time between the two brands.
@@
     ) %>%
-    filter(vax1_az==1)
+    filter(vax1_az==1) %>%
+    left_join(
+      hazard %>% select(outcome, model, tstop, hr, hr.ll, hr.ul), 
+      by=c("outcome", "model", "tstop")
+    ) %>% 
+    mutate(
+      diff.ll = (1-survival) - (1-survival)*hr.ll,
+      diff.ul = (1-survival) + (1-survival)*hr.ul
+    ) #-(surv-lag(surv))/lag(surv)
 
@@
     geom_step(aes(x=tstop, y=diff*1000), colour='orchid4')+
-    # geom_rect(aes(xmin=lag(tstop, 1, 0), xmax=tstop, ymin=diff.ll*1000, ymax=diff.ul*1000), alpha=0.1, colour="transparent", fill='orchid4')+
+    geom_rect(aes(xmin=lag(tstop, 1, 0), xmax=tstop, ymin=diff.ll*1000, ymax=diff.ul*1000), alpha=0.1, colour="transparent", fill='orchid4')+
     facet_grid(cols=vars(outcome_descr))+
@@
 
-A number of studies estimate COVID-19 vaccine effectiveness in observational data using unvaccinated controls [@cabezas2021; @vasileiou2021]. [describe pfizer versus AZ findings]
+A number of studies estimate COVID-19 vaccine effectiveness in observational data using unvaccinated controls <!--# [@cabezas2021; @vasileiou2021]  -->. [describe pfizer versus AZ findings]
 

### 2021-09-01 analysis/report/draft-manuscript.Rmd
put limits on hazard and risk diff plots

@@
     scale_y_log10(
+      limits = c(0.1, 10),
       #breaks=c(0.125, 0.167, 0.25, 0.333, 0.5, 0.667, 1, 1.5, 2, 3, 4, 6, 8),
@@
     scale_y_continuous(
+      limits = c(-5, 5),
       expand = expansion(0),

### 2021-09-01 analysis/report/draft-manuscript.Rmd
remove dodgy CIs for risk difference

@@
     ) %>% 
+    group_by(model, model_name, outcome, outcome_descr) %>%
+    arrange(model, outcome, vax1_az, tstop) %>%
     mutate(
+      hazard1 = -(survival - lag(survival, 1, 1)) / lag(survival, 1, 1),
+      hazard0 = hazard1/hr,
+      hazard0.ll = hazard1/hr.ul,
+      hazard0.ul = hazard1/hr.ll,
+
+      survival0 = exp(-cumsum(if_else(is.na(hazard0), 0, hazard0))),
+      survival0.ll = exp(-cumsum(if_else(is.na(hazard0.ul), 0, hazard0.ul))),
+      survival0.ul = exp(-cumsum(if_else(is.na(hazard0.ll), 0, hazard0.ul))),
+
+      diff2 = (1-survival)-(1-survival0),
+
       diff.ll = (1-survival) - (1-survival)*hr.ll,
-      diff.ul = (1-survival) + (1-survival)*hr.ul
-    ) #-(surv-lag(surv))/lag(surv)
+      diff.ul = (1-survival) + (1-survival)*hr.ul,
+      #diff_CI = paste0(label_number(0.01, scale=1000)(diff), " (", diff.ll, " ,", diff.ul, ")")
+    )
 
@@
     geom_step(aes(x=tstop, y=diff*1000), colour='orchid4')+
-    geom_rect(aes(xmin=lag(tstop, 1, 0), xmax=tstop, ymin=diff.ll*1000, ymax=diff.ul*1000), alpha=0.1, colour="transparent", fill='orchid4')+
+    # geom_rect(aes(xmin=lag(tstop, 1, 0), xmax=tstop, ymin=diff.ll*1000, ymax=diff.ul*1000), alpha=0.1, colour="transparent", fill='orchid4')+
     facet_grid(cols=vars(outcome_descr))+
@@
 
-```{r comp,echo=FALSE, message=FALSE, out.width='90%', fig.height=8, results='asis'}
+```{r comp,echo=FALSE, message=FALSE, warnings=FALSE, out.width='90%', fig.height=8, results='asis'}
 patchwork::wrap_plots(plot_cmlinc, plot_riskdiff, plot_hazard, ncol=1)

### 2021-09-01 analysis/report/draft-manuscript.Rmd
typo

@@
 
-```{r comp,echo=FALSE, message=FALSE, warnings=FALSE, out.width='90%', fig.height=8, results='asis'}
+```{r comp,echo=FALSE, message=FALSE, warning=FALSE, out.width='90%', fig.height=8, results='asis'}
 patchwork::wrap_plots(plot_cmlinc, plot_riskdiff, plot_hazard, ncol=1)

### 2021-09-01 analysis/report/draft-manuscript.Rmd
formatting

@@
 
-A total of `r flowchart[1, "n"]` HCWs aged 18-64 receiving a first dose of BNT162b2 or ChAdOx1 between 4 January and 28 February 2021 and actively registered at at TPP practice were identified, with `r flowchart[[nrow(flowchart), "n"]]` (`r label_percent(0.1)(flowchart[[nrow(flowchart), "pct_all"]])`) meeting the study eligibility criteria.
-`r baseline_pfizer$n` (`r label_percent(0.1)(baseline_pfizer$n/baseline$n)`) were vaccinated with BNT162b2, contributing `r label_number(0.1)(baseline_pfizer$fu_years)` person-years of follow-up.
-For ChAdOx1, this was `r baseline_az$n` (`r label_percent(0.1)(baseline_az$n/baseline$n)`) and `r label_number(0.1)(baseline_az$fu_years)` person-years.
+A total of `r label_number(1, big.mark=",")(flowchart[1, "n"])` HCWs aged 18-64 receiving a first dose of BNT162b2 or ChAdOx1 between 4 January and 28 February 2021 and actively registered at at TPP practice were identified, with `r label_number(1)(flowchart[[nrow(flowchart), "n"]])` (`r label_percent(0.1)(flowchart[[nrow(flowchart), "pct_all"]])`) meeting the study eligibility criteria.
+`r label_number(1, big.mark=",")(baseline_pfizer$n)` (`r label_percent(0.1)(baseline_pfizer$n/baseline$n)`) were vaccinated with BNT162b2, contributing `r label_number(0.1)(baseline_pfizer$fu_years)` person-years of follow-up.
+For ChAdOx1, this was `r label_number(1, big.mark=",")(baseline_az$n)` (`r label_percent(0.1)(baseline_az$n/baseline$n)`) and `r label_number(0.1)(baseline_az$fu_years)` person-years.
 

### 2021-09-01 analysis/report/draft-manuscript.Rmd
bug

@@
 
-A total of `r label_number(1, big.mark=",")(flowchart[1, "n"])` HCWs aged 18-64 receiving a first dose of BNT162b2 or ChAdOx1 between 4 January and 28 February 2021 and actively registered at at TPP practice were identified, with `r label_number(1)(flowchart[[nrow(flowchart), "n"]])` (`r label_percent(0.1)(flowchart[[nrow(flowchart), "pct_all"]])`) meeting the study eligibility criteria.
+A total of `r label_number(1, big.mark=",")(flowchart[[1, "n"]])` HCWs aged 18-64 receiving a first dose of BNT162b2 or ChAdOx1 between 4 January and 28 February 2021 and actively registered at at TPP practice were identified, with `r label_number(1)(flowchart[[nrow(flowchart), "n"]])` (`r label_percent(0.1)(flowchart[[nrow(flowchart), "pct_all"]])`) meeting the study eligibility criteria.
 `r label_number(1, big.mark=",")(baseline_pfizer$n)` (`r label_percent(0.1)(baseline_pfizer$n/baseline$n)`) were vaccinated with BNT162b2, contributing `r label_number(0.1)(baseline_pfizer$fu_years)` person-years of follow-up.

### 2021-09-01 analysis/report/draft-manuscript.Rmd
bounds

@@
     scale_y_log10(
-      limits = c(0.1, 10),
       #breaks=c(0.125, 0.167, 0.25, 0.333, 0.5, 0.667, 1, 1.5, 2, 3, 4, 6, 8),
@@
     )+
+    coord_cartesian(ylim= c(0.1, 10))+
     labs(
@@
     scale_y_continuous(
-      limits = c(-5, 5),
       expand = expansion(0),
@@
     )+
+    coord_cartesian(ylim=limits = c(-5, 5))+
     scale_colour_brewer(type="qual", palette="Set1", na.value="grey")+

### 2021-09-02 analysis/report/draft-manuscript.Rmd
limits

@@
     )+
-    coord_cartesian(ylim=limits = c(-5, 5))+
+    coord_cartesian(ylim= c(-5, 5))+
     scale_colour_brewer(type="qual", palette="Set1", na.value="grey")+

### 2021-09-02 analysis/models/report_plr.R
add variance for risk difference

@@ -258,7 +258,7 @@ group_by(vax1_az, tstop) %>%
   ) %>%
   arrange(model, vax1_az, tstop)
 
-cml_inc <- ggplot(curves %>% filter(model=="3"))+
+cmlinc <- ggplot(curves %>% filter(model=="3"))+
   geom_step(aes(x=tstop, y=1-survival, group=vax1_az_descr, colour=vax1_az_descr))+
   scale_x_continuous(
     breaks = seq(0,7*52,by=14),
@@ -282,8 +282,8 @@ cml_inc <- ggplot(curves %>% filter(model=="3"))+
 
 write_rds(curves, here("output", "models", outcome, timescale, glue("reportplr_cmlinccurves_pw.rds")), compress="gz")
 
-ggsave(filename=here("output", "models", outcome, timescale, glue("reportplr_cmlincplot_pw.svg")), cml_inc, width=20, height=15, units="cm")
-ggsave(filename=here("output", "models", outcome, timescale, glue("reportplr_cmlincplot_pw.png")), cml_inc, width=20, height=15, units="cm")
+ggsave(filename=here("output", "models", outcome, timescale, glue("reportplr_cmlincplot_pw.svg")), cmlinc, width=20, height=15, units="cm")
+ggsave(filename=here("output", "models", outcome, timescale, glue("reportplr_cmlincplot_pw.png")), cmlinc, width=20, height=15, units="cm")
 
 
 
@@ -446,9 +446,69 @@ cmlinc_variance <- function(model, vcov, newdata, id, time, weights){
 
 #test<-cmlinc_variance(plrmod3, vcov3, mutate(data_plr, vax1_az=1L), data_plr$patient_id, data_plr$tstop, data_plr$sample_weights)
 
+
+## risk-adjusted survival curves ----
+
+riskdiff_variance <- function(model, vcov, newdata0, newdata1, id, time, weights){
+
+  # calculate variance of adjusted survival / adjusted cumulative incidence
+  # needs model object, cluster-vcov, input data, model weights, and indices for patient and time
+
+  tt <- terms(model) # this helpfully grabs the correct spline basis from the model, rather than recalculating based on `newdata`
+  Terms <- delete.response(tt)
+  m.mat0 <- model.matrix(Terms, data=newdata0)
[…]

### 2021-09-02 create-project.R
remove unused outcomes

@@ -419,7 +419,7 @@ actions_list <- splice(
       "descr_km", "descr_vaxdate",
       as.list(
         glue(
-          outcome = c("test", "postest", "emergency", "admitted", "covidadmitted"),
+          outcome = c("postest", "emergency", "covidadmitted"),
           "report_{outcome}_timesincevax_plr"
         )
       )

### 2021-09-02 analysis/report/draft-manuscript.Rmd
clarify risk diff flow

@@
 
-  riskdifference <- 
-    cmlinc  %>%
-    group_by(model, model_name, outcome, outcome_descr, tstop) %>%
-    arrange(tstop, model, outcome, vax1_az) %>%
+  riskdiff <-
+    map_dfr(
+      outcomes,
+      ~read_rds(here("output", "models", .x, "timesincevax", glue("reportplr_adjusteddiff_ns.rds"))),
+      .id="outcome"
+    ) %>%
+    left_join(metadata_outcomes, by="outcome") %>%
     mutate(
-      diff = (1-survival) - first((1-survival)),
-      diff_format = label_number(0.01, scale=1000)(diff)
+      outcome_descr = fct_inorder(outcome_descr),
+      model_name = fct_inorder(str_wrap(model_name, 15)),
+      diff_format = label_number(0.01, scale=1000)(diff),
     ) %>%
-    filter(vax1_az==1) %>%
-    left_join(
-      hazard %>% select(outcome, model, tstop, hr, hr.ll, hr.ul), 
-      by=c("outcome", "model", "tstop")
-    ) %>% 
-    group_by(model, model_name, outcome, outcome_descr) %>%
-    arrange(model, outcome, vax1_az, tstop) %>%
-    mutate(
-      hazard1 = -(survival - lag(survival, 1, 1)) / lag(survival, 1, 1),
-      hazard0 = hazard1/hr,
-      hazard0.ll = hazard1/hr.ul,
-      hazard0.ul = hazard1/hr.ll,
+    filter(model==3)
 
-      survival0 = exp(-cumsum(if_else(is.na(hazard0), 0, hazard0))),
-      survival0.ll = exp(-cumsum(if_else(is.na(hazard0.ul), 0, hazard0.ul))),
-      survival0.ul = exp(-cumsum(if_else(is.na(hazard0.ll), 0, hazard0.ul))),
-
-      diff2 = (1-survival)-(1-survival0),
-
-      diff.ll = (1-survival) - (1-survival)*hr.ll,
-      diff.ul = (1-survival) + (1-survival)*hr.ul,
-      #diff_CI = paste0(label_number(0.01, scale=1000)(diff), " (", diff.ll, " ,", diff.ul, ")")
-    )
+  # riskdiff <- 
+  #   cmlinc  %>%
+  #   group_by(model, model_name, outcome, outcome_descr, tstop) %>%
+  #   arrange(tstop, model, outcome, vax1_az) %>%
+  #   mutate(
+  #     diff = (1-survival) - first((1-survival)),
+  #     
+  #   ) %>%
+  #   filter(vax1_az==1) %>%
+  #   left_join(
+  #     hazard %>% select(outcome, model, tstop, hr, hr.ll, hr.ul), 
+  #     by=c("outcome", "model", "tstop")
+  #   ) %>% 
+  #   group_by(model, model_name, outcome, outcome_descr) %>%
+  #   arrange(model, outcome, vax1_az, tstop) %>%
+  #   mutate(
+  #     hazard1 = -(survival - lag(survival, 1, 1)) / lag(survival, 1, 1),
+  #     hazard0 = hazard1/hr,
+  #     hazard0.ll = hazard1/hr.ul,
+  #     hazard0.ul = hazard1/hr.ll,
+  #     
+  #     survival0 = exp(-cumsum(if_else(is.na(hazard0), 0, hazard0))),
+  #     survival0.ll = exp(-cumsum(if_else(is.na(hazard0.ul), 0, hazard0.ul))),
+  #     survival0.ul = exp(-cumsum(if_else(is.na(hazard0.ll), 0, hazard0.ul))),
+  #     
+  #     diff2 = (1-survival)-(1-survival0),
+  #     
+  #     diff.ll = (1-survival) - (1-survival)*hr.ll,
+  #     diff.ul = (1-survival) + (1-survival)*hr.ul,
+  #     #diff_CI = paste0(label_number(0.01, scale=1000)(diff), " (", diff.ll, " ,", diff.ul, ")")
+  #   )
 
   plot_riskdiff <-
-    ggplot(riskdifference)+
+    ggplot(riskdiff)+
     geom_hline(aes(yintercept=0), colour='black')+
     geom_step(aes(x=tstop, y=diff*1000), colour='orchid4')+
-    # geom_rect(aes(xmin=lag(tstop, 1, 0), xmax=tstop, ymin=diff.ll*1000, ymax=diff.ul*1000), alpha=0.1, colour="transparent", fill='orchid4')+
+    geom_rect(aes(xmin=lag(tstop, 1, 0), xmax=tstop, ymin=diff.ll*1000, ymax=diff.ul*1000), alpha=0.1, colour="transparent", fill='orchid4')+
     facet_grid(cols=vars(outcome_descr))+
@@
 
-At 6 weeks post-vaccination, the absolute risk difference (ChAdOx1 - BNT162b2) for a positive SARS-CoV-2 test was `r filter(riskdifference, outcome=="postest", tstop==7*6)$diff_format` per 1000 people, for A&E attendances was `r filter(riskdifference, outcome=="emergency", tstop==7*6)$diff_format` and for COVID-19 hospital admissions was `r filter(riskdifference, outcome=="covidadmitted", tstop==7*6)$diff_format` per 1000 people.
+At 6 weeks post-vaccination, the absolute risk difference (ChAdOx1 - BNT162b2) for a positive SARS-CoV-2 test was `r filter(riskdiff, outcome=="postest", tstop==7*6)$diff_format` per 1000 people, for A&E attendances was `r filter(riskdiff, outcome=="emergency", tstop==7*6)$diff_format` and for COVID-19 hospital admissions was `r filter(riskdiff, outcome=="covidadmitted", tstop==7*6)$diff_format` per 1000 people.
 

### 2021-09-06 analysis/report/effectiveness_report.Rmd
add covid-19 emergency attendances outcome; remove tests; remove diagnoses

@@
 
-* registered at a GP practice using TPP's SystmOne clinical information system on the day of they received their first dose of BNT162b2 or ChAdOx1; 
-* the date of vaccination was between `r format(as.Date(gbl_vars$start_date_az), "%d %B")` and `r format(as.Date(gbl_vars$lastvax_date), "%d %B %Y")` (`r as.Date(gbl_vars$lastvax_date) - as.Date(gbl_vars$start_date_az)` days), a period when both vaccines were being administered widely;
-* aged between 18 and 64 inclusive; 
-* not classed as Clinically Extremely Vulnerable, as set out by government guidance, at the time of vaccination; 
-* information on sex, ethnicity, deprivation, and geographical region was known.
+-   registered at a GP practice using TPP's SystmOne clinical information system on the day of they received their first dose of BNT162b2 or ChAdOx1;
+-   the date of vaccination was between `r format(as.Date(gbl_vars$start_date_az), "%d %B")` and `r format(as.Date(gbl_vars$lastvax_date), "%d %B %Y")` (`r as.Date(gbl_vars$lastvax_date) - as.Date(gbl_vars$start_date_az)` days), a period when both vaccines were being administered widely;
+-   aged between 18 and 64 inclusive;
+-   not classed as Clinically Extremely Vulnerable, as set out by government guidance, at the time of vaccination;
+-   information on sex, ethnicity, deprivation, and geographical region was known.
 
@@
 
-* death or deregistration;
-* Second vaccine dose;
-* The study end date, `r format(as.Date(gbl_vars$end_date), "%d %B %Y")`.
+-   death or deregistration;
+-   Second vaccine dose;
+-   The study end date, `r format(as.Date(gbl_vars$end_date), "%d %B %Y")`.
 
@@
 
-* SARS-CoV-2 infection, as identified via SGSS records. Both PCR and LFT tests are included.
-* Unplanned COVID-19-related hospitalisation via HES records. ICD-10 codes are used to identify COVID-19-related admissions. 
-* Unplanned COVID-19-related ICU admission via HES records. ICD-10 codes are used to identify COVID-19-related admissions. 
-* COVID-19-related death via death registration records. With COVID-19 ICD-10 codes anywhere on the death certificate (i.e., as an underlying or contributing cause of death)
+-   SARS-CoV-2 infection, as identified via SGSS records. Both PCR and LFT tests are included.
+-   Unplanned COVID-19-related hospitalisation via HES records. ICD-10 codes are used to identify COVID-19-related admissions.
+-   Unplanned COVID-19-related ICU admission via HES records. ICD-10 codes are used to identify COVID-19-related admissions.
+-   COVID-19-related death via death registration records. With COVID-19 ICD-10 codes anywhere on the death certificate (i.e., as an underlying or contributing cause of death)
 
@@
 
-Patient characteristics used for adjustment include: age, sex, deprivation, ethnicity, NHS region, clinically “at risk” (but not clinically extremely vulnerable) as per national prioritisation guidelines, asthma, number of previous SARS-CoV-2 tests (via SGSS), rurality, evidence of prior covid infection (positive test or COVID-19 hospitalisation), number of recorded comorbidities, severe mental illness.
+Patient characteristics used for adjustment include: age, sex, deprivation, ethnicity, NHS region, clinically "at risk" (but not clinically extremely vulnerable) as per national prioritisation guidelines, asthma, number of previous SARS-CoV-2 tests (via SGSS), rurality, evidence of prior covid infection (positive test or COVID-19 hospitalisation), number of recorded comorbidities, severe mental illness.
 All characteristics are ascertained as at the time of vaccination.
@@
 ### Time-dependent Cox models
+
 Time-dependent Cox models are used to estimate the time-varying effects for each vaccine type.
@@
 
-* *ChAdOx1* versus *BNT162b2* hazard ratio splines
-* Risk-adjusted survival curves for *ChAdOx1* and *BNT162b2*
+-   *ChAdOx1* versus *BNT162b2* hazard ratio splines
+-   Risk-adjusted survival curves for *ChAdOx1* and *BNT162b2*
 
@@
 cat("  \n\n")
+print(glue("### COVID-19 A&E attendance"), "  \n")
+wrap_plots(plots_spline("covidemergency"), plots_adjsurv("covidemergency"), nrow=2)
+
+cat("  \n\n")
 print(glue("### Any unplanned hospital admission"), "  \n")

### 2021-09-06 analysis/descriptive/km.R

@@ -97,6 +97,9 @@ data_tte <- data_cohort %>%
     tte_emergency = tte(vax1_date-1, emergency_date, censor_date, na.censor=FALSE),
     ind_emergency = censor_indicator(emergency_date, censor_date),
 
+    tte_covidemergency = tte(vax1_date-1, emergency_covid_date, censor_date, na.censor=FALSE),
+    ind_covidemergency = censor_indicator(emergency_covid_date, censor_date),
+
     tte_admitted = tte(vax1_date-1, admitted_date, censor_date, na.censor=FALSE),
     ind_admitted = censor_indicator(admitted_date, censor_date),
 
@@ -245,6 +248,7 @@ metadata_outcomes <- tribble(
   "test", "SARS-CoV-2 test",
   "postest", "Positive test",
   "emergency", "A&E attendance",
+  "covidemergency", "COVID-19 A&E attendance",
   "admitted", "Unpanned admission",
   "covidadmitted", "COVID-19 admission",
   "covidcc", "COVID-19 critical care",

### 2021-09-06 analysis/descriptive/table_irr.R

@@ -73,6 +73,9 @@ data_tte <- data_cohort %>%
     tte_emergency = tte(vax1_date-1, emergency_date, censor_date, na.censor=TRUE),
     ind_emergency = censor_indicator(emergency_date, censor_date),
 
+    tte_covidemergency = tte(vax1_date-1, emergency_covid_date, censor_date, na.censor=TRUE),
+    ind_covidemergency = censor_indicator(emergency_covid_date, censor_date),
+
     tte_admitted = tte(vax1_date-1, admitted_date, censor_date, na.censor=TRUE),
     ind_admitted = censor_indicator(admitted_date, censor_date),
 
@@ -119,6 +122,7 @@ data_cox_split <- tmerge(
   tstop = tte_censor,
   test = event(tte_test),
   postest = event(tte_postest),
+  covidemergency = event(tte_covidemergency),
   emergency = event(tte_emergency),
   admitted = event(tte_admitted),
   covidadmitted = event(tte_covidadmitted),
@@ -131,6 +135,7 @@ data_cox_split <- tmerge(
   status_test = tdc(tte_test),
   status_postest = tdc(tte_postest),
   status_emergency = tdc(tte_emergency),
+  status_covidemergency = tdc(tte_covidemergency),
   status_admitted = tdc(tte_admitted),
   status_covidadmitted = tdc(tte_covidadmitted),
   status_covidcc = tdc(tte_covidcc),
@@ -248,16 +253,17 @@ data_summary <- local({
   temp1 <- pt_summary(data_cox_split, "test")
   temp2 <- pt_summary(data_cox_split, "postest")
   temp3 <- pt_summary(data_cox_split, "emergency")
-  temp4 <- pt_summary(data_cox_split, "admitted")
-  temp5 <- pt_summary(data_cox_split, "covidadmitted")
-  temp6 <- pt_summary(data_cox_split, "covidcc")
-  temp7 <- pt_summary(data_cox_split, "coviddeath")
-  temp8 <- pt_summary(data_cox_split, "noncoviddeath")
-  temp9 <- pt_summary(data_cox_split, "death")
+  temp4 <- pt_summary(data_cox_split, "covidemergency")
+  temp5 <- pt_summary(data_cox_split, "admitted")
+  temp6 <- pt_summary(data_cox_split, "covidadmitted")
+  temp7 <- pt_summary(data_cox_split, "covidcc")
[…]

### 2021-09-06 analysis/process/data_process.R

@@ -24,12 +24,12 @@ source(here("analysis", "lib", "utility_functions.R"))
 
 # import globally defined repo variables from
 study_dates <- jsonlite::read_json(
-  path=here("data","data","metadata_study-dates.json")
+  path=here("output","data","metadata_study-dates.json")
 )
 
 ## load A&E diagnosis column names
 diagnosis_codes <- jsonlite::read_json(
-  path=here("analysis","lib,","diagnosis_groups.json")
+  path=here("analysis","lib","diagnosis_groups.json")
 )
 diagnosis_col_names <- paste0("emergency_", names(diagnosis_codes), "_date")
 diagnosis_short <- str_remove(str_remove(diagnosis_col_names, "emergency_"), "_date")
@@ -105,7 +105,8 @@ data_processed <- data_extract %>%
     start_date_pfizer = as.Date(study_dates$start_date_pfizer),
     start_date_az = as.Date(study_dates$start_date_az),
     lastvax_date = as.Date(study_dates$lastvax_date),
-    end_date = as.Date(study_dates$end_date),
+    end_date14 = as.Date(study_dates$end_date14),
+    end_date20 = as.Date(study_dates$end_date20),
 
     ageband = cut(
       age,
@@ -184,24 +185,6 @@ data_processed <- data_extract %>%
 
     prior_covid_infection = !is.na(prior_positive_test_date) | !is.na(prior_covidadmitted_date) | !is.na(prior_primary_care_covid_case_date),
 
-    vax1_type = case_when(
-      pmin(covid_vax_az_1_date, as.Date("2030-01-01"), na.rm=TRUE) <= pmin(covid_vax_pfizer_1_date, covid_vax_moderna_1_date, as.Date("2030-01-01"), na.rm=TRUE) ~ "az",
-      pmin(covid_vax_pfizer_1_date, as.Date("2030-01-01"), na.rm=TRUE) <= pmin(covid_vax_az_1_date, covid_vax_moderna_1_date, as.Date("2030-01-01"), na.rm=TRUE) ~ "pfizer",
-      pmin(covid_vax_moderna_1_date, as.Date("2030-01-01"), na.rm=TRUE) <= pmin(covid_vax_pfizer_1_date, covid_vax_az_1_date, as.Date("2030-01-01"), na.rm=TRUE) ~ "moderna",
-      TRUE ~ NA_character_
-    ),
-
-    vax1_type_descr = fct_case_when(
[…]

### 2021-09-06 analysis/process/design.R

@@ -15,13 +15,13 @@ fs::dir_create(here("output", "data"))
 # define key dates ----
 
 study_dates <- list(
-  start_date = "2020-12-08",
+  start_date = "2020-12-08", #inclusive
   start_date_pfizer = "2020-12-08",
   start_date_az = "2021-01-04",
   start_date_moderna = "2021-03-04",
-  lastvax_date = "2021-02-28",
-  end_date14 = "2021-04-25",
-  end_date20 = "2021-04-25"
+  lastvax_date = "2021-02-28", #inclusive
+  end_date14 = "2021-04-25", #inclusive
+  end_date20 = "2021-06-06" # inclusive
 )
 
 jsonlite::write_json(study_dates, path = here("output", "data", "metadata_study-dates.json"), auto_unbox=TRUE, pretty =TRUE)
@@ -33,12 +33,14 @@ metadata_outcomes <- tribble(
   "test", "covid_test_date", "SARS-CoV-2 test",
   "postest", "positive_test_date", "Positive SARS-CoV-2 test",
   "emergency", "emergency_date", "A&E attendance",
+  "covidemergency", "emergency_covid_date", "COVID-19 A&E attendance",
   "admitted", "admitted_date", "Unplanned hospitalisation",
   "covidadmitted", "covidadmitted_date", "COVID-19 hospitalisation",
   "covidcc", "covidcc_date", "COVID-19 critical care",
   "coviddeath", "coviddeath_date", "COVID-19 death",
   "noncoviddeath", "noncoviddeath_date", "Non-COVID-19 death",
   "death", "death_date", "Any death",
+  "seconddose", "covid_vax_any_2_date", "Second dose",
 )
 
 write_rds(metadata_outcomes, here("output", "data", "metadata_outcomes.rds"))
@@ -65,8 +67,8 @@ formula_all_rhsvars <- update(1 ~ 1, formula_exposure) %>%
   update(formula_secular) %>%
   update(formula_secular_region)
 
-postvaxcuts <- c(0, 7, 14, 21, 28, 35, 42, 70)
-postvaxcuts_2week <- c(0, 14, 28, 42, 70, 84, 98)
+postvaxcuts <- c(0, 7, 14, 21, 28, 35, 42, 56, 84, 112)
+postvaxcuts_2week <- c(0, 14, 28, 42, 70, 84, 98, 112, 136, 140)
 
 lastfupday <- 14*7
 
[…]

### 2021-09-06 create-project.R

@@ -221,8 +221,7 @@ actions_list <- splice(
     run = "r:latest analysis/process/data_process.R",
     needs = list("extract"),
     highly_sensitive = list(
-      processed = "output/data/data_processed.rds",
-      diagnoses = "output/data/data_diagnoses.rds"
+      processed = "output/data/data_processed.rds"
     )
   ),
 
@@ -315,15 +314,15 @@ actions_list <- splice(
 
   comment("# # # # # # # # # # # # # # # # # # #", "Models", "# # # # # # # # # # # # # # # # # # #"),
 
-  comment("###  SARS-CoV-2 Test"),
-  action_model("test", "timesincevax", "cox"),
-  action_report("test", "timesincevax", "cox"),
-  action_model("test", "calendar", "cox"),
-  action_report("test", "calendar", "cox"),
-  action_model("test", "timesincevax", "plr", 50000),
-  action_report("test", "timesincevax", "plr"),
-  action_model("test", "calendar", "plr", 50000),
-  action_report("test", "calendar", "plr"),
+  # comment("###  SARS-CoV-2 Test"),
+  # action_model("test", "timesincevax", "cox"),
+  # action_report("test", "timesincevax", "cox"),
+  # action_model("test", "calendar", "cox"),
+  # action_report("test", "calendar", "cox"),
+  # action_model("test", "timesincevax", "plr", 50000),
+  # action_report("test", "timesincevax", "plr"),
+  # action_model("test", "calendar", "plr", 50000),
+  # action_report("test", "calendar", "plr"),
 
   comment("###  Positive SARS-CoV-2 Test"),
   action_model("postest", "timesincevax", "cox"),
@@ -344,6 +343,16 @@ actions_list <- splice(
   action_report("emergency", "timesincevax", "plr"),
   action_model("emergency", "calendar", "plr", 50000),
   action_report("emergency", "calendar", "plr"),
+
+  comment("###  COVID-19 A&E attendence"),
+  action_model("covidemergency", "timesincevax", "cox"),
+  action_report("covidemergency", "timesincevax", "cox"),
+  action_model("covidemergency", "calendar", "cox"),
+  action_report("covidemergency", "calendar", "cox"),
+  action_model("covidemergency", "timesincevax", "plr", 50000),
+  action_report("covidemergency", "timesincevax", "plr"),
+  action_model("covidemergency", "calendar", "plr", 50000),
+  action_report("covidemergency", "calendar", "plr"),
 
   comment("### Unplanned Hospital admission"),
[…]

### 2021-09-09 analysis/descriptive/seconddose.R
improve supplementary materials

@@ -29,17 +29,26 @@ list_formula <- read_rds(here("output", "data", "metadata_formulas.rds"))
 list2env(list_formula, globalenv())
 lastfupday <- lastfupday20
 
+metadata_outcomes <- read_rds(here("output", "data", "metadata_outcomes.rds"))
+
+
+
 ## create output directory ----
 fs::dir_create(here("output", "descriptive", "seconddose"))
 
-
+threshold <- 5
 
 ceiling_any <- function(x, to=1){
   # round to nearest 100 millionth to avoid floating point errors
   ceiling(plyr::round_any(x/to, 1/100000000))*to
 }
 
-data_cohort <- read_rds(here::here("output", "data", "data_cohort.rds")) %>%
+data_cohort <- read_rds(here::here("output", "data", "data_cohort.rds"))
+
+
+
+data_tte <-
+  data_cohort %>%
   mutate(
 
     patient_id,
@@ -55,65 +64,136 @@ data_cohort <- read_rds(here::here("output", "data", "data_cohort.rds")) %>%
     vaxpfizer2_date = if_else(vax1_type=="pfizer", vax2_date, as.Date(NA)),
     vaxaz2_date = if_else(vax1_type=="az", vax2_date, as.Date(NA)),
     vaxmoderna2_date = if_else(vax1_type=="moderna", vax2_date, as.Date(NA)),
+)
 
-    # assume vaccination occurs at the start of the day, and all other events occur at the end of the day.
-    # so use vax1_date - 1
-    censor_date = pmin(
-      vax1_date - 1 + lastfupday,
-      dereg_date,
-      death_date,
-      end_date,
-      na.rm=TRUE
-    ),
-    tte_censor = tte(vax1_date-1, censor_date, censor_date),
 
-    tte_vaxpfizer2 = tte(vax1_date-1, vaxpfizer2_date, censor_date),
-    ind_vaxpfizer2 = censor_indicator(vaxpfizer2_date, censor_date),
+seconddose <- function(outcome){
 
-    tte_vaxaz2 = tte(vax1_date-1, vaxaz2_date, censor_date),
-    ind_vaxaz2 = censor_indicator(vaxaz2_date, censor_date),
+  outcome_var <- metadata_outcomes$outcome_var[metadata_outcomes$outcome==outcome]
 
[…]

### 2021-09-09 analysis/descriptive/table1_allvax.R

@@ -57,15 +57,16 @@ data_cohort <- read_rds(here("output", "data", "data_cohort_allvax.rds")) %>%
 ## baseline variables
 tab_summary_baseline <- data_cohort %>%
   select(
-    all_of(names(var_labels)),
+    all_of(names(var_labels)), cev,
     -age, -stp, -vax1_type, -fu_days12, -fu_days20
   ) %>%
   tbl_summary(
     by = vax1_type_descr,
-    label=unname(var_labels[names(.)]),
-    missing = "ifany",
+    label=unname(splice(var_labels, list(cev = cev ~ "Clinically Extremely Vulnerable"))[names(.)]),
+    missing = "ifany"
   )  %>%
   modify_footnote(starts_with("stat_") ~ NA)
+
 
 tab_summary_baseline$inputs$data <- NULL
 

### 2021-09-09 create-project.R

@@ -488,6 +488,30 @@ actions_list <- splice(
   ),
 
   action(
+    name = "rmd_supplement",
+    run = glue(
+      "r:latest -e {q}",
+      q = single_quote('rmarkdown::render("analysis/report/draft-supplement.Rmd",  knit_root_dir = "/workspace",  output_dir = "/workspace/output/report", output_format = c("html_document")   )')
+    ),
+    needs = splice(
+      "design", "data_selection",
+      "descr_table1_allvax", "descr_irr",
+      "descr_km", "descr_vaxdate",
+      "descr_seconddose",
+      as.list(
+        glue(
+          outcome = c("postest", "emergency", "covidemergency", "covidadmitted"),
+          "report_{outcome}_timesincevax_1_plr"
+        )
+      )
+    ),
+    moderately_sensitive = list(
+      html = "output/report/draft-supplement.html",
+      md = "output/report/draft-supplement.md"
+    )
+  ),
+
+  action(
     name = "rmd_report_timescales",
     run = glue(
       "r:latest -e {q}",
@@ -500,11 +524,12 @@ actions_list <- splice(
       as.list(
         glue_data(
           .x = expand_grid(
-            outcome = c("postest", "emergency", "covidemergency", "admitted", "covidadmitted"),
+            outcome = c("postest", "covidemergency", "covidadmitted"),
             modeltype = c("cox", "plr"),
-            timescale = c("timesincevax", "calendar")
+            timescale = c("timesincevax", "calendar"),
+            censor_seconddose = c("0", "1")
           ),
-          "report_{outcome}_{timescale}_1_{modeltype}"
+          "report_{outcome}_{timescale}_{censor_seconddose}_{modeltype}"
         )
       )
     ),

### 2021-09-09 analysis/report/effectiveness_report_comparetimescales.Rmd
add cumulative incidence difference

@@
 
-print_plots_effects <- function(outcome, modeltype){
+print_plots_effects <- function(outcome, modeltype, censor_seconddose){
 
@@
 
-  calendar_effects <- read_rds(here("output", "models", outcome, "calendar", glue("report{modeltype}_effects{suffix}.rds"))) %>%
+  calendar_effects <- read_rds(here("output", "models", outcome, "calendar", censor_seconddose, glue("report{modeltype}_effects{suffix}.rds"))) %>%
     mutate(timescale = "Calendar time")
-  timesincevax_effects <- read_rds(here("output", "models", outcome, "timesincevax", glue("report{modeltype}_effects{suffix}.rds"))) %>%
+  timesincevax_effects <- read_rds(here("output", "models", outcome, "timesincevax", censor_seconddose, glue("report{modeltype}_effects{suffix}.rds"))) %>%
     mutate(timescale = "Vaccinated time")
@@
       breaks = seq(0, 7*52, 14), 
-      limits = c(min(tidy_effects$term_left), 14*7), 
+      limits = c(min(tidy_effects$term_left), 20*7), 
       expand = c(0, 0)
@@
 
-print_plots_spline <- function(outcome){
+print_plots_spline <- function(outcome, censor_seconddose){
 
-  calendar_effects <- read_rds(here("output", "models", outcome, "calendar", glue("reportplr_effects_ns.rds"))) %>%
+  calendar_effects <- read_rds(here("output", "models", outcome, "calendar", censor_seconddose, glue("reportplr_effects_ns.rds"))) %>%
     mutate(timescale = "Calendar time")
-  timesincevax_effects <- read_rds(here("output", "models", outcome, "timesincevax", glue("reportplr_effects_ns.rds"))) %>%
+  timesincevax_effects <- read_rds(here("output", "models", outcome, "timesincevax", censor_seconddose, glue("reportplr_effects_ns.rds"))) %>%
     mutate(timescale = "Vaccinated time")
@@
 
-print_plots_adjsurv <- function(outcome){
+print_plots_adjsurv <- function(outcome, censor_seconddose){
 
-  calendar_curve <- read_rds(here("output", "models", outcome, "calendar", glue("reportplr_adjustedsurvival_ns.rds"))) %>%
+  calendar_curve <- read_rds(here("output", "models", outcome, "calendar", censor_seconddose, glue("reportplr_adjustedsurvival_ns.rds"))) %>%
     mutate(timescale = "Calendar time")
-  timesincevax_curve <- read_rds(here("output", "models", outcome, "timesincevax", glue("reportplr_adjustedsurvival_ns.rds"))) %>%
+  timesincevax_curve <- read_rds(here("output", "models", outcome, "timesincevax", censor_seconddose, glue("reportplr_adjustedsurvival_ns.rds"))) %>%
     mutate(timescale = "Vaccinated time")
@@
     ggplot(curves %>% filter(model=="3"))+
-    geom_step(aes(x=tstop, y=survival, group=vax1_az_descr, colour=vax1_az_descr))+
-    geom_rect(aes(xmin=lag(tstop, 1, 0), xmax=tstop, ymin=survival.ll, ymax=survival.ul, group=vax1_az_descr, fill=vax1_az_descr), alpha=0.1)+
+    geom_step(aes(x=tstop, y=1-survival, group=vax1_az_descr, colour=vax1_az_descr))+
+    geom_rect(aes(xmin=lag(tstop, 1, 0), xmax=tstop, ymin=1-survival.ll, ymax=1-survival.ul, group=vax1_az_descr, fill=vax1_az_descr), alpha=0.1)+
     facet_grid(cols=vars(timescale))+
@@
     theme(
-      legend.position=c(0.05,.95),
-      legend.justification = c(0,1),
+      legend.position=c(0.95,.05),
+      legend.justification = c(1,0),
       axis.text.x.top=element_text(hjust=0),
@@
 
+print_plots_diff <- function(outcome, censor_seconddose){
+  calendar_curve <- read_rds(here("output", "models", outcome, "calendar", censor_seconddose, glue("reportplr_adjusteddiff_ns.rds"))) %>%
+    mutate(timescale = "Calendar time")
+  timesincevax_curve <- read_rds(here("output", "models", outcome, "timesincevax", censor_seconddose, glue("reportplr_adjusteddiff_ns.rds"))) %>%
+    mutate(timescale = "Vaccinated time")
+
+  curves <- bind_rows(calendar_curve, timesincevax_curve)
+
+  plot_diff <-
+    ggplot(curves)+
+    geom_hline(aes(yintercept=0, colour=model_name))+
+    geom_step(aes(x=tstop, y=diff, colour=model_name))+
+    geom_rect(aes(xmin=lag(tstop, 1, 0), xmax=tstop, ymin=diff.ll, ymax=diff.ul, fill=model_name), alpha=0.1, colour="transparent")+
+    facet_grid(cols=vars(timescale))+
+    scale_x_continuous(
+      limits = c(0, NA),
+      breaks = seq(0,7*52,by=14),
+      expand = expansion(0)
+    )+
+    scale_y_continuous(
+      expand = expansion(0),
+      sec.axis = dup_axis(name="Lower risk after \n <- BNT162b2 / ChAdOx1 ->", breaks = NULL)
+    )+
+    #coord_cartesian(ylim= c(-5, 5))+
+    scale_colour_brewer(type="qual", palette="Set1", na.value="grey")+
+    scale_fill_brewer(type="qual", palette="Set1", guide="none", na.value="grey")+
+    labs(
+      x=NULL,
+      y="Difference\nin cumulative incidence",
+      colour=NULL,
+      fill=NULL
+    )+
+    theme_bw()+
+    theme(
+      panel.border = element_blank(),
+      axis.line.y.left = element_line(colour = "black"),
+
+      panel.grid.minor.x = element_blank(),
+      panel.grid.minor.y = element_blank(),
+      strip.background = element_blank(),
+      strip.text = element_blank(),
+
+      panel.spacing = unit(0.8, "lines"),
+
+      plot.title = element_text(hjust = 0),
+      plot.title.position = "plot",
+      plot.caption.position = "plot",
+      plot.caption = element_text(hjust = 0, face= "italic"),
+
+     ) +
+   NULL
+
+  print(plot_diff)
   cat("  \n\n")
-print(glue("### SARS-CoV-2 test"), "  \n")
-print(glue("#### Unadjusted"), "  \n")
-print_plots_km("test")
-print(glue("#### Cox"), "  \n")
-print_plots_effects("test", "cox")
-print(glue("#### PLR, piecewise"), "  \n")
-print_plots_effects("test", "plr")
-print(glue("#### PLR, spline"), "  \n")
-print_plots_spline("test")
-print_plots_adjsurv("test")
+}
+
+# cat("  \n\n")
+# print(glue("### SARS-CoV-2 test"), "  \n")
+# print(glue("#### Unadjusted"), "  \n")
+# print_plots_km("test")
+# print(glue("#### Cox"), "  \n")
+# print_plots_effects("test", "cox", "0")
+# print(glue("#### PLR, piecewise"), "  \n")
+# print_plots_effects("test", "plr", "0")
+# print(glue("#### PLR, spline"), "  \n")
+# print_plots_spline("test", "0")
+# print_plots_adjsurv("test", "0")
+# print_plots_diff("test", "0")
 
@@
 print(glue("#### Cox"), "  \n")
-print_plots_effects("postest", "cox")
+print_plots_effects("postest", "cox", "0")
 print(glue("#### PLR, piecewise"), "  \n")
-print_plots_effects("postest", "plr")
+print_plots_effects("postest", "plr", "0")
 print(glue("#### PLR, spline"), "  \n")
-print_plots_spline("postest")
-print_plots_adjsurv("postest")
+print_plots_spline("postest", "0")
+print_plots_adjsurv("postest", "0")
+print_plots_diff("postest", "0")
+
+# cat("  \n\n")
+# print(glue("### A&E attendance"), "  \n")
+# print(glue("#### Unadjusted"), "  \n")
+# print_plots_km("emergency")
+# print(glue("#### Cox"), "  \n")
+# print_plots_effects("emergency", "cox", "0")
+# print(glue("#### PLR, piecewise"), "  \n")
+# print_plots_effects("emergency", "plr", "0")
+# print(glue("#### PLR, spline"), "  \n")
+# print_plots_spline("emergency", "0")
+# print_plots_adjsurv("emergency", "0")
+# print_plots_diff("postest", "0")
 
 cat("  \n\n")
-print(glue("### A&E attendance"), "  \n")
+print(glue("### COVID-19 A&E attendance"), "  \n")
 print(glue("#### Unadjusted"), "  \n")
-print_plots_km("emergency")
+print_plots_km("covidemergency")
 print(glue("#### Cox"), "  \n")
-print_plots_effects("emergency", "cox")
+print_plots_effects("covidemergency", "cox", "0")
 print(glue("#### PLR, piecewise"), "  \n")
-print_plots_effects("emergency", "plr")
+print_plots_effects("covidemergency", "plr", "0")
 print(glue("#### PLR, spline"), "  \n")
-print_plots_spline("emergency")
-print_plots_adjsurv("emergency")
+print_plots_spline("covidemergency", "0")
+print_plots_adjsurv("covidemergency", "0")
+print_plots_diff("covidemergency", "0")
 
-cat("  \n\n")
-print(glue("### Any unplanned hospital admission"), "  \n")
-print(glue("#### Unadjusted"), "  \n")
-print_plots_km("admitted")
-print(glue("#### Cox"), "  \n")
-print_plots_effects("admitted", "cox")
-print(glue("#### PLR, piecewise"), "  \n")
-print_plots_effects("admitted", "plr")
-print(glue("#### PLR, spline"), "  \n")
-print_plots_spline("admitted")
-print_plots_adjsurv("admitted")
+# cat("  \n\n")
+# print(glue("### Any unplanned hospital admission"), "  \n")
+# print(glue("#### Unadjusted"), "  \n")
+# print_plots_km("admitted")
+# print(glue("#### Cox"), "  \n")
+# print_plots_effects("admitted", "cox", "0")
+# print(glue("#### PLR, piecewise"), "  \n")
+# print_plots_effects("admitted", "plr", "0")
+# print(glue("#### PLR, spline"), "  \n")
+# print_plots_spline("admitted", "0")
+# print_plots_adjsurv("admitted", "0")
+# print_plots_diff("admitted", "0")
 
@@
 print(glue("#### Cox"), "  \n")
-print_plots_effects("covidadmitted", "cox")
+print_plots_effects("covidadmitted", "cox", "0")
 print(glue("#### PLR, piecewise"), "  \n")
-print_plots_effects("covidadmitted", "plr")
+print_plots_effects("covidadmitted", "plr", "0")
 print(glue("#### PLR, spline"), "  \n")
-print_plots_spline("covidadmitted")
-print_plots_adjsurv("covidadmitted")
+print_plots_spline("covidadmitted", "0")
+print_plots_adjsurv("covidadmitted", "0")
+print_plots_diff("covidadmitted", "0")
 

### 2021-09-09 analysis/report/effectiveness_report_comparemodels.Rmd
update model comparison report

--- 2021-09-09 analysis/report/effectiveness_report_comparetimescales.Rmd
@@
     ggplot(data = tidy_effects) +
-    geom_point(aes(y=exp(estimate), x=term_midpoint, colour=model_name), position = position_dodge(width = 1.8))+
-    geom_linerange(aes(ymin=exp(conf.low), ymax=exp(conf.high), x=term_midpoint, colour=model_name), position = position_dodge(width = 1.8))+
+    geom_point(aes(y=exp(estimate), x=term_midpoint), position = position_dodge(width = 1.8))+
+    geom_linerange(aes(ymin=exp(conf.low), ymax=exp(conf.high), x=term_midpoint), position = position_dodge(width = 1.8))+
     geom_hline(aes(yintercept=1), colour='grey')+
-    facet_grid(cols=vars(timescale))+
+    facet_grid(rows=vars(timescale), cols=vars(model_name))+
     scale_x_continuous(
@@
   )+
-    scale_colour_brewer(type="qual", palette="Set2", guide=guide_legend(ncol=2))+
     labs(
@@
     geom_hline(aes(yintercept=1), colour='grey')+
-    geom_line(aes(x=tstop-1, y=hr, colour=model_name))+
-    geom_ribbon(aes(x=tstop-1, ymin=hr.ll, ymax=hr.ul, fill=model_name), alpha=0.2, colour="transparent")+
-    facet_grid(cols=vars(timescale))+
+    geom_line(aes(x=tstop-1, y=hr))+
+    geom_ribbon(aes(x=tstop-1, ymin=hr.ll, ymax=hr.ul), alpha=0.2, colour="transparent")+
+    facet_grid(rows=vars(timescale), cols=vars(model_name))+
     scale_x_continuous(
@@
     )+
-    scale_colour_brewer(type="qual", palette="Set2", guide=guide_legend(ncol=1))+
-    scale_fill_brewer(type="qual", palette="Set2", guide="none")+
     labs(
@@
     geom_rect(aes(xmin=lag(tstop, 1, 0), xmax=tstop, ymin=1-survival.ll, ymax=1-survival.ul, group=vax1_az_descr, fill=vax1_az_descr), alpha=0.1)+
-    facet_grid(cols=vars(timescale))+
+    facet_grid(rows=vars(timescale), cols=vars(model_name))+
     scale_x_continuous(
@@
     ggplot(curves)+
-    geom_hline(aes(yintercept=0, colour=model_name))+
-    geom_step(aes(x=tstop, y=diff, colour=model_name))+
-    geom_rect(aes(xmin=lag(tstop, 1, 0), xmax=tstop, ymin=diff.ll, ymax=diff.ul, fill=model_name), alpha=0.1, colour="transparent")+
-    facet_grid(cols=vars(timescale))+
+    geom_hline(aes(yintercept=0))+
+    geom_step(aes(x=tstop, y=diff))+
+    geom_rect(aes(xmin=lag(tstop, 1, 0), xmax=tstop, ymin=diff.ll, ymax=diff.ul), alpha=0.1, colour="transparent")+
+    facet_grid(rows=vars(timescale), cols=vars(model_name))+
     scale_x_continuous(
@@
     )+
-    #coord_cartesian(ylim= c(-5, 5))+
-    scale_colour_brewer(type="qual", palette="Set1", na.value="grey")+
-    scale_fill_brewer(type="qual", palette="Set1", guide="none", na.value="grey")+
     labs(
       x=NULL,
-      y="Difference\nin cumulative incidence",
-      colour=NULL,
-      fill=NULL
+      y="Difference\nin cumulative incidence"
     )+
@@
       plot.caption = element_text(hjust = 0, face= "italic"),
+      legend.position = "bottom"
 
@@
 cat("  \n\n")
-print(glue("### COVID-19 A&E attendance"), "  \n")
-print(glue("#### Unadjusted"), "  \n")
-print_plots_km("covidemergency")
-print(glue("#### Cox"), "  \n")
-print_plots_effects("covidemergency", "cox", "0")
-print(glue("#### PLR, piecewise"), "  \n")
-print_plots_effects("covidemergency", "plr", "0")
-print(glue("#### PLR, spline"), "  \n")
-print_plots_spline("covidemergency", "0")
-print_plots_adjsurv("covidemergency", "0")
-print_plots_diff("covidemergency", "0")
+# print(glue("### COVID-19 A&E attendance"), "  \n")
+# print(glue("#### Unadjusted"), "  \n")
+# print_plots_km("covidemergency")
+# print(glue("#### Cox"), "  \n")
+# print_plots_effects("covidemergency", "cox", "0")
+# print(glue("#### PLR, piecewise"), "  \n")
+# print_plots_effects("covidemergency", "plr", "0")
+# print(glue("#### PLR, spline"), "  \n")
+# print_plots_spline("covidemergency", "0")
+# print_plots_adjsurv("covidemergency", "0")
+# print_plots_diff("covidemergency", "0")
 

### 2021-09-09 create-project.R

@@ -512,10 +512,10 @@ actions_list <- splice(
   ),
 
   action(
-    name = "rmd_report_timescales",
+    name = "rmd_report_models",
     run = glue(
       "r:latest -e {q}",
-      q = single_quote('rmarkdown::render("analysis/report/effectiveness_report_comparetimescales.Rmd",  knit_root_dir = "/workspace",  output_dir = "/workspace/output/report", output_format = c("rmarkdown::github_document")   )')
+      q = single_quote('rmarkdown::render("analysis/report/effectiveness_report_comparemodels.Rmd",  knit_root_dir = "/workspace",  output_dir = "/workspace/output/report", output_format = c("rmarkdown::github_document")   )')
     ),
     needs = splice(
       "design", "data_selection",
@@ -534,8 +534,8 @@ actions_list <- splice(
       )
     ),
     moderately_sensitive = list(
-      html = "output/report/effectiveness_report_comparetimescales.html",
-      md = "output/report/effectiveness_report_comparetimescales.md",
+      html = "output/report/effectiveness_report_comparemodels.html",
+      md = "output/report/effectiveness_report_comparemodels.md",
       figures = "output/report/figures_timescales/*.png"
     )
   )

### 2021-09-09 analysis/report/effectiveness_report_comparemodels.Rmd
remove second dose censoring

@@
 cat("  \n\n")
-# print(glue("### COVID-19 A&E attendance"), "  \n")
-# print(glue("#### Unadjusted"), "  \n")
-# print_plots_km("covidemergency")
-# print(glue("#### Cox"), "  \n")
-# print_plots_effects("covidemergency", "cox", "0")
-# print(glue("#### PLR, piecewise"), "  \n")
-# print_plots_effects("covidemergency", "plr", "0")
-# print(glue("#### PLR, spline"), "  \n")
-# print_plots_spline("covidemergency", "0")
-# print_plots_adjsurv("covidemergency", "0")
-# print_plots_diff("covidemergency", "0")
+print(glue("### COVID-19 A&E attendance"), "  \n")
+print(glue("#### Unadjusted"), "  \n")
+print_plots_km("covidemergency")
+print(glue("#### Cox"), "  \n")
+print_plots_effects("covidemergency", "cox", "0")
+print(glue("#### PLR, piecewise"), "  \n")
+print_plots_effects("covidemergency", "plr", "0")
+print(glue("#### PLR, spline"), "  \n")
+print_plots_spline("covidemergency", "0")
+print_plots_adjsurv("covidemergency", "0")
+print_plots_diff("covidemergency", "0")
 

### 2021-09-09 analysis/descriptive/table1.R

@@ -40,13 +40,13 @@ var_labels <- read_rds(here("output", "data", "metadata_labels.rds"))
 
 list_formula <- read_rds(here("output", "data", "metadata_formulas.rds"))
 list2env(list_formula, globalenv())
-lastfupdaay <- lastfupday20
+lastfupday <- lastfupday20
 
 ## Import processed data ----
 data_cohort <- read_rds(here("output", "data", "data_cohort.rds")) %>%
   mutate(
-    censor_date12 = pmin(vax1_date - 1 + lastfupday12, end_date, dereg_date, death_date, vax2_date, na.rm=TRUE),
-    censor_date20 = pmin(vax1_date - 1 + lastfupday20, end_date, dereg_date, death_date, vax2_date, na.rm=TRUE),
+    censor_date12 = pmin(vax1_date - 1 + lastfupday12, end_date, dereg_date, death_date, na.rm=TRUE),
+    censor_date20 = pmin(vax1_date - 1 + lastfupday20, end_date, dereg_date, death_date, na.rm=TRUE),
     tte_censor12 = tte(vax1_date-1, censor_date12, censor_date12, na.censor=TRUE),
     tte_censor20 = tte(vax1_date-1, censor_date20, censor_date20, na.censor=TRUE),
     fu_days12 = tte_censor12,

### 2021-09-09 analysis/report/draft-manuscript.Rmd
update manuscript

@@
 list2env(list_formula, globalenv())
+lastfupday <- lastfupday20
 ```
@@
 Patient-facing health and social care workers (HCWs) were amongst the first groups eligible for vaccination due to the high occupational exposure to the SARS-CoV-2 virus, and many were vaccinated during the period where both vaccines were widely used <!--#  @collaborative2021 -->.
-This study assesses the comparative effectiveness of one dose of ChAdOx1 with one dose of BNT162b2 in HCWs, using the OpenSAFELY-TPP linked primary care database covering around 40% of England's population.
+This study assesses the effectiveness of ChAdOx1 compared with BNT162b2 in HCWs, including second dose effectiveness, using the OpenSAFELY-TPP linked primary care database covering around 40% of England's population.
 
@@
 
-Vaccinated HCWs were included in the study if: they were registered at a GP practice using TPP's SystmOne clinical information system on the day of they received their first dose of BNT162b2 or ChAdOx1; the date of vaccination was between `r format(as.Date(gbl_vars$start_date_az), "%d %B")` and `r format(as.Date(gbl_vars$lastvax_date), "%d %B %Y")` (`r as.integer(as.Date(gbl_vars$lastvax_date) - as.Date(gbl_vars$start_date_az)) + 1` days), a period when both vaccines were being administered widely; they aged between 18 and 64 inclusive; not classed as Clinically Extremely Vulnerable, as set out by government guidance, at the time of vaccination; information on sex, ethnicity, deprivation, and geographical region was known.
+Vaccinated HCWs were included in the study if: they were registered at a GP practice using TPP's SystmOne clinical information system on the day that they received their first dose of BNT162b2 or ChAdOx1; the date of vaccination was between `r format(as.Date(gbl_vars$start_date_az), "%d %B")` and `r format(as.Date(gbl_vars$lastvax_date), "%d %B %Y")` (`r as.integer(as.Date(gbl_vars$lastvax_date) - as.Date(gbl_vars$start_date_az)) + 1` days), a period when both vaccines were being administered widely; they were aged between 18 and 64 inclusive; not classed as Clinically Extremely Vulnerable, as set out by government guidance, at the time of vaccination; information on sex, ethnicity, deprivation, and geographical region was known.
 
-Study participants were followed up for no more than 14 weeks from the day of the first dose.
-Follow-up was censored at: the study end date, `r format(as.Date(gbl_vars$end_date), "%d %B %Y")`; a second vaccine dose; death or de-registration.
+Study participants were followed up for no more than 20 weeks from the day of the first dose, including time after their second dose.
+Follow-up was censored earlier than this at `r format(as.Date(gbl_vars$end_date), "%d %B %Y")`, death, or de-registration.
 
@@
 Positive SARS-CoV-2 tests were identified using SGSS records and based on swab date.
-Both polymerase chain reaction (PCR) and lateral flow tests are included.
-We did not differentiate between symptomatic and asymptomatic infection.
-A&E attendances were identified using HES emergency care records.
+Both polymerase chain reaction (PCR) and lateral flow tests are included, without differentiation between symptomatic and asymptomatic infection.
+COVID-19 A&E attendances were identified using HES emergency care records.
 COVID-19 hospital admissions were identified using HES in-patient hospital records.
-Admissions where the ICD-10 coded primary or non-primary "reason for admission" included U07.1 ("COVID-19, virus identified") and U07.2 ( "COVID-19, virus not identified") were included <!--# @emergenc --> .
+Admissions where the ICD-10 coded primary or non-primary "reason for admission" included U07.1 ("COVID-19, virus identified") or U07.2 ("COVID-19, virus not identified") were included <!--# @emergenc --> .
 
 Although severe disease (such as requirement for intensive or critical care) and mortality were of interest there were too few events to investigate these outcomes fully.
-Unadjusted rates for these outcomes are reported descriptively.
-COVID-19 critical care admissions were identified via HES.
-COVID-19 deaths were identified using linked death registration data.
-Death with COVID-19 ICD-10 codes (as above) mentioned anywhere on the death certificate (i.e., as an underlying or contributing cause of death) were included.
+Unadjusted incidence of COVID-19 deaths are reported descriptively.
+These were identified using linked death registration data.
+Deaths COVID-19 ICD-10 codes (as above) mentioned anywhere on the death certificate (i.e., as an underlying or contributing cause of death) were included.
 
@@
 
-Participant characteristics used describe the cohort and for confounder adjustment include: age, sex (male or female), English Index of Multiple Deprivation (IMD, grouped by quintiles), ethnicity (Black, Mixed, South Asian, White, Other, as per the UK census), NHS region (East of England, Midlands, London, North East and Yorkshire, North West, South East, South West), number of conditions in the clinically "at risk" (but not clinically extremely vulnerable) classification, as per national prioritisation guidelines, <!--number of previous SARS-CoV-2 tests (via SGSS),--> rurality, evidence of prior SARS-CoV-2 infection (positive test or COVID-19 hospitalisation), learning disabilities, severe mental illness.
+Participant characteristics used describe the cohort and for confounder adjustment include: age, sex (male or female), English Index of Multiple Deprivation (IMD, grouped by quintiles), ethnicity (Black, Mixed, South Asian, White, Other, as per the UK census), NHS region (East of England, Midlands, London, North East and Yorkshire, North West, South East, South West), number of conditions in the clinically "at risk" (but not clinically extremely vulnerable) classification, as per national prioritisation guidelines, <!--number of previous SARS-CoV-2 tests (via SGSS),--> rurality (urban conurbation, urban city or town, rural town or village), evidence of prior SARS-CoV-2 infection (positive test or COVID-19 hospitalisation), learning disabilities, severe mental illness.
 All characteristics were ascertained as at the time of vaccination.
@@
 
-We estimated the relative effectiveness of a first dose of ChAdOx1 versus BNT162b2 using pooled logistic regression (PLR) models <!--# [@cupples1988a; @ngwa2016] --> with the outcome risk estimated at daily intervals.
+We compared the effectiveness of a first dose of ChAdOx1 versus BNT162b2 using pooled logistic regression (PLR), <!--# [@cupples1988a; @ngwa2016] --> with the outcome risk estimated at daily intervals.
 The effect is permitted to vary by time since vaccination to account for the potential differences in vaccine protection over time between the two brands.
-The PLR models approximate Cox models with time-varying effects, and enables the estimation of risk-adjusted survival rates for each vaccine type using the parametric g-formula.
-This is one minus the average of the cumulative incidence for each day of follow-up predicted by the PLR model, under the (counterfactual) assumption that every participant received the BNT162b2 vaccine or that every participant received the ChAdOx1 vaccine.
-Assuming appropriate parameterisation of risk (i.e., adequate confounder adjustment), this replicates the cumulative incidence that would have been observed in an RCT comparing the two vaccines in the population under consideration.
-Standard errors for the PLR model were obtained using the HC0 clustered sandwich estimator to account for within-participant clustering.
-Confidence intervals for the risk-adjusted cumulative incidence were obtained using the delta method (i.e, a first-order Taylor series approximation of the variance).
+A PLR model can be used to approximate Cox models with time-varying treatment effects on the time-since-vaccination timescale, and enables the estimation of risk-adjusted cumulative incidence rates for each vaccine type.
+This is the average over all participants of the cumulative incidence for each day of follow-up predicted by the PLR model, under the (counterfactual) assumption that everyone received the BNT162b2 vaccine or that everyone received the ChAdOx1 vaccine.
+Assuming adequate confounder adjustment, this estimates the vaccine-specific cumulative incidence that would have been observed in an RCT comparing the two vaccines in the population under consideration.
+Standard errors for the PLR model were obtained using the clustered sandwich estimator to account for within-participant clustering.
+Confidence intervals for the vaccine-specific marginal cumulative incidence, and their difference, were obtained using the delta method (i.e, a first-order Taylor series approximation of the variance).
 
-Four models were fit for each outcome, with progressive adjustment for confounders: (1) An unadjusted model using a restricted cubic spline term with 4 degrees of freedom for time since vaccination, vaccine type, and their interaction; (2) Additionally adjusting for region-specific calendar-time effects, by including a restricted cubic spline term with 4 degrees of freedom for the date of vaccination and its interaction with region; (3) additionally adjusting for demographic characteristics; (4) additionally adjusting for clinical characteristics.
+The PLR model included vaccine type, and a vaccine-specific 3-knot restricted cubic spline for the timescale (time since vaccination).
+Knot locations were based on quartiles of the event times.
+Three models were fit for each outcome, with progressive adjustment for confounders: (1) adjusting for region-specific calendar-time effects, by including a 3-knot restricted cubic spline for the date of vaccination and its interaction with region; (2) additionally adjusting for demographic characteristics; (3) additionally adjusting for clinical characteristics.
+
+During the study period, the recommended dosing interval for both vaccines under the national vaccination programme was 12 weeks.
+Under an intention-to-treat approach, comparative effectiveness estimates beyond 14 weeks were considered to be second-dose effects.
+We report the actual timing of second doses to assess the extent of any deviation from the this treatment strategy.
+
+<!-- In a sensitivity analysis, we censored participants who did not receive a second dose by week 14 or who received a dose of a vaccine that was not the same type as their first dose. -->
 
@@
 PLR models are computationally expensive to fit, as the input dataset must be arranged as one row per person per day of follow-up.
-To manage this, a sampling strategy was used such that all those who experienced the event of interest are selected, and a random sample of 50000 who did not experience the event are selected.
-The models are weighted to recover the characteristics of the complete cohort.
-Sampling was based on a ranked hash of the person identifier to ensure the sampled non-events overlapped as much as possible across outcomes.
+To manage this, a sampling strategy was used such that all those who experienced the event of interest are selected (to retain statistical power), and a random sample of 50,000 event-free participants are selected.
+Person-time is weighted by the inverse of the sampling probability to recover the characteristics of the complete cohort.
+<!--Sampling was based on a ranked hash of the person identifier to ensure the sampled non-events overlapped as much as possible across outcomes.-->
 
@@
   mutate(
-    censor_date = pmin(vax1_date - 1 + lastfupday, end_date, dereg_date, death_date, covid_vax_any_2_date, na.rm=TRUE),
+    censor_date = pmin(vax1_date - 1 + lastfupday, end_date, dereg_date, death_date, vax2_date, na.rm=TRUE),
     tte_censor = tte(vax1_date-1, censor_date, censor_date, na.censor=TRUE),
@@
     emergency = censor_indicator(emergency_date, censor_date),
+    covidemergency = censor_indicator(emergency_covid_date, censor_date),
     admitted = censor_indicator(admitted_date, censor_date),
@@
     postest = sum(postest),
+    covidemergency = sum(covidemergency),
     emergency = sum(emergency),
@@
 
-A total of `r label_number(1, big.mark=",")(flowchart[[1, "n"]])` HCWs aged 18-64 receiving a first dose of BNT162b2 or ChAdOx1 between 4 January and 28 February 2021 and actively registered at at TPP practice were identified, with `r label_number(1)(flowchart[[nrow(flowchart), "n"]])` (`r label_percent(0.1)(flowchart[[nrow(flowchart), "pct_all"]])`) meeting the study eligibility criteria.
-`r label_number(1, big.mark=",")(baseline_pfizer$n)` (`r label_percent(0.1)(baseline_pfizer$n/baseline$n)`) were vaccinated with BNT162b2, contributing `r label_number(0.1)(baseline_pfizer$fu_years)` person-years of follow-up.
-For ChAdOx1, this was `r label_number(1, big.mark=",")(baseline_az$n)` (`r label_percent(0.1)(baseline_az$n/baseline$n)`) and `r label_number(0.1)(baseline_az$fu_years)` person-years.
+A total of `r label_number(1, big.mark=",")(flowchart[[1, "n"]])` HCWs aged 18-64 receiving a first dose of BNT162b2 or ChAdOx1 between 4 January and 28 February 2021 and actively registered at at TPP practice were identified, with `r label_number(1, big.mark=",")(flowchart[[nrow(flowchart), "n"]])` (`r label_percent(0.1)(flowchart[[nrow(flowchart), "pct_all"]])`) meeting the study eligibility criteria.
 
@@
 
-Baseline characteristics are reported in Table 1 and were largely well-balanced between recipients of each vaccine, though regional and temporal differences in the distribution of each vaccine are notable.
-BNT162b2 was on average administered earlier than ChAdOx1 (median day of vaccination `r format(baseline_pfizer$median_vaxday, "%d %B")` for BNT162b2, `r format(baseline_az$median_vaxday, "%d %B")` for ChadOx1).
+#### Figure 1: Vaccination dates
+
+```{r, vaxdate, echo=FALSE, message=FALSE, warning=FALSE, out.width='80%', results='asis'}
+vaxdate_plot <- read_rds(here("output", "descriptive", "vaxdate", "plot_vaxdate_stack.rds"))
+print(vaxdate_plot)
+cat("  \n\n")
+```
+
+In total, `r label_number(1, big.mark=",")(baseline_pfizer$n)` (`r label_percent(0.1)(baseline_pfizer$n/baseline$n)`) were vaccinated with BNT162b2, contributing `r label_number(1, big.mark=",")(baseline_pfizer$fu_years)` person-years of potential follow-up (including all person-time prior to a censoring event, and possibly after an outcome event).
+For ChAdOx1, this was `r label_number(1, big.mark=",")(baseline_az$n)` (`r label_percent(0.1)(baseline_az$n/baseline$n)`) and `r label_number(1, big.mark=",")(baseline_az$fu_years)` person-years.
+
+Characteristics were largely well-balanced between recipients of each vaccine, though regional and temporal differences in the distribution of each vaccine are notable (Table 1).
+BNT162b2 was on average administered earlier than ChAdOx1 (Figure 1; median day of vaccination `r format(baseline_pfizer$median_vaxday, "%d %B")` for BNT162b2, `r format(baseline_az$median_vaxday, "%d %B")` for ChadOx1).
 BNT162b2 was relatively more likely to be administered in the South and East of England, and ChAdOx1 the Midlands and Northern England.
+Evidence of prior SARS-CoV-2 infection was higher in ChAdOx1 recipients (`r label_percent(0.1)(baseline_pfizer$priorinfection)` for BNT162b2, `r label_percent(0.1)(baseline_az$priorinfection)` for ChAdOx1), consistent with ChAdOx1 recipients being vaccinated later on average.
 The proportion of each clinical condition is slightly higher in ChAdOx1 recipients, though consistently under a 0.6% percent-point difference.
@@
 
-The cumulative proportion of vaccinated study participants by calendar time is reported in Figure 1.
-
-#### Figure 1: Vaccination dates
-
-```{r, vaxdate, echo=FALSE, message=FALSE, warning=FALSE, out.width='80%', results='asis'}
-vaxdate_plot <- read_rds(here("output", "descriptive", "vaxdate", "plot_vaxdate_stack.rds"))
-print(vaxdate_plot)
-cat("  \n\n")
-```
-
 ## Event rates
 
-Table 2 shows event rates by vaccine type, and the crude incidence.
-Over the duration of `r label_number(0.1)(baseline$fu_years)` person-years of follow-up there were `r baseline$postest` positive SARS-CoV-2 tests, `r baseline$emergency` A&E attendances, `r baseline$covidadmitted` COVID-19 hospital admissions, and `r baseline$coviddeath` COVID-19 deaths.
+Over the duration of `r label_number(1, big.mark=",")(baseline$fu_years)` person-years of follow-up there were `r baseline$postest` positive SARS-CoV-2 tests, `r label_number(1, big.mar=",")(baseline$covidemergency)` COVID-19 A&E attendances, `r label_number(1, big.mar=",")(baseline$covidadmitted)` COVID-19 hospital admissions, and `r if_else(baseline$coviddeath<6, "<6", label_number(1)(baseline$coviddeath))` COVID-19 deaths (Table 2).
 
@@
       "postest",
-      "emergency",
+      "covidemergency",
       "covidadmitted"
@@
 
+  ## cumulative incidence
   cmlinc <-
@@
       outcomes,
-      ~read_rds(here("output", "models", .x, "timesincevax", glue("reportplr_adjustedsurvival_ns.rds"))),
+      ~read_rds(here("output", "models", .x, "timesincevax", "0", glue("reportplr_adjustedsurvival_ns.rds"))),
       .id="outcome"
@@
     ggplot(cmlinc)+
-    geom_step(aes(x=tstop, y=survival, group=vax1_az_descr, colour=vax1_az_descr))+
-    geom_rect(aes(xmin=lag(tstop, 1, 0), xmax=tstop, ymin=survival.ll, ymax=survival.ul, group=vax1_az_descr, fill=vax1_az_descr), alpha=0.1)+
-    geom_hline(aes(yintercept=1), colour='black')+
+    geom_step(aes(x=tstop, y=(1-survival)*1000, group=vax1_az_descr, colour=vax1_az_descr))+
+    geom_rect(aes(xmin=lag(tstop, 1, 0), xmax=tstop, ymin=(1-survival.ll)*1000, ymax=(1-survival.ul)*1000, group=vax1_az_descr, fill=vax1_az_descr), alpha=0.1)+
+    geom_hline(aes(yintercept=0), colour='black')+
     facet_grid(cols=vars(outcome_descr))+
     scale_x_continuous(
-      breaks = seq(0,7*52,by=14),
+      breaks = seq(0,7*52,by=28),
       expand = expansion(0)
@@
       expand = expansion(0),
-      limits = c(NA,1L)
+      limits = c(0L,NA)
     )+
@@
       x=NULL,
-      y="1 - cumulative incidence",
+      y="Marginalised\ncumulative incidence\nper 1000",
       colour=NULL
@@
     theme(
-      legend.position=c(0.9,0.05),
-      legend.justification = c(1,0),
+      legend.position=c(0.05,0.95),
+      legend.justification = c(0,1),
       legend.direction = "vertical",
@@
 
-  hazard <-
-    map_dfr(
-      outcomes,
-      ~read_rds(here("output", "models", .x, "timesincevax", glue("reportplr_effects_ns.rds"))),
-      .id="outcome"
-    ) %>%
-    left_join(metadata_outcomes, by="outcome") %>%
-    mutate(
-      outcome_descr = fct_inorder(outcome_descr),
-      model_name = fct_inorder(str_wrap(model_name, 15))
-    ) %>%
-    filter(model==3)
-
-  plot_hazard <-
-    ggplot(hazard)+
-    geom_hline(aes(yintercept=1), colour='black')+
-    geom_line(aes(x=tstop, y=hr), colour='orchid4')+
-    geom_ribbon(aes(x=tstop, ymin=hr.ll, ymax=hr.ul), alpha=0.1, colour="transparent", fill='orchid4')+
-    facet_grid(cols=vars(outcome_descr))+
-    scale_x_continuous(
-      limits = c(0, NA),
-      breaks = seq(0,7*52,by=14),
-      expand = expansion(0)
-    )+
-    scale_y_log10(
-      #breaks=c(0.125, 0.167, 0.25, 0.333, 0.5, 0.667, 1, 1.5, 2, 3, 4, 6, 8),
-      sec.axis = dup_axis(name="Lower hazard after \n <- BNT162b2 / ChAdOx1 ->", breaks = NULL)
-    )+
-    coord_cartesian(ylim= c(0.1, 10))+
-    labs(
-      x = "Days since first dose",
-      y = "Hazard ratio"
-    )+
-    theme_bw()+
-    theme(
-      panel.border = element_blank(),
-      axis.line.y.left = element_line(colour = "black"),
-
-      panel.grid.minor.x = element_blank(),
-      panel.grid.minor.y = element_blank(),
-      strip.background = element_blank(),
-      strip.text = element_blank(),
-
-      panel.spacing = unit(0.8, "lines"),
-
-      plot.title = element_text(hjust = 0),
-      plot.title.position = "plot",
-      plot.caption.position = "plot",
-      plot.caption = element_text(hjust = 0, face= "italic")
-
-     ) +
-   NULL
+  ## difference in cumulative incidence
 
@@
       outcomes,
-      ~read_rds(here("output", "models", .x, "timesincevax", glue("reportplr_adjusteddiff_ns.rds"))),
+      ~read_rds(here("output", "models", .x, "timesincevax", "0", glue("reportplr_adjusteddiff_ns.rds"))),
       .id="outcome"
@@
       diff_format = label_number(0.01, scale=1000)(diff),
+      diff_CI = paste0("(" , label_number(0.01, scale=1000)(diff.ll), ", ", label_number(0.01, scale=1000)(diff.ul), ")")
     ) %>%
@@
       limits = c(0, NA),
-      breaks = seq(0,7*52,by=14),
+      breaks = seq(0,7*52,by=28),
       expand = expansion(0)
@@
       x=NULL,
-      y="Risk difference\nper 1000",
+      y="Difference\nin cumulative incidence\nper 1000",
       colour=NULL
+    )+
+    theme_bw()+
+    theme(
+      panel.border = element_blank(),
+      axis.line.y.left = element_line(colour = "black"),
+
+      panel.grid.minor.x = element_blank(),
+      panel.grid.minor.y = element_blank(),
+      strip.background = element_blank(),
+      strip.text = element_blank(),
+
+      panel.spacing = unit(0.8, "lines"),
+
+      plot.title = element_text(hjust = 0),
+      plot.title.position = "plot",
+      plot.caption.position = "plot",
+      plot.caption = element_text(hjust = 0, face= "italic")
+
+     ) +
+   NULL
+
+  ## hazard ratio
+
+  hazard <-
+    map_dfr(
+      outcomes,
+      ~read_rds(here("output", "models", .x, "timesincevax", "0", glue("reportplr_effects_ns.rds"))),
+      .id="outcome"
+    ) %>%
+    left_join(metadata_outcomes, by="outcome") %>%
+    mutate(
+      outcome_descr = fct_inorder(outcome_descr),
+      model_name = fct_inorder(str_wrap(model_name, 15))
+    ) %>%
+    filter(model==3)
+
+  plot_hazard <-
+    ggplot(hazard)+
+    geom_hline(aes(yintercept=1), colour='black')+
+    geom_line(aes(x=tstop, y=hr), colour='orchid4')+
+    geom_ribbon(aes(x=tstop, ymin=hr.ll, ymax=hr.ul), alpha=0.1, colour="transparent", fill='orchid4')+
+    facet_grid(cols=vars(outcome_descr))+
+    scale_x_continuous(
+      limits = c(0, NA),
+      breaks = seq(0,7*52,by=28),
+      expand = expansion(0)
+    )+
+    scale_y_log10(
+      breaks=c(0.125, 0.25, 0.5, 1, 2, 4, 8),
+      sec.axis = dup_axis(name="Lower hazard after \n <- BNT162b2 / ChAdOx1 ->", breaks = NULL)
+    )+
+    coord_cartesian(ylim= c(0.1, 10))+
+    labs(
+      x = "Days since first dose",
+      y = "Hazard ratio"
     )+
@@
 
-Figure 3 shows, for each outcome based on the fully adjusted model, the time-dependent hazard ratio for ChAdOx1 versus BNT162b2, and one minus the risk-adjusted cumulative incidence, and the absolute risk difference.
+Figure 2 shows, for each outcome based on the fully adjusted model, the marginal cumulative incidence for ChAdOx1 and BNT162b2, their difference, and the hazard ratio.
 Similar figures for the non-fully-adjusted models are presented in supplementary materials.
 
-At 6 weeks post-vaccination, the absolute risk difference (ChAdOx1 - BNT162b2) for a positive SARS-CoV-2 test was `r filter(riskdiff, outcome=="postest", tstop==7*6)$diff_format` per 1000 people, for A&E attendances was `r filter(riskdiff, outcome=="emergency", tstop==7*6)$diff_format` and for COVID-19 hospital admissions was `r filter(riskdiff, outcome=="covidadmitted", tstop==7*6)$diff_format` per 1000 people.
+At 6 weeks post-vaccination, the ChAdOx1 versus BNT162b2 absolute risk difference per 1000 people (95% CI) for a positive SARS-CoV-2 test was `r filter(riskdiff, outcome=="postest", tstop==7*6)$diff_format` `r filter(riskdiff, outcome=="postest", tstop==7*6)$diff_CI`, for COVID-19 A&E attendances was `r filter(riskdiff, outcome=="covidemergency", tstop==7*6)$diff_format` `r filter(riskdiff, outcome=="covidemergency", tstop==7*6)$diff_CI` and for COVID-19 hospital admissions was `r filter(riskdiff, outcome=="covidadmitted", tstop==7*6)$diff_format` `r filter(riskdiff, outcome=="covidadmitted", tstop==7*6)$diff_CI`.
 
-#### Figure 2: Comparative effectivevess
+At 18 weeks post-vaccination, the absolute risk difference for a positive SARS-CoV-2 test was `r filter(riskdiff, outcome=="postest", tstop==7*18)$diff_format` `r filter(riskdiff, outcome=="postest", tstop==7*18)$diff_CI`), for COVID-19 A&E attendances was `r filter(riskdiff, outcome=="covidemergency", tstop==7*18)$diff_format` (`r filter(riskdiff, outcome=="covidemergency", tstop==7*18)$diff_CI`) and for COVID-19 hospital admissions was `r filter(riskdiff, outcome=="covidadmitted", tstop==7*18)$diff_format` `r filter(riskdiff, outcome=="covidadmitted", tstop==7*18)$diff_CI`.
+
+#### Figure 2: Comparative effectiveness
 
@@
 
-This observational study of almost 1/3 million health and social care workers living in England found no significant differences in rates of COVID-19-related events between those receiving a first dose of BNT162b2 or ChAdOx1.
-For hospitalisation, it was not possible to rule-out large differences in the risk due to uncertainty in the effect estimates.
-There is a clear leveling-off of the event rates after around 3 weeks, consistent with the expected time-to-onset of vaccine-induced immunity of around 2 weeks plus the delay from infection to a positive test.
+This observational study of almost 1/3 million health and social care workers living in England found no significant differences in the incidence of COVID-19-related events between those receiving BNT162b2 or ChAdOx1 at 6 weeks.
+There is a clear leveling-off of the incidence of positive tests after around 3 weeks, consistent with the expected time-to-onset of vaccine-induced immunity of around 2 weeks plus the delay from infection to a positive test.
+After
 
@@
 Despite reasonable balance of vaccine allocation across baseline characteristics and adjustment for a range of potential confounders, the possibility of unmeasured confounding remains.
-The cold-chain storage requirements of BNT162b2 meant that it is more likely to have be administered in Acute NHS Trusts and other large vaccination centres, and is thus a potential confounder for instance due to differences in exposure to risk across these settings.
-Though we adjusted for region, rurality, and deprivation, we were unable to directly account for occupational differences that may affect exposure risk and vaccine type.
+The cold-chain storage requirements of BNT162b2 meant that it is more likely to have be administered in Acute NHS Trusts and other large vaccination centres, and is thus a potential confounder for instance due to differences in viral exposure across these settings.
+Though we adjusted for region, rurality, and deprivation, we were unable to directly account for occupational differences that may affect both exposure risk and vaccine type.
 
@@
 
-A number of studies estimate COVID-19 vaccine effectiveness in observational data using unvaccinated controls <!--# [@cabezas2021; @vasileiou2021]  -->. [describe pfizer versus AZ findings]
+A number of studies estimate COVID-19 vaccine effectiveness in observational data using unvaccinated controls <!--# [@cabezas2021; @vasileiou2021]  -->.
 
-Such designs are extremely vulnerable to confounding for example due to vaccine prioritisation and eligiblity policies, as well as vaccine access and acceptance, that cause substantial imbalance between vaccinated and unvaccinated groups.
+Such designs are extremely vulnerable to confounding for example due to vaccine prioritisation and eligibility policies, as well as vaccine access and acceptance, that cause substantial imbalance between vaccinated and unvaccinated groups.
 Many of these biases can be bypassed when making direct comparisons between recipients of different vaccine types.

### 2021-09-09 analysis/report/draft-supplement.Rmd
add supplement

---
title: "Comparative vaccine effectiveness of ChAdOx1 versus BNT162b2 in Health and Social Care workers in England: Supplementary materials"
output:
  html_document: 
    keep_md: yes
    self_contained: TRUE
  word_document: default
  bookdown::html_document2:
    number_sections: false
    toc: false
#bibliography: references.bib
#csl: nature.csl # from https://raw.githubusercontent.com/citation-style-language/styles/master/nature.csl
#link-citations: yes
#zotero: true
---

```{r setup, include=FALSE}

library('tidyverse')
library('here')
library('glue')
library('lubridate')
library('gt')
library('patchwork')
library('scales')
source(here("analysis", "lib", "utility_functions.R"))

output_dir <- here("output")

fs::dir_create(output_dir, "report", "figures")

knitr::opts_chunk$set(
  echo = TRUE,
  fig.path = paste0(fs::path(output_dir, "report", "figures"), "/")
)

gbl_vars <- jsonlite::fromJSON(
  txt=here("analysis", "global-variables.json")
)

data_cohort <- read_rds(here("output", "data", "data_cohort.rds"))

flowchart <- read_csv(here("output", "data", "flowchart.csv"))

metadata_outcomes <- read_rds(here("output", "data", "metadata_outcomes.rds"))

list_formula <- read_rds(here("output", "data", "metadata_formulas.rds"))
list2env(list_formula, globalenv())
lastfupday <- lastfupday20
```

```{r characteristics, echo=FALSE, warning=FALSE, message=FALSE}

data_cohort <- 
  data_cohort %>%
  mutate(
    censor_date = pmin(vax1_date - 1 + lastfupday, end_date, dereg_date, death_date, vax2_date, na.rm=TRUE),
    tte_censor = tte(vax1_date-1, censor_date, censor_date, na.censor=TRUE),

    postest = censor_indicator(positive_test_date, censor_date),
    emergency = censor_indicator(emergency_date, censor_date),
    covidemergency = censor_indicator(emergency_covid_date, censor_date),
    admitted = censor_indicator(admitted_date, censor_date),
    covidadmitted = censor_indicator(covidadmitted_date, censor_date),
    covidcc = censor_indicator(covidcc_date, censor_date),
    coviddeath = censor_indicator(coviddeath_date, censor_date),
    death = censor_indicator(death_date, censor_date),
  )

baseline <- 
  data_cohort %>%
  summarise(
    n = n(),
    age_median = median(age),
    age_Q1 = quantile(age, 0.25),
    age_Q3 = quantile(age, 0.75),
    female = mean(sex=="Female"),
    fu_years = sum(tte_censor)/365.25,
    fu_median = median(tte_censor),
    priorinfection = mean(prior_covid_infection),

    postest = sum(postest),
    emergency = sum(emergency),
    admitted = sum(admitted),
    covidadmitted = sum(covidadmitted),
    coviddeath = sum(coviddeath),
    death = sum(death),
  )

baseline_az <- 
  data_cohort %>%
  filter(vax1_type=="az") %>%
  summarise(
    n = n(),
    age_median = median(age),
    age_Q1 = quantile(age, 0.25),
    age_Q3 = quantile(age, 0.75),
    female = mean(sex=="Female"),
    fu_years = sum(tte_censor)/365.25,
    fu_median = median(tte_censor),
    priorinfection = mean(prior_covid_infection),
    median_vaxday = first(start_date_az) + floor(median(vax1_day)) -1
  )

baseline_pfizer <- 
  data_cohort %>%
  filter(vax1_type=="pfizer") %>%
  summarise(
    n = n(),
    age_median = median(age),
    age_Q1 = quantile(age, 0.25),
    age_Q3 = quantile(age, 0.75),
    female = mean(sex=="Female"),
    fu_years = sum(tte_censor)/365.25,
    fu_median = median(tte_censor),
    priorinfection = mean(prior_covid_infection),
    median_vaxday = first(start_date_az) + floor(median(vax1_day)) - 1
  )
```

```{r, effects, echo=FALSE, message=FALSE, warning=FALSE}

  outcomes <- 
    c(
      "postest",
      "emergency",
      "covidadmitted"
    ) %>% 
    set_names(., .)

  cmlinc <-
    map_dfr(
      outcomes,
      ~read_rds(here("output", "models", .x, "timesincevax", "0", glue("reportplr_adjustedsurvival_ns.rds"))),
      .id="outcome"
    ) %>%
    left_join(metadata_outcomes, by="outcome") %>%
    mutate(
      outcome_descr = fct_inorder(outcome_descr),
      model_name = fct_inorder(str_wrap(model_name, 15))
    ) %>%
    filter(model %in% c(1,2))

  plot_cmlinc <- function(model){
    modell <- model

    cmlinc %>%
      filter(model == modell) %>%
      ggplot(cmlinc)+
      geom_step(aes(x=tstop, y=(1-survival)*1000, group=vax1_az_descr, colour=vax1_az_descr))+
      geom_rect(aes(xmin=lag(tstop, 1, 0), xmax=tstop, ymin=(1-survival.ll)*1000, ymax=(1-survival.ul)*1000, group=vax1_az_descr, fill=vax1_az_descr), alpha=0.1)+
      geom_hline(aes(yintercept=0), colour='black')+
      facet_grid(cols=vars(outcome_descr))+
      scale_x_continuous(
        breaks = seq(0,7*52,by=14),
        expand = expansion(0)
      )+
      scale_y_continuous(
        expand = expansion(0),
        limits = c(0L,NA)
      )+
      scale_colour_brewer(type="qual", palette="Set1", na.value="grey")+
      scale_fill_brewer(type="qual", palette="Set1", guide="none", na.value="grey")+
      labs(
        x=NULL,
        y="Marginalised cumulative incidence\nper 1000",
        colour=NULL
      )+
      theme_minimal()+
      theme(
        legend.position=c(0.1,0.95),
        legend.justification = c(0,1),
        legend.direction = "vertical",
        axis.text.x.top=element_text(hjust=0),

        panel.border = element_blank(),
        axis.line.y = element_line(colour = "black"),
        axis.line.x.top = element_line(colour = "black"),

        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        strip.background = element_blank(),

        panel.spacing = unit(0.8, "lines"),

      )+
      NULL
  }

  hazard <-
    map_dfr(
      outcomes,
      ~read_rds(here("output", "models", .x, "timesincevax", glue("reportplr_effects_ns.rds"))),
      .id="outcome"
    ) %>%
    left_join(metadata_outcomes, by="outcome") %>%
    mutate(
      outcome_descr = fct_inorder(outcome_descr),
      model_name = fct_inorder(str_wrap(model_name, 15))
    ) %>%
    filter(model %in% c(1,2))

  plot_hazard <- function(model){
    modell <- model

    hazard %>% 
      filter(model == modell) %>%
      ggplot()+
      geom_hline(aes(yintercept=1), colour='black')+
      geom_line(aes(x=tstop, y=hr), colour='orchid4')+
      geom_ribbon(aes(x=tstop, ymin=hr.ll, ymax=hr.ul), alpha=0.1, colour="transparent", fill='orchid4')+
      facet_grid(cols=vars(outcome_descr))+
      scale_x_continuous(
        limits = c(0, NA),
        breaks = seq(0,7*52,by=14),
        expand = expansion(0)
      )+
      scale_y_log10(
        #breaks=c(0.125, 0.167, 0.25, 0.333, 0.5, 0.667, 1, 1.5, 2, 3, 4, 6, 8),
        sec.axis = dup_axis(name="Lower hazard after \n <- BNT162b2 / ChAdOx1 ->", breaks = NULL)
      )+
      coord_cartesian(ylim= c(0.1, 10))+
      labs(
        x = "Days since first dose",
        y = "Hazard ratio"
      )+
      theme_bw()+
      theme(
        panel.border = element_blank(),
        axis.line.y.left = element_line(colour = "black"),

        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        strip.background = element_blank(),
        strip.text = element_blank(),

        panel.spacing = unit(0.8, "lines"),

        plot.title = element_text(hjust = 0),
        plot.title.position = "plot",
        plot.caption.position = "plot",
        plot.caption = element_text(hjust = 0, face= "italic")

       ) +
     NULL
  }

  riskdiff <-
    map_dfr(
      outcomes,
      ~read_rds(here("output", "models", .x, "timesincevax", "0", glue("reportplr_adjusteddiff_ns.rds"))),
      .id="outcome"
    ) %>%
    left_join(metadata_outcomes, by="outcome") %>%
    mutate(
      outcome_descr = fct_inorder(outcome_descr),
      model_name = fct_inorder(str_wrap(model_name, 15)),
      diff_format = label_number(0.01, scale=1000)(diff),
      diff_CI = paste0("(" , label_number(0.01, scale=1000)(diff.ll), ", ", label_number(0.01, scale=1000)(diff.ul), ")")
    ) %>%
    filter(model %in% c(1,2))

  plot_riskdiff <- function(model){

    modell <- model
    riskdiff %>%
      filter(model==modell) %>%
      ggplot(riskdiff)+
      geom_hline(aes(yintercept=0), colour='black')+
      geom_step(aes(x=tstop, y=diff*1000), colour='orchid4')+
      geom_rect(aes(xmin=lag(tstop, 1, 0), xmax=tstop, ymin=diff.ll*1000, ymax=diff.ul*1000), alpha=0.1, colour="transparent", fill='orchid4')+
      facet_grid(cols=vars(outcome_descr))+
      scale_x_continuous(
        limits = c(0, NA),
        breaks = seq(0,7*52,by=14),
        expand = expansion(0)
      )+
      scale_y_continuous(
        expand = expansion(0),
        sec.axis = dup_axis(name="Lower risk after \n <- BNT162b2 / ChAdOx1 ->", breaks = NULL)
      )+
      coord_cartesian(ylim= c(-5, 5))+
      scale_colour_brewer(type="qual", palette="Set1", na.value="grey")+
      scale_fill_brewer(type="qual", palette="Set1", guide="none", na.value="grey")+
      labs(
        x=NULL,
        y="Difference\nin cumulative incidence\nper 1000",
        colour=NULL
      )+
      theme_bw()+
      theme(
        panel.border = element_blank(),
        axis.line.y.left = element_line(colour = "black"),

        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        strip.background = element_blank(),
        strip.text = element_blank(),

        panel.spacing = unit(0.8, "lines"),

        plot.title = element_text(hjust = 0),
        plot.title.position = "plot",
        plot.caption.position = "plot",
        plot.caption = element_text(hjust = 0, face= "italic")

       ) +
     NULL
  }

```

#### Supplmentary Figure 1a: Comparative effectiveness using a minimally adjusted model

```{r comp1, echo=FALSE, message=FALSE, warning=FALSE, out.width='90%', fig.height=8, results='asis'}
patchwork::wrap_plots(plot_cmlinc(1), plot_riskdiff(1), plot_hazard(1), ncol=1)
```

#### Supplmentary Figure 1b: Comparative effectiveness using a demographically adjusted model

```{r comp2, echo=FALSE, message=FALSE, warning=FALSE, out.width='90%', fig.height=8, results='asis'}
patchwork::wrap_plots(plot_cmlinc(2), plot_riskdiff(2), plot_hazard(2), ncol=1)
```

#### Supplmentary Figure 2: Second dose interval by vaccine type

Second doses are censored by study end date, de-registration, death, and additionally by outcome-specific events.

```{r seconddose_postest, echo=FALSE, message=FALSE, warning=FALSE, out.width='90%', fig.height=8, results='asis'}
plot_seconddose_all <- read_rds(here("output", "descriptive", "secondose", "plot_seconddose_all.rds"))
plot_seconddose_all
```

#### Supplmentary Table 1: Baseline characteristics before eligibility exclusions

Characteristics for all HCWs aged 18-64 receiving a first dose of BNT162b2 or ChAdOx1 between 4 January and 28 February 2021 and actively registered at at TPP practice are reported, before exclusions due to missing demographic characteristics and Clinically Extremely Vulnerable status.

```{r table1_allvax, echo=FALSE, warning=FALSE, message=FALSE}
read_rds(here("output", "descriptive", "tables", "table1_allvax.rds"))
```

### 2021-09-09 analysis/scrapheap/model_ssm.R
add scripts to scrapheap

@@ -1,0 +1,218 @@
+
+# # # # # # # # # # # # # # # # # # # # #
+# This script:
+# imports processed data and restricts it to patients in "cohort"
+# fits some marginal structural models for vaccine effectiveness, with different adjustment sets
+# saves model summaries (tables and figures)
+# "tte" = "time-to-event"
+#
+# The script should be run via an action in the project.yaml
+# The script must be accompanied by four arguments,
+# 1. the name of the cohort defined in data_define_cohorts.R
+# 2. the name of the outcome defined in data_define_cohorts.R
+# 4. the stratification variable. Use "all" if no stratification
+# # # # # # # # # # # # # # # # # # # # #
+
+# Preliminaries ----
+
+
+# import command-line arguments ----
+
+args <- commandArgs(trailingOnly=TRUE)
+
+
+if(length(args)==0){
+  # use for interactive testing
+  removeobs <- FALSE
+  outcome <- "admitted"
+  timescale <- "timesincevax"
+} else {
+  removeobs <- TRUE
+  outcome <- args[[1]]
+  timescale <- args[[2]]
+}
+
+
+
+## Import libraries ----
+library('tidyverse')
+library('here')
+library('glue')
+library('survival')
+library('splines')
+library('rstpm2')
+
+## Import custom user functions from lib
+
+source(here("analysis", "lib", "utility_functions.R"))
+source(here("analysis", "lib", "redaction_functions.R"))
+source(here("analysis", "lib", "survival_functions.R"))
+
+
+# create output directories ----
+fs::dir_create(here("output", "models", outcome, timescale))
+
+## create special log file ----
+cat(glue("## script info for {outcome} ##"), "  \n", file = here("output", "models", outcome, timescale, glue("modelcox_log.txt")), append = FALSE)
+## function to pass additional log text
+logoutput <- function(...){
+  cat(..., file = here("output", "models", outcome, timescale, glue("modelcox_log.txt")), sep = "\n  ", append = TRUE)
+  cat("\n", file = here("output", "models", outcome, timescale, glue("modelcox_log.txt")), sep = "\n  ", append = TRUE)
+}
+
+## import metadata ----
+metadata_outcomes <- read_rds(here("output", "data", "metadata_outcomes.rds"))
[…]

### 2021-09-09 analysis/scrapheap/trackdown-commands.R

@@ -1,0 +1,15 @@
+# upload file if it doesn't already exist in google drive
+
+trackdown::upload_file(
+  file = "analysis/report/effectiveness_report.Rmd",
+  gpath = "Research/Vaccine comparative effectiveness",
+  shared_drive = "OpenSAFELY",
+  hide_code = TRUE
+)
+
+#
+trackdown::update_file(
+  file = "analysis/report/effectiveness_report.Rmd",
+  gpath = "analysis/report/effectiveness_report.Rmd",
+  hide_code = TRUE
+)

### 2021-09-09 analysis/scrapheap/varfunctions.R

@@ -1,0 +1,115 @@
+sdfsf
+
+get_varp <- function(model, vcov, newdata){
+  ## function to get (some of the) variance of the cumulative incidence
+  ## this only works because the treatment variable is numeric (0/1)
+  ## if it were a factor, then model.matrix would throw an error about contrasts (because treatment var is set to all one variable)
+
+  # # select only rows for given patient ID and day
+  # newdata <- newdata %>%
+  #   filter(
+  #     tstop <= day,
+  #     patient_id == patient_idd
+  #   )
+
+  tt <- terms(model) # this helpfully grabs the correct spline basis from the model, rather than recalculating based on `newdata`
+  Terms <- delete.response(tt)
+  m.mat <- model.matrix(Terms, data=newdata) # t x k
+  m.coef <- model$coef # 1 x k
+
+  len <- dim(m.mat)[1]
+
+  nu <- m.coef%*%t(m.mat) # t x 1
+
+  logit2nu <- (exp(nu)/((1+exp(nu))^2)) # t x 1
+
+  deriv_nu <- t(m.mat) %*% c(logit2nu)  # k x n
+
+  stopifnot("term indices are not aligned" = rownames(vcov)==names(deriv_nu))
+
+  var <- as.vector(t(deriv_nu) %*% vcov %*% deriv_nu)
+  var
+}
+
+
+get_partialderiv <- function(model, vcov, newdata){
+  ## function to get the partial derivative of the cumulative incidence for a single patient
+
+  tt <- terms(model) # this helpfully grabs the correct spline basis from the model, rather than recalculating based on `newdata`
+  Terms <- delete.response(tt)
+  m.mat <- model.matrix(Terms, data=newdata) # t x k
+  m.coef <- model$coef # 1 x k
+
+  # max follow up time
+  len <- dim(m.mat)[1] # scalar
+
+  # log-odds, nu_t, at time t
+  nu <- m.coef %*% t(m.mat) # T x 1
+  # part of partial derivative
+  pdc <- (exp(nu)/((1+exp(nu))^2)) # T x 1
+  # partial derivative for P_t(theta_t), for time t and term k
+  pderiv_nu <- t(m.mat) %*% c(pdc)  # k x T
+
+  pderiv_nu
+
+  # # "covariance" of var[P_t(theta_t), P_t(theta_t)]
[…]

### 2021-09-09 analysis/scrapheap/variance testing.R

@@ -1,0 +1,142 @@
+
+
+
+
+
+
+if(missing(newdata)){ newdata <- model$model }
+tt <- terms(model) # this helpfully grabs the correct spline basis from the model, rather than recalculating based on `newdata`
+Terms <- delete.response(tt)
+m.mat <- model.matrix(Terms, data=newdata)
+m.coef <- model$coef
+
+
+id <- c(1,1,1,1,2,2,2)
+time <- c(1,2,3,4,1,2,3)
+weights <- c(1,1,1,1,2,2,2)
+
+m.mat <- cbind(
+  time,
+  c(1,1,1,1,2,2,2),
+  c(3,3,3,3,2,2,2)
+)
+m.coef <- c(1,0.4,0.5)
+
+vcov <- cbind(
+  c(1,0.3,0.3),
+  c(0.2,0.5,0.2),
+  c(0.7,0.3,0.7)
+)
+
+
+N <- nrow(m.mat)
+K <- length(m.coef)
+
+# log-odds, nu_t, at time t
+nu <- m.coef %*% t(m.mat) # t_i x 1
+# part of partial derivative
+pdc <- (exp(nu)/((1+exp(nu))^2)) # t_i x 1
+# summand for partial derivative of P_t(theta_t | X_t), for each time t and term k
+
+#summand <- crossprod(diag(as.vector(pdc)), m.mat)    # t_i  x k
+summand <- matrix(0, nrow=N, ncol=K)
+for (k in seq_len(K)){
+  summand[,k] <- m.mat[,k] * as.vector(pdc)
+}
+
+# cumulative sum of summand, by patient_id  # t_i x k
+cmlsum <- matrix(0, nrow=N, ncol=K)
+for (k in seq_len(K)){
+  cmlsum[,k] <- ave(summand[,k], id, FUN=cumsum)
+}
+
+## multiply by model weights (weights are normalised here so we can use `sum` later, not `weighted.mean`)
+normweights <- weights / ave(weights, time, FUN=sum) # t_i x 1
+
+#wgtcmlsum <- crossprod(diag(normweights), cmlsum ) # t_i x k
+wgtcmlsum <- matrix(0, nrow=N, ncol=K)
+for (k in seq_len(K)){
+  wgtcmlsum[,k] <- cmlsum[,k] * normweights
+}
+
[…]

### 2021-09-09 codelists/format_codelists.R
script to format produce table of codelists

@@ -1,0 +1,29 @@
+library('tidyverse')
+library('here')
+library('gt')
+
+# import codelists from json
+codelists <- jsonlite::read_json(
+  path=here("codelists", "codelists.json"),
+)
+
+
+# reformat
+codelists_formatted <- enframe(codelists[[1]]) %>% unnest_wider(value) %>%
+  mutate(
+    file = name,
+    name= str_extract(id, "(?<=/)(.+)(?=/)"),
+    downloaded_at = as.Date(downloaded_at, "%Y-%m-%d")
+  )
+
+# output to html
+codelists_formatted %>%
+  select(name, url, downloaded_at) %>%
+  gt() %>%
+    cols_label(
+      name = "Name",
+      url = "URL",
+      downloaded_at = "Accessed on"
+    ) %>%
+    gtsave(here("output", "codelists.html"))
+

### 2021-09-09 create-project.R
correct dependencies

@@ -475,9 +475,12 @@ actions_list <- splice(
       "descr_table1", "descr_irr",
       "descr_km", "descr_vaxdate",
       as.list(
-        glue(
-          outcome = c("postest", "emergency", "covidemergency", "covidadmitted"),
-          "report_{outcome}_timesincevax_1_plr"
+        glue_data(
+          .x = expand_grid(
+            outcome = c("postest", "emergency", "covidemergency", "covidadmitted"),
+            censor_seconddose = c("0", "1")
+          ),
+          "report_{outcome}_timesincevax_{censor_seconddose}_plr"
         )
       )
     ),
@@ -499,9 +502,12 @@ actions_list <- splice(
       "descr_km", "descr_vaxdate",
       "descr_seconddose",
       as.list(
-        glue(
-          outcome = c("postest", "emergency", "covidemergency", "covidadmitted"),
-          "report_{outcome}_timesincevax_1_plr"
+        glue_data(
+          .x = expand_grid(
+            outcome = c("postest", "emergency", "covidemergency", "covidadmitted"),
+            censor_seconddose = c("0", "1")
+          ),
+          "report_{outcome}_timesincevax_{censor_seconddose}_plr"
         )
       )
     ),

### 2021-09-09 analysis/report/draft-supplement.Rmd
correct directories; plot ticks

@@
       outcomes,
-      ~read_rds(here("output", "models", .x, "timesincevax", glue("reportplr_effects_ns.rds"))),
+      ~read_rds(here("output", "models", .x, "timesincevax", "0", glue("reportplr_effects_ns.rds"))),
       .id="outcome"

### 2021-09-09 analysis/report/effectiveness_report_comparemodels.Rmd
correct directories; plot ticks

@@
     scale_x_continuous(
-      breaks = seq(0, 7*52, 14), 
+      breaks = seq(0, 7*52, 28), 
       limits = c(min(tidy_effects$term_left), 20*7), 
@@
     scale_x_continuous(
-      breaks = seq(0,7*52,by=14),
+      breaks = seq(0,7*52,by=28),
       expand = expansion(0)
@@
   plot_cmlinc <-
-    ggplot(curves %>% filter(model=="3"))+
+    ggplot(curves)+
     geom_step(aes(x=tstop, y=1-survival, group=vax1_az_descr, colour=vax1_az_descr))+
@@
 # print(glue("#### PLR, spline"), "  \n")
-# print_plots_spline("test", "0")
 # print_plots_adjsurv("test", "0")
 # print_plots_diff("test", "0")
+# print_plots_spline("test", "0")
 
@@
 print(glue("#### PLR, spline"), "  \n")
-print_plots_spline("postest", "0")
 print_plots_adjsurv("postest", "0")
 print_plots_diff("postest", "0")
+print_plots_spline("postest", "0")
 
@@
 # print(glue("#### PLR, spline"), "  \n")
-# print_plots_spline("emergency", "0")
 # print_plots_adjsurv("emergency", "0")
 # print_plots_diff("postest", "0")
+# print_plots_spline("emergency", "0")
 
@@
 print(glue("#### PLR, spline"), "  \n")
-print_plots_spline("covidemergency", "0")
 print_plots_adjsurv("covidemergency", "0")
 print_plots_diff("covidemergency", "0")
+print_plots_spline("covidemergency", "0")
 
@@
 # print(glue("#### PLR, spline"), "  \n")
-# print_plots_spline("admitted", "0")
 # print_plots_adjsurv("admitted", "0")
 # print_plots_diff("admitted", "0")
+# print_plots_spline("admitted", "0")
 
@@
 print(glue("#### PLR, spline"), "  \n")
-print_plots_spline("covidadmitted", "0")
 print_plots_adjsurv("covidadmitted", "0")
 print_plots_diff("covidadmitted", "0")
+print_plots_spline("covidadmitted", "0")
 

### 2021-09-09 analysis/report/draft-supplement.Rmd
fix mapping

@@
       filter(model == modell) %>%
-      ggplot(cmlinc)+
+      ggplot()+
       geom_step(aes(x=tstop, y=(1-survival)*1000, group=vax1_az_descr, colour=vax1_az_descr))+
@@
       scale_x_continuous(
-        breaks = seq(0,7*52,by=14),
+        breaks = seq(0,7*52,by=28),
         expand = expansion(0)
@@
         limits = c(0, NA),
-        breaks = seq(0,7*52,by=14),
+        breaks = seq(0,7*52,by=28),
         expand = expansion(0)
@@
       filter(model==modell) %>%
-      ggplot(riskdiff)+
+      ggplot()+
       geom_hline(aes(yintercept=0), colour='black')+

### 2021-09-09 analysis/report/draft-supplement.Rmd
fix path

@@
 ```{r seconddose_postest, echo=FALSE, message=FALSE, warning=FALSE, out.width='90%', fig.height=8, results='asis'}
-plot_seconddose_all <- read_rds(here("output", "descriptive", "secondose", "plot_seconddose_all.rds"))
+plot_seconddose_all <- read_rds(here("output", "descriptive", "seconddose", "plot_seconddose_all.rds"))
 plot_seconddose_all

### 2021-09-09 analysis/descriptive/seconddose.R

@@ -231,7 +231,6 @@ surv_plot <- survtidyall %>%
     strip.text.y = element_text(angle=0)
   )
 
-surv_plot
 
 write_rds(surv_plot, here("output", "descriptive", "seconddose", glue("plot_seconddose_all.rds")))
 ggsave(surv_plot, filename=glue("plot_seconddose_all.png"), path=here("output", "descriptive", "seconddose"))

### 2021-09-09 analysis/report/draft-manuscript.Rmd
separate same/different second dose types; fix in-line code

@@
     theme(
-      legend.position=c(0.05,0.95),
-      legend.justification = c(0,1),
+      legend.position=c(0.95,0.95),
+      legend.justification = c(1,1),
       legend.direction = "vertical",
@@
     )+
-    coord_cartesian(ylim= c(-5, 5))+
+    coord_cartesian(ylim= c(-2.5, 2.5))+
     scale_colour_brewer(type="qual", palette="Set1", na.value="grey")+
@@
 
-At 18 weeks post-vaccination, the absolute risk difference for a positive SARS-CoV-2 test was `r filter(riskdiff, outcome=="postest", tstop==7*18)$diff_format` `r filter(riskdiff, outcome=="postest", tstop==7*18)$diff_CI`), for COVID-19 A&E attendances was `r filter(riskdiff, outcome=="covidemergency", tstop==7*18)$diff_format` (`r filter(riskdiff, outcome=="covidemergency", tstop==7*18)$diff_CI`) and for COVID-19 hospital admissions was `r filter(riskdiff, outcome=="covidadmitted", tstop==7*18)$diff_format` `r filter(riskdiff, outcome=="covidadmitted", tstop==7*18)$diff_CI`.
+At 18 weeks post-vaccination, the absolute risk difference for a positive SARS-CoV-2 test was `r filter(riskdiff, outcome=="postest", tstop==7*18)$diff_format` `r filter(riskdiff, outcome=="postest", tstop==7*18)$diff_CI`, for COVID-19 A&E attendances was `r filter(riskdiff, outcome=="covidemergency", tstop==7*18)$diff_format` `r filter(riskdiff, outcome=="covidemergency", tstop==7*18)$diff_CI` and for COVID-19 hospital admissions was `r filter(riskdiff, outcome=="covidadmitted", tstop==7*18)$diff_format` `r filter(riskdiff, outcome=="covidadmitted", tstop==7*18)$diff_CI`.
 

### 2021-09-09 analysis/report/draft-supplement.Rmd
separate same/different second dose types; fix in-line code

--- 2021-09-09 analysis/report/draft-supplement.Rmd
@@
 
-#### Supplmentary Figure 2: Second dose interval by vaccine type
+#### Supplmentary Figure 2a: Second dose interval, where second dose vaccine is same as first dose vaccine
 
@@
 
-```{r seconddose_postest, echo=FALSE, message=FALSE, warning=FALSE, out.width='90%', fig.height=8, results='asis'}
-plot_seconddose_all <- read_rds(here("output", "descriptive", "secondose", "plot_seconddose_all.rds"))
-plot_seconddose_all
+```{r seconddose_same, echo=FALSE, message=FALSE, warning=FALSE, out.width='90%', fig.height=8, results='asis'}
+read_rds(here("output", "descriptive", "seconddose", "plot_seconddose_samebrand.rds"))
+```
+
+#### Supplmentary Figure 2b: Second dose interval, where second dose vaccine is different than the first dose vaccine
+
+Second doses are censored by study end date, de-registration, death, and additionally by outcome-specific events.
+
+```{r seconddose_diff, echo=FALSE, message=FALSE, warning=FALSE, out.width='90%', fig.height=8, results='asis'}
+read_rds(here("output", "descriptive", "seconddose", "plot_seconddose_diffbrand.rds"))
 ```

### 2021-09-09 analysis/descriptive/seconddose.R

@@ -148,8 +148,8 @@ seconddose <- function(outcome){
     facet_wrap(vars(vax1_type_descr), strip.position="top", ncol=1)+
     scale_colour_brewer(type="qual", palette="Set1", na.value="grey")+
     scale_fill_brewer(type="qual", palette="Set1", guide="none", na.value="grey")+
-    scale_x_continuous(breaks = seq(0,lastfupday,14), expand=expansion(mult=c(0,0.01)))+
-    scale_y_continuous(expand = expansion(mult=c(0,0.01)))+
+    scale_x_continuous(breaks = seq(0,lastfupday,14), expand=expansion(mult=c(0)))+
+    scale_y_continuous(limits=c(0,1), expand = expansion(mult=c(0)))+
     coord_cartesian(xlim=c(0, lastfupday) ,ylim=c(0,NA))+
     labs(
       x="Days since vaccination",
@@ -193,30 +193,27 @@ survtidyall <-
   )
 
 
-surv_plot <- survtidyall %>%
-  mutate(
-    vax1_type_descr = paste0("First dose ", vax1_type_descr)
+surv_plot_samebrand <- survtidyall %>%
+  filter(
+    vax1_type_descr == vax2_type_descr
   ) %>%
   ggplot(
-    aes(group=paste(vax1_type_descr, vax2_type_descr), colour=vax2_type_descr, fill=vax2_type_descr)
+    aes(group=vax1_type_descr, colour=vax1_type_descr, fill=vax1_type_descr),
   ) +
+  geom_hline(aes(yintercept=0), colour='black')+
+  geom_hline(aes(yintercept=1), colour='black')+
   geom_step(aes(x=time, y=1-surv))+
   geom_rect(aes(xmin=time, xmax=leadtime, ymin=1-surv.ll, ymax=1-surv.ul), alpha=0.1, colour="transparent")+
-  geom_hline(aes(yintercept=0), colour='black')+
-  facet_grid(
-    rows=vars(outcome_descr),
-    cols=vars(vax1_type_descr),
-    switch="y"
-  )+
+  facet_wrap(vars(outcome_descr), ncol=1)+
   scale_colour_brewer(type="qual", palette="Set1", na.value="grey")+
   scale_fill_brewer(type="qual", palette="Set1", guide="none", na.value="grey")+
-  scale_x_continuous(breaks = seq(0,lastfupday,14), expand=expansion(mult=c(0,0.01)))+
-  scale_y_continuous(expand = expansion(mult=c(0,0.01)))+
[…]

### 2021-09-10 analysis/report/draft-supplement.Rmd
assume second dose occurs at the start of the day

@@
 
-#### Supplmentary Figure 1a: Comparative effectiveness using a minimally adjusted model
+#### Figure S1a: Comparative effectiveness adjusting for region-specific calendar-time
 
@@
 
-#### Supplmentary Figure 1b: Comparative effectiveness using a demographically adjusted model
+#### Figure S1b: Comparative effectiveness adjusting for region-specific calendar-time and demographic characteristics
 
@@
 
-#### Supplmentary Figure 2a: Second dose interval, where second dose vaccine is same as first dose vaccine
+#### Figure S2: Second dose interval
 
-Second doses are censored by study end date, de-registration, death, and additionally by outcome-specific events.
+Second dose events are censored by study end date, de-registration, death, and additionally by outcome-specific events.
 
@@
 
-#### Supplmentary Figure 2b: Second dose interval, where second dose vaccine is different than the first dose vaccine
+<!-- #### Figure S2b: Second dose interval, where second dose vaccine is different than the first dose vaccine -->
 
-Second doses are censored by study end date, de-registration, death, and additionally by outcome-specific events.
+<!-- Second doses are censored by study end date, de-registration, death, and additionally by outcome-specific events. -->
 
-```{r seconddose_diff, echo=FALSE, message=FALSE, warning=FALSE, out.width='90%', fig.height=8, results='asis'}
-read_rds(here("output", "descriptive", "seconddose", "plot_seconddose_diffbrand.rds"))
-```
+<!-- ```{r seconddose_diff, echo=FALSE, message=FALSE, warning=FALSE, out.width='90%', fig.height=8, results='asis'} -->
 
-#### Supplmentary Table 1: Baseline characteristics before eligibility exclusions
+<!-- read_rds(here("output", "descriptive", "seconddose", "plot_seconddose_diffbrand.rds")) -->
+
+<!-- ``` -->
+
+#### Table S1: Baseline characteristics before eligibility exclusions
 

### 2021-09-10 analysis/descriptive/seconddose.R

@@ -88,14 +88,14 @@ seconddose <- function(outcome){
       ),
       tte_censor = tte(vax1_date-1, censor_date, censor_date),
 
-      tte_vaxpfizer2 = tte(vax1_date-1, vaxpfizer2_date, censor_date),
-      ind_vaxpfizer2 = censor_indicator(vaxpfizer2_date, censor_date),
+      tte_vaxpfizer2 = tte(vax1_date-1, vaxpfizer2_date-1, censor_date),
+      ind_vaxpfizer2 = censor_indicator(vaxpfizer2_date-1, censor_date),
 
-      tte_vaxaz2 = tte(vax1_date-1, vaxaz2_date, censor_date),
-      ind_vaxaz2 = censor_indicator(vaxaz2_date, censor_date),
+      tte_vaxaz2 = tte(vax1_date-1, vaxaz2_date-1, censor_date),
+      ind_vaxaz2 = censor_indicator(vaxaz2_date-1, censor_date),
 
-      tte_vaxmoderna2 = tte(vax1_date-1, vaxmoderna2_date, censor_date),
-      ind_vaxmoderna2 = censor_indicator(vaxmoderna2_date, censor_date),
+      tte_vaxmoderna2 = tte(vax1_date-1, vaxmoderna2_date-1, censor_date),
+      ind_vaxmoderna2 = censor_indicator(vaxmoderna2_date-1, censor_date),
     )
 
 

### 2021-09-10 analysis/models/model_cox.R

@@ -88,7 +88,7 @@ data_tte <- data_cohort %>%
       vax1_date - 1 + lastfupday,
       dereg_date,
       death_date,
-      if_else(rep(censor_seconddose, n())==1, vax2_date, as.Date(Inf)),
+      if_else(rep(censor_seconddose, n())==1, vax2_date-1, as.Date(Inf)),
       end_date,
       na.rm=TRUE
     ),

### 2021-09-10 analysis/models/model_plr.R

@@ -91,7 +91,7 @@ data_tte <- data_cohort %>%
       vax1_date - 1 + lastfupday,
       dereg_date,
       death_date,
-      if_else(rep(censor_seconddose, n())==1, vax2_date, as.Date(Inf)),
+      if_else(rep(censor_seconddose, n())==1, vax2_date-1, as.Date(Inf)),
       end_date,
       na.rm=TRUE
     ),

### 2021-09-10 analysis/descriptive/table_irr.R
don't censor second dose

@@ -58,7 +58,7 @@ data_tte <- data_cohort %>%
     vax1_type,
     end_date,
 
-    censor_date = pmin(vax1_date - 1 + lastfupday, end_date, dereg_date, death_date, vax2_date, na.rm=TRUE),
+    censor_date = pmin(vax1_date - 1 + lastfupday, end_date, dereg_date, death_date, na.rm=TRUE),
 
     # time to last follow up day
     tte_enddate = tte(vax1_date-1, end_date, end_date),

### 2021-09-10 analysis/report/draft-manuscript.Rmd
add second dose to irr table; remove fu time from table1;

@@
   mutate(
-    censor_date = pmin(vax1_date - 1 + lastfupday, end_date, dereg_date, death_date, vax2_date, na.rm=TRUE),
+    censor_date = pmin(vax1_date - 1 + lastfupday, end_date, dereg_date, death_date, na.rm=TRUE),
     tte_censor = tte(vax1_date-1, censor_date, censor_date, na.censor=TRUE),
+    tte_seconddose = tte(vax1_date-1, vax2_date-1, censor_date, na.censor=TRUE),
 
@@
     death = censor_indicator(death_date, censor_date),
+
+    seconddose = censor_indicator(vax2_date-1, censor_date)
   )
@@
     priorinfection = mean(prior_covid_infection),
+    seconddose_12week = mean(replace_na(tte_seconddose<=12*7, FALSE)),
 
@@
     death = sum(death),
+    seconddose = sum(seconddose),
   )
@@
     priorinfection = mean(prior_covid_infection),
-    median_vaxday = first(start_date_az) + floor(median(vax1_day)) -1
+    median_vaxday = first(start_date_az) + floor(median(vax1_day)) -1,
+    seconddose_12week = mean(replace_na(tte_seconddose<=12*7, FALSE)),
+
   )
@@
     priorinfection = mean(prior_covid_infection),
-    median_vaxday = first(start_date_az) + floor(median(vax1_day)) - 1
+    median_vaxday = first(start_date_az) + floor(median(vax1_day)) - 1,
+    seconddose_12week = mean(replace_na(tte_seconddose<=12*7, FALSE)),
   )
@@
 Over the duration of `r label_number(1, big.mark=",")(baseline$fu_years)` person-years of follow-up there were `r baseline$postest` positive SARS-CoV-2 tests, `r label_number(1, big.mar=",")(baseline$covidemergency)` COVID-19 A&E attendances, `r label_number(1, big.mar=",")(baseline$covidadmitted)` COVID-19 hospital admissions, and `r if_else(baseline$coviddeath<6, "<6", label_number(1)(baseline$coviddeath))` COVID-19 deaths (Table 2).
+
+At 12 weeks, `r label_percent(0.1)(baseline_pfizer$seconddose_12week)` of BNT162b2 recipients had received a second dose, compared with `r label_percent(0.1)(baseline_az$seconddose_12week)` of ChAdOx1 recipients.
 
@@
 This observational study of almost 1/3 million health and social care workers living in England found no significant differences in the incidence of COVID-19-related events between those receiving BNT162b2 or ChAdOx1 at 6 weeks.
-There is a clear leveling-off of the incidence of positive tests after around 3 weeks, consistent with the expected time-to-onset of vaccine-induced immunity of around 2 weeks plus the delay from infection to a positive test.
-After
+There is a clear leveling-off of the incidence of positive tests after around 3-4 weeks, consistent with the expected time-to-onset of vaccine-induced immunity of around 2 weeks plus the delay from infection to a positive test.
 
@@
 Though we adjusted for region, rurality, and deprivation, we were unable to directly account for occupational differences that may affect both exposure risk and vaccine type.
+
+Second-dose effects were considered using an intention-to-treat approach, which does not account for potential differences in treatment adherence between vaccine types.
+However, we found that by 12 weeks more BNT162b2 recipients had received a second dose than ChAdOx1 recipients, which would bias effect estimates in favour of BNT162b2 assuming higher protection after a second-dose of either vaccine.
+Either way, there was insufficient power to reliably estimate comparative second-dose effects as event incidence at 12 weeks and beyond was low for both vaccines, with fewer than 10 positive tests per 1000-person-years, and far fewer for COVID-19 A&E attendances and hospital admissions during this period.
+The need for many thousands of second-dose person-years to detect differences in second-dose effects is encouraging to the extent that it suggests extremely high protective effects for both vaccines, though the fact that background case rates in England were declining over the duration of the study period should not be ignored <!-- covid dashboard -->.
+We also cannot rule-out potential waning effects, which we did not study.
 

### 2021-09-10 analysis/descriptive/table1.R

@@ -58,7 +58,7 @@ data_cohort <- read_rds(here("output", "data", "data_cohort.rds")) %>%
 tab_summary_baseline <- data_cohort %>%
   select(
     all_of(names(var_labels)),
-    -age, -stp, -vax1_type,
+    -age, -stp, -vax1_type, -fu_days12, -fu_days20
   ) %>%
   tbl_summary(
     by = vax1_type_descr,

### 2021-09-10 analysis/descriptive/table_irr.R

@@ -96,6 +96,9 @@ data_tte <- data_cohort %>%
     tte_death = tte(vax1_date-1, death_date, censor_date, na.censor=TRUE),
     ind_death = censor_indicator(death_date, censor_date),
 
+    tte_seconddose = tte(vax1_date-1, vax2_date-1, censor_date, na.censor=TRUE),
+    ind_seconddose = censor_indicator(vax2_date-1, censor_date),
+
   ) %>%
   filter(
     # TDOD remove once study def rerun with new dereg date
@@ -132,6 +135,7 @@ data_cox_split <- tmerge(
   coviddeath = event(tte_coviddeath),
   noncoviddeath = event(tte_noncoviddeath),
   death = event(tte_death),
+  seconddose = event(tte_seconddose),
 
 
   status_test = tdc(tte_test),
@@ -143,7 +147,8 @@ data_cox_split <- tmerge(
   status_covidcc = tdc(tte_covidcc),
   status_coviddeath = tdc(tte_coviddeath),
   status_noncoviddeath = tdc(tte_noncoviddeath),
-  status_death = tdc(tte_death)
+  status_death = tdc(tte_death),
+  status_seconddose = tdc(tte_seconddose),
 
 ) %>%
   tmerge( # create treatment timescale variables
@@ -262,10 +267,11 @@ data_summary <- local({
   temp8 <- pt_summary(data_cox_split, "coviddeath")
   temp9 <- pt_summary(data_cox_split, "noncoviddeath")
   temp10 <- pt_summary(data_cox_split, "death")
+  temp11 <- pt_summary(data_cox_split, "seconddose")
   bind_rows(
     temp1, temp2, temp3, temp4,
     temp5, temp6, temp7, temp8,
-    temp9, temp10
+    temp9, temp10, temp11
   )
 }) %>%
 left_join(
@@ -327,11 +333,11 @@ tab_summary_simple <-
   filter(
     event %in% c(
       "postest",
-      "emergency",
       "covidemergency",
       "covidadmitted",
       "coviddeath",
-      "death"
+      "death",
+      "seconddose"
     )
   ) %>%
   transmute(

### 2021-10-08 analysis/lib/redaction_functions.R
integer threshold

@@ -240,7 +240,7 @@ redacted_summary_catcat <- function(
   variable2,
   .missing_name = "(missing)",
   .redacted_name="redacted",
-  redaction_threshold=5,
+  redaction_threshold=5L,
   redaction_accuracy=1L,
   .total_name=NULL
 ){
@@ -269,7 +269,6 @@ redacted_summary_catcat <- function(
       pct_nonmiss = (n_nonmiss/sum(n_nonmiss, na.rm=TRUE)),
     ) %>%
     select(-n_nonmiss)
-
 
 
   dat_freq_redact0 <- dat_freq %>%

### 2021-10-08 analysis/descriptive/seconddose.R
merge crosss-brand  second doses

@@ -49,7 +49,6 @@ ceiling_any <- function(x, to=1){
 data_cohort <- read_rds(here::here("output", "data", "data_cohort.rds"))
 
 
-
 data_tte <-
   data_cohort %>%
   mutate(
@@ -71,12 +70,13 @@ data_tte <-
 
 data_tab <- data_tte %>%
   mutate(
-    vax2_type_descr =if_else(vax2_day-vax1_day<20*7, vax2_type_descr, factor(NA_character_))
+    vax2_type_descr =if_else((vax2_day-vax1_day)<=20*7, vax2_type_descr, factor(NA_character_)),
+    vaxmatch = vax2_type_descr == vax1_type_descr
   )
 
 dose_tab <- redacted_summary_catcat(
-  data_tab$vax2_type_descr, data_tab$vax1_type_descr,
-  .missing_name="No second dose",
+  data_tab$vaxmatch, data_tab$vax1_type_descr,
+  .missing_name="no second dose at 20 weeks",
   .redacted_name="REDACTED",
   redaction_threshold=5L,
   redaction_accuracy=1L,
@@ -84,7 +84,7 @@ dose_tab <- redacted_summary_catcat(
 ) %>%
   select(
     `First dose`=.level2,
-    `Second dose, up to 20 weeks`=.level1,
+    `Matching second dose`=.level1,
     everything()
   )
 

### 2021-10-08 analysis/descriptive/seconddose.R
compare properly if different factor levels

@@ -71,7 +71,7 @@ data_tte <-
 data_tab <- data_tte %>%
   mutate(
     vax2_type_descr =if_else((vax2_day-vax1_day)<=20*7, vax2_type_descr, factor(NA_character_)),
-    vaxmatch = vax2_type_descr == vax1_type_descr
+    vaxmatch = as.character(vax2_type_descr) == as.character(vax1_type_descr)
   )
 
 dose_tab <- redacted_summary_catcat(

### 2021-10-08 analysis/report/effectiveness_report.Rmd
simplify intro

@@
 
-The COVID-19 global pandemic has prompted the rapid development and delivery of vaccines to combat the disease.
-Following demonstration of high safety and efficacy against symptomatic and severe disease in phase-III randomised clinical trials (RCTs), two vaccines have been approved and widely administered as part of the national vaccination programme in the United Kingdom: the Pfizer-BioNTech BNT162b2 mRNA COVID-19 vaccine (*BNT162b2*) and the Oxford-AstraZeneca ChAdOx1 nCoV-19 vaccine (*ChAdOx1*).
-To date, there have been no RCTs that have directly compared these two vaccines to estimate the relative efficacy against COVID-19 infection and disease in the same cohort.
-The concurrent roll-out of BNT162b2 and ChAdOx1 within the same UK population provides an opportunity to emulate such a trial using observational data.
-
-COVID-19 vaccination in the UK has been prioritised based on the risk of infection and subsequent severity of disease.
-Patient-facing health and social care workers (HCWs) were amongst the first groups eligible for vaccination due to the high occupational exposure to the SARS-CoV-2 virus, and many were vaccinated during the period where both vaccines were widely used (`r format(as.Date(gbl_vars$start_date_az), "%d %B %Y")` onwards).
 This study assessed the comparative effectiveness of one dose of ChAdOx1 with one dose of BNT162b2 in HCWs, using the OpenSAFELY-TPP database covering around 40% of England's population.
@@
 
-Study participants were followed up for 14 weeks from the date of the first dose, or until the first of:
+Study participants were followed up for 20 weeks from the date of the first dose, or until the first of:
 
 -   death or deregistration;
--   Second vaccine dose;
 -   The study end date, `r format(as.Date(gbl_vars$end_date), "%d %B %Y")`.

### 2021-10-13 manuscript/draft-manuscript.Rmd
add manuscript files

---
title: "Comparative effectiveness of ChAdOx1 versus BNT162b2 COVID-19 vaccines in Health and Social Care workers in England: a cohort study using OpenSAFELY"
output:
  pdf_document:
    toc: no
  html_document:
    keep_md: yes
    self_contained: yes
  bookdown::html_document2:
    number_sections: no
    toc: no
  word_document: default
bibliography: references.bib
csl: vancouver.csl
link-citations: yes
zotero: yes
header-includes:
- \usepackage{float} 
- \floatplacement{figure}{H}
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
- \captionsetup[table]{labelformat=empty}
---

```{r setup, include=FALSE}

library('tidyverse')
library('here')
library('glue')
library('lubridate')
library('gt')
library('patchwork')
library('scales')
source(here("analysis", "lib", "utility_functions.R"))

# where are the outputs (ie the inputs for this manuscript!) saved?
#output_dir_os <- here("output")
output_dir_os <- here("released_outputs", "output")

# where should we put the objects created within this rmd script?
output_dir_rmd <- here("manuscript", "figures")
fs::dir_create(output_dir_rmd)

# only applicable if `self-contained: yes` option is enables in yaml header
knitr::opts_chunk$set(
  echo = TRUE,
  fig.path = paste0(output_dir_rmd, "/"),
  fig.pos = 'H' #to stop figures floating around when rendering to pdf
)

gbl_vars <- jsonlite::read_json(
  path=here("output", "data", "metadata_study-dates.json")
)

flowchart <- read_csv(fs::path(output_dir_os, "data", "flowchart.csv"))

```

```{r characteristics, echo=FALSE, warning=FALSE, message=FALSE}

baseline <- read_csv(fs::path(output_dir_os, "report", "objects", "baseline.csv"))
baseline_az <- read_csv(fs::path(output_dir_os, "report", "objects", "baseline_az.csv"))
baseline_pfizer <- read_csv(fs::path(output_dir_os, "report", "objects", "baseline_pfizer.csv"))

seconddose <- read_csv(fs::path(output_dir_os, "descriptive", "seconddose", "seconddose.csv"))

samebrand_n <- seconddose %>% filter(`Matching second dose`=="yes") %>% pull(n) %>% sum
noseconddose_n <- seconddose %>% filter(`Matching second dose`=="no second dose at 20 weeks") %>% pull(n) %>% sum
crossbrand_n <- seconddose %>% filter(`Matching second dose`=="no") %>% pull(n) %>% sum
crossbrand_prop <- crossbrand_n/baseline$n
```

```{r, effects, echo=FALSE, message=FALSE, warning=FALSE}

  outcomes <- 
    c(
      "postest",
      "covidemergency",
      "covidadmitted"
    ) %>% 
    set_names(., .)

cmlinc_ns <- read_csv(fs::path(output_dir_os, "report", "objects", "cmlinc_ns.csv"))  %>%
  mutate(
    outcome_descr = fct_inorder(outcome_descr),
    model_name = fct_inorder(str_wrap(model_name, 15)),
    inc_CI_to = paste0(label_number(0.1, scale=1000)(1-survival.ul), " to ", label_number(0.1, scale=1000)(1-survival.ll))
  )
cmlinc_pw <- read_csv(fs::path(output_dir_os, "report", "objects", "cmlinc_pw.csv")) %>%
  mutate(
    outcome_descr = fct_inorder(outcome_descr),
    model_name = fct_inorder(str_wrap(model_name, 15)),
    inc_CI_to = paste0(label_number(0.1, scale=1000)(1-survival.ul), " to ", label_number(0.1, scale=1000)(1-survival.ll))
  )

riskdiff_ns <- read_csv(fs::path(output_dir_os, "report", "objects", "riskdiff_ns.csv")) %>%
  mutate(
    outcome_descr = fct_inorder(outcome_descr),
    model_name = fct_inorder(str_wrap(model_name, 15)),
    diff_CI_to = paste0(label_number(0.01, scale=1000)(diff.ll), " to ", label_number(0.01, scale=1000)(diff.ul))
  )
riskdiff_pw <- read_csv(fs::path(output_dir_os, "report", "objects", "riskdiff_pw.csv")) %>%
  mutate(
    outcome_descr = fct_inorder(outcome_descr),
    model_name = fct_inorder(str_wrap(model_name, 15)),
    diff_CI_to = paste0(label_number(0.01, scale=1000)(diff.ll), " to ", label_number(0.01, scale=1000)(diff.ul))
  )

hazard_ns <- read_csv(fs::path(output_dir_os, "report", "objects", "hazard_ns.csv")) %>%
  mutate(
    outcome_descr = fct_inorder(outcome_descr),
    model_name = fct_inorder(str_wrap(model_name, 15))
  )
hazard_pw <- read_csv(fs::path(output_dir_os, "report", "objects", "hazard_pw.csv")) %>%
  mutate(
    outcome_descr = fct_inorder(outcome_descr),
    model_name = fct_inorder(str_wrap(model_name, 15))
  )

cmlinc_ns3 <- cmlinc_ns %>% filter(model==3)
cmlinc_pw3 <- cmlinc_pw %>% filter(model==3)
riskdiff_ns3 <- riskdiff_ns %>% filter(model==3)
riskdiff_pw3 <- riskdiff_pw %>% filter(model==3)
hazard_ns3 <- hazard_ns %>% filter(model==3)
hazard_pw3 <- hazard_pw %>% filter(model==3)

```

William J Hulme^1^, Elizabeth J Williamson^2^, Amelia Green^1^, Krishnan Bhaskaran^2^, Helen I McDonald^2^, Christopher T Rentsch^2^, Anna Schultze^2^, John Tazare^2^, Helen J Curtis^1^, Alex J Walker^1^, Laurie Tomlinson^2^, Tom Palmer^4,5^, Elsie Horne^4,5^, Brian MacKenna^1^, Caroline E Morton^1^, Amir Mehrkar^1^, Louis Fisher^1^, Seb Bacon^1^, Dave Evans^1^, Peter Inglesby^1^, George Hickman^1^, Simon Davy^1^, Tom Ward ^1^, Richard Croker^1^, Rosalind M Eggo^2^, Angel YS Wong^2^, Rohini Mathur^2^, Kevin Wing^2^, Harriet Forbes^2^, Daniel Grint^2^, Ian J Douglas^2^, Stephen JW Evans^2^, Liam Smeeth^2^, Chris Bates^3^, Jonathan Cockburn^3^, John Parry^3^, Frank Hester^3^, Sam Harper^3^, Jonathan AC Sterne^4,5^, Miguel Hernán^6,7^, Ben Goldacre^1^

1\. The DataLab, Nuffield Department of Primary Care Health Sciences, University of Oxford, OX26GG, UK

2\. London School of Hygiene and Tropical Medicine, Keppel Street, London, WC1E 7HT, UK

3\. TPP, TPP House, 129 Low Lane, Horsforth, Leeds, LS18 5PX

4\. Population Health Sciences, University of Bristol, Oakfield House, Oakfield Grove, Bristol, BS8 2BN, UK

5\. NIHR Bristol, Biomedical Research Centre, Bristol, UK

6\. CAUSALab, Harvard T.H. Chan School of Public Health, Boston, MA 02115

7\. Departments of Epidemiology and Biostatistics, Harvard T.H. Chan School of Public Health, Boston, MA 02115

\newpage

## Abstract

<!-- ### Standard -->

<!-- *Background* -->

<!-- The UK COVID-19 vaccination programme delivered both the BNT162b2 mRNA (Pfizer-BioNTech) and the ChAdOx1 (Oxford-AstraZeneca) vaccines during overlapping periods, providing a rare opportunity to emulate a trial that directly compares both vaccines using routinely-collected NHS data. Frontline Health and Social Care workers comprise a useful population to assess comparative effectiveness due to early vaccine eligibility and relatively high post-vaccination transmission risk due to occupational exposure. -->

<!-- *Methods* -->

<!-- With the approval of NHS England we used the OpenSAFELY-TPP database, covering 40% of GP practices in England and linked to national coronavirus surveillance, hospital episodes, and death registry data, to compare the effectiveness of ChAdOx1 versus BNT162b2 in `r label_number(1, big.mark=",")(flowchart[[nrow(flowchart), "n"]])` health and social care workers vaccinated between 4 January and 28 February 2021. Recipients were followed-up for 20 weeks. Second-dose effects were estimated under an intention-to-treat strategy. Primary outcomes were recorded SARS-CoV-2 infection, COVID-19-related accident and emergency attendance, and unplanned COVID-19-related hospital admission. -->

<!-- *Results* -->

<!-- The cumulative incidence of each outcome was similar for both vaccines during the first 20 weeks post-vaccination. The cumulative incidence of recorded SARS-CoV-2 infection 6 weeks after vaccination with BNT162b2 was `r filter(cmlinc_ns3, outcome=="postest", vax1_az==0, tstop==7*6)$inc_format` per 1000 people (95%CI `r filter(cmlinc_ns3, outcome=="postest", vax1_az==0, tstop==7*6)$inc_CI_to`) and with ChAdOx1 was `r filter(cmlinc_ns3, outcome=="postest", vax1_az==1, tstop==7*6)$inc_format` (95%CI `r filter(cmlinc_ns3, outcome=="postest", vax1_az==1, tstop==7*6)$inc_CI_to`), representing a difference of `r filter(riskdiff_ns3, outcome=="postest", tstop==7*6)$diff_format` per 1000 people (95%CI `r filter(riskdiff_ns3, outcome=="postest", tstop==7*6)$diff_CI_to`). The difference in the cumulative incidence per 1000 people of COVID-19 accident and emergency attendance at 6 weeks was `r filter(riskdiff_ns3, outcome=="covidemergency", tstop==7*6)$diff_format` per 1000 people (95%CI `r filter(riskdiff_ns3, outcome=="covidemergency", tstop==7*6)$diff_CI_to`). For COVID-19 hospital admission, this difference was `r filter(riskdiff_ns3, outcome=="covidadmitted", tstop==7*6)$diff_format` per 1000 people (95%CI `r filter(riskdiff_ns3, outcome=="covidadmitted", tstop==7*6)$diff_CI_to`). -->

<!-- *Conclusion* -->

<!-- In this cohort of healthcare workers where we would not anticipate vaccine type to be related to health status, we found no substantial differences in the incidence of SARS-CoV-2 infection or COVID-19 disease up to 20 weeks after vaccination. Incidence dropped sharply after 3-4 weeks and there were very few COVID-19 hospital attendance and admission events after this period. This is in line with expected onset of vaccine-induced immunity, and suggests strong protection against COVID-19 disease for both vaccines. -->

<!-- **Keywords** COVID-19; Vaccine effectiveness; Target trial; EHR data -->

*Objectives* To compare the effectiveness of the BNT162b2 mRNA (Pfizer-BioNTech) and the ChAdOx1 (Oxford-AstraZeneca) COVID-19 vaccines against infection and COVID-19 disease in health and social care workers.

*Design* Cohort study, emulating a comparative effectiveness trial.

*Setting* Linked primary care, hospital, and COVID-19 surveillance records available within the OpenSAFELY-TPP research platform.

*Participants* `r label_number(1, big.mark=",")(flowchart[[nrow(flowchart), "n"]])` health and social care workers vaccinated between 4 January and 28 February 2021, registered with a GP practice using the TPP SystmOne clinical information system in England, and not clinically extremely vulnerable.

*Interventions* Vaccination with either BNT162b2 or ChAdOx1 administered as part of the national COVID-19 vaccine roll-out.

*Main outcome measures* Recorded SARS-CoV-2 positive test, or COVID-19 related Accident and Emergency attendance or hospital admission occurring within 20 weeks of vaccination.

*Results* The cumulative incidence of each outcome was similar for both vaccines during the first 20 weeks post-vaccination. The cumulative incidence of recorded SARS-CoV-2 infection 6 weeks after vaccination with BNT162b2 was `r filter(cmlinc_ns3, outcome=="postest", vax1_az==0, tstop==7*6)$inc_format` per 1000 people (95%CI `r filter(cmlinc_ns3, outcome=="postest", vax1_az==0, tstop==7*6)$inc_CI_to`) and with ChAdOx1 was `r filter(cmlinc_ns3, outcome=="postest", vax1_az==1, tstop==7*6)$inc_format` (95%CI `r filter(cmlinc_ns3, outcome=="postest", vax1_az==1, tstop==7*6)$inc_CI_to`), representing a difference of `r filter(riskdiff_ns3, outcome=="postest", tstop==7*6)$diff_format` per 1000 people (95%CI `r filter(riskdiff_ns3, outcome=="postest", tstop==7*6)$diff_CI_to`). The difference in the cumulative incidence per 1000 people of COVID-19 accident and emergency attendance at 6 weeks was `r filter(riskdiff_ns3, outcome=="covidemergency", tstop==7*6)$diff_format` per 1000 people (95%CI `r filter(riskdiff_ns3, outcome=="covidemergency", tstop==7*6)$diff_CI_to`). For COVID-19 hospital admission, this difference was `r filter(riskdiff_ns3, outcome=="covidadmitted", tstop==7*6)$diff_format` per 1000 people (95%CI `r filter(riskdiff_ns3, outcome=="covidadmitted", tstop==7*6)$diff_CI_to`).

*Conclusions* In this cohort of healthcare workers where we would not anticipate vaccine type to be related to health status, we found no substantial differences in the incidence of SARS-CoV-2 infection or COVID-19 disease up to 20 weeks after vaccination. Incidence dropped sharply after 3-4 weeks and there were very few COVID-19 hospital attendance and admission events after this period. This is in line with expected onset of vaccine-induced immunity, and suggests strong protection against COVID-19 disease for both vaccines.

\newpage

# Introduction

The COVID-19 global pandemic has prompted the rapid development and delivery of vaccines to combat the disease. Following demonstration of high safety and efficacy against symptomatic and severe disease in phase-III randomised controlled trials (RCTs), two vaccines have been approved and widely administered as part of the national vaccination programme in the United Kingdom: the Pfizer-BioNTech BNT162b2 mRNA COVID-19 vaccine @polack2020 (*BNT162b2*) and the Oxford-AstraZeneca ChAdOx1 nCoV-19 vaccine @voysey2021 (*ChAdOx1*). Post-authorisation assessment of vaccine effectiveness using observational data is necessary to monitor the success of such programmes as, invariably, target populations and settings differ substantially from those of trials. To date, there have been no RCTs that have directly compared the BNT162b2 and ChAdOx1 vaccines to estimate the relative efficacy against COVID-19 infection and disease in the same population. The concurrent roll-out of these vaccines across the UK @collaborative2021, combined with the country's well-developed infrastructure for conducting research using routinely-collected primary care health data, provides a rare opportunity to emulate such a trial using observational data.

COVID-19 vaccination in the UK has been prioritised based on the risk of infection and subsequent severity of disease @covid19. Health and social care workers (HCWs) were amongst the first groups eligible for vaccination due to the high occupational exposure to the SARS-CoV-2 virus, making HCW status an important confounder of the effect of vaccination on infection and post-infection disease. By studying this group in isolation, this confounding is mitigated, and vaccine effects in a healthy, high-exposure group can be estimated.

This cohort study used a target-trial design to assess the effectiveness of ChAdOx1 compared with BNT162b2 in HCWs, including second dose effectiveness, using the OpenSAFELY-TPP linked primary care database covering around 40% of England's population.

# Methods

## Data source

The OpenSAFELY-TPP database (<https://opensafely.org>) covers 24 million people registered at GP practices that use TPP SystmOne electronic health record software. It includes pseudonymized data such as coded diagnoses, medications and physiological parameters. No free text data are included. This primary care data is linked (via NHS numbers) with A&E attendance and in-patient hospital spell records via NHS Digital's Hospital Episode Statistics (HES), national coronavirus testing records via the Second Generation Surveillance System (SGSS), and national death registry records from the Office for National Statistics (ONS). Vaccination status is available in the GP record directly via the National Immunisation Management System (NIMS). HCW status is recorded for all vaccine recipients at the time of vaccination, and this information is sent to OpenSAFELY-TPP from NHS Digital's COVID-19 data store.

## Study population

We studied health and social care workers in England vaccinated with either BNT162b2 or ChAdOx1. This group was prioritised for vaccination at the start of the vaccine roll-out due to the high occupational exposure to the SARS-CoV-2 virus, and many were vaccinated during the period where both vaccines were widely used.

Vaccinated HCWs were included in the study if: they were registered at a GP practice using TPP's SystmOne clinical information system on the day that they received their first dose of BNT162b2 or ChAdOx1; the date of vaccination was between `r format(as.Date(gbl_vars$start_date_az), "%d %B")` (when ChAdOx1 was first administered in England) and `r format(as.Date(gbl_vars$lastvax_date), "%d %B %Y")` , a `r as.integer(as.Date(gbl_vars$lastvax_date) - as.Date(gbl_vars$start_date_az)) + 1`-day period during which both vaccines were being administered widely; they were aged between 18 and 64 inclusive; not classed as Clinically Extremely Vulnerable, as set out by government guidance @covid19, at the time of vaccination; information on sex, ethnicity, deprivation, and geographical region was known.

Study participants were followed up for no more than 20 weeks from the day of the first dose, including time after their second dose. Follow-up was censored earlier than this at `r format(as.Date(gbl_vars$end_date), "%d %B %Y")`, death, or de-registration.

## Outcomes

Three outcomes were defined: positive SARS-CoV-2 test; COVID-19 A&E attendance; and unplanned COVID-19 hospital admission. Positive SARS-CoV-2 tests were identified using SGSS records and based on swab date. Both polymerase chain reaction (PCR) and lateral flow tests are included, without differentiation between symptomatic and asymptomatic infection. COVID-19 A&E attendances were identified using HES emergency care records with U07.1 ("COVID-19, virus identified") or U07.2 ("COVID-19, virus not identified") ICD-10 diagnosis codes @emergenc . Unplanned COVID-19 hospital admissions were identified using HES in-patient hospital records with U07.1 or U07.2 reason for admission ICD-10 codes.

Although severe disease (such as requirement for intensive or critical care) and mortality were of interest there were too few events to investigate these outcomes fully. Unadjusted incidence of COVID-19 deaths are reported descriptively. These were identified using linked death registration data. Deaths COVID-19 ICD-10 codes (as above) mentioned anywhere on the death certificate (i.e., as an underlying or contributing cause of death) were included.

## Additional variables

Participant characteristics used describe the cohort and for confounder adjustment include: age, sex (male or female), English Index of Multiple Deprivation (IMD, grouped by quintiles), ethnicity (Black, Mixed, South Asian, White, Other, as per the UK census), NHS region (East of England, Midlands, London, North East and Yorkshire, North West, South East, South West), number of conditions in the clinically "at risk" (but not clinically extremely vulnerable) classification, as per national prioritisation guidelines, the number of SARS-CoV-2 tests in the 90 days prior to the study start date (via SGSS), rurality (urban conurbation, urban city or town, rural town or village), evidence of prior SARS-CoV-2 infection (positive test or COVID-19 hospitalisation), learning disabilities, severe mental illness. All characteristics were ascertained as at the time of vaccination.

## Statistical Analysis

We compared the effectiveness of a first dose of ChAdOx1 versus BNT162b2 using pooled logistic regression @ngwa2016 @efron1988 (PLR) on the time since vaccination timescale with the outcome risk estimated each day. The effect is permitted to vary over the timescale to account for the potential time-varying differences in vaccine protection between the two brands. A PLR model can be used to approximate Cox models with time-varying treatment effects on the time-since-vaccination timescale, and enables the estimation of risk-adjusted cumulative incidence for each vaccine type. This is the average over all participants of the cumulative incidence for each day of follow-up predicted by the PLR model, under the (counterfactual) assumption that everyone received the BNT162b2 vaccine or that everyone received the ChAdOx1 vaccine. Assuming adequate confounder adjustment, this estimates the vaccine-specific cumulative incidence that would have been observed in an RCT comparing the two vaccines in the population under consideration. Standard errors for the PLR model were obtained using the clustered sandwich estimator to account for within-participant clustering. Confidence intervals for the vaccine-specific marginal cumulative incidence, and their difference, were obtained using the delta method (i.e, a first-order Taylor series approximation of the variance).

The PLR model included vaccine type, and a vaccine-specific 3-knot restricted cubic spline for time since vaccination. Knot locations were based on quartiles of the event times. Three models were fit for each outcome, with progressive adjustment for confounders: (1) adjusting for region-specific calendar-time effects, by including a 2-knot restricted cubic spline for the date of vaccination and its interaction with region; (2) additionally adjusting for demographic characteristics; (3) additionally adjusting for clinical characteristics.

During the study period, the advice from the UK Chief Medical Officers was that "second doses of both vaccines will be administered towards the end of the recommended vaccine dosing schedule of 12 weeks" @statemena. Using an intention-to-treat approach, comparative effectiveness estimates beyond 14 weeks were considered to be second-dose effects. We report the actual timing of second doses to assess the extent of any deviation from this recommended treatment strategy, frequency of any cross-brand second doses, and how these may differ between vaccine types.

<!-- Events occurring on the same day as vaccination are included in follow-up. -->

PLR models are computationally expensive to fit, as the input dataset must be arranged as one row per person per day of follow-up. To manage this, a sampling strategy was used such that all those who experienced the event of interest are selected (to retain statistical power), and a random sample of 50,000 event-free participants are selected. Person-time is weighted by the inverse of the sampling probability to recover the characteristics of the complete cohort. <!--Sampling was based on a ranked hash of the person identifier to ensure the sampled non-events overlapped as much as possible across outcomes.-->

## Missing data

A complete-case approach was used to deal with missing values. After exclusions for missing values on demographic variables (see exclusion criteria), there were no missing values in remaining variables as they were defined by the presence or absence of clinical codes or events.

## Software, code, and reproducibility

Data management and analyses were conducted in Python 3.8 and R version 4.0.2. All code is available at <https://github.com/opensafely/comparative-ve-research>, and is shared openly for review and re-use under MIT open license. Codelists are available at <https://www.opencodelists.org/>. No person-level data is shared. Any reported figures based on counts below 6 are redacted or rounded for disclosure control.

This study followed the STROBE-RECORD reporting guidelines.

## Patient and public involvement

We have developed a publicly available website <https://opensafely.org/> through which we invite any patient or member of the public to contact us regarding this study or the broader OpenSAFELY project.

# Results

## Study population

A total of `r label_number(1, big.mark=",")(flowchart[[1, "n"]])` HCWs aged 18-64 receiving a first dose of BNT162b2 or ChAdOx1 between 4 January and 28 February 2021 and actively registered at at TPP practice were identified, with `r label_number(1, big.mark=",")(flowchart[[nrow(flowchart), "n"]])` (`r label_percent(0.1)(flowchart[[nrow(flowchart), "pct_all"]])`) meeting the study eligibility criteria (supplementary Tables S1 and S2).

In total, `r label_number(1, big.mark=",")(baseline_pfizer$n)` (`r label_percent(0.1)(baseline_pfizer$n/baseline$n)`) were vaccinated with BNT162b2, contributing `r label_number(1, big.mark=",")(baseline_pfizer$fu_years)` person-years of potential follow-up (including all person-time prior to a censoring event, and possibly after an outcome event). For ChAdOx1, this was `r label_number(1, big.mark=",")(baseline_az$n)` (`r label_percent(0.1)(baseline_az$n/baseline$n)`) and `r label_number(1, big.mark=",")(baseline_az$fu_years)` person-years.

Characteristics were largely well-balanced between recipients of each vaccine, though regional and temporal differences in the distribution of each vaccine are notable (Table 1). BNT162b2 was on average administered earlier than ChAdOx1 (Figure 1; median day of vaccination `r format(baseline_pfizer$median_vaxday, "%d %B")` for BNT162b2, `r format(baseline_az$median_vaxday, "%d %B")` for ChadOx1). BNT162b2 was relatively more likely to be administered in the South and East of England, and ChAdOx1 the Midlands and Northern England. Evidence of prior SARS-CoV-2 infection was higher in ChAdOx1 recipients (`r label_percent(0.1)(baseline_pfizer$priorinfection)` for BNT162b2, `r label_percent(0.1)(baseline_az$priorinfection)` for ChAdOx1), consistent with ChAdOx1 recipients being vaccinated later on average. The proportion of each clinical condition is slightly higher in ChAdOx1 recipients, though consistently under a 0.6% percent-point difference.

\newpage

#### Figure 1: Vaccination dates

Cumulative enrollment of study participants by date.

```{r, vaxdate, echo=FALSE, message=FALSE, warning=FALSE, out.width='60%', out.height='50%', results='asis', out.extra=''}
#read_rds(fs::path(output_dir_os, "descriptive", "vaxdate", "plot_vaxdate_stack.rds"))
knitr::include_graphics(fs::path(output_dir_os, "descriptive", "vaxdate", "plot_vaxdate_stack.png"))
cat("  \n\n")
```

#### Table 1: Baseline characteristics

Participant characteristics as on the day of vaccination.

```{r table1, echo=FALSE, warning=FALSE, message=FALSE}
table1_body <- read_csv(fs::path(output_dir_os, "descriptive", "tables", "table1.csv"))
table1_by <- read_csv(fs::path(output_dir_os, "descriptive", "tables", "table1_by.csv"))

table1 <- 
  table1_body %>%
  group_by(variable) %>%
  mutate(
    label = if_else(row_number()==1, label, str_c("&nbsp;&nbsp;&nbsp;", label)),
  ) %>%
  mutate(
    across(
      starts_with("stat_"),
      ~if_else(is.na(.x) & row_number()==1, "", .x)
    )
  ) %>%
  ungroup() %>%
  mutate(
    order = cumsum(lag(variable, 1, first(variable))!=variable)
  ) %>%
  select(
    variable, var_label, label, starts_with("stat_"), order
  ) %>%
  rename(!!!set_names(table1_by$by_col, table1_by$by_chr))

table1 %>%
  gt() %>%
  fmt_markdown(columns = "label") %>%
  tab_style(
    style = list(
      cell_fill(color = "gray96")
    ),
    locations = cells_body(
      rows = order%%2 == 0
    )
  ) %>%
  tab_style(
    style = list(
      cell_text(weight = "bold")
    ),
    locations = cells_column_labels(everything())
  ) %>%
  cols_label(
     label=md("Characteristic")
  ) %>%
  cols_hide(c("variable", "var_label", "order"))
```

\newpage

## Events

Over the duration of `r label_number(1, big.mark=",")(baseline$fu_years)` person-years of follow-up there were `r label_number(1, big.mar=",")(baseline$postest)` positive SARS-CoV-2 tests, `r label_number(1, big.mar=",")(baseline$covidemergency)` COVID-19 A&E attendances, and `r label_number(1, big.mar=",")(baseline$covidadmitted)` COVID-19 hospital admissions. While not a primary outcome, there were also `r if_else(is.na(baseline$death), " fewer than six", label_number(1)(as.integer(baseline$death)))` deaths from any cause, of which `r if_else(is.na(baseline$coviddeath), " fewer than six", label_number(1)(as.integer(baseline$coviddeath)))` were COVID-19 related (supplementary Table S3).

At 12 weeks, `r label_percent(0.1)(baseline_pfizer$seconddose_12week)` of BNT162b2 recipients had received a second dose, compared with `r label_percent(0.1)(baseline_az$seconddose_12week)` of ChAdOx1 recipients. There were `r label_number(1)(crossbrand_n)` (`r label_percent(0.01)(crossbrand_prop)`) participants who received a second dose within 20 weeks that was not the same brand as the first dose, including Moderna mRNA-1273 vaccine second doses (supplementary Figure S3, Table S4).

## Comparative effectiveness

At 6 weeks post-vaccination, the ChAdOx1 versus BNT162b2 absolute risk difference per 1000 people (Figure 2) for a positive SARS-CoV-2 test was `r filter(riskdiff_ns3, outcome=="postest", tstop==7*6)$diff_format` (95%CI `r filter(riskdiff_ns3, outcome=="postest", tstop==7*6)$diff_CI_to`), for COVID-19 A&E attendances was `r filter(riskdiff_ns3, outcome=="covidemergency", tstop==7*6)$diff_format` (`r filter(riskdiff_ns3, outcome=="covidemergency", tstop==7*6)$diff_CI_to`) and for COVID-19 hospital admissions was `r filter(riskdiff_ns3, outcome=="covidadmitted", tstop==7*6)$diff_format` (`r filter(riskdiff_ns3, outcome=="covidadmitted", tstop==7*6)$diff_CI_to`).

At 18 weeks post-vaccination, the absolute risk difference per 1000 people for a positive SARS-CoV-2 test was `r filter(riskdiff_ns3, outcome=="postest", tstop==7*18)$diff_format` (95%CI `r filter(riskdiff_ns3, outcome=="postest", tstop==7*18)$diff_CI_to`), for COVID-19 A&E attendances was `r filter(riskdiff_ns3, outcome=="covidemergency", tstop==7*18)$diff_format` (`r filter(riskdiff_ns3, outcome=="covidemergency", tstop==7*18)$diff_CI_to`) and for COVID-19 hospital admissions was `r filter(riskdiff_ns3, outcome=="covidadmitted", tstop==7*18)$diff_format` (`r filter(riskdiff_ns3, outcome=="covidadmitted", tstop==7*18)$diff_CI_to`).

The estimated log hazard ratios were non-zero at certain periods (Figure 2), but the underlying absolute risk was small enough that this did not manifest in substantial differences in the cumulative incidence beyond the fourth week.

\newpage

#### Figure 2: Comparative effectiveness

For each outcome based on the fully adjusted model, the marginal cumulative incidence for ChAdOx1 and BNT162b2, their difference, and the hazard ratio are shown. Models that assumed piecewise-constant hazards gave similar effect estimates (supplementary Figure S2). The models with less extensive confounder adjustment gave very similar estimates (supplementary Figure S1) suggesting that recipients of each vaccine were similar after accounting for differences in vaccine allocation over space and time (as did all models).

```{r plot_effects_ns, echo=FALSE, message=FALSE, warning=FALSE, out.extra='', fig.height=7.5, out.width='90%', results='asis'}
plot_cmlinc_ns <- cmlinc_ns3 %>%
    ggplot()+
    geom_step(aes(x=tstop, y=(1-survival)*1000, group=vax1_az_descr, colour=vax1_az_descr))+
    geom_rect(aes(xmin=tstop, xmax=lead_tstop, ymin=(1-survival.ll)*1000, ymax=(1-survival.ul)*1000, group=vax1_az_descr, fill=vax1_az_descr), alpha=0.1)+
    geom_hline(aes(yintercept=0), colour='black')+
    facet_wrap(facets=vars(outcome_descr), nrow=1, scales="free_y")+
    scale_x_continuous(
      breaks = seq(0,7*52,by=28),
      expand = expansion(0)
    )+
    scale_y_continuous(
      expand = expansion(0),
      #limits = c(0L,NA)
    )+
    scale_colour_brewer(type="qual", palette="Set1", na.value="grey")+
    scale_fill_brewer(type="qual", palette="Set1", guide="none", na.value="grey")+
    labs(
      x=NULL,
      y="Marginalised\ncumul. incidence\nper 1000",
      colour=NULL
    )+
    theme_minimal(base_size = 9)+
    theme(
      legend.position=c(0.05,0.1),
      legend.justification = c(0,0),
      legend.direction = "vertical",
      axis.text.x.top=element_text(hjust=0),

      panel.border = element_blank(),
      axis.line.y = element_line(colour = "black"),
      axis.line.x.top = element_line(colour = "black"),

      panel.grid.minor.x = element_blank(),
      panel.grid.minor.y = element_blank(),
      strip.background = element_blank(),

      panel.spacing = unit(0.8, "lines"),

    )+
    NULL

  ## difference in cumulative incidence

  plot_riskdiff_ns <-
    riskdiff_ns3 %>%
    ggplot()+
    geom_hline(aes(yintercept=0), colour='black')+
    geom_step(aes(x=tstop, y=diff*1000), colour='orchid4')+
    geom_rect(aes(xmin=tstop, xmax=lead_tstop, ymin=diff.ll*1000, ymax=diff.ul*1000), alpha=0.1, colour="transparent", fill='orchid4')+
    facet_wrap(facets=vars(outcome_descr), nrow=1, scales="free_y")+
    scale_x_continuous(
      limits = c(0, NA),
      breaks = seq(0,7*52,by=28),
      expand = expansion(0)
    )+
    scale_y_continuous(
      expand = expansion(0),
      sec.axis = dup_axis(name="Higher incidence after \n <- ChAdOx1 / BNT162b2 ->", breaks = NULL)
    )+
    coord_cartesian(ylim= c(-3, 3))+
    scale_colour_brewer(type="qual", palette="Set1", na.value="grey")+
    scale_fill_brewer(type="qual", palette="Set1", guide="none", na.value="grey")+
    labs(
      x=NULL,
      y="Difference\nin cumul. incidence\nper 1000",
      colour=NULL
    )+
    theme_minimal(base_size = 9)+
    theme(
      axis.line.y.left = element_line(colour = "black"),

      panel.grid.minor.x = element_blank(),
      panel.grid.minor.y = element_blank(),
      strip.background = element_blank(),
      strip.text = element_blank(),

      panel.spacing = unit(0.8, "lines"),

      plot.title = element_text(hjust = 0),
      plot.title.position = "plot",
      plot.caption.position = "plot",
      plot.caption = element_text(hjust = 0, face= "italic")

     ) +
   NULL

  ## hazard ratio
  plot_hazard_ns <-
    hazard_ns3 %>%
    ggplot()+
    geom_hline(aes(yintercept=1), colour='black')+
    geom_step(aes(x=tstop, y=hr), colour='orchid4', direction="vh")+
    geom_rect(aes(xmin=lag_tstop, xmax=tstop, ymin=hr.ll, ymax=hr.ul), alpha=0.1, colour="transparent", fill='orchid4')+
    facet_wrap(facets=vars(outcome_descr), nrow=1, scales="free_y")+
    scale_x_continuous(
      limits = c(0, NA),
      breaks = seq(0,7*52,by=28),
      expand = expansion(0)
    )+
    scale_y_log10(
      breaks=c(0.1, 0.2, 0.5, 1, 2, 5, 10),
      sec.axis = dup_axis(name="Higher hazard after \n <- ChAdOx1 / BNT162b2 ->", breaks = NULL)
    )+
    coord_cartesian(ylim= c(0.1, 10))+
    labs(
      x = "Days since first dose",
      y = "Hazard ratio"
    )+
    theme_minimal(base_size = 9)+
    theme(
      axis.line.y.left = element_line(colour = "black"),

      panel.grid.minor.x = element_blank(),
      panel.grid.minor.y = element_blank(),
      strip.background = element_blank(),
      strip.text = element_blank(),

      panel.spacing = unit(0.8, "lines"),

      plot.title = element_text(hjust = 0),
      plot.title.position = "plot",
      plot.caption.position = "plot",
      plot.caption = element_text(hjust = 0, face= "italic")

     ) +
   NULL

  plot_mainresults <- patchwork::wrap_plots(plot_cmlinc_ns, plot_riskdiff_ns, plot_hazard_ns, ncol=1)
  ggsave(
    filename="mainresults.pdf",
    path=output_dir_rmd,
    plot=plot_mainresults,
  )

  plot_mainresults

```

\newpage

# Discussion

This observational study of `r label_number(1, big.mark=",")(flowchart[[nrow(flowchart), "n"]])` health and social care workers living in England found similar outcomes for those receiving the BNT162b2 or ChAdOx1 COVID-19 vaccines during the first 20 weeks post-vaccination. There is a clear leveling-off of the incidence of positive tests after around 3-4 weeks for both vaccines, consistent with the expected time-to-onset of vaccine-induced immunity of around 2 weeks plus the delay from infection to a positive test. At 20 weeks, the difference in the cumulative incidence of positive test was `r filter(riskdiff_ns3, outcome=="postest", tstop==7*20)$diff_format` per 1000 people `r filter(riskdiff_ns3, outcome=="postest", tstop==7*20)$diff_CI` higher for ChAdOx1 compared with BNT162b2, and extrapolating beyond this time would suggest a small advantage for BNT162b2. However, BNT162b2 recipients were more likely to receive their second dose sooner, conferring a greater protective effect that is not accounted for in this estimate.

There were very few COVID-19 hospital attendance and admission events after this period, and the difference in the cumulative incidence (and their CIs) was well below 1 event per 1000 people in either direction.

## Strengths and weaknesses

We used routinely-collected health records with comprehensive coverage of primary care, hospital admissions, COVID-19 testing, COVID-19 vaccination, and death registrations to study vaccinated health and social care workers. This group were eligible for vaccination at the start of the UK's vaccination programme due to exposure to higher viral loads and the need to reduce enforced absences in essential healthcare workers during a global pandemic. As such, HCW status is an important determinant of both vaccination time and SARS-CoV-2 infection, and a certain source of confounding in studies of vaccine effectiveness which may be reduced by studying this group in isolation. Although it is not typically possible to identify HCWs in NHS primary care records, this information was comprehensively collected at the time of COVID-19 vaccination and so is known for all vaccine recipients. HCWs are the only early vaccinees who are relatively young and healthy, and were vaccinated during a period where infection rates were high and both vaccines were being widely administered @collaborative2021. Overall, this provides a rare opportunity to study comparative effectiveness under conditions that, to some extent, approximate random vaccine allocation. However, some limitations remain.

Despite reasonable balance of vaccine allocation across baseline characteristics and adjustment for a range of potential confounders, the possibility of unmeasured confounding remains. The cold storage requirements of BNT162b2 meant that it was more likely to have be administered in Acute NHS Trusts and other large vaccination centres, and is thus a potential confounder for instance due to differences in viral exposure across these settings. Though we adjusted for region, rurality, and deprivation, we were unable to directly account for occupational differences that may affect both exposure risk and vaccine type. We note also that the GP practices covered in this study are not geographically representative, as TPP SystmOne software is not widely-used in some regions such as London and the North West.

We were unable to fully investigate differences in protection against severe disease, in large part due to clear protective benefits of both vaccines, reducing the absolute numbers of events, and therefore statistical power, in the studied cohort.

Second-dose effects were considered using an intention-to-treat approach, which does not account for potential differences in the timing of the second dose between vaccine types. However, we found that by 12 weeks more BNT162b2 recipients had received a second dose than ChAdOx1 recipients, which would bias effect estimates in favour of BNT162b2 assuming higher protection after a second-dose of either vaccine. Regardless, there was insufficient power to reliably estimate comparative second-dose effects as event incidence at 12 weeks and beyond was low for both vaccines, with fewer than 10 positive tests per 1000-person-years, and far fewer for COVID-19 A&E attendances and hospital admissions during this period. The need for many thousands more person-years to detect differences in second-dose effects is encouraging to the extent that it suggests extremely high protective effects for both vaccines, though the fact that background case rates in England were declining over the duration of the study period should not be ignored.

## Findings in context

A number of studies have estimated COVID-19 vaccine effectiveness in observational data @vasileiou2021 @cabezas2021 @effectiv @hall2021. Vaccine protection is typically assessed by comparing vaccinated against unvaccinated person-time, or against the early post-vaccination period before anticipated onset of vaccine-induced immunity. Such designs are vulnerable to confounding and selection bias, for example due to vaccine prioritisation and eligibility policies. In addition, differences in vaccine access and acceptance can cause substantial imbalance between vaccinated and unvaccinated groups that may be impossible to account for. Many of these biases can be bypassed when making direct comparisons between recipients of different vaccine types, if the relative performance of two vaccines is of interest. To our knowledge, the present study is the first to assess effectiveness of the BNT162b2 and ChAdOx1 vaccines in a head-to-head comparison in either an experimental or observational setting.

Our study suggests that there is little to differentiate the vaccines with respect to their respective protective effects against infection and hospital-related COVID-19 outcomes. This pushes decisions about which vaccine to favour onto other attributes such as safety profile, logistics, and cost. However, we cannot rule-out larger differences in effectiveness in other settings. For instance: results seen in this cohort of predominantly healthy HCWs may not reflect comparative effectiveness in more vulnerable groups, such as the elderly or immunosuppressed; the dominant circulating variant during the period of study was the Alpha variant but has now been replaced by Delta variant against which vaccines may be less effective @bernal2021; longer-term effectiveness, and in particular the potential immunological waning, may differ.

<!--# Inferring comparative effectiveness from these studies is possible by calculating the ratio of 1 minus the vaccine effectiveness for each vaccine: EAVE-II estimated VE in 18-64 age group in 5th week to be 92% for BNT162b2 and 100% for ChAdOx1, = HR=0.00; EAVE-II estimated VE in 65-79 age group in 4th week to be 91% for BNT162b2 and 68% for ChAdOx1, = HR=3.56; EAVE-II estimated VE in 80+ age group in 4th week to be 88% for BNT162b2 and 81% for ChAdOx1, = HR=1.58.-->

## Conclusion

This study found no substantial differences in the incidence of SARS-CoV-2 infection or COVID-19-related hospital events following vaccination with BNT162b2 or ChAdOx1 in a cohort of health and social care workers in England. The incidence of COVID-19-related A&E attendances or hospital admissions four weeks after the first dose of either vaccine was fewer than one event in 10,000-person-years, suggesting strong protective effects for both vaccines but nevertheless limiting power to reliably estimate comparative effectiveness with respect to severe COVID-19 outcomes. Further studies are needed to assess comparative effectiveness in newer, more prevalent variants.

\newpage

# Summary

### What is already known

-   Phase III randomised controlled trials show clear evidence of efficacy of the BNT162b2 mRNA and the ChAdOx1 vaccines
-   Observational studies have provided further evidence of effectiveness in mass vaccine rollout settings.
-   No head-to-head comparison of these vaccines has been made in randomised trials, and such comparisons are difficult in observational data without administration of both vaccines across the same population at the same time.

### What this study adds

-   By exploiting the concurrent rollout of both vaccines in health and social care workers in England, we were able to make direct comparisons of first and second dose effectiveness between the vaccines.
-   We found similar incidence rates for infection and COVID-19-related hospital attendance and admission.

\newpage

## Acknowledgements

We are very grateful for all the support received from the TPP Technical Operations team throughout this work, and for generous assistance from the information governance and database teams at NHS England / NHSX.

## Conflicts of Interest

All authors have completed the ICMJE uniform disclosure form and declare the following: BG has received research funding from Health Data Research UK (HDRUK), the Laura and John Arnold Foundation, the Wellcome Trust, the NIHR Oxford Biomedical Research Centre, the NHS National Institute for Health Research School of Primary Care Research, the Mohn-Westlake Foundation, the Good Thinking Foundation, the Health Foundation, and the World Health Organisation; he also receives personal income from speaking and writing for lay audiences on the misuse of science. IJD holds shares in GlaxoSmithKline (GSK).

## Funding

This work was supported by the Medical Research Council MR/V015737/1 and the Longitudinal Health and wellbeing strand of the National Core Studies programme. The OpenSAFELY platform is funded by the Wellcome Trust. TPP provided technical expertise and infrastructure within their data centre pro bono in the context of a national emergency. BGs work on clinical informatics is supported by the NIHR Oxford Biomedical Research Centre and the NIHR Applied Research Collaboration Oxford and Thames Valley. BG's work on better use of data in healthcare more broadly is currently funded in part by: NIHR Oxford Biomedical Research Centre, NIHR Applied Research Collaboration Oxford and Thames Valley, the Mohn-Westlake Foundation, NHS England, and the Health Foundation; all DataLab staff are supported by BG's grants on this work. LS reports grants from Wellcome, MRC, NIHR, UKRI, British Council, GSK, British Heart Foundation, and Diabetes UK outside this work. KB holds a Wellcome Senior Research Fellowship (220283/Z/20/Z). HIM is funded by the NIHR Health Protection Research Unit in Immunisation, a partnership between Public Health England and London School of Hygiene & Tropical Medicine. AYSW holds a fellowship from the British Heart Foundation. EJW holds grants from MRC. RM holds a Sir Henry Wellcome Fellowship funded by the Wellcome Trust (201375/Z/16/Z). HF holds a UKRI fellowship. IJD holds grants from NIHR and GSK.

Funders had no role in the study design, collection, analysis, and interpretation of data; in the writing of the report and in the decision to submit the article for publication.

The views expressed are those of the authors and not necessarily those of the NIHR, NHS England, Public Health England or the Department of Health and Social Care.

For the purpose of Open Access, the author has applied a CC BY public copyright licence to any Author Accepted Manuscript (AAM) version arising from this submission.

## Information governance and ethical approval

NHS England is the data controller; TPP is the data processor; and the key researchers on OpenSAFELY are acting with the approval of NHS England. This implementation of OpenSAFELY is hosted within the TPP environment which is accredited to the ISO 27001 information security standard and is NHS IG Toolkit compliant @betad @datasec ; Patient data has been pseudonymised for analysis and linkage using industry standard cryptographic hashing techniques; all pseudonymised datasets transmitted for linkage onto OpenSAFELY are encrypted; access to the platform is via a virtual private network (VPN) connection, restricted to a small group of researchers; the researchers hold contracts with NHS England and only access the platform to initiate database queries and statistical models; all database activity is logged; only aggregate statistical outputs leave the platform environment following best practice for anonymisation of results such as statistical disclosure control for low cell counts @isb1523 . The OpenSAFELY research platform adheres to the obligations of the UK General Data Protection Regulation (GDPR) and the Data Protection Act 2018. In March 2020, the Secretary of State for Health and Social Care used powers under the UK Health Service (Control of Patient Information) Regulations 2002 (COPI) to require organisations to process confidential patient information for the purposes of protecting public health, providing healthcare services to the public and monitoring and managing the COVID-19 outbreak and incidents of exposure; this sets aside the requirement for patient consent @coronavi2020 . Taken together, these provide the legal bases to link patient datasets on the OpenSAFELY platform. GP practices, from which the primary care data are obtained, are required to share relevant health information to support the public health response to the pandemic, and have been informed of the OpenSAFELY analytics platform.

This study was approved by the Health Research Authority (REC reference 20/LO/0651) and by the LSHTM Ethics Board (reference 21863).

## Guarantor

BG is guarantor.

\newpage

## References

### 2021-10-13 manuscript/draft-supplement.Rmd
add manuscript files

---
title: "Comparative vaccine effectiveness of ChAdOx1 versus BNT162b2 in Health and Social Care workers in England: Supplementary materials"
output:
  pdf_document:
    toc: no
  word_document: default
  bookdown::html_document2:
    number_sections: no
    toc: no
  html_document:
    keep_md: yes
    self_contained: yes
header-includes:
- \usepackage{float} 
- \floatplacement{figure}{H}
bibliography: references.bib
---

```{r setup, include=FALSE}

library('tidyverse')
library('here')
library('glue')
library('lubridate')
library('gt')
library('patchwork')
library('scales')
source(here("analysis", "lib", "utility_functions.R"))

#output_dir <- here("output")
output_dir <- here("released_outputs", "output")

fs::dir_create(output_dir, "report", "figures")

knitr::opts_chunk$set(
  echo = TRUE,
  fig.path = paste0(fs::path(output_dir, "report", "figures"), "/"),
  fig.pos = 'H' #to stop figures floating around when rendering to pdf
)

gbl_vars <- jsonlite::read_json(
  path=here("output", "data", "metadata_study-dates.json")
)

flowchart <- read_csv(fs::path(output_dir, "data", "flowchart.csv"))

metadata_outcomes <- read_rds(here("output", "data", "metadata_outcomes.rds"))

list_formula <- read_rds(here("output", "data", "metadata_formulas.rds"))
list2env(list_formula, globalenv())
lastfupday <- lastfupday20
```

```{r characteristics, echo=FALSE, warning=FALSE, message=FALSE}
baseline <- read_csv(fs::path(output_dir, "report", "objects", "baseline.csv"))
baseline_az <- read_csv(fs::path(output_dir, "report", "objects", "baseline_az.csv"))
baseline_pfizer <- read_csv(fs::path(output_dir, "report", "objects", "baseline_pfizer.csv"))
```

```{r, effects, echo=FALSE, message=FALSE, warning=FALSE}

  outcomes <- 
    c(
      "postest",
      "covidemergency",
      "covidadmitted"
    ) %>% 
    set_names(., .)

  cmlinc_ns <- read_csv(fs::path(output_dir, "report", "objects", "cmlinc_ns.csv")) %>%
  mutate(
    outcome_descr = fct_inorder(outcome_descr),
    model_name = fct_inorder(str_wrap(model_name, 15))
  )
  cmlinc_pw <- read_csv(fs::path(output_dir, "report", "objects", "cmlinc_pw.csv")) %>%
  mutate(
    outcome_descr = fct_inorder(outcome_descr),
    model_name = fct_inorder(str_wrap(model_name, 15))
  )

  riskdiff_ns <- read_csv(fs::path(output_dir, "report", "objects", "riskdiff_ns.csv")) %>%
  mutate(
    outcome_descr = fct_inorder(outcome_descr),
    model_name = fct_inorder(str_wrap(model_name, 15))
  )
  riskdiff_pw <- read_csv(fs::path(output_dir, "report", "objects", "riskdiff_pw.csv")) %>%
  mutate(
    outcome_descr = fct_inorder(outcome_descr),
    model_name = fct_inorder(str_wrap(model_name, 15))
  )

  hazard_ns <- read_csv(fs::path(output_dir, "report", "objects", "hazard_ns.csv")) %>%
  mutate(
    outcome_descr = fct_inorder(outcome_descr),
    model_name = fct_inorder(str_wrap(model_name, 15))
  )
  hazard_pw <- read_csv(fs::path(output_dir, "report", "objects", "hazard_pw.csv")) %>%
  mutate(
    outcome_descr = fct_inorder(outcome_descr),
    model_name = fct_inorder(str_wrap(model_name, 15))
  )

  plot_cmlinc <- function(model){
    modell <- model

    cmlinc_ns %>%
      filter(model == modell) %>%
      ggplot()+
      geom_step(aes(x=tstop, y=(1-survival)*1000, group=vax1_az_descr, colour=vax1_az_descr))+
      geom_rect(aes(xmin=tstop, xmax=lead_tstop, ymin=(1-survival.ll)*1000, ymax=(1-survival.ul)*1000, group=vax1_az_descr, fill=vax1_az_descr), alpha=0.1)+
      geom_hline(aes(yintercept=0), colour='black')+
      facet_wrap(facets=vars(outcome_descr), nrow=1, scales="free_y")+
      scale_x_continuous(
        breaks = seq(0,7*52,by=28),
        expand = expansion(0)
      )+
      scale_y_continuous(
        expand = expansion(0),
        #limits = c(0L,NA)
      )+
      scale_colour_brewer(type="qual", palette="Set1", na.value="grey")+
      scale_fill_brewer(type="qual", palette="Set1", guide="none", na.value="grey")+
      labs(
        x=NULL,
        y="Marginalised\ncumul. incidence\nper 1000",
        colour=NULL
      )+
      theme_minimal(base_size = 9)+
      theme(
        legend.position=c(0.95,0.95),
        legend.justification = c(1,1),
        legend.direction = "vertical",
        axis.text.x.top=element_text(hjust=0),

        panel.border = element_blank(),
        axis.line.y = element_line(colour = "black"),
        axis.line.x.top = element_line(colour = "black"),

        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        strip.background = element_blank(),

        panel.spacing = unit(0.8, "lines"),

      )+
      NULL
  }

   plot_riskdiff <- function(model){

    modell <- model
    riskdiff_ns %>%
      filter(model==modell) %>%
      ggplot()+
      geom_hline(aes(yintercept=0), colour='black')+
      geom_step(aes(x=tstop, y=diff*1000), colour='orchid4')+
      geom_rect(aes(xmin=tstop, xmax=lead_tstop, ymin=diff.ll*1000, ymax=diff.ul*1000), alpha=0.1, colour="transparent", fill='orchid4')+
      facet_wrap(facets=vars(outcome_descr), nrow=1, scales="free_y")+
      scale_x_continuous(
        limits = c(0, NA),
        breaks = seq(0,7*52,by=28),
        expand = expansion(0)
      )+
      scale_y_continuous(
        expand = expansion(0),
        sec.axis = dup_axis(name="Lower risk after \n <- BNT162b2 / ChAdOx1 ->", breaks = NULL)
      )+
      coord_cartesian(ylim= c(-3, 3))+
      scale_colour_brewer(type="qual", palette="Set1", na.value="grey")+
      scale_fill_brewer(type="qual", palette="Set1", guide="none", na.value="grey")+
      labs(
        x=NULL,
        y="Difference\nin cumul. incidence\nper 1000",
        colour=NULL
      )+
      theme_minimal(base_size = 9)+
      theme(
        axis.line.y.left = element_line(colour = "black"),

        #panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        strip.background = element_blank(),
        strip.text = element_blank(),

        panel.spacing = unit(0.8, "lines"),

        plot.title = element_text(hjust = 0),
        plot.title.position = "plot",
        plot.caption.position = "plot",
        plot.caption = element_text(hjust = 0, face= "italic")

       ) +
     NULL
  }

  plot_hazard <- function(model){
    modell <- model

    hazard_ns %>% 
      filter(model == modell) %>%
      ggplot()+
      geom_hline(aes(yintercept=1), colour='black')+
      geom_step(aes(x=tstop, y=hr), colour='orchid4', direction="vh")+
      geom_rect(aes(xmin=lag_tstop, xmax=tstop, ymin=hr.ll, ymax=hr.ul), alpha=0.1, colour="transparent", fill='orchid4')+
      facet_wrap(facets=vars(outcome_descr), nrow=1, scales="free_y")+
      scale_x_continuous(
        limits = c(0, NA),
        breaks = seq(0,7*52,by=28),
        expand = expansion(0)
      )+
      scale_y_log10(
        breaks=c(0.1, 0.2, 0.5, 1, 2, 5, 10),
        sec.axis = dup_axis(name="Lower hazard after \n <- BNT162b2 / ChAdOx1 ->", breaks = NULL)
      )+
      coord_cartesian(ylim= c(0.1, 10))+
      labs(
        x = "Days since first dose",
        y = "Hazard ratio"
      )+
      theme_minimal(base_size = 9)+
      theme(
        axis.line.y.left = element_line(colour = "black"),

        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        strip.background = element_blank(),
        strip.text = element_blank(),

        panel.spacing = unit(0.8, "lines"),

        plot.title = element_text(hjust = 0),
        plot.title.position = "plot",
        plot.caption.position = "plot",
        plot.caption = element_text(hjust = 0, face= "italic")

       ) +
     NULL
  }

```

\newpage

### Figure S1a: Comparative effectiveness adjusting for region-specific calendar-time

```{r comp1, echo=FALSE, message=FALSE, warning=FALSE, out.width='90%', fig.height=7, out.extra=''}
patchwork::wrap_plots(plot_cmlinc(1), plot_riskdiff(1), plot_hazard(1), ncol=1)
```

\newpage

### Figure S1b: Comparative effectiveness adjusting for region-specific calendar-time and demographic characteristics

```{r comp2, echo=FALSE, message=FALSE, warning=FALSE, out.width='90%', fig.height=7, out.extra=''}
patchwork::wrap_plots(plot_cmlinc(2), plot_riskdiff(2), plot_hazard(2), ncol=1)
```

\newpage

### Figure S2: Fully adjusted comparative effectiveness using piecewise-constant estimates of the hazard

```{r plot_effects_pw, echo=FALSE, message=FALSE, warning=FALSE, out.width='90%', fig.height=7, results='asis', out.extra=''}

  plot_cmlinc_pw <- 
    cmlinc_pw %>%
    filter(model==3) %>%
    ggplot()+
    geom_step(aes(x=tstop, y=(1-survival)*1000, group=vax1_az_descr, colour=vax1_az_descr))+
    geom_rect(aes(xmin=tstop, xmax=lead_tstop, ymin=(1-survival.ll)*1000, ymax=(1-survival.ul)*1000, group=vax1_az_descr, fill=vax1_az_descr), alpha=0.1)+
    geom_hline(aes(yintercept=0), colour='black')+
    facet_wrap(facets=vars(outcome_descr), nrow=1, scales="free_y")+
    scale_x_continuous(
      breaks = seq(0,7*52,by=28),
      expand = expansion(0)
    )+
    scale_y_continuous(
      expand = expansion(0),
      #limits = c(0L,NA)
    )+
    scale_colour_brewer(type="qual", palette="Set1", na.value="grey")+
    scale_fill_brewer(type="qual", palette="Set1", guide="none", na.value="grey")+
    labs(
      x=NULL,
      y="Marginalised\ncumul. incidence\nper 1000",
      colour=NULL
    )+
    theme_minimal(base_size = 9)+
    theme(
      legend.position=c(0.05,0.1),
      legend.justification = c(0,0),
      legend.direction = "vertical",
      axis.text.x.top=element_text(hjust=0),

      panel.border = element_blank(),
      axis.line.y = element_line(colour = "black"),
      axis.line.x.top = element_line(colour = "black"),

      panel.grid.minor.x = element_blank(),
      panel.grid.minor.y = element_blank(),
      strip.background = element_blank(),

      panel.spacing = unit(0.8, "lines"),

    )+
    NULL

  ## difference in cumulative incidence

  plot_riskdiff_pw <-
    riskdiff_pw %>%
    filter(model==3) %>%
    ggplot()+
    geom_hline(aes(yintercept=0), colour='black')+
    geom_step(aes(x=tstop, y=diff*1000), colour='orchid4')+
    geom_rect(aes(xmin=tstop, xmax=lead_tstop, ymin=diff.ll*1000, ymax=diff.ul*1000), alpha=0.1, colour="transparent", fill='orchid4')+
    facet_wrap(facets=vars(outcome_descr), nrow=1, scales="free_y")+
    scale_x_continuous(
      limits = c(0, NA),
      breaks = seq(0,7*52,by=28),
      expand = expansion(0)
    )+
    scale_y_continuous(
      expand = expansion(0),
      sec.axis = dup_axis(name="Higher incidence after \n <- ChAdOx1 / BNT162b2 ->", breaks = NULL)
    )+
    coord_cartesian(ylim= c(-3, 3))+
    scale_colour_brewer(type="qual", palette="Set1", na.value="grey")+
    scale_fill_brewer(type="qual", palette="Set1", guide="none", na.value="grey")+
    labs(
      x=NULL,
      y="Difference\nin cumul. incidence\nper 1000",
      colour=NULL
    )+
    theme_minimal(base_size = 9)+
    theme(
      axis.line.y.left = element_line(colour = "black"),

      panel.grid.minor.x = element_blank(),
      panel.grid.minor.y = element_blank(),
      strip.background = element_blank(),
      strip.text = element_blank(),

      panel.spacing = unit(0.8, "lines"),

      plot.title = element_text(hjust = 0),
      plot.title.position = "plot",
      plot.caption.position = "plot",
      plot.caption = element_text(hjust = 0, face= "italic")

     ) +
   NULL

  ## hazard ratio
  plot_hazard_pw <-
    hazard_pw %>%
    filter(model==3) %>%
    ggplot()+
    geom_hline(aes(yintercept=1), colour='black')+
    geom_step(aes(x=tstop, y=hr), colour='orchid4', direction="vh")+
    geom_rect(aes(xmin=lag_tstop, xmax=tstop, ymin=hr.ll, ymax=hr.ul), alpha=0.1, colour="transparent", fill='orchid4')+
    facet_wrap(facets=vars(outcome_descr), nrow=1, scales="free_y")+
    scale_x_continuous(
      limits = c(0, NA),
      breaks = seq(0,7*52,by=28),
      expand = expansion(0)
    )+
    scale_y_log10(
      breaks=c(0.1, 0.2, 0.5, 1, 2, 5, 10),
      sec.axis = dup_axis(name="Higher hazard after \n <- ChAdOx1 / BNT162b2 ->", breaks = NULL)
    )+
    coord_cartesian(ylim= c(0.1, 10))+
    labs(
      x = "Days since first dose",
      y = "Hazard ratio"
    )+
    theme_minimal(base_size = 9)+
    theme(
      axis.line.y.left = element_line(colour = "black"),

      panel.grid.minor.x = element_blank(),
      panel.grid.minor.y = element_blank(),
      strip.background = element_blank(),
      strip.text = element_blank(),

      panel.spacing = unit(0.8, "lines"),

      plot.title = element_text(hjust = 0),
      plot.title.position = "plot",
      plot.caption.position = "plot",
      plot.caption = element_text(hjust = 0, face= "italic")

     ) +
   NULL

patchwork::wrap_plots(plot_cmlinc_pw, plot_riskdiff_pw, plot_hazard_pw, ncol=1)
```

\newpage

### Figure S3: Second dose interval

Second dose events are censored by study end date, de-registration, death, and additionally by outcome-specific events.

```{r seconddose, echo=FALSE, warning=FALSE, message=FALSE}
knitr::include_graphics(fs::path(output_dir, "descriptive", "seconddose", "plot_seconddose_all.png"))
```

\newpage

### Figure S4: Event rates by calendar time

```{r eventrates, echo=FALSE, warning=FALSE, message=FALSE}
knitr::include_graphics(fs::path(output_dir, "descriptive", "eventdate", "plot_eventdate_step.png"))
```

\newpage

### Table S1: Inclusion criteria

```{r flowchart, echo=FALSE, warning=FALSE, message=FALSE}
flowchart %>%
  transmute(
    Participants = str_wrap(criteria, 40),
    N = n,
    `N excluded` = n_exclude,
    #`% excluded` = pct_exclude,
    `% remaining` = pct_all
  ) %>%
  mutate(
    Participants=if_else(row_number()==1,"HCWs aged 18-64 receiving BNT162b2 or ChAdOx1 between 4 Jan and 28 Feb",  Participants)
  ) %>%
  gt() %>%
  fmt_percent(
    columns = starts_with(c("%")),
    decimals = 1
  ) %>%
  fmt_number(
    columns = starts_with(c("N")),
    decimals= 0
  ) %>%
  fmt_missing(
    everything(),
    missing_text="-"
  ) %>%
  tab_style(
    style=cell_text(whitespace ="pre-wrap"),
    #style=cell_text(style ="italic"),
    locations = cells_body(columns = "Participants")
  )
```

\newpage

### Table S2: Baseline characteristics before eligibility exclusions

Characteristics for all HCWs aged 18-64 receiving a first dose of BNT162b2 or ChAdOx1 between 4 January and 28 February 2021 and actively registered at at TPP practice are reported, before exclusions due to missing demographic characteristics and Clinically Extremely Vulnerable status.

```{r table1_allvax, echo=FALSE, warning=FALSE, message=FALSE, out.extra=''}
table1_body <- read_csv(fs::path(output_dir, "descriptive", "tables", "table1_allvax.csv"))
table1_by <- read_csv(fs::path(output_dir, "descriptive", "tables", "table1_allvax_by.csv"))

table1 <- 
  table1_body %>%
  group_by(variable) %>%
  mutate(
    label = if_else(row_number()==1, label, str_c("&nbsp;&nbsp;&nbsp;", label)),
  ) %>%
  mutate(
    across(
      starts_with("stat_"),
      ~if_else(is.na(.x) & row_number()==1, "", .x)
    )
  ) %>%
  ungroup() %>%
  mutate(
    order = cumsum(lag(variable, 1, first(variable))!=variable)
  ) %>%
  select(
    variable, var_label, label, starts_with("stat_"), order
  ) %>%
  rename(!!!set_names(table1_by$by_col, table1_by$by_chr))

table1 %>%
  gt() %>%
  fmt_markdown(columns = "label") %>%
  tab_style(
    style = list(
      cell_fill(color = "gray96")
    ),
    locations = cells_body(
      rows = order%%2 == 0
    )
  ) %>%
  tab_style(
    style = list(
      cell_text(weight = "bold")
    ),
    locations = cells_column_labels(everything())
  ) %>%
  cols_label(
     label=md("Characteristic")
  ) %>%
  cols_hide(c("variable", "var_label", "order"))
```

\newpage

### Table S3: Event rates

```{r table_irr, echo=FALSE, message=FALSE, warning=FALSE}
table_irr <- read_csv(here("output", "descriptive", "tables", "table_irr.csv"))

table_irr %>%
  filter(
    event %in% c(
      "postest",
      "covidemergency",
      "covidadmitted",
      "coviddeath",
      "seconddose"
    )
  ) %>%
  transmute(
    outcome_descr, timesincevax,
    pfizer_q,
    pfizer_rate_fmt = if_else(pfizer_rate<0.001, "<0.001", number_format(0.001)(pfizer_rate)),
    pfizer_rate_fmt = if_else(pfizer_rate==0, "0", pfizer_rate_fmt),
    az_q,
    az_rate_fmt = if_else(az_rate<0.001, "<0.001", number_format(0.001)(az_rate)),
    az_rate_fmt = if_else(az_rate==0, "0", az_rate_fmt),
  ) %>%
  gt(
    groupname_col = "outcome_descr",
  ) %>%
  cols_label(
    outcome_descr = "Outcome",
    timesincevax = "Days since first dose",

    pfizer_q = "Events / person-years",
    pfizer_rate_fmt = "Incidence",

    az_q   = "Events / person-years",
    az_rate_fmt = "Incidence",
  ) %>%
  fmt_missing(
    everything(),
    missing_text="--"
  ) %>%
  cols_align(
    align = "right",
    columns = everything()
  ) %>%
  cols_align(
    align = "left",
    columns = "timesincevax"
  ) %>% 
  tab_spanner(
    label = "BNT162b2",
    columns = starts_with("pfizer")
  ) %>%
  tab_spanner(
    label = "ChAdOx1",
    columns = starts_with("az")
  )
```

### Table S4: Vaccine type for second doses

```{r seconddose_table, echo=FALSE, message=FALSE, warning=FALSE}
seconddose <- read_csv(fs::path(output_dir, "descriptive", "seconddose", "seconddose.csv"))
seconddose %>%
  select(-REDACTED, -pct_nonmiss) %>%
  gt(
    #groupname_col="First dose"
  ) %>%
  fmt_percent(
    columns = starts_with("pct"),
    decimals = 1
  ) %>%
  fmt_number(
    columns = starts_with(c("n")),
    decimals= 0
  ) %>%
  fmt_missing(
    everything(),
    missing_text="(REDACTED)"
  ) %>%
  cols_label(
    n = "N",
    pct = "%"
  )
```

### 2021-10-13 manuscript/draft-manuscript.Rmd
update manuscript, readme

@@
 output:
-  pdf_document:
-    toc: no
+  word_document: default
   html_document:
@@
     toc: no
-  word_document: default
+  pdf_document:
+    toc: no
 bibliography: references.bib
@@
 The incidence of COVID-19-related A&E attendances or hospital admissions four weeks after the first dose of either vaccine was fewer than one event in 10,000-person-years, suggesting strong protective effects for both vaccines but nevertheless limiting power to reliably estimate comparative effectiveness with respect to severe COVID-19 outcomes.
-Further studies are needed to assess comparative effectiveness in newer, more prevalent variants.
+Further studies are needed to assess comparative effectiveness in newer, more prevalent variants and to assess longer-term effectiveness.
 

### 2022-03-30 analysis/models/model_plr.R
add sensitivity excluding prior covid infection

@@ -27,14 +27,21 @@ if(length(args)==0){
   timescale <- "timesincevax"
   censor_seconddose <- as.integer("0")
   samplesize_nonoutcomes_n <- 5000
+  exclude_prior_infection <- as.integer("1")
 } else {
   removeobs <- TRUE
   outcome <- args[[1]]
   timescale <- args[[2]]
   censor_seconddose <- as.integer(args[[3]])
   samplesize_nonoutcomes_n <- as.integer(args[[4]])
+  if (length(args)<5) {exclude_prior_infection <- 0L} else {exclude_prior_infection <- as.integer(args[[5]])}
 }
 
+if(exclude_prior_infection==1L){
+  sens_path <- paste0(censor_seconddose,"/","1")
+} else {
+  sens_path <- censor_seconddose
+}
 
 
 ## Import libraries ----
@@ -52,15 +59,18 @@ source(here("analysis", "lib", "redaction_functions.R"))
 source(here("analysis", "lib", "survival_functions.R"))
 
 
+
+
+
 # create output directories ----
-fs::dir_create(here("output", "models", outcome, timescale, censor_seconddose))
+fs::dir_create(here("output", "models", outcome, timescale, sens_path))
 
 ## create special log file ----
-cat(glue("## script info for {outcome} ##"), "  \n", file = here("output", "models", outcome, timescale, censor_seconddose, glue("modelplr_log_{outcome}.txt")), append = FALSE)
+cat(glue("## script info for {outcome} ##"), "  \n", file = here("output", "models", outcome, timescale, sens_path, glue("modelplr_log_{outcome}.txt")), append = FALSE)
 ## function to pass additional log text
 logoutput <- function(...){
-  cat(..., file = here("output", "models", outcome, timescale, censor_seconddose, glue("modelplr_log_{outcome}.txt")), sep = "\n  ", append = TRUE)
-  cat("\n", file = here("output", "models", outcome, timescale, censor_seconddose, glue("modelplr_log_{outcome}.txt")), sep = "\n  ", append = TRUE)
+  cat(..., file = here("output", "models", outcome, timescale, sens_path, glue("modelplr_log_{outcome}.txt")), sep = "\n  ", append = TRUE)
+  cat("\n", file = here("output", "models", outcome, timescale, sens_path, glue("modelplr_log_{outcome}.txt")), sep = "\n  ", append = TRUE)
 }
 
 ## import metadata ----
[…]

### 2022-03-30 analysis/report/report-sensitivity-objects.R

@@ -1,0 +1,115 @@
+# load libraries ----
+
+library('tidyverse')
+library('here')
+library('glue')
+library('lubridate')
+library('gt')
+library('patchwork')
+library('scales')
+source(here("analysis", "lib", "utility_functions.R"))
+
+
+data_cohort <- read_rds(here("output", "data", "data_cohort.rds"))
+
+metadata_outcomes <- read_rds(here("output", "data", "metadata_outcomes.rds"))
+
+list_formula <- read_rds(here("output", "data", "metadata_formulas.rds"))
+list2env(list_formula, globalenv())
+lastfupday <- lastfupday20
+
+## create directory ----
+
+fs::dir_create(here("output", "report", "objects"))
+
+## baseline information ----
+### model plots information -----
+
+outcomes <-
+  c(
+    "postest",
+    "covidemergency",
+    "covidadmitted"
+  ) %>%
+  set_names(., .)
+
+# cumulative incidence
+cmlinc_ns <-
+  map_dfr(
+    outcomes,
+    ~read_rds(here("output", "models", .x, "timesincevax", "0", "1", glue("reportplr_adjustedsurvival_ns.rds"))),
+    .id="outcome"
+  ) %>%
+  left_join(metadata_outcomes, by="outcome") %>%
+  mutate(
+    inc_format = label_number(0.1, scale=1000)(1-survival),
+    inc_CI = paste0("(" , label_number(0.1, scale=1000)(1-survival.ul), ", ", label_number(0.1, scale=1000)(1-survival.ll), ")")
+  )
+
+write_csv(cmlinc_ns, here("output", "report", "objects", "sensitivity", "exclude_prior_infection", "cmlinc_ns.csv"))
+
+cmlinc_pw <-
+  map_dfr(
+    outcomes,
+    ~read_rds(here("output", "models", .x, "timesincevax", "0", "1", glue("reportplr_adjustedsurvival_pw.rds"))),
+    .id="outcome"
+  ) %>%
+  left_join(metadata_outcomes, by="outcome") %>%
+  mutate(
+    inc_format = label_number(0.1, scale=1000)(1-survival),
+    inc_CI = paste0("(" , label_number(0.1, scale=1000)(1-survival.ul), ", ", label_number(0.1, scale=1000)(1-survival.ll), ")")
+  )
+
[…]

### 2022-03-30 create-project.R

@@ -66,19 +66,30 @@ action <- function(
 
 ## model action function ----
 action_model<- function(
-  outcome, timescale, censor_seconddose, modeltype, samplesize_nonoutcomes_n=NULL
+  outcome, timescale, censor_seconddose, modeltype, samplesize_nonoutcomes_n=NULL, exclude_prior_infection=NULL
 ){
+
+  if(is.null(exclude_prior_infection)){
+    exclude_prior_infection_path<-""
+    exclude_prior_infection_needs<-""
+  } else{
+    if(exclude_prior_infection=="1"){
+      exclude_prior_infection_path<-"/1"
+      exclude_prior_infection_needs<-"_1"
+    }
+  }
+
   action(
-    name = glue("model_{outcome}_{timescale}_{censor_seconddose}_{modeltype}"),
+    name = glue("model_{outcome}_{timescale}_{censor_seconddose}{exclude_prior_infection_needs}_{modeltype}"),
     run = glue("r:latest analysis/models/model_{modeltype}.R"),
-    arguments = c(outcome, timescale, censor_seconddose, samplesize_nonoutcomes_n),
+    arguments = c(outcome, timescale, censor_seconddose, samplesize_nonoutcomes_n, exclude_prior_infection),
     needs = list("design", "data_selection"),
     highly_sensitive = lst(
-      rds = glue("output/models/{outcome}/{timescale}/{censor_seconddose}/model{modeltype}*.rds")
+      rds = glue("output/models/{outcome}/{timescale}/{censor_seconddose}{exclude_prior_infection_path}/model{modeltype}*.rds")
     ),
     moderately_sensitive = lst(
-      txt = glue("output/models/{outcome}/{timescale}/{censor_seconddose}/model{modeltype}*.txt"),
-      csv = glue("output/models/{outcome}/{timescale}/{censor_seconddose}/model{modeltype}*.csv")
+      txt = glue("output/models/{outcome}/{timescale}/{censor_seconddose}{exclude_prior_infection_path}/model{modeltype}*.txt"),
+      csv = glue("output/models/{outcome}/{timescale}/{censor_seconddose}{exclude_prior_infection_path}/model{modeltype}*.csv")
     )
   )
 }
@@ -86,20 +97,30 @@ action_model<- function(
 
 ## report action function ----
 action_report <- function(
-  outcome, timescale, censor_seconddose, modeltype
[…]

### 2022-03-30 manuscript/draft-manuscript-v1-1.Rmd
some revisions

--- 2021-10-13 manuscript/draft-manuscript.Rmd
@@
 
-*Setting* Linked primary care, hospital, and COVID-19 surveillance records available within the OpenSAFELY-TPP research platform.
+*Setting* Linked primary care, hospital, and COVID-19 surveillance records available within the OpenSAFELY-TPP research platform, covering a period where the SARS-CoV-2 Alpha variant was dominant.
 
@@
 
-*Main outcome measures* Recorded SARS-CoV-2 positive test, or COVID-19 related Accident and Emergency attendance or hospital admission occurring within 20 weeks of vaccination.
+*Main outcome measures* Recorded SARS-CoV-2 positive test, or COVID-19 related Accident and Emergency attendance or hospital admission occurring within 20 weeks of receipt of the first vaccine dose.
 
 *Results* The cumulative incidence of each outcome was similar for both vaccines during the first 20 weeks post-vaccination.
-The cumulative incidence of recorded SARS-CoV-2 infection 6 weeks after vaccination with BNT162b2 was `r filter(cmlinc_ns3, outcome=="postest", vax1_az==0, tstop==7*6)$inc_format` per 1000 people (95%CI `r filter(cmlinc_ns3, outcome=="postest", vax1_az==0, tstop==7*6)$inc_CI_to`) and with ChAdOx1 was `r filter(cmlinc_ns3, outcome=="postest", vax1_az==1, tstop==7*6)$inc_format` (95%CI `r filter(cmlinc_ns3, outcome=="postest", vax1_az==1, tstop==7*6)$inc_CI_to`), representing a difference of `r filter(riskdiff_ns3, outcome=="postest", tstop==7*6)$diff_format` per 1000 people (95%CI `r filter(riskdiff_ns3, outcome=="postest", tstop==7*6)$diff_CI_to`).
-The difference in the cumulative incidence per 1000 people of COVID-19 accident and emergency attendance at 6 weeks was `r filter(riskdiff_ns3, outcome=="covidemergency", tstop==7*6)$diff_format` per 1000 people (95%CI `r filter(riskdiff_ns3, outcome=="covidemergency", tstop==7*6)$diff_CI_to`).
-For COVID-19 hospital admission, this difference was `r filter(riskdiff_ns3, outcome=="covidadmitted", tstop==7*6)$diff_format` per 1000 people (95%CI `r filter(riskdiff_ns3, outcome=="covidadmitted", tstop==7*6)$diff_CI_to`).
+The cumulative incidence of recorded SARS-CoV-2 infection 20 weeks after first-dose vaccination with BNT162b2 was `r filter(cmlinc_ns3, outcome=="postest", vax1_az==0, tstop==7*20)$inc_format` per 1000 people (95%CI `r filter(cmlinc_ns3, outcome=="postest", vax1_az==0, tstop==7*20)$inc_CI_to`) and with ChAdOx1 was `r filter(cmlinc_ns3, outcome=="postest", vax1_az==1, tstop==7*20)$inc_format` (95%CI `r filter(cmlinc_ns3, outcome=="postest", vax1_az==1, tstop==7*20)$inc_CI_to`), representing a difference of `r filter(riskdiff_ns3, outcome=="postest", tstop==7*20)$diff_format` per 1000 people (95%CI `r filter(riskdiff_ns3, outcome=="postest", tstop==7*20)$diff_CI_to`).
+The difference in the cumulative incidence per 1000 people of COVID-19 accident and emergency attendance at 20 weeks was `r filter(riskdiff_ns3, outcome=="covidemergency", tstop==7*20)$diff_format` per 1000 people (95%CI `r filter(riskdiff_ns3, outcome=="covidemergency", tstop==7*20)$diff_CI_to`).
+For COVID-19 hospital admission, this difference was `r filter(riskdiff_ns3, outcome=="covidadmitted", tstop==7*20)$diff_format` per 1000 people (95%CI `r filter(riskdiff_ns3, outcome=="covidadmitted", tstop==7*20)$diff_CI_to`).
 
 *Conclusions* In this cohort of healthcare workers where we would not anticipate vaccine type to be related to health status, we found no substantial differences in the incidence of SARS-CoV-2 infection or COVID-19 disease up to 20 weeks after vaccination.
-Incidence dropped sharply after 3-4 weeks and there were very few COVID-19 hospital attendance and admission events after this period.
-This is in line with expected onset of vaccine-induced immunity, and suggests strong protection against COVID-19 disease for both vaccines.
+Incidence dropped sharply at 3-4 weeks post-vaccination and there were very few COVID-19 hospital attendance and admission events after this period.
+This is in line with expected onset of vaccine-induced immunity, and suggests strong protection against Alpha-variant COVID-19 disease for both vaccines in this relatively young and healthy population of HCWs.
 
@@
 The COVID-19 global pandemic has prompted the rapid development and delivery of vaccines to combat the disease.
-Following demonstration of high safety and efficacy against symptomatic and severe disease in phase-III randomised controlled trials (RCTs), two vaccines have been approved and widely administered as part of the national vaccination programme in the United Kingdom: the Pfizer-BioNTech BNT162b2 mRNA COVID-19 vaccine @polack2020 (*BNT162b2*) and the Oxford-AstraZeneca ChAdOx1 nCoV-19 vaccine @voysey2021 (*ChAdOx1*).
+Following demonstration of high safety and efficacy against symptomatic and severe disease in phase-III randomised controlled trials (RCTs), two vaccines have been approved and widely administered as part of the national vaccination programme in the United Kingdom: the Pfizer-BioNTech BNT162b2 mRNA COVID-19 vaccine @polack2020 (*BNT162b2*) and the Oxford-AstraZeneca ChAdOx1 nCoV-19 viral vector vaccine @voysey2021 (*ChAdOx1*).
 Post-authorisation assessment of vaccine effectiveness using observational data is necessary to monitor the success of such programmes as, invariably, target populations and settings differ substantially from those of trials.
@@
 COVID-19 vaccination in the UK has been prioritised based on the risk of infection and subsequent severity of disease @covid19.
-Health and social care workers (HCWs) were amongst the first groups eligible for vaccination due to the high occupational exposure to the SARS-CoV-2 virus, making HCW status an important confounder of the effect of vaccination on infection and post-infection disease.
-By studying this group in isolation, this confounding is mitigated, and vaccine effects in a healthy, high-exposure group can be estimated.
+Health and social care workers (HCWs) were amongst the first groups eligible for vaccination due to the potentially high occupational exposure to the SARS-CoV-2 virus, making HCW status an important confounder of the effect of vaccination on infection and post-infection disease.
+By studying this group in isolation, this confounding is mitigated.
+Further, unlike many other young people vaccinated later, HCWs were vaccinated during a period of high community infection, increasing event rates and therefore statistical power.
 
-This cohort study used a target-trial design to assess the effectiveness of ChAdOx1 compared with BNT162b2 in HCWs, including second dose effectiveness, using the OpenSAFELY-TPP linked primary care database covering around 40% of England's population.
+This cohort study used a target-trial design to assess the effectiveness of ChAdOx1 compared with BNT162b2 against COVID-19-related outcomes in HCWs, including second dose effectiveness, using the OpenSAFELY-TPP linked primary care database covering around 40% of England's population.
+Effectiveness was assessed primarily in an era where the SARS-CoV-2 Alpha variant was dominant.
 
@@
 
-The OpenSAFELY-TPP database (<https://opensafely.org>) covers 24 million people registered at GP practices that use TPP SystmOne electronic health record software.
+The OpenSAFELY-TPP database (<https://opensafely.org>) covers 24 million people registered at National Health Service (NHS) General Practices using TPP SystmOne electronic health record software.
 It includes pseudonymized data such as coded diagnoses, medications and physiological parameters.
 No free text data are included.
-This primary care data is linked (via NHS numbers) with A&E attendance and in-patient hospital spell records via NHS Digital's Hospital Episode Statistics (HES), national coronavirus testing records via the Second Generation Surveillance System (SGSS), and national death registry records from the Office for National Statistics (ONS).
+This primary care data is linked, via NHS numbers, with Accident & Emergency (A&E) attendance and in-patient hospital spell records via NHS Digital's Hospital Episode Statistics (HES), national coronavirus testing records via the Second Generation Surveillance System (SGSS), and national death registry records from the Office for National Statistics (ONS).
 Vaccination status is available in the GP record directly via the National Immunisation Management System (NIMS).
-HCW status is recorded for all vaccine recipients at the time of vaccination, and this information is sent to OpenSAFELY-TPP from NHS Digital's COVID-19 data store.
+Self-designated HCW status is recorded for all vaccine recipients at the time of vaccination, and this information is sent to OpenSAFELY-TPP from NHS Digital's COVID-19 data store.
 
@@
 We studied health and social care workers in England vaccinated with either BNT162b2 or ChAdOx1.
-This group was prioritised for vaccination at the start of the vaccine roll-out due to the high occupational exposure to the SARS-CoV-2 virus, and many were vaccinated during the period where both vaccines were widely used.
+This group was prioritised for vaccination at the start of the vaccine roll-out due to the potentially high occupational exposure to the SARS-CoV-2 virus, and many were vaccinated during the period where both vaccines were widely used.
 
@@
 Positive SARS-CoV-2 tests were identified using SGSS records and based on swab date.
-Both polymerase chain reaction (PCR) and lateral flow tests are included, without differentiation between symptomatic and asymptomatic infection.
+Both polymerase chain reaction (PCR) and lateral flow positive tests were included, without differentiation between symptomatic and asymptomatic infection.
+Test provenance, for example patient-initiated tests versus employee-screening, was not known.
 COVID-19 A&E attendances were identified using HES emergency care records with U07.1 ("COVID-19, virus identified") or U07.2 ("COVID-19, virus not identified") ICD-10 diagnosis codes @emergenc .
@@
 
-At 12 weeks, `r label_percent(0.1)(baseline_pfizer$seconddose_12week)` of BNT162b2 recipients had received a second dose, compared with `r label_percent(0.1)(baseline_az$seconddose_12week)` of ChAdOx1 recipients.
-There were `r label_number(1)(crossbrand_n)` (`r label_percent(0.01)(crossbrand_prop)`) participants who received a second dose within 20 weeks that was not the same brand as the first dose, including Moderna mRNA-1273 vaccine second doses (supplementary Figure S3, Table S4).
+## Second doses
+
+At 12 weeks (72 days), `r label_percent(0.1)(baseline_pfizer$seconddose_12week)` of BNT162b2 recipients had received a second dose, compared with `r label_percent(0.1)(baseline_az$seconddose_12week)` of ChAdOx1 recipients.
+There were `r label_number(1)(crossbrand_n)` (`r label_percent(0.01)(crossbrand_prop)`) participants who received a second dose within 20 weeks (140 days) that was not the same brand as the first dose, including Moderna mRNA-1273 vaccine second doses (supplementary Figure S3, Table S4).
 
@@
 
-At 6 weeks post-vaccination, the ChAdOx1 versus BNT162b2 absolute risk difference per 1000 people (Figure 2) for a positive SARS-CoV-2 test was `r filter(riskdiff_ns3, outcome=="postest", tstop==7*6)$diff_format` (95%CI `r filter(riskdiff_ns3, outcome=="postest", tstop==7*6)$diff_CI_to`), for COVID-19 A&E attendances was `r filter(riskdiff_ns3, outcome=="covidemergency", tstop==7*6)$diff_format` (`r filter(riskdiff_ns3, outcome=="covidemergency", tstop==7*6)$diff_CI_to`) and for COVID-19 hospital admissions was `r filter(riskdiff_ns3, outcome=="covidadmitted", tstop==7*6)$diff_format` (`r filter(riskdiff_ns3, outcome=="covidadmitted", tstop==7*6)$diff_CI_to`).
+By 6 weeks (42 days) post-vaccination, prior to receipt of the second dose for the vast majority of participants, the ChAdOx1 versus BNT162b2 absolute risk difference per 1000 people (Figure 2) for a positive SARS-CoV-2 test was `r filter(riskdiff_ns3, outcome=="postest", tstop==7*6)$diff_format` (95%CI `r filter(riskdiff_ns3, outcome=="postest", tstop==7*6)$diff_CI_to`), for COVID-19 A&E attendances was `r filter(riskdiff_ns3, outcome=="covidemergency", tstop==7*6)$diff_format` (`r filter(riskdiff_ns3, outcome=="covidemergency", tstop==7*6)$diff_CI_to`) and for COVID-19 hospital admissions was `r filter(riskdiff_ns3, outcome=="covidadmitted", tstop==7*6)$diff_format` (`r filter(riskdiff_ns3, outcome=="covidadmitted", tstop==7*6)$diff_CI_to`).
 
-At 18 weeks post-vaccination, the absolute risk difference per 1000 people for a positive SARS-CoV-2 test was `r filter(riskdiff_ns3, outcome=="postest", tstop==7*18)$diff_format` (95%CI `r filter(riskdiff_ns3, outcome=="postest", tstop==7*18)$diff_CI_to`), for COVID-19 A&E attendances was `r filter(riskdiff_ns3, outcome=="covidemergency", tstop==7*18)$diff_format` (`r filter(riskdiff_ns3, outcome=="covidemergency", tstop==7*18)$diff_CI_to`) and for COVID-19 hospital admissions was `r filter(riskdiff_ns3, outcome=="covidadmitted", tstop==7*18)$diff_format` (`r filter(riskdiff_ns3, outcome=="covidadmitted", tstop==7*18)$diff_CI_to`).
+At 20 weeks (140 days) post-vaccination, at which time most participants had received their second dose, the absolute risk difference per 1000 people for a positive SARS-CoV-2 test was `r filter(riskdiff_ns3, outcome=="postest", tstop==7*20)$diff_format` (95%CI `r filter(riskdiff_ns3, outcome=="postest", tstop==7*20)$diff_CI_to`), for COVID-19 A&E attendances was `r filter(riskdiff_ns3, outcome=="covidemergency", tstop==7*20)$diff_format` (`r filter(riskdiff_ns3, outcome=="covidemergency", tstop==7*20)$diff_CI_to`) and for COVID-19 hospital admissions was `r filter(riskdiff_ns3, outcome=="covidadmitted", tstop==7*20)$diff_format` (`r filter(riskdiff_ns3, outcome=="covidadmitted", tstop==7*20)$diff_CI_to`).
+We note that extrapolating beyond 20 weeks for positive SARS-CoV-2 tests, we would infer a small advantage for BNT162b2 in terms of absolute risk.
 
+In general, events rates are much higher in the first few weeks after vaccination, and then there is a clear leveling-off after around 3-4 weeks for both vaccines, beyond which the event rates were very low.
+For COVID-19 hospital attendance and admission events in particular, the difference in the cumulative incidence (and their CIs) was well below 1 event per 1000 people in either direction.
 The estimated log hazard ratios were non-zero at certain periods (Figure 2), but the underlying absolute risk was small enough that this did not manifest in substantial differences in the cumulative incidence beyond the fourth week.
@@
       expand = expansion(0),
-      sec.axis = dup_axis(name="Higher incidence after \n <- ChAdOx1 / BNT162b2 ->", breaks = NULL)
+      sec.axis = dup_axis(name="Higher incidence after \n <- ChAdOx1 | BNT162b2 ->", breaks = NULL)
     )+
@@
       breaks=c(0.1, 0.2, 0.5, 1, 2, 5, 10),
-      sec.axis = dup_axis(name="Higher hazard after \n <- ChAdOx1 / BNT162b2 ->", breaks = NULL)
+      sec.axis = dup_axis(name="Higher hazard after \n <- ChAdOx1 | BNT162b2 ->", breaks = NULL)
     )+
@@
 
-This observational study of `r label_number(1, big.mark=",")(flowchart[[nrow(flowchart), "n"]])` health and social care workers living in England found similar outcomes for those receiving the BNT162b2 or ChAdOx1 COVID-19 vaccines during the first 20 weeks post-vaccination.
-There is a clear leveling-off of the incidence of positive tests after around 3-4 weeks for both vaccines, consistent with the expected time-to-onset of vaccine-induced immunity of around 2 weeks plus the delay from infection to a positive test.
-At 20 weeks, the difference in the cumulative incidence of positive test was `r filter(riskdiff_ns3, outcome=="postest", tstop==7*20)$diff_format` per 1000 people `r filter(riskdiff_ns3, outcome=="postest", tstop==7*20)$diff_CI` higher for ChAdOx1 compared with BNT162b2, and extrapolating beyond this time would suggest a small advantage for BNT162b2.
-However, BNT162b2 recipients were more likely to receive their second dose sooner, conferring a greater protective effect that is not accounted for in this estimate.
+## Key findings
 
-There were very few COVID-19 hospital attendance and admission events after this period, and the difference in the cumulative incidence (and their CIs) was well below 1 event per 1000 people in either direction.
+This observational study of `r label_number(1, big.mark=",")(flowchart[[nrow(flowchart), "n"]])` adult health and social care workers living in England found similar outcomes for those receiving the BNT162b2 or ChAdOx1 COVID-19 vaccines during the first 20 weeks post-vaccination in the Alpha variant era.
+Consistent with the expected time-to-onset of vaccine-induced immunity at around 2 weeks post-vaccination, plus the delay from infection to the event of interest, event rates tapered after the first few weeks.
+There were very few COVID-19 hospital attendance and admission events after this period, suggesting strong protective effects for both vaccines but nevertheless limiting power to reliably estimate comparative effectiveness with respect to severe COVID-19 outcomes.
+
+There is some evidence that, compared with ChAdOx1 recipients, there was better longer-term protection for BNT162b2 recipients, but they were more also likely to receive their second dose sooner, likely conferring a greater protective effect that is not accounted for in the intention-to-treat effectiveness estimates.
 
@@
 Though we adjusted for region, rurality, and deprivation, we were unable to directly account for occupational differences that may affect both exposure risk and vaccine type.
-We note also that the GP practices covered in this study are not geographically representative, as TPP SystmOne software is not widely-used in some regions such as London and the North West.
+For instance, we had no means to quantify any differences in HCW specialty, working hours, or patient contact time between the treatment groups.
+The primary care practices covered in this study are not geographically representative, as TPP SystmOne software is not widely-used in some regions such as London and the North West.
 
@@
 Regardless, there was insufficient power to reliably estimate comparative second-dose effects as event incidence at 12 weeks and beyond was low for both vaccines, with fewer than 10 positive tests per 1000-person-years, and far fewer for COVID-19 A&E attendances and hospital admissions during this period.
-The need for many thousands more person-years to detect differences in second-dose effects is encouraging to the extent that it suggests extremely high protective effects for both vaccines, though the fact that background case rates in England were declining over the duration of the study period should not be ignored.
+The need for many thousands more person-years to detect differences in second-dose effects is encouraging to the extent that it suggests extremely high protective effects for both vaccines, though declining background case rates in England over the duration of the study period (Figure S4) would have also contributed to reduced event rates in both treatment groups.
 
@@
 
-A number of studies have estimated COVID-19 vaccine effectiveness in observational data @vasileiou2021 @cabezas2021 @effectiv @hall2021.
+A number of studies have estimated first-dose COVID-19 vaccine effectiveness in observational data @Iheanacho2021, including BNT162b2 and ChAdOx1 vaccines in similar populations @vasileiou2021 @cabezas2021 @effectiv @bernal2021 @Shrotri2021 and in healthcare workers @Paris2021 @hall2021.
 Vaccine protection is typically assessed by comparing vaccinated against unvaccinated person-time, or against the early post-vaccination period before anticipated onset of vaccine-induced immunity.
-Such designs are vulnerable to confounding and selection bias, for example due to vaccine prioritisation and eligibility policies.
+Such designs are vulnerable to confounding and selection bias, for example due to vaccine prioritisation and eligibility policies and risk or health-seeking behaviours.
 In addition, differences in vaccine access and acceptance can cause substantial imbalance between vaccinated and unvaccinated groups that may be impossible to account for.
-Many of these biases can be bypassed when making direct comparisons between recipients of different vaccine types, if the relative performance of two vaccines is of interest.
+Many of these biases can be bypassed when making direct comparisons between recipients of different vaccine types, if the relative performance of two vaccines is of interest \@@Dickerman2022.
 To our knowledge, the present study is the first to assess effectiveness of the BNT162b2 and ChAdOx1 vaccines in a head-to-head comparison in either an experimental or observational setting.
 
-Our study suggests that there is little to differentiate the vaccines with respect to their respective protective effects against infection and hospital-related COVID-19 outcomes.
+Our study suggests that there is little to differentiate the vaccines with respect to their protective effects against infection and hospital-related COVID-19 outcomes within the first 20 weeks after the first dose in adult HCWs under 65 years old.
 This pushes decisions about which vaccine to favour onto other attributes such as safety profile, logistics, and cost.
@@
 
-This study found no substantial differences in the incidence of SARS-CoV-2 infection or COVID-19-related hospital events following vaccination with BNT162b2 or ChAdOx1 in a cohort of health and social care workers in England.
-The incidence of COVID-19-related A&E attendances or hospital admissions four weeks after the first dose of either vaccine was fewer than one event in 10,000-person-years, suggesting strong protective effects for both vaccines but nevertheless limiting power to reliably estimate comparative effectiveness with respect to severe COVID-19 outcomes.
-Further studies are needed to assess comparative effectiveness in newer, more prevalent variants and to assess longer-term effectiveness.
+This study found no substantial differences in the incidence of SARS-CoV-2 infection or COVID-19-related hospital events following vaccination with BNT162b2 or ChAdOx1 in a cohort of health and social care workers in England in the Alpha era.
+There were very few severe COVID-19 outcomes after the expected onset of vaccine-induced immunity at around 2 weeks, reducing power to reliably compare effectiveness against these outcomes.
+Further studies are needed to assess comparative effectiveness against newer, more prevalent variants and to assess longer-term effectiveness.
 

### 2022-03-30 manuscript/draft-supplement-v1-1.Rmd
some revisions

--- 2021-10-13 manuscript/draft-supplement.Rmd
@@
         expand = expansion(0),
-        sec.axis = dup_axis(name="Lower risk after \n <- BNT162b2 / ChAdOx1 ->", breaks = NULL)
+        sec.axis = dup_axis(name="Higher incidence after \n <- ChAdOx1 | BNT162b2 ->", breaks = NULL)
       )+
@@
         breaks=c(0.1, 0.2, 0.5, 1, 2, 5, 10),
-        sec.axis = dup_axis(name="Lower hazard after \n <- BNT162b2 / ChAdOx1 ->", breaks = NULL)
+        sec.axis = dup_axis(name="Higher hazard after \n <- ChAdOx1 | BNT162b2 ->", breaks = NULL)
       )+

### 2022-03-30 create-project.R
rename sensitivity outputs

@@ -546,7 +546,7 @@ actions_list <- splice(
 
   action(
     name = "report_objects_sensitivity",
-    run = "r:latest analysis/report/report-objects.R",
+    run = "r:latest analysis/report/report-objects-sensitivity.R",
     arguments = NULL,
     needs = splice(
       "design", "data_selection",
